### 2018-09-18 Docs/CCS18/Communication/ccs2018.tex
scripts

\documentclass[english,11pt]{beamer}

\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Proba}{\mathbb{P}}

\newcommand{\Covb}[2]{\ensuremath{\Cov\!\left[#1,#2\right]}}
\newcommand{\Eb}[1]{\ensuremath{\E\!\left[#1\right]}}
\newcommand{\Pb}[1]{\ensuremath{\Proba\!\left[#1\right]}}
\newcommand{\Varb}[1]{\ensuremath{\Var\!\left[#1\right]}}

% norm
\newcommand{\norm}[1]{\| #1 \|}

\newcommand{\indep}{\rotatebox[origin=c]{90}{$\models$}}

\usepackage{mathptmx,amsmath,amssymb,graphicx,bibentry,bbm,babel,ragged2e}

\makeatletter

\newcommand{\noun}[1]{\textsc{#1}}
\newcommand{\jitem}[1]{\item \begin{justify} #1 \end{justify} \vfill{}}
\newcommand{\sframe}[2]{\frame{\frametitle{#1} #2}}

\newenvironment{centercolumns}{\begin{columns}[c]}{\end{columns}}
%\newenvironment{jitem}{\begin{justify}\begin{itemize}}{\end{itemize}\end{justify}}

\usetheme{Warsaw}
\setbeamertemplate{footline}[text line]{}
\setbeamercolor{structure}{fg=purple!50!blue, bg=purple!50!blue}

\setbeamersize{text margin left=15pt,text margin right=15pt}

\setbeamercovered{transparent}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{multirow}

\makeatother

\begin{document}

\title{A systematic comparison of interaction models for systems of cities}

\author{J.~Raimbault$^{1,2,\ast}$\\
\texttt{juste.raimbault@polytechnique.edu}
}

\institute{$^{1}$Complex Systems Institute, Paris, UPS CNRS 3611 ISC-PIF\\
$^{2}$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}

\date{CCS 2018\\\smallskip
Thessaloniki\\\smallskip
September 25th 2018
}

\frame{\maketitle}

% Understanding patterns of growth for cities is a crucial issue for its practical and policy implications e.g. regarding sustainability, but also from a theoretical point of view regarding the validation of theories for urban systems. A particular entry is taken by the Evolutive Urban Theory [1] which postulates interactions between cities as main drivers of their growth. Within this framework, numerous concurrent models have been introduced for the evolution of systems of cities, and applied on diverse case studies, but there is to the best of our knowledge no systematic comparison of performances across models and application cases. This contribution makes a first step towards such a systematic benchmark. We consider simple models simulating population of cities only, but including heterogenous underlying processes and assumptions. More precisely, we compare (i) the Favaro-Pumain model for the diffusion of innovation [2]; (ii) the Marius multi-model based on economic exchanges [3]; and (iii) a model including flows within abstract physical networks parametrized on elevation data [4]. These models are calibrated on a large scale harmonized dataset [5] including the European, former Soviet Union, Chinese, Brazilian, South-African, Indian and USA systems of cities on a time period covering 1960-2010. Calibrations are done for different versions of each model including more or less parameters, using distributed genetic algorithms on a computation grid through the intermediary of the OpenMOLE model exploration software [6]. Calibration results show that: (i) at a fixed number of parameters, no model performs particularly better on all urban systems, suggesting the complementary of the innovation, economic, and network dimensions taken into account by the different models; and (ii) when adjusting for the number of parameters with an empirical information criterion [4], we find that additional components generally improve performances for all models, what reveals a high effective dimensionality of urban systems. This work confirms the need for multiple complementary approaches to model urban systems and sketches a framework allowing systematic benchmarks of concurrent models for complex urban systems.

\sframe{Urban systems}{

\centering

	%\includegraphics[width=\textwidth]{figures/}

	\footnotesize
\textit{Source: }

}

\sframe{Modeling urban growth}{

\centering

}

\sframe{Urban systems interaction models}{

\centering

Comparison of three approaches based on the evolutionary urban theory \cite{pumain1997pour} capturing different dimensions of urban systems:
\begin{itemize}
	\item The Favaro-Pumain model for the diffusion of innovation \cite{favaro2011gibrat}
	\item 
\end{itemize}

}

\sframe{Model calibration}{

Computationally intensive: high-dimensional parameter space and possible spatial setup.

\medskip

$\rightarrow$ use of grid computing, made smooth with the OpenMOLE software \url{https://next.openmole.org/}

\centering

\smallskip

\includegraphics[height=0.35\textheight]{figures/openmole.png}

\smallskip

\raggedright\justify

\footnotesize \textit{OpenMOLE: (i) embed any model as a black box; (ii) transparent access to main High Performance Computing environments; (iii) model exploration and calibration methods.}

}

\sframe{Discussion}{

\justify

\vspace{-1cm}

\textbf{Implications}

$\rightarrow$

\smallskip

$\rightarrow$

\bigskip

\textbf{Developments}

$\rightarrow$

\smallskip

$\rightarrow$

\smallskip

$\rightarrow$ More elaborated method to compare models in a ``fair'' way (correcting for additional parameters, open question for models of simulation).

}

\sframe{Conclusion}{

\justify

\vspace{-1cm}

$\rightarrow$ % model coupling / benchmarking

\medskip

$\rightarrow$ Multiple perspectives on urban systems ? \textbf{Need for more interdisciplinarity.}

\bigskip

\footnotesize

\textbf{Related works}

Raimbault, J. (2018). Indirect evidence of network effects in a system of cities. Environment and Planning B: Urban Analytics and City Science, 2399808318774335.

\url{https://halshs.archives-ouvertes.fr/halshs-01788559}

\medskip

Raimbault, J. (2018). Modeling the co-evolution of cities and networks. \textit{Forthcoming in Handbook of cities and networks, Rozenblat C., Niel Z., eds.} arXiv:1804.09430.

\medskip

Raimbault, J. (2018). Caractérisation et modélisation de la co-évolution des réseaux de transport et des territoires (Doctoral dissertation, Université Paris 7 Denis Diderot). \url{https://halshs.archives-ouvertes.fr/tel-01857741}

\medskip

\textbf{Open repository} (code, data and results) at\\\texttt{https://github.com/JusteRaimbault/UrbanGrowth}\\\medskip
\textbf{Acknowledgments} : I thank the \textit{European Grid Infrastructure} and its \textit{National Grid Initiatives} (\textit{France-Grilles} in particular) to give the technical support and the infrastructure.

}

\sframe{Reserve slides}{

\centering

\Large

\textbf{Reserve Slides}

}

%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{/Users/juste/ComplexSystems/CityNetwork/Biblio/Bibtex/CityNetwork,biblio}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2018-09-18 Models/urbangrowth/openmole/runCalib.sh

@@ -1,0 +1,1 @@
+cat runs| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G"}'|sh

### 2018-09-19 Docs/CCS18/Communication/ccs2018.tex
slides
copyres

@@
 
+% meso / micro
+
+% macro
+
+}
+
+\sframe{An evolutionary urban theory}{
+\centering
+
+\footnotesize
+\textit{\cite{raimbault2017applied} Citation network analysis of core publications in the evolutionary urban theory}
+
+}
+
+\sframe{Towards a systematic model comparison}{
+
+\centering
+
+% research question
+
+% before coupling models, benchmarking them !
+
+\textbf{Research objective : }
+
 }
@@
 	\item The Favaro-Pumain model for the diffusion of innovation \cite{favaro2011gibrat}
-	\item 
+	\item The Marius model family based on economic exchanges \cite{cottineau2014evolution}
+	\item An interaction model including physical transportation networks \cite{raimbault2018indirect}
 \end{itemize}
+
+}
+
+\sframe{Datasets}{
+
+\centering
+
+% geodivercity harmonization of datasets
+
+}
+
+\sframe{Data preprocessing}{
+\centering
+
+Remove small cities (\cite{adam2006medium} for definition of medium-sized)
+
+}
+
+\sframe{Stylized facts}{
+\centering
+
+% - fit distributions of growth rates
+% - cluster by type of distrib ?
 
@@
 \textbf{Developments}
+
+% future perspective : towards integrative theories / coupling of these models for multi-dimensional grasp of urban systems (more work needed to understand model coupling and role of parameters in fit improvement)
+% Compare these models on synthetic city systems / with suited indicators (here relatively data-driven approach)
 

### 2018-09-19 Models/DataProcessing/clean.R

@@ -22,7 +22,9 @@ colnames(cities)<-c("ID","Name","Long","Lat","X1961","X1971","X1981","X1991","X2
 #cities[apply(cities,1,function(r){length(which(is.na(r)))>0}),]
 # remove nas (20 rows)
 cities=cities[apply(cities,1,function(r){length(which(is.na(r)))==0}),]
-cities = cities[cities$X2011>=median(cities$X2011),]
+#cities = cities[cities$X2011>=median(cities$X2011),]
+# for memory purposes, diminish significantly the number of cities
+cities = cities[cities$X2011>=quantile(cities$X2011,0.75),]
 write.table(cities,file=paste0(targetDir,'Europe.csv'),row.names = F,sep=",")
 
 

### 2018-09-19 Models/urbangrowth/openmole/copyRes.py

@@ -1,0 +1,21 @@
+import sys,os,shutil
+
+# copy latest generation for each run in a folder
+
+SOURCE=sys.argv[1]
+TARGET=sys.argv[2]
+
+def getLatestGen(dir):
+    files = os.listdir(dir)
+    return(map(lambda s : int(s.split('population')[1].split('.')[0]),files).sort(reverse=T)[0])
+
+
+sourcedirs = os.listdir(SOURCE)
+targetdirs = os.listdir(TARGET)
+
+for dir in sourcedirs :
+    if not os.path.isdir(TARGET+'/'+dir):
+        os.mkdir(TARGET+'/'+dir)
+    gen = getLatestGen(dir)
+    if not os.path.isfile(TARGET+'/'+dir+'/population'+gen+'.csv'):
+        shutil.copy(SOURCE+'/'+dir+'/population'+gen+'.csv',TARGET+'/'+dir+'/population'+gen+'.csv')

### 2018-09-19 Models/urbangrowth/openmole/copyRes.py
res

@@ -7,7 +7,8 @@ TARGET=sys.argv[2]
 
 def getLatestGen(dir):
     files = os.listdir(dir)
-    return(map(lambda s : int(s.split('population')[1].split('.')[0]),files).sort(reverse=T)[0])
+    print(sorted(map(lambda s:int(s.split('population')[1].split('.')[0]),files),reverse=True)[0])
+    return(str(sorted(map(lambda s : int(s.split('population')[1].split('.')[0]),files),reverse=True)[0]))
 
 
 sourcedirs = os.listdir(SOURCE)
@@ -16,6 +17,6 @@ targetdirs = os.listdir(TARGET)
 for dir in sourcedirs :
     if not os.path.isdir(TARGET+'/'+dir):
         os.mkdir(TARGET+'/'+dir)
-    gen = getLatestGen(dir)
+    gen = getLatestGen(SOURCE+'/'+dir)
     if not os.path.isfile(TARGET+'/'+dir+'/population'+gen+'.csv'):
         shutil.copy(SOURCE+'/'+dir+'/population'+gen+'.csv',TARGET+'/'+dir+'/population'+gen+'.csv')

### 2018-09-19 Models/Analysis/calibration.R
all_runs

@@ -1,0 +1,78 @@
+
+setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/urbangrowth/openmole'))
+
+library(dplyr)
+library(ggplot2)
+
+source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+
+# parameters : where calibration results are stored and where to store result figures
+sourcedir = 'calibration/'
+resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Calibration/')
+
+latestgen <- function(dir){max(as.integer(sapply(strsplit(sapply(strsplit(list.files(dir),"population"),function(s){s[2]}),".csv"),function(s){s[1]})))}
+
+# population dirs listed by hand
+
+models = c('intgib')
+systems = c('ZA','CN','US')
+
+popdirs = list(
+  'intgib_ZA'='CALIB_intgib_ZA_20180919_052829',
+  'intgib_CN'='CALIB_intgib_CN_20180918_231457',
+  'intgib_US'='CALIB_intgib_US_20180918_175315'
+)
+
+populations = list()
+
+for(popname in names(popdirs)){
+  populations[[popname]] = read.csv(paste0(sourcedir,popdirs[[popname]],'/population',latestgen(paste0(sourcedir,popdirs[[popname]])),'.csv'))
+}
+
+
+# aggregate with model and system
+pop = data.frame()
+for(model in models){
+  for(system in systems){
+    currentdata = populations[[paste0(model,'_',system)]]
+    if(!is.null(currentdata)){
+      pop=rbind(pop,cbind(currentdata,model=rep(model,nrow(currentdata)),system=rep(system,nrow(currentdata))))
+    }
+  }
+}
+pop=as.tbl(pop)
+
+
+#######
+
+# compare all models and systems
+g = ggplot(pop,aes(x = mselog,y=logmse,color=as.character(gravityWeight>0)))
+g+geom_point()+facet_wrap(~system,scales = 'free')
+
+# -> checking when baseline is better than interaction model
+
+
+
+## exploration of influence of parameters
+
+g=ggplot(pop[pop$gravityWeight>0,],aes(x=gravityWeight))
[…]

### 2018-09-19 Models/DataProcessing/clean.R

@@ -24,7 +24,7 @@ colnames(cities)<-c("ID","Name","Long","Lat","X1961","X1971","X1981","X1991","X2
 cities=cities[apply(cities,1,function(r){length(which(is.na(r)))==0}),]
 #cities = cities[cities$X2011>=median(cities$X2011),]
 # for memory purposes, diminish significantly the number of cities
-cities = cities[cities$X2011>=quantile(cities$X2011,0.75),]
+cities = cities[cities$X2011>=quantile(cities$X2011,0.85),]
 write.table(cities,file=paste0(targetDir,'Europe.csv'),row.names = F,sep=",")
 
 
@@ -39,7 +39,8 @@ write.table(cities,file=paste0(targetDir,cityfiles[[countrycode]],'.csv'),row.na
 countrycode = 'BR'
 cities <- read.csv(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Data/Geodivercity/data/',cityfiles[[countrycode]],'.csv'))
 cities = cities[,c("ID","Name","Long","Lat","X1960","X1970","X1980","X1991","X2000","X2010")]
-cities = cities[cities$X2010>=median(cities$X2010),]
+#cities = cities[cities$X2010>=median(cities$X2010),]
+cities = cities[cities$X2010>=quantile(cities$X2010,0.7),]
 write.table(cities,file=paste0(targetDir,cityfiles[[countrycode]],'.csv'),row.names = F,sep=",")
 
 
@@ -48,7 +49,8 @@ countrycode = 'IN'
 cities <- read.csv(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Data/Geodivercity/data/',cityfiles[[countrycode]],'.csv'))
 cities = cities[,c("ID","Name","Long","Lat","X1961","X1981","X1991","X2001","X2011")]
 cities=cities[apply(cities[,c("X1961","X1981","X1991","X2001","X2011")],1,function(r){length(which(is.na(r)))==0}),]
-cities = cities[cities$X2011>=median(cities$X2011),]
+#cities = cities[cities$X2011>=median(cities$X2011),]
+cities = cities[cities$X2011>=quantile(cities$X2011,0.75),]
 write.table(cities,file=paste0(targetDir,cityfiles[[countrycode]],'.csv'),row.names = F,sep=",")
 
 
[…]

### 2018-09-19 Models/DataProcessing/distMats.R

@@ -13,7 +13,8 @@ countrycodes = c('EU','BR','CN','IN','RU','ZA','US')
 # 'FR'='France'
 cityfiles = list('EU'='Europe','BR'='Brazil','CN'='China','IN'='India','RU'='Russia','ZA'='South-Africa','US'='USA')
 
-withGraph = F
+#withGraph = F
+withGraph = T
 
 if(withGraph){
   dem <- raster(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Data/DEM/raw/DEM_geotiff/alwdgg.tif'))

### 2018-09-19 Models/DataProcessing/functions.R

@@ -5,8 +5,9 @@ library(rgdal)
 
 extractRaster <- function(dem,countries,countrycode){
   country = countries[countries$CNTR_ID==countrycode,]
-  country = SpatialPolygons(list(Polygons(list(country@polygons[[1]]@Polygons[[1]]),ID="id")),proj4string = country@proj4string)
+  if(countrycode!='EU'){country = SpatialPolygons(list(Polygons(list(country@polygons[[1]]@Polygons[[1]]),ID="id")),proj4string = country@proj4string)}
   ext <- extent(country)
+  if(countrycode=='EU'){ext = extent(matrix(c(-17,28,33,66 ),nrow=2))}
   
   rows=rowFromY(dem,ext@ymax):rowFromY(dem,ext@ymin)
   cols=colFromX(dem,ext@xmin):colFromX(dem,ext@xmax)

### 2018-09-19 Models/urbangrowth/src/main/scala/urbangrowth/models/innovation/Innovation.scala

@@ -89,7 +89,10 @@ case class Innovation(
 
   override def run(): Result = Innovation.run(this)
 
-  override def toString: String = "Innovation model"
+  override def toString: String = "Innovation model with parameters"+
+     "\n\tgrowthRate = "+growthRate+"\n\tinnovationWeight = "+innovationWeight+"\n\tgravityDecay = "+gravityDecay+
+     "\n\tinnovationDecay = "+innovationDecay+"\n\tinnovationUtility = "+innovationUtility+"\n\tinnovationUtilityGrowth = "+innovationUtilityGrowth+
+     "\n\tearlyAdoptersRate = "+earlyAdoptersRate+"\n\tnewInnovationHierarchy = "+newInnovationHierarchy+"\n\tnewInnovationPopulationProportion = "+newInnovationPopulationProportion
 
 }
 

### 2018-09-20 Models/Analysis/calibration.R
rungib

@@ -14,38 +14,61 @@ latestgen <- function(dir){max(as.integer(sapply(strsplit(sapply(strsplit(list.f
 
 # population dirs listed by hand
 
-models = c('intgib')
-systems = c('ZA','CN','US')
+models = c('intgib','innovation')
+systems = c('ZA','CN','US','BR','EU','IN','RU')
 
 popdirs = list(
   'intgib_ZA'='CALIB_intgib_ZA_20180919_052829',
   'intgib_CN'='CALIB_intgib_CN_20180918_231457',
-  'intgib_US'='CALIB_intgib_US_20180918_175315'
+  'intgib_US'='CALIB_intgib_US_20180918_175315',
+  'intgib_BR'='CALIB_intgib_BR_20180919_143214',
+  'intgib_EU'='CALIB_intgib_EU_20180919_163321',
+  'intgib_IN'='CALIB_intgib_IN_20180919_183427',
+  'intgib_RU'='CALIB_intgib_RU_20180919_203608',
+  'innovation_ZA'='CALIB_innovation_ZA_20180920_064141',
+  'innovation_CN'='CALIB_innovation_CN_20180920_003846',
+  'innovation_US'='CALIB_innovation_US_20180920_084300',
+  'innovation_BR'='CALIB_innovation_BR_20180919_223752',
+  #'innovation_EU'=''
+  'innovation_IN'='CALIB_innovation_IN_20180920_023945',
+  'innovation_RU'='CALIB_innovation_RU_20180920_044043'
 )
 
 populations = list()
 
+popcolnames=c()
 for(popname in names(popdirs)){
   populations[[popname]] = read.csv(paste0(sourcedir,popdirs[[popname]],'/population',latestgen(paste0(sourcedir,popdirs[[popname]])),'.csv'))
+  popcolnames=append(popcolnames,colnames(populations[[popname]]))
 }
-
+popcolnames=unique(c(popcolnames,'model','system'))
 
 # aggregate with model and system
-pop = data.frame()
+pop = data.frame(matrix(rep(0,length(popcolnames)),ncol=length(popcolnames)))
+colnames(pop)=popcolnames
 for(model in models){
   for(system in systems){
     currentdata = populations[[paste0(model,'_',system)]]
     if(!is.null(currentdata)){
-      pop=rbind(pop,cbind(currentdata,model=rep(model,nrow(currentdata)),system=rep(system,nrow(currentdata))))
[…]

### 2018-09-20 Models/urbangrowth/openmole/runCalib.sh
.

@@ -1,1 +1,1 @@
-cat runs| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G"}'|sh
+cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G"}'|sh

### 2018-09-20 Models/Analysis/calibration.R
calib_marius

@@ -14,7 +14,7 @@ latestgen <- function(dir){max(as.integer(sapply(strsplit(sapply(strsplit(list.f
 
 # population dirs listed by hand
 
-models = c('intgib','innovation')
+models = c('intgib','innovation','gibrat')
 systems = c('ZA','CN','US','BR','EU','IN','RU')
 
 popdirs = list(
@@ -31,7 +31,18 @@ popdirs = list(
   'innovation_BR'='CALIB_innovation_BR_20180919_223752',
   #'innovation_EU'=''
   'innovation_IN'='CALIB_innovation_IN_20180920_023945',
-  'innovation_RU'='CALIB_innovation_RU_20180920_044043'
+  'innovation_RU'='CALIB_innovation_RU_20180920_044043',
+  'gibrat_BR'='CALIB_gibrat_BR_20180920'
+)
+
+systembounds=list(
+  'BR'=c('logmse'=40,'mselog'=1500),
+  'IN'=c('logmse'=33,'mselog'=300),
+  'ZA'=c('logmse'=40,'mselog'=25000),
+  'US'=c('logmse'=40,'mselog'=21000),
+  'RU'=c('logmse'=50,'mselog'=65000),
+  'EU'=c('logmse'=40,'mselog'=1000),
+  'CN'=c('logmse'=40,'mselog'=2000)
 )
 
 populations = list()
@@ -54,6 +65,8 @@ for(model in models){
       show(dim(currentdata))
       for(col in popcolnames[!popcolnames%in%colnames(currentdata)]){currentdata[,col]=rep(NA,nrow(currentdata))}
       currentdata[,"system"]=rep(system,nrow(currentdata));currentdata[,"model"]=rep(model,nrow(currentdata))
+      #filter with system bounds
+      currentdata = currentdata[currentdata$logmse<systembounds[[system]][['logmse']]&currentdata$mselog<systembounds[[system]][['mselog']],]
       #pop=rbind(pop,cbind(currentdata,model=rep(model,nrow(currentdata)),system=rep(system,nrow(currentdata))))
       pop=rbind(pop,currentdata)
     }

### 2018-09-20 Models/urbangrowth/openmole/runCalib.sh

@@ -1,1 +1,1 @@
-cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G"}'|sh
+cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; #openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G"}'|sh

### 2018-09-20 Models/urbangrowth/openmole/copyRes.py
copyresmultilple

@@ -4,6 +4,7 @@ import sys,os,shutil
 
 SOURCE=sys.argv[1]
 TARGET=sys.argv[2]
+MULTIPLE=sys.argv[3]
 
 def getLatestGen(dir):
     files = os.listdir(dir)
@@ -18,5 +19,9 @@ for dir in sourcedirs :
     if not os.path.isdir(TARGET+'/'+dir):
         os.mkdir(TARGET+'/'+dir)
     gen = getLatestGen(SOURCE+'/'+dir)
-    if not os.path.isfile(TARGET+'/'+dir+'/population'+gen+'.csv'):
-        shutil.copy(SOURCE+'/'+dir+'/population'+gen+'.csv',TARGET+'/'+dir+'/population'+gen+'.csv')
+    gens = [gen]
+    if len(MULTIPLE) > 0 :
+        gens = range((gen - 1000),gen,step = 100)
+    for gen in gens :
+        if not os.path.isfile(TARGET+'/'+dir+'/population'+gen+'.csv'):
+            shutil.copy(SOURCE+'/'+dir+'/population'+gen+'.csv',TARGET+'/'+dir+'/population'+gen+'.csv')

### 2018-09-20 Models/urbangrowth/openmole/copyRes.py
a_bit_more_file(convergence)

@@ -4,7 +4,9 @@ import sys,os,shutil
 
 SOURCE=sys.argv[1]
 TARGET=sys.argv[2]
-MULTIPLE=sys.argv[3]
+MULTIPLE=False
+if len(sys.argv) > 3:
+    MULTIPLE=True
 
 def getLatestGen(dir):
     files = os.listdir(dir)
@@ -20,8 +22,8 @@ for dir in sourcedirs :
         os.mkdir(TARGET+'/'+dir)
     gen = getLatestGen(SOURCE+'/'+dir)
     gens = [gen]
-    if len(MULTIPLE) > 0 :
-        gens = range((gen - 1000),gen,step = 100)
+    if MULTIPLE :
+        gens = range((int(gen) - 1000),int(gen),100)
     for gen in gens :
-        if not os.path.isfile(TARGET+'/'+dir+'/population'+gen+'.csv'):
-            shutil.copy(SOURCE+'/'+dir+'/population'+gen+'.csv',TARGET+'/'+dir+'/population'+gen+'.csv')
+        if not os.path.isfile(TARGET+'/'+dir+'/population'+str(gen)+'.csv') and os.path.isfile(SOURCE+'/'+dir+'/population'+str(gen)+'.csv'):
+            shutil.copy(SOURCE+'/'+dir+'/population'+str(gen)+'.csv',TARGET+'/'+dir+'/population'+str(gen)+'.csv')

### 2018-09-20 Docs/CCS18/Communication/ccs2018.tex
runmarius

@@
 
+\sframe{Network interaction model}{
+
+% increasing complexity for the presentation of models
+
+}
+
+\sframe{The Favaro-Pumain model}{
+
+}
+
 \sframe{Datasets}{
@@
 % geodivercity harmonization of datasets
+
+Harmonized dataset, explored in particular by \cite{pumain2015multilevel} (built by
 
@@
 \footnotesize \textit{OpenMOLE: (i) embed any model as a black box; (ii) transparent access to main High Performance Computing environments; (iii) model exploration and calibration methods.}
+
+}
+
+\sframe{Model calibration}{
+
+% number of runs (equivalent / year) ; caracs of NSGA2 ; convergence for each model/system
+
+}
+
+\sframe{Results: comparison of models on the different systems}{
+
+% global overview : Pareto plots / tabular of % effective pareto front for each system
+% (requires the extraction of effective fronts)
+
+}
+
+\sframe{Results: particular cases}{
+
+% systems on which the behavior is particularly interesting ?
+
+}
+
+%\sframe{Results: influence of parameters}{
+% add if cool results
+%}
+
+\sframe{Results: empirical AIC}{
+% TODO : select optimal points / run empirical AIC
+% OR : run on all points of the Pareto front ? -> local OK
 
@@
 (2018).
-Caractérisation et modélisation de la co-évolution des réseaux de transport et des territoires (Doctoral dissertation, Université Paris 7 Denis Diderot).
+Caract{\'e}risation et mod{\'e}lisation de la co-{\'e}volution des r{\'e}seaux de transport et des territoires (Doctoral dissertation, Université Paris 7 Denis Diderot).
 \url{https://halshs.archives-ouvertes.fr/tel-01857741}

### 2018-09-20 Models/Analysis/calibration.R

@@ -6,6 +6,8 @@ library(ggplot2)
 
 source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 
+source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
+
 # parameters : where calibration results are stored and where to store result figures
 sourcedir = 'calibration/'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Calibration/')
@@ -44,6 +46,17 @@ systembounds=list(
   'EU'=c('logmse'=40,'mselog'=1000),
   'CN'=c('logmse'=40,'mselog'=2000)
 )
+
+### CV check
+
+for(popdir in paste0(sourcedir,popdirs)){
+  fdiff = frontDiffs(popdir)
+  plot(fdiff$gens,fdiff$dists)
+}
+
+
+
+### Analysis
 
 populations = list()
 

### 2018-09-20 Models/Analysis/functions.R

@@ -1,0 +1,24 @@
+
+library(sp)
+
+#'
+#' Get successive distances between pareto fronts
+#'
+frontDiffs <- function(popdirectory,objectives=c('logmse','mselog')){
+  files = sort(list.files(popdirectory))
+  #show(files)
+  gens = sapply(strsplit(files,'population'),function(l){sapply(strsplit(l[2],split = '.',fixed = T),function(l){as.numeric(l[1])})})
+  dists = c()
+  for(i in 2:length(files)){
+    prevfront=read.csv(file=paste0(popdirectory,'/',files[i-1]))
+    currentfront=read.csv(file=paste0(popdirectory,'/',files[i]))
+    #dists = append(dists,sum(c(dist(as.matrix(prevfront[,objectives]),as.matrix(currentfront[,objectives])))))
+    owndistprev = mean(c(dist(as.matrix(prevfront[,objectives]))))
+    owndistcurrent = mean(c(dist(as.matrix(currentfront[,objectives]))))
+    betweendist = mean(c(spDists(as.matrix(prevfront[,objectives]),as.matrix(currentfront[,objectives]))))
+    dists = append(dists,2*betweendist/(owndistprev+owndistcurrent))
+  }
+  return(list(gens=gens[2:length(gens)],dists=dists))
+}
+
+

### 2018-09-20 Models/urbangrowth/openmole/runCalib.sh

@@ -1,1 +1,1 @@
-cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; #openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G"}'|sh
+cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G"}'|sh

### 2018-09-21 Models/urbangrowth/openmole/runCalib.sh
innovext

@@ -1,1 +1,1 @@
-cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G"}'|sh
+cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 512G --logger-level ALL"}'|sh

### 2018-09-21 Models/Analysis/calibration.R
debug

@@ -16,7 +16,7 @@ latestgen <- function(dir){max(as.integer(sapply(strsplit(sapply(strsplit(list.f
 
 # population dirs listed by hand
 
-models = c('intgib','innovation','gibrat')
+models = c('intgib','innovation','gibrat','innovationext')
 systems = c('ZA','CN','US','BR','EU','IN','RU')
 
 popdirs = list(
@@ -31,10 +31,23 @@ popdirs = list(
   'innovation_CN'='CALIB_innovation_CN_20180920_003846',
   'innovation_US'='CALIB_innovation_US_20180920_084300',
   'innovation_BR'='CALIB_innovation_BR_20180919_223752',
-  #'innovation_EU'=''
+  #'innovation_EU'='',
   'innovation_IN'='CALIB_innovation_IN_20180920_023945',
   'innovation_RU'='CALIB_innovation_RU_20180920_044043',
-  'gibrat_BR'='CALIB_gibrat_BR_20180920'
+  'gibrat_ZA'='CALIB_gibrat_ZA_20180920',
+  #'gibrat_CN'='',
+  'gibrat_US'='CALIB_gibrat_US_20180920',
+  'gibrat_BR'='CALIB_gibrat_BR_20180920',
+  'gibrat_EU'='CALIB_gibrat_EU_20180920',
+  'gibrat_IN'='CALIB_gibrat_IN_20180920',
+  'gibrat_RU'='CALIB_gibrat_RU_20180920',
+  'innovationext_ZA'='CALIB_innovationext_ZA_20180921_081251',
+  'innovationext_CN'='CALIB_innovationext_CN_20180921_020951',
+  #'innovationext_US'='',
+  'innovationext_BR'='CALIB_innovationext_BR_20180921_000851',
+  #'innovationext_EU'='',
+  'innovationext_IN'='CALIB_innovationext_IN_20180921_041049',
+  'innovationext_RU'='CALIB_innovationext_RU_20180921_061152'
 )
 
 systembounds=list(

### 2018-09-21 Models/urbangrowth/src/main/scala/urbangrowth/indicators/Indicators.scala

@@ -29,6 +29,8 @@ object Indicators {
     * @return
     */
   def logmse(m: Matrix,target: Matrix): Double = {
+    println(m.getColumnDimension)
+    println(m.getRowDimension)
     val sqdiff = m.minus(target).arrayTimes(m.minus(target))
     return Math.log(sqdiff.getArray().flatten.sum)
   }

### 2018-09-21 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/Marius.scala

@@ -118,7 +118,7 @@ trait Marius {
     }.takeWhileInclusive {
       _ match {
         case Success(s) => !ended(s)
-        case Failure(_) => false
+        case Failure(e) => {println(e.getMessage);false}
       }
     }
 
@@ -128,7 +128,7 @@ trait Marius {
     * @param state a state of the model
     * @return true if the simulation is finished
     */
-  def ended(state: State) = state.step >= maxStep
+  def ended(state: State): Boolean = state.step >= maxStep
 
   /**
     * The next state of the model
@@ -145,6 +145,7 @@ trait Marius {
             population = updatedPopulation(city, updatedWealth,configuration.dates(state.step + 1) - configuration.dates(state.step))
           )
       }
+    //println(state.step)
     state.copy(cities = updatedCities, step = state.step + 1)
   }
 
@@ -176,11 +177,17 @@ trait Marius {
     * @return the updated population
     */
   def updatedPopulation(city: City, updatedWealth: Double,deltaT: Double): Double = {
-    assert(updatedWealth >= 0, s"Negative wealth before conversion toPop $updatedWealth")
-    val deltaPopulation = deltaT*(wealthToPopulation(updatedWealth) - wealthToPopulation(city.wealth)) / economicMultiplier
+    //assert(updatedWealth >= 0, s"Negative wealth before conversion toPop $updatedWealth")
+    // negative wealth can not happen (cf updatedWealth function)
+    val wealthpopupdated = wealthToPopulation(updatedWealth)
+    val deltaPopulation = deltaT*( - wealthToPopulation(city.wealth)) / economicMultiplier
     val updatedPopulation = city.population + deltaPopulation
-    assert(updatedPopulation >= 0, s"Negative population $updatedPopulation")
-    updatedPopulation
+    //assert(updatedPopulation >= 0, s"Negative population $updatedPopulation")
+    //this does not make sense, take 0 if negative. Then some conservation equations may not hold ?
+    //assert(!updatedPopulation.isInfinite,s"Infinite pop $updatedPopulation")
+    assert(updatedPopulation<1e9,s"divergent population : updatedWealth $updatedWealth wealthtopop updatedWealth")
+
+    if (updatedPopulation >= 0.0) updatedPopulation else 0.0
   }
 
   /**
@@ -189,7 +196,11 @@ trait Marius {
     * @param population the population the city
     * @return the supply
     */
[…]

### 2018-09-21 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/TestModel.scala

@@ -6,15 +6,15 @@ import java.io.File
 object TestModel {
 
   def testModel: SimpleModel = SimpleModel(
-    new File("data/processed/FR_dist.csv"),
-    new File( "data/processed/FR_pops.csv"),
-    new File("data/processed/FR_dates.csv"),
-     0.34380934416368,
-    1.00175638801509,
-    1.07926078029368,
-    0.672263161474518,
-    0.380435604357353,
-    1.08660127543912
+    new File("data/processed/BR_dist.csv"),
+    new File( "data/processed/BR_pops.csv"),
+    new File("data/processed/BR_dates.csv"),
+    economicMultiplier= 5.9059803,
+    sizeEffectOnSupply =3.60315698509,
+    sizeEffectOnDemand=3.30481631293628,
+    distanceDecay=1.698876633327205,
+    wealthToPopulationExponent=1.18720491738353,
+    populationToWealthExponent=1.1140844912
   )
 
   /*

### 2018-09-21 Models/urbangrowth/src/main/scala/urbangrowth/test/Test.scala

@@ -12,9 +12,9 @@ import scala.collection.mutable.ArrayBuffer
 
 object Test extends App {
 
-  TestModels.testInnovation()
+  //TestModels.testInnovation()
 
-  //TestModels.testMarius()
+  TestModels.testMarius()
 
   //TestModels.testIntGib()
 

### 2018-09-21 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/Marius.scala
test_jav8_ubuntu18.04

@@ -180,12 +180,13 @@ trait Marius {
     //assert(updatedWealth >= 0, s"Negative wealth before conversion toPop $updatedWealth")
     // negative wealth can not happen (cf updatedWealth function)
     val wealthpopupdated = wealthToPopulation(updatedWealth)
-    val deltaPopulation = deltaT*( - wealthToPopulation(city.wealth)) / economicMultiplier
+    val wealthpop = wealthToPopulation(city.wealth)
+    val deltaPopulation = deltaT*(wealthpopupdated - wealthpop) / economicMultiplier
     val updatedPopulation = city.population + deltaPopulation
     //assert(updatedPopulation >= 0, s"Negative population $updatedPopulation")
     //this does not make sense, take 0 if negative. Then some conservation equations may not hold ?
     //assert(!updatedPopulation.isInfinite,s"Infinite pop $updatedPopulation")
-    assert(updatedPopulation<1e9,s"divergent population : updatedWealth $updatedWealth wealthtopop updatedWealth")
+    assert(updatedPopulation<1e9,s"divergent population : updatedWealth $updatedWealth wealthtopop $wealthpop wealthpopupdated $wealthpopupdated")
 
     if (updatedPopulation >= 0.0) updatedPopulation else 0.0
   }

### 2018-09-23 Models/StylizedFacts/stylized.R
correlation_plots

@@ -127,82 +127,101 @@ table(data.frame(distrs,systems,years))
 ######
 # correlation of growth rates as a function of distance - in time
 
-#countrycode = 'US'
+#countrycode = 'ZA'
+countrycodes = c('EU','BR','CN','IN','RU','ZA','US')
+#countrycodes = c('BR','CN','IN','RU','ZA','US')
 
 # hand def colsnames (different number and names for europe)
-popcolnames=list('EU'=c("Pop_1961","Pop_1971","Pop_1981","Pop_1991","Pop_2001","Pop_2011"),
+popcolnames=list('EU'=c("X1961","X1971","X1981","X1991","X2001","X2011"),
                  'US'=c("X1960","X1970","X1980","X1990","X2000","X2010"),
-                 'BR'=c("X1960","X1970","X1980","X1990","X2000","X2010"),
-                 'CN'=c("X1964","X1982","X1990","X2000"),
                  'BR'=c("X1960","X1970","X1980","X1991","X2000","X2010"),
+                 'CN'=c("X1964","X1982","X1990","X2000"),
+                 'IN'=c("X1951","X1961","X1981","X1991","X2001","X2011"),
                  'RU'=c("X1959","X1970","X1979","X1989","X2002","X2010"),
-                 'ZA'=c("X1959","X1970","X1979","X1989","X2002","X2010")
+                 'ZA'=c("X1960","X1970","X1980","X1991","X1996","X2001")
                  )
+
+corresdir=paste0(resdir,'corrs/');dir.create(corresdir)
+
+generatePeriods<-function(cols){
+  res=list()
+  for(l in 1:length(cols)){
+    currentcols=list()
+    for(i in 1:(length(cols)-l+1)){
+      currentcols[[substring(cols[(i+l-1)],2)]]=cols[i:(i+l-1)]
+    }
+    res[[as.character(l)]]=currentcols
+  }
+  return(res)
+}
 
 for(countrycode in countrycodes){
   show(countrycode)
   if(countrycode!='EU'){
-     cities <- read.csv(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Data/Geodivercity/data/',cityfiles[[countrycode]],'.csv'))
+    cities <- read.csv(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Data/Geodivercity/data/',cityfiles[[countrycode]],'.csv'))
   }else{
[…]

### 2018-09-23 Docs/CCS18/Communication/ccs2018.tex
v1 slides

@@
 Thessaloniki\\\smallskip
-September 25th 2018
+September 24th 2018
 }
@@
 
-\sframe{Urban systems}{
+\sframe{The long time memory of urban systems}{
 
@@
 
-	%\includegraphics[width=\textwidth]{figures/}
+	\includegraphics[width=0.52\textwidth]{figures/rotundagalerius.jpg}
+	\hspace{0.2cm}
+	\includegraphics[width=0.42\textwidth]{figures/whitetower.png}
 
 	\footnotesize
-\textit{Source: }
+\textit{Source: Wikipedia}
 
@@
 
-% meso / micro
+% meso / micro - macro
 
-% macro
+\begin{columns}
+	\begin{column}{0.5\textwidth}
+		Macroscopic urban system
+		\medskip
+		%\centering
+		\includegraphics[width=\textwidth]{figures/macrogrowth.png}
+		\footnotesize
+		\justify
+		Raimbault, J. (2018). Indirect evidence of network effects in a system of cities. \textit{Environment and Planning B: Urban Analytics and City Science}, 2399808318774335.
+		\nocite{raimbault2018indirect}
+	\end{column}
+	\begin{column}{0.5\textwidth}
+		Mesoscopic territorial systems
+		\medskip
+		%\centering
+		\includegraphics[width=\textwidth]{figures/mesogrowth.png}
+
+		\footnotesize
+		\justify
+		Raimbault, J. (2018). Calibration of a density-based model of urban morphogenesis. PloS one, 13(9), e0203516.
+		\nocite{raimbault2018calibration}
+	\end{column}
+\end{columns}
 
@@
 \sframe{An evolutionary urban theory}{
+
+From \cite{pumain1997pour} to \cite{pumain2018evolutionary}: systems of cities as co-evolutive systems in which interactions are crucial
+
 \centering
+
+\includegraphics[height=0.7\textheight]{figures/evolurbantheory.pdf}
 
@@
 
-\centering
+\justify
+
+$\rightarrow$ several models in this context have been introduced, but never compared (e.g. in terms of explanative power)
+
+\medskip
+
+$\rightarrow$ multidimensionality of urban systems and the potential complementarity between very different processes
 
@@
 
-\textbf{Research objective : }
+\bigskip
+\bigskip
+
+\textbf{Research objective : } \textit{Benchmark interaction models for systems of cities developed in the frame of the evolutionary urban theory, on several comparable systems of cities.}
 
@@
 
-\centering
+%\centering
+\justify
 
 Comparison of three approaches based on the evolutionary urban theory \cite{pumain1997pour} capturing different dimensions of urban systems:
+
+\bigskip
+
 \begin{itemize}
@@
 
-\sframe{Network interaction model}{
+\sframe{Description of models}{
 
@@
 
-}
+\begin{columns}
+	\begin{column}{0.33\textwidth}
+		\textbf{Network interaction model}
 
-\sframe{The Favaro-Pumain model}{
+		\medskip
+
+		\begin{itemize}
+			\item Endogenous growth
+			\item Interactions inducing growth through gravity potential
+			\item Static physical network taken into account (geographical shortest path with topography)
+		\end{itemize}
+
+	\end{column}
+	\vrule{}
+	\begin{column}{0.33\textwidth}
+		%\vspace{-1cm}
+		\textbf{Favaro-Pumain model}
+
+		\medskip
+
+		\begin{itemize}
+			\item Endogenous growth
+			\item Innovation emerge and diffuse in cities
+			\item Growth rates adapted according to utility of innovation and level of adaptation
+		\end{itemize}
+
+	\end{column}
+	\vrule{}
+	\begin{column}{0.33\textwidth}
+		\textbf{Marius model}
+
+		\medskip
+
+		\begin{itemize}
+			\item Cities produce economic goods
+			\item Economic exchanges are estimated according to gravity flows
+			\item Populations grow depending on final economic balances
+		\end{itemize}
+
+	\end{column}
+\end{columns}
 
@@
 
-\centering
+%\centering
 
@@
 
-Harmonized dataset, explored in particular by \cite{pumain2015multilevel} (built by
+\justify
 
-}
+\textbf{Dataset}
 
-\sframe{Data preprocessing}{
-\centering
+Harmonized dataset, explored in particular by \cite{pumain2015multilevel} (built in the context of Geodivercity ERC)
 
-Remove small cities (\cite{adam2006medium} for definition of medium-sized)
+\bigskip
+
+\textbf{Data preprocessing}
+
+\begin{itemize}
+	\item Remove small cities and medium-sized cities (\cite{adam2006medium} for definition of medium-sized) for scalability of simulations
+	\item Shortest paths matrices (direct and taking into account topography, using the EEA world DEM) computed a priori and cached (\textit{no evolution of the network})
+\end{itemize}
 
@@
 \sframe{Stylized facts}{
-\centering
+%\centering
 
@@
 % - cluster by type of distrib ?
+
+\textit{Population distributions mainly log-normal (75\% with KS-test, all with AIC); consistent hierarchies in time.}
+
+\medskip
+
+\includegraphics[width=0.49\textwidth]{figures/popdistribs.png}
+\includegraphics[width=0.49\textwidth]{figures/popranksize.png}
+
+}
+
+\sframe{Stylized facts}{
+
+\textit{Growth rates distributions evolve considerably in time}
+
+\medskip
+
+\includegraphics[width=\textwidth]{figures/growthrates_byyear.png}
+
+}
+
+\sframe{Stylized facts}{
+
+\textit{Correlation as a function of distance: long-range correlations in several systems}
+
+\medskip
+
+\includegraphics[width=0.49\textwidth]{figures/corrsdist_US_quantiles21_periods5.png}
+\includegraphics[width=0.49\textwidth]{figures/corrsdist_EU_quantiles21_periods5.png}
+
+%\smallskip
+
+%$\rightarrow$ \textit{non-linear models good candidates}
 
@@
 
-Computationally intensive: high-dimensional parameter space and possible spatial setup.
-
-\medskip
-
-$\rightarrow$ use of grid computing, made smooth with the OpenMOLE software \url{https://next.openmole.org/}
+Bi-objective calibration of 7 models for 7 city systems $\rightarrow$ use of genetic algorithms on grid, made smooth with the OpenMOLE software \url{https://next.openmole.org/}
 
@@
 
-}
+\bigskip
 
-\sframe{Model calibration}{
-
-% number of runs (equivalent / year) ; caracs of NSGA2 ; convergence for each model/system
+\textbf{Come to the Satellite on Wednesday, and apply to the summer school !} 
+\\(\texttt{https://exmodelo.org/})
 
@@
 
-\sframe{Results: comparison of models on the different systems}{
+\sframe{Comparison of models on all systems}{
 
@@
 
+\centering
+
+\includegraphics[width=\textwidth]{figures/allmodels_allsystems.png}
+
+\footnotesize
+\textit{Systems driven by economic exchanges: EU, US, RU, ZA}
+
 }
 
-\sframe{Results: particular cases}{
+\sframe{Indian urban system: direct interactions}{
 
 % systems on which the behavior is particularly interesting ?
+
+\centering
+\includegraphics[width=0.9\textwidth]{figures/IN_zoomed.png}
+
+}
+
+\sframe{Brazilian urban system: multiple factors}{
+
+\centering
+\includegraphics[width=0.9\textwidth,height=0.8\textheight]{figures/BR_zoomed.png}
+
+\footnotesize
+\textit{Importance of topography; innovation processes mostly.}
+
+}
+
+\sframe{China: a tight competition}{
+
+\centering
+\includegraphics[width=0.9\textwidth,height=0.8\textheight]{figures/CN_zoomed.png}
+
+\footnotesize
+\textit{No clear best model: other processes in play ? (strong top-down planning)}
 
@@
 
+\textit{Comparing Gibrat and Network model by correcting for the number of parameters \cite{raimbault2018indirect}: additional parameters actually improve the fit (but not always)}
+
+\bigskip
+
+\centering
+
+% gib - intgib
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+System & $\Delta AIC$ (log) & $\Delta BIC$ (log) & $\Delta AIC$ (mse) & $\Delta BIC$ (mse) \\\hline
+ZA & 33.82658 & 29.40496 &  -838.07184 &  -842.49347\\
+CN & 20.67713 & 16.78273 & 13.40767 & 9.51327\\
+BR & -263.27003 & -267.71575 & -407.55220 & -411.99792\\
+IN & -85.01949 & -89.24206 & -188.16703 & -192.38960\\
+RU & 50.60535 & 46.24233 & 104.94806 &  100.58503\\\hline
+\end{tabular}
+
+% marius - mariusrestr
+
 }
@@
 
-$\rightarrow$
+$\rightarrow$ Complementarity of economic, innovation and direct interaction processes
 
@@
 
-$\rightarrow$
+$\rightarrow$ High dimensionality of urban systems
 
@@
 % Compare these models on synthetic city systems / with suited indicators (here relatively data-driven approach)
+% - nonstat in time for some city systems ?
+% - calib at the second order on correlation curves (-> new methodo approach - link with correlated synthetic data !) (many things are connected)
+% saturation of models ?
+% Rq : issue in calib algos (restricted vs full models : strange behavior ex the bifurcation in intgib % gibrat
 
-$\rightarrow$
+$\rightarrow$ Still points on robustness to be investigated: influence of stochasticity, convergence of GA, saturation of parameters, number of cities.
 
@@
 
-$\rightarrow$
+$\rightarrow$ Towards integrated models coupling these different components ?
 
@@
 
-$\rightarrow$ More elaborated method to compare models in a ``fair'' way (correcting for additional parameters, open question for models of simulation).
+$\rightarrow$ Compare on synthetic systems of cities with appropriate indicators \cite{raimbault2018unveiling}.
+
+\smallskip
+
+$\rightarrow$ Calibrate at the second order (correlations), non-stationary in time.
+
+\smallskip
+
+$\rightarrow$ More elaborated method to compare models in a ``fair'' way (correcting for additional parameters, open question for models of simulation): model reduction ?
 
@@
 
-$\rightarrow$ % model coupling / benchmarking
+$\rightarrow$ First step towards systematic benchmarks and multi-modeling. \textbf{Need for more systematic model exploration.} % model coupling / benchmarking
 
-\medskip
+\smallskip
+
+$\rightarrow$ Model integration and multi-scalarity ? \textbf{Need for more integrated models.}
+
+\smallskip
 
@@
 
-\medskip
+\smallskip
 
@@
 
-\medskip
+\smallskip
 
-\textbf{Open repository} (code, data and results) at\\\texttt{https://github.com/JusteRaimbault/UrbanGrowth}\\\medskip
-\textbf{Acknowledgments} : I thank the \textit{European Grid Infrastructure} and its \textit{National Grid Initiatives} (\textit{France-Grilles} in particular) to give the technical support and the infrastructure.
+\textbf{Open repository} at \texttt{https://github.com/JusteRaimbault/UrbanGrowth}\\\smallskip
+\textbf{Acknowledgments}: thanks to the \textit{EGI} for access to the infrastructure.
 
@@
 
+\sframe{Model calibration}{
+
+% number of runs (equivalent / year) ; caracs of NSGA2 ; convergence for each model/system
+
+Genetic algorithm NSGA2:
+
+\bigskip
+
+\begin{itemize}
+	\item Around 40000 generations (4e6 model runs)
+	\item Population $\mu = 100$
+	\item Convergence tested with hypervolume relative variation
+\end{itemize}
+
+}
+
 %%%%%%%%%%%%%%%%%%%%%

### 2018-09-23 Models/StylizedFacts/stylized.R
calib_results

@@ -227,6 +227,7 @@ for(countrycode in countrycodes){
 
 
 ##### Summary stats (pop in time)
+# -> ok done in geodivercity papers
 
 ## developments : - nonstat in time for some city systems ?
 # - calib at the second order on correlation curves (-> new methodo approach - link with correlated synthetic data !)

### 2018-09-23 Models/urbangrowth/src/main/scala/urbangrowth/utils/io/FileUtils.scala

@@ -1,7 +1,8 @@
 
 package urbangrowth.utils.io
 
-import java.io.{BufferedReader, File, FileReader}
+import java.io._
+
 import Jama.Matrix
 
 
@@ -22,7 +23,7 @@ object FileUtils {
     res.toArray
   }
 
-  def parseSimple(f: File) = {
+  def parseSimple(f: File): Array[Double] = {
     val reader = new BufferedReader(new FileReader(f))
     var currentLine = reader.readLine()
     var res = List(currentLine.toDouble)
@@ -31,6 +32,15 @@ object FileUtils {
       if (currentLine != null) res = res :+ currentLine.toDouble
     }
     res.toArray
+  }
+
+  def exportCSV(f: File,m: Matrix,delimiter: String = ","): Unit ={
+    val writer = new BufferedWriter(new FileWriter(f))
+    m.getArray.foreach{
+      case row =>
+        writer.write(row.map{case d => String.format("%f",d.asInstanceOf[AnyRef])}.mkString(delimiter));writer.newLine()
+    }
+    writer.close()
   }
 
 }

### 2018-10-02 Models/urbangrowth/src/main/scala/urbangrowth/utils/io/FileUtils.scala
.

@@ -3,6 +3,7 @@ package urbangrowth.utils.io
 
 import java.io.{BufferedReader, File, FileReader}
 import Jama.Matrix
+import scala.io._
 
 
 object FileUtils {
@@ -12,14 +13,16 @@ object FileUtils {
   }
 
   def parseCSV(f: File, delimiter: String): Array[Array[Double]] = {
-    val reader = new BufferedReader(new FileReader(f))
+    Source.fromFile(f).getLines.map{_.split(delimiter).map{_.toDouble}.toArray}.toArray
+
+    /*val reader = new BufferedReader(new FileReader(f))
     var currentLine = reader.readLine()
     var res = List((currentLine.split(delimiter).map { s => s.toDouble })) //= new List[Array[Double]] {}
     while (currentLine != null) {
       currentLine = reader.readLine()
       if (currentLine != null) res = res :+ (currentLine.split(delimiter).map { s => s.toDouble }) //+: res
     }
-    res.toArray
+    res.toArray*/
   }
 
   def parseSimple(f: File) = {

### 2018-10-02 Docs/CCS18/Communication/ccs2018.tex

@@
 \date{CCS 2018\\\smallskip
+\textit{Session Infrastructure, Planning and Environment}\\\smallskip
 Thessaloniki\\\smallskip
@@
 
-% Understanding patterns of growth for cities is a crucial issue for its practical and policy implications e.g.
-regarding sustainability, but also from a theoretical point of view regarding the validation of theories for urban systems.
-A particular entry is taken by the Evolutive Urban Theory [1] which postulates interactions between cities as main drivers of their growth.
-Within this framework, numerous concurrent models have been introduced for the evolution of systems of cities, and applied on diverse case studies, but there is to the best of our knowledge no systematic comparison of performances across models and application cases.
-This contribution makes a first step towards such a systematic benchmark.
-We consider simple models simulating population of cities only, but including heterogenous underlying processes and assumptions.
-More precisely, we compare (i) the Favaro-Pumain model for the diffusion of innovation [2]; (ii) the Marius multi-model based on economic exchanges [3]; and (iii) a model including flows within abstract physical networks parametrized on elevation data [4].
-These models are calibrated on a large scale harmonized dataset [5] including the European, former Soviet Union, Chinese, Brazilian, South-African, Indian and USA systems of cities on a time period covering 1960-2010.
-Calibrations are done for different versions of each model including more or less parameters, using distributed genetic algorithms on a computation grid through the intermediary of the OpenMOLE model exploration software [6].
-Calibration results show that: (i) at a fixed number of parameters, no model performs particularly better on all urban systems, suggesting the complementary of the innovation, economic, and network dimensions taken into account by the different models; and (ii) when adjusting for the number of parameters with an empirical information criterion [4], we find that additional components generally improve performances for all models, what reveals a high effective dimensionality of urban systems.
-This work confirms the need for multiple complementary approaches to model urban systems and sketches a framework allowing systematic benchmarks of concurrent models for complex urban systems.
+\sframe{The long time memory of urban systems}{
 
-\sframe{The long time memory of urban systems}{
+% Working in geography has this advantage that we always situate in an interesting object of study, and reflexivity is therin inevitable.
+% I have taken the habit to open my talks with some local context, and this one will not fail this rule, especially because of the beauty and hostory of Thessaloniki.
+% Urban systems are fascinating as they are truly multi-scalar, both in time and space.
+In particular, they have a very long temporcal memory, partly because of the pphysical artifats : here the Rotunda Galerius which was later a chruch and a Mosque, or the white tower constructed by the ottoman empire ; but also because of non-ergodic dynamics.
+% In this geo-hostorical context, understanding urban growth is a real challenge : to what extent the different relation of a given city influence its current growth and its future trajectory ?
 
@@
 \sframe{Modeling urban growth}{
+
+% We will switch to shorter time scales, but still remain at macroscopic and mesoscopic scales.
+
+% Understanding patterns of growth for cities is a crucial issue for its practical and policy implications e.g.
+regarding sustainability, but also from a theoretical point of view regarding the validation of theories for urban systems.
 
@@
 
+% A particular entry is taken by the Evolutive Urban Theory [1] which postulates interactions between cities as main drivers of their growth.
+
 From \cite{pumain1997pour} to \cite{pumain2018evolutionary}: systems of cities as co-evolutive systems in which interactions are crucial
@@
 \sframe{Towards a systematic model comparison}{
+
+% Within this framework, numerous concurrent models have been introduced for the evolution of systems of cities, and applied on diverse case studies, but there is to the best of our knowledge no systematic comparison of performances across models and application cases.
+This contribution makes a first step towards such a systematic benchmark.
 
@@
 \sframe{Urban systems interaction models}{
+
+%  We consider simple models simulating population of cities only, but including heterogenous underlying processes and assumptions.
+More precisely, we compare (i) the Favaro-Pumain model for the diffusion of innovation [2]; (ii) the Marius multi-model based on economic exchanges [3]; and (iii) a model including flows within abstract physical networks parametrized on elevation data [4].
 
@@
 
+% These models are calibrated on a large scale harmonized dataset [5] including the European, former Soviet Union, Chinese, Brazilian, South-African, Indian and USA systems of cities on a time period covering 1960-2010.
+
 %\centering
@@
 
-Harmonized dataset, explored in particular by \cite{pumain2015multilevel} (built in the context of Geodivercity ERC)
+Harmonized dataset, explored in particular by \cite{pumain2015multilevel} (built in the context of Geodivercity ERC): Urban systems for Europe, United States, Brazil, China, India, South Africa, Russia, consistent and comparable between 1960 and 2010 each 10 years.
 
@@
 
+% Calibrations are done for different versions of each model including more or less parameters, using distributed genetic algorithms on a computation grid through the intermediary of the OpenMOLE model exploration software [6].
+
 Bi-objective calibration of 7 models for 7 city systems $\rightarrow$ use of genetic algorithms on grid, made smooth with the OpenMOLE software \url{https://next.openmole.org/}
@@
 \sframe{Comparison of models on all systems}{
+
+%Calibration results show that: (i) at a fixed number of parameters, no model performs particularly better on all urban systems, suggesting the complementary of the innovation, economic, and network dimensions taken into account by the different models.
 
@@
 \sframe{Results: empirical AIC}{
-% TODO : select optimal points / run empirical AIC
+
+% (ii) when adjusting for the number of parameters with an empirical information criterion [4], we find that additional components generally improve performances for all models, what reveals a high effective dimensionality of urban systems.
+
+% - select optimal points / run empirical AIC
 % OR : run on all points of the Pareto front ? -> local OK
@@
 \sframe{Discussion}{
+
+% This work confirms the need for multiple complementary approaches to model urban systems and sketches a framework allowing systematic benchmarks of concurrent models for complex urban systems.
 
@@
 
+\sframe{Models settings}{
+
+$\rightarrow$ Work under Gibrat independence assumptions, i.e.
+$\Covb{P_i(t)}{P_j(t)}=0$.
+If $\vec{P}(t+1)=\mathbf{R}\cdot \vec{P}(t)$ where $\mathbf{R}$ is also independent, then $\Eb{\vec{P}(t+1)}=\Eb{\mathbf{R}}\cdot\Eb{\vec{P}}(t)$.
+Consider expectancies only (higher moments computable similarly)
+
+\medskip
+
+$\rightarrow$ With $\vec{\mu}(t)=\Eb{\vec{P}(t)}$, we generalize this approach by taking $\vec{\mu}(t+1)=f(\vec{\mu}(t))$
+
+}
+
+\sframe{Network model}{
+
+Direct network interaction model \cite{raimbault2018indirect}:
+
+\medskip
+
+Let $\vec{\mu}(t)=\Eb{\vec{P}(t)}$ cities population and $(d_{ij})$ distance matrix
+
+\medskip
+
+Model specified by
+
+\[
+f(\vec{\mu}) = r_0\cdot \mathbf{Id}\cdot \vec{\mu} + \mathbf{G}\cdot \mathbf{1} + \mathbf{N}
+\]
+
+ with 
+\begin{itemize}
+\item $G_{ij} = w_G\cdot \frac{V_{ij}}{<V_{ij}>}$ and $V_{ij} = \left(\frac{\mu_i\mu_j}{\sum{\mu_k}^2}\right)^{\gamma_G} \exp{(-d_{ij}/d_G)}$
+\item $N_{i} = w_N \cdot \sum_{kl} \left(\frac{\mu_k\mu_l}{\sum\mu}\right)^{\gamma_N}\exp{(-d_{kl,i})/d_N}$ where $d_{kl,i}$ is distance to shortest path between $k,l$ computed with slope impedance ($Z=\left(1+\alpha/\alpha_0\right)^{n_0}$ with $\alpha_0\simeq 3$)
+\end{itemize}
+
+}
+
+\sframe{Innovation diffusion}{
+
+Favaro-Pumain model \cite{favaro2011gibrat}:
+
+\medskip
+
+1) Diffuse innovations according to 
+\[
+\delta_{c,i,t} = \frac{\sum_j p_{c,j,t-1}^{s_c} \exp{(-\lambda_s d_{ij})}}{\sum_c \sum_j p_{c,j,t-1}^{s_c} \exp{(-\lambda_s d_{ij})}}
+\]
+
+\medskip
+
+2) Update population with $G_{ij}$ (see network model) such that
+\[
+V_{ij}= \frac{p_i p_j}{(\sum_k p_k)^2} \exp{(-\lambda_m d_{ij} \prod_c \delta_{c,i}^{\phi_c})}
+\]
+with $\phi_c = \sum_i p_{i,c}/\sum_{i,c} p_{i,c}$
+
+\medskip
+
+3) Introduce innovation with utility $s_{c+1} = g_0 \cdot s_c$ in a randomly chosen city with a hierarchy parameter $\alpha_I$, if global adoption share $\phi_c$ is larger than a threshold $\theta_I$.
+Initial utility $s_0$ is a parameter.
+New innovation has an initial penetration rate $r_I$ in the city.
+
+}
+
+\sframe{Economic exchanges}{
+
+Marius model \cite{10.1371/journal.pone.0138212}:
+
+\medskip
+
+Initial wealth as a power law of population (exponent $\alpha_W$)
+
+\medskip
+
+1) Update supply and demands as superlinear functions of population (exponents $\alpha_S,\alpha_D$)
+
+\medskip
+
+2) Exchange goods according to a gravity potential of interaction (distance decay $d_M$), supplies and demands; update wealth accordingly
+
+\medskip
+
+3) Update population such that population difference is a power law of wealth difference (economic multiplier $e_M$ and exponent $\alpha_P$)
+
+}
+
+\sframe{Benchmarked models}{
+
+\begin{enumerate}
+	\item Gibrat model: 1 param. $r_0$
+	\item Direct interaction model (geographical distance): 4 param. $r_0,w_G,\gamma_G,d_G$
+	\item Physical network interaction model (topographical distance: 4 param. $r_0,w_G,\gamma_G,d_G$
+	\item Innovation diffusion model (simplified): 4 param. $r_0,w_I,\lambda_s,\lambda_m$ (other parameters at default values from \cite{favaro2011gibrat})
+	\item Innovation diffusion model (full): 9 param. $r_0,w_I,\lambda_s,\lambda_m,s_0,g_0,r_I,\alpha_I,\theta_I$
+	\item Restricted Marius model: 4 param. $e_M,\alpha_S,\alpha_D,d_M$
+	\item Marius model: 6 param. $e_M,\alpha_S,\alpha_D,d_M,\alpha_W,\alpha_P$
+\end{enumerate}
+
+}
+
 \sframe{Model calibration}{

### 2019-02-12 Docs/Communications/XTerM2019/extendedAbstract.tex
skeleton_abstract

\documentclass[a4paper,twocolumn,twoside,10pt]{article}
\usepackage{graphicx,amsmath,amssymb,amsthm}

\makeatletter \oddsidemargin-.25in \evensidemargin-.25in
\makeatother \topmargin-0.75in \textwidth7in \textheight9.5in

% Place your Latex macros here -----------------------------
% THEOREMS -------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}{Corollary}[section]
\newtheorem{prob}{Problem}[section]
\newtheorem{lem}{Lemma}[section]
\newtheorem{prop}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{rem}{Remark}[section]
\newtheorem{ex}{Example}[section]

% MATH -----------------------------------------------------------
\newcommand{\norm}[1]{\left\Vert#1\right\Vert}
\newcommand{\trnorm}[1]{\left\Vert#1\right\Vert_{\mathbf{tr}}}
\newcommand{\abs}[1]{\left\vert#1\right\vert}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\Real}{\mathbb R}
\newcommand{\eps}{\varepsilon}
\newcommand{\To}{\longrightarrow}
\newcommand{\BX}{\mathbf{B}(X)}
\newcommand{\tr}[1]{\text{tr}\left[#1\right]}
\newcommand{\maxeig}[1]{\mathbf{\lambda_{\max}}\left(#1\right)}
\newcommand{\mineig}[1]{\mathbf{\lambda_{\min}}\left(#1\right)}
\newcommand{\e}[1]{\text{E}\left[#1\right]}
\newcommand{\diag}[1]{\text{diag}\left\{#1\right\}}
\newcommand{\A}{\mathcal{A}}

% Running Head----------------------------------------------------------------
\markboth{$~$ \hfill {\rm Author(s) Name} \hfill $~$} {$~$ \hfill
{\rm Short Title of the Paper} \hfill$~$}

\pagestyle{myheadings} 
%\setcounter{page}{151}

\begin{document}
\def \thepage {}
\date{}

% Title----------------------------------------------------------------
\title{\begin{flushleft}
\noindent {\small {\it Proceedings XTerM 2019
\\Normandie Univ - Le Havre, France - June 26-28, 2019\\[5mm]
 }}
% \noindent {\small {\it Journal of Nonlinear Systems and Applications
% (2009)
% 151--154\\
%  Copyright $\copyright$ 2009 Watam Press \hfill http://www.watam.org/JNSA/\\[5.0mm]}}
\end{flushleft}\Large\bf \uppercase{A multi-scale model of urban growth coupling dynamics of systems of cities with morphological processes} }
%\author{\Large\bf Author Index}
\author{J. Raimbault
 \thanks{J. Raimbault is with UPS CNRS 3611 ISC-PIF, France; CASA, UCL, UK; and UMR CNRS 8504 G{\'e}ographie-cit{\'e}s. E-mail:
juste.raimbault@polytechnique.edu}
%\thanks{Manuscript received April 19, 2009; revised January 11, 2010.}
}
 \maketitle

% Abstract----------------------------------------------------------------

{\footnotesize \noindent {\bf Abstract.} We introduce a multi-scale model of urban growth. Focusing on the spatial structure of processes rather than on their intrinsic multi-dimensionality, we take into account population only and couple a macroscopic interaction model for a system of cities with urban morphogenesis models simulating the spatial distribution of population in metropolitan areas based on reaction-diffusion processes. Different coupling regimes between scales are explored on synthetic systems of cities, and the model is partly calibrated for a weak coupling with a consistent database for the European urban system. This work paves the way towards more complex and operational multi-scale models of urban growth.\\
{\bf Keywords.} Urban growth; Multi-scale; Urban morphology; Systems of cities; Model calibration.}

\vskip.2in

% Contents----------------------------------------------------------------

% ----------------------------------------------------------------
\section{Introduction}

The modeling of urban growth is a crucial issue for the design of sustainable territorial policies, through the understanding of past urbanization processes and the forecasting of future urban trajectories. Models have been proposed at different scales and integrating different dimensions of urban systems. At the scale of a metropolitan area, Land-use Transport Interaction models \cite{wegener2004land} are for example a privileged tool to anticipate the answer of spatial distributions of activities (mostly residential location and economic activities) to an evolution of the accessibility landscape permitted by new transportation infrastructures. At the same scale, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}. At the smaller scale of the system of cities, macroscopic models of urban growth have focused on reproducing the distribution of city sizes, either through economic processes as e.g. \cite{gabaix1999zipf}, or from a geographical point of view focusing on interactions between cities \cite{favaro2011gibrat}.

Territorial dynamics, and more particularly urban dynamics, have according to \cite{pumain1997pour} an intrinsic multi-scalar nature, with successive autonomous levels of emergence from individual microscopic agents to the mesoscopic scale of the city and the macroscopic scale of the system of cities. Furthermore, the need for sustainable territorial policies would imply the construction of multi-scalar models to take into account issues associated to each relevant scale \cite{Rozenblat2018}.

This contribution contributes to that open question by introducing a multi-scale model of urban growth which focuses on the spatial structure of processes rather than on their multi-dimensionality. Therefore, we take into account only population variables, but both at the macroscopic scale of the system of cities in the legacy of \cite{pumain2017urban} and at the mesoscopic scale of the metropolitan area with an urban morphogenesis model. The coupling of these scales is a crucial novel feature of our model. We describe in the following stylized facts justifying the approach, describe the model, and summarize preliminary results from its exploration and calibration.

\section{Empirical stylized facts}

We use the dataset of urban systems dynamics on long time provided by \cite{pumain2015multilevel}, and in particular for the European urban system spanning 1850-2000. Regarding the precise distribution of population, we use the static Eurostat population grid in 2010 aggregated at a 500m resolution. Empirical data analysis provide the two following crucial stylized facts: (i) the spatial distribution of morphological indicators computed on spatial moving windows of size 50km has a high spatial non-stationarity, recovering the results obtained by \cite{raimbault2018urban} for the interaction between population distribution and the road network; (ii) population growth rates of urban areas, shown in Fig.~\ref{fig:empcorrs} as a function of distance, exhibit as expected short-range positive correlations but also negative long-range correlations. The first point justifies differentiating mesoscopic models of morphogenesis in space, and the second to take into account the interactions between these urban areas at the macroscopic scale.

\begin{figure}[htp]
\centering
\includegraphics[width=0.48\textwidth]{figures/corrsdist_EU_quantiles21_periods5.png}
\caption{Growth rate correlations as a function of distance between urban areas, for the European urban system.}
\label{fig:empcorrs}
\end{figure}

\section{Multi-scalar model}

\cite{raimbault2018indirect}
\cite{raimbault2018calibration}

\begin{figure}[htp]
\centering
\includegraphics[width=0.48\textwidth]{figures/multiscale_morph.pdf}
\caption{Theoretical flowchart of the multi-scalar model.}
\label{fig:model}
\end{figure}

\cite{reuillon2013openmole}

\section{Discussion}

%\section*{Acknowledgements}
%The research for this work was supported, in part, by the Natural
%Sciences and Engineering Research Council of Canada.

\footnotesize

% Bibliography----------------------------------------------------------------

\bibliographystyle{unsrt}
\bibliography{biblio.bib}

\end{document}
% ----------------------------------------------------------------

### 2019-02-12 Models/Analysis/profile.R

@@ -13,12 +13,13 @@ sourcedir = 'PROFILE_GRID_intgib_BR_20181221_103649/'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Calibration/',sourcedir);dir.create(resdir)
 
 
-res=as.tbl(read.csv(file=paste0(sourcedir,'population6899.csv')))
+#res=as.tbl(read.csv(file=paste0(sourcedir,'population6899.csv')))
+res=as.tbl(read.csv(file=paste0(sourcedir,'population20000.csv')))
 
 # g=ggplot(res[res$gravityDecay<=1000,],aes(x=gravityDecay,y=logmse))
 g=ggplot(res,aes(x=gravityDecay,y=logmse))
 g+geom_point()+geom_line()+stdtheme
-ggsave(file=paste0(resdir,'profile_logmse-gravityDecay.png'),width=15,height = 10,units='cm')
+ggsave(file=paste0(resdir,'profile_logmse-gravityDecay_gen20000.png'),width=15,height = 10,units='cm')
 
 # Q : 
 #  - more precise profile in the [0,100] interval ? -> relaunch more precise

### 2019-04-12 Docs/Communications/ignored/XTerM2019/extendedAbstract.tex
abstract_ectqg

--- 2019-02-12 Docs/Communications/XTerM2019/extendedAbstract.tex
@@
 
-\cite{raimbault2018indirect}
-\cite{raimbault2018calibration}
+The generic multi-scalar model is based on a coupling of several instances of local urban morphogenesis models, and more particularly the reaction-diffusion model studied by \cite{raimbault2018calibration}, through a macroscopic model of interaction between these areas in the spirit of \cite{raimbault2018indirect}.
+It is therefore able to (i) integrate long range interactions in systems of cities and (ii) simulate the local urban form.
+
+We assume model ontologies similar to the original ones and operate the coupling through a potential dependency of parameters between scales.
+Let the index $\mu$ denote the macro scale and $M$ the meso scale.
+Then we assume the parameters of reaction-diffusion instances to depend on time and the macro-scale: $(\alpha,\beta,N_G/P_m)\left[t,\mu\right]$ and similarly for the parameters of the interaction model $(g_0,\vec{gamma},d_G)\left[t,M\right]$.
+% TODO Note : here a clarification is needed : shouldnt the macro ontology be a bit modified to include short range and long range interactions ? then cannot be fully generic ?
+
+We are interested in the following coupling specifications: weak coupling $\mu \rightarrow M$ (no upward causation, study of the consequence on local form of global dynamics), weak coupling $M \rightarrow \mu$ (no downward causation, aggregation of macro trajectories based on macro dynamics to e.g.
+study their performance compared to the model alone) and strong coupling $M \leftrightarrow \mu$ which can implemented in different ways, but which implies upward and downward causations.
+The comparison between these coupling is in itself an important knowledge.
 
@@
 
-\cite{reuillon2013openmole}
+The model is implemented in scala for performance purposes and an easier integration into the model exploration software OpenMOLE \cite{reuillon2013openmole}.
 

### 2019-04-30 Docs/Communications/2019/CCS2019/ccs2019.tex
explo_res

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%     LaTeX template for CCS 2019     %%
%%        abstract submissions         %%
%%                                     %%
%%          <30 March 2019>            %%
%%                                     %%
%% Author: Liu Wenyuan                 %%
%% Eamil: liuw0037@ntu.edu.sg          %%
%% Please contact author if you have   %%
%% any problem using this template     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[11pt, a4paper]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[top=3.17cm, bottom=2.54cm, left=2.54cm, right=2.54cm, includehead, includefoot,]{geometry}
\usepackage{authblk}
\renewcommand{\refname}{References}
%Only use following two lines if you want to use Times New Roman font
%Only work with XeLaTex or LuaLaTex
%\usepackage{fontspec}
%\setmainfont{Times New Roman}

\begin{document}
\title{\vspace{-3.0cm}\large\textbf{A multi-scalar model for system of cities}}

\author[1,2,3]{\footnotesize Juste Raimbault}
\affil[1]{\footnotesize UPS CNRS 3611 ISC-PIF, Paris, France, \url{juste.raimbault@polytechnique.edu}}
\affil[2]{\footnotesize CASA, UCL, London, UK}
\affil[3]{\footnotesize UMR CNRS 8504 Géographie-cités, Paris, France}
\date{\vspace{-5ex}}

\maketitle
\thispagestyle{empty}

\noindent

The modeling of urban growth is a crucial issue for the design of sustainable territorial policies, through the understanding of past urbanization processes and the forecasting of future urban trajectories. Several models have been proposed at different scales and integrating different dimensions of urban systems, such as land-use transport interaction models \cite{wegener2004land} or systems of cities models \cite{pumain2017urban}. While multi-scalar models are recognized as crucial for the study of such systems \cite{Rozenblat2018}, they remain in practice unexplored.

%At the scale of a metropolitan area, Land-use Transport Interaction models \cite{wegener2004land} are for example a privileged tool to anticipate the answer of spatial distributions of activities (mostly residential location and economic activities) to an evolution of the accessibility landscape permitted by new transportation infrastructures. At the same scale, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}. At the smaller scale of the system of cities, macroscopic models of urban growth have focused on reproducing the distribution of city sizes, either through economic processes as e.g. \cite{gabaix1999zipf}, or from a geographical point of view focusing on interactions between cities \cite{favaro2011gibrat}.
%Territorial dynamics, and more particularly urban dynamics, have according to \cite{pumain1997pour} an intrinsic multi-scalar nature, with successive autonomous levels of emergence from individual microscopic agents to the mesoscopic scale of the city and the macroscopic scale of the system of cities. Furthermore, the need for sustainable territorial policies would imply the construction of multi-scalar models to take into account issues associated to each relevant scale \cite{Rozenblat2018}.
%This contribution contributes to that open question by introducing a multi-scale model of urban growth which focuses on the spatial structure of processes rather than on their multi-dimensionality. Therefore, we take into account only population variables, but both at the macroscopic scale of the system of cities in the legacy of \cite{pumain2017urban} and at the mesoscopic scale of the metropolitan area with an urban morphogenesis model. The coupling of these scales is a crucial novel feature of our model. We describe in the following stylized facts justifying the approach, describe the model, and summarize preliminary results from its exploration and calibration.

This contribution introduces a parsimonious multi-scalar model for systems of cities, based on simple dimensions (mainly populations) with stylized processes, but yielding an effective strong coupling between the metropolitan mesoscopic scale and the macroscopic scale of the system of cities. The model couples the spatial interaction model of \cite{raimbault2018indirect} for the macro scale with the reaction-diffusion model for urban form studied by \cite{raimbault2018calibration}. More precisely, urban areas viewed as a population grid are embedded into the macroscopic interaction model. To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved at a given speed conditionally to the population variation; (iv) changes in urban form influence macroscopic interaction range (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.

The model is applied on synthetic systems of cities typical of a continental range (500km, hierarchy around 1, 20 cities), with initial local population grid configurations as monocentric. Parameter space is explored with the OpenMOLE model exploration software \cite{reuillon2013openmole}, eased by the implementation of the model in scala \cite{model}. First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario

%\begin{figure}[!htp]
%  \centering
%  %\includegraphics[width=0.6\textwidth]{Fig1.png}
%  \caption{Poincare set of the Duffing equation, for a=0.1, b=0.3}\label{fig:1}
%\end{figure}

%\noindent
%The sections for acknowledgements and references should appear at the end of the abstract (Times New Roman 11pt).

%\section*{Acknowledgements (optional)}
%The acknowledgements should be written here. Thank you for spending your time in reading the instructions. Feel free to contact us for any further information.

\footnotesize

\bibliographystyle{unsrt}
\bibliography{biblio.bib}

\end{document}

### 2019-04-30 Models/Analysis/functions.R

@@ -1,7 +1,15 @@
 
 library(sp)
-library(hypervolume)
+#library(hypervolume)
 
+
+hierarchy<-function(x){
+  xx = x[x>0]
+  if(length(xx)==0){return(0)}else{
+    reg = lm(data=data.frame(logx = sort(log(xx),decreasing = T),logrank=log(1:length(xx))),logx~logrank)
+    return(reg$coefficients[2])
+  }
+}
 
 runmodel<-function(args){
   system(paste0('sbt "run ',args,'"'))

### 2019-04-30 Models/Analysis/multiscale.R

@@ -1,0 +1,62 @@
+
+setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/urbangrowth/openmole'))
+
+library(dplyr)
+library(ggplot2)
+
+source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
+
+
+resPrefix = '20190429_193158_MULTISCALE_GRID_GRID'
+resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Multiscale/',resPrefix);dir.create(resdir)
+
+res <- as.tbl(read.csv(paste0('exploration/',resPrefix,'.csv')))
+
+params = c("macroGrowthRate","macroInteractionDecay","macroInteractionGamma",
+           "mesoAlpha","mesoBeta","mesoTimeSteps",
+           "macroMesoAlphaUpdateMax","macroMesoBetaUpdateMax","mesoMacroCongestionCost","mesoMacroDecayUpdateMax"
+           )
+
+macroindics = c("macroAccessibilities","macroClosenesses","macroPopulations")
+mesoindics = c("mesoSlopes","mesoMorans","mesoEntropy","mesoDistances")
+
+ncities = 20
+tsteps = 5
+
+for(macroindic in macroindics){
+  res[[paste0("deltaHierarchy",macroindic)]] = apply(res[,paste0(macroindic,0:(ncities-1))],2,hierarchy) - apply(res[,paste0(macroindic,(ncities*(tsteps-1)):(ncities*tsteps - 1))],2,hierarchy)
+}
+
+for(mesoindic in mesoindics){
+  res[[paste0("delta",mesoindic)]] = apply(res[,paste0(mesoindic,0:(ncities-1))],2,mean) - apply(res[,paste0(mesoindic,(ncities*(tsteps-1)):(ncities*tsteps - 1))],2,mean)
+}
+
+macroindics=paste0("deltaHierarchy",macroindics)
+mesoindics=paste0("delta",mesoindics)
+
+
+sres = res %>% group_by(
+  macroGrowthRate,macroInteractionDecay,macroInteractionGamma,mesoAlpha,mesoBeta,mesoTimeSteps,
[…]

### 2019-04-30 Models/urbangrowth/src/main/scala/urbangrowth/models/multiscale/MultiscaleModel.scala

@@ -185,11 +185,11 @@ object MultiscaleModel {
     * @param call
     * @param threshold
     */
-  def ensureConsistence(macroState: InteractionMacroState,mesoStates: Vector[ReactionDiffusionMesoState],call: String = "",threshold: Double = 100.0): InteractionMacroState = {
+  def ensureConsistence(macroState: InteractionMacroState,mesoStates: Vector[ReactionDiffusionMesoState],call: String = "",threshold: Double = 10000.0): InteractionMacroState = {
     val deltaPopLevels = macroState.populations.zip(mesoStates).map{case (p,ms)=>math.abs(p - ms.populationGrid.flatten.sum)}
-    assert(deltaPopLevels.sum/deltaPopLevels.size<threshold,call+" - incoherence between levels : "+deltaPopLevels+"\n"+macroState.populations+"\n"+
+    /*assert(deltaPopLevels.sum/deltaPopLevels.size<threshold,call+" - incoherence between levels : "+deltaPopLevels+"\n"+macroState.populations+"\n"+
       mesoStates.map(_.populationGrid.flatten.sum)
-    )
+    )*/
     log("total consistence adj = "+deltaPopLevels.map(math.abs).sum)
     macroState.copy(populations = mesoStates.map(_.populationGrid.flatten.sum))
   }

### 2019-04-30 Models/urbangrowth/src/main/scala/urbangrowth/models/multiscale/MultiscaleState.scala

@@ -33,8 +33,14 @@ case class MultiScaleResult(
                            mesoEntropy: Vector[Double],
                            mesoSlopes: Vector[Double],
                            mesoSlopeRsquared: Vector[Double],
-                           mesoCongestedFlows: Vector[Double]
-                           )
+                           mesoCongestedFlows: Vector[Double],
+                           mesoMissingPopulations: Vector[Double]
+                           ) {
+  def asArrayTuple =  (macroPopulations.toArray,macroClosenesses.toArray,macroAccessibilities.toArray,
+    mesoMorans.toArray,mesoDistances.toArray,mesoEntropy.toArray,mesoSlopes.toArray,mesoSlopeRsquared.toArray,
+    mesoCongestedFlows.toArray,mesoMissingPopulations.toArray
+  )
+}
 
 object MultiScaleResult {
 
@@ -51,9 +57,12 @@ object MultiScaleResult {
       morphologies.map(_._3),
       morphologies.map(_._4),
       morphologies.map(_._5),
-      rawStates.flatMap(_.macroState.congestedFlows)
+      rawStates.flatMap(_.macroState.congestedFlows),
+      rawStates.flatMap(_.mesoStates.map(_.missingPopulation))
     )
   }
+
+
 
 }
 

### 2019-04-30 Models/urbangrowth/src/main/scala/urbangrowth/models/multiscale/ReactionDiffusionMeso.scala

@@ -18,11 +18,14 @@ object ReactionDiffusionMeso {
   def mesoStep(states: Vector[ReactionDiffusionMesoState])(implicit rng: Random): Vector[ReactionDiffusionMesoState] = {
     states.map{
       s => {
+        val toadd = s.growthRate*s.mesoTimeSteps
+        val prevpop = s.populationGrid.flatten.sum
         var state = s
         (0 until s.mesoTimeSteps).foreach{_ =>
           state = state.copy(time=state.time+1,populationGrid=reactionDiffusionStep(s.alpha,s.beta,s.ndiff,s.growthRate)(state.populationGrid))
         }
-        state
+        val newpop = state.populationGrid.flatten.sum
+        state.copy(missingPopulation = toadd-(newpop-prevpop))
       }
     }
   }

### 2019-04-30 Models/urbangrowth/src/main/scala/urbangrowth/models/multiscale/ReactionDiffusionMesoState.scala

@@ -13,7 +13,8 @@ case class ReactionDiffusionMesoState(
                                      beta: Double,
                                      ndiff: Int,
                                      growthRate: Double,
-                                     mesoTimeSteps: Int
+                                     mesoTimeSteps: Int,
+                                     missingPopulation: Double
                                      ) extends MesoState {
 
 }
@@ -31,7 +32,7 @@ object ReactionDiffusionMesoState {
                             mesoTimeSteps: Int)(implicit rng: Random): ReactionDiffusionMesoState = ReactionDiffusionMesoState(
     0,
     ExpMixtureGenerator(gridSize,1,centerDensity,kernelRadius,false,Seq((gridSize/2,gridSize/2))).generateGrid.map{_.toVector}.toVector,
-    alpha,beta,ndiff,growthRate,mesoTimeSteps
+    alpha,beta,ndiff,growthRate,mesoTimeSteps,0.0
   )
 
 }

### 2019-04-30 Models/urbangrowth/src/main/scala/urbangrowth/test/Test.scala

@@ -34,9 +34,9 @@ object TestModels {
 
     val n = 20
 
-    val model = MultiscaleModel(timeSteps = 5,macroNcities = n,macroInitialHierarchy = 1.0,macroInitialMaxPop = 10000,macroRange=500,
+    val model = MultiscaleModel(timeSteps = 20,macroNcities = n,macroInitialHierarchy = 1.0,macroInitialMaxPop = 100000,macroRange=500,
       macroGrowthRate = 0.01,macroInteractionDecay = 200.0,macroInteractionWeight=0.001,macroInteractionGamma=1.5,
-      mesoGridSize=50,mesoCenterDensity=100,mesoAlpha = 1.1,mesoBeta=0.05,mesoNdiff = 1,mesoTimeSteps = 5,
+      mesoGridSize=50,mesoCenterDensity=1000,mesoAlpha = 1.1,mesoBeta=0.05,mesoNdiff = 1,mesoTimeSteps = 5,
       macroMesoAlphaUpdateMax = 0.1,macroMesoBetaUpdateMax = 0.01,mesoMacroCongestionCost = 1.0,mesoMacroDecayUpdateMax = 10.0
     )
 

### 2019-09-12 Docs/Papers/Multiscale/paper.tex
multiscale_paper

\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[margin=2cm]{geometry}
\usepackage{multirow}

\begin{document}

% Systems of cities
%Multi-scalar model
%Spatial interaction
%Reaction-diffusion

\section{Introduction}

The modeling of urban growth is a crucial issue for the design of sustainable territorial policies, through the understanding of past urbanization processes and the forecasting of future urban trajectories. Several models have been proposed at different scales and integrating different dimensions of urban systems, such as land-use transport interaction models \cite{wegener2004land} or systems of cities models \cite{pumain2017urban}. While multi-scalar models are recognized as crucial for the study of such systems \cite{Rozenblat2018}, they remain in practice unexplored.

At the scale of a metropolitan area, Land-use Transport Interaction models \cite{wegener2004land} are for example a privileged tool to anticipate the answer of spatial distributions of activities (mostly residential location and economic activities) to an evolution of the accessibility landscape permitted by new transportation infrastructures. At the same scale, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}. At the smaller scale of the system of cities, macroscopic models of urban growth have focused on reproducing the distribution of city sizes, either through economic processes as e.g. \cite{gabaix1999zipf}, or from a geographical point of view focusing on interactions between cities \cite{favaro2011gibrat}.

Territorial dynamics, and more particularly urban dynamics, have according to \cite{pumain1997pour} an intrinsic multi-scalar nature, with successive autonomous levels of emergence from individual microscopic agents to the mesoscopic scale of the city and the macroscopic scale of the system of cities. Furthermore, the need for sustainable territorial policies would imply the construction of multi-scalar models to take into account issues associated to each relevant scale \cite{Rozenblat2018}.

This paper contributes to that open question by introducing a multi-scale model of urban growth which focuses on the spatial structure of processes rather than on their multi-dimensionality. Therefore, we take into account only population variables, but both at the macroscopic scale of the system of cities in the legacy of \cite{pumain2017urban} and at the mesoscopic scale of the metropolitan area with an urban morphogenesis model. The coupling of these scales is a crucial novel feature of our model. We describe in the following stylized facts justifying the approach, describe the model, and summarize preliminary results from its exploration and calibration.

\section{Multi-scale model}

\subsection{Rationale}

This contribution introduces a parsimonious multi-scalar model for systems of cities, based on simple dimensions (mainly populations) with stylized processes, but yielding an effective strong coupling between the metropolitan mesoscopic scale and the macroscopic scale of the system of cities. The model couples the spatial interaction model of \cite{raimbault2018indirect} for the macro scale with the reaction-diffusion model for urban form studied by \cite{raimbault2018calibration}. More precisely, urban areas viewed as a population grid are embedded into the macroscopic interaction model. To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved with the reaction-diffusion models at a given speed conditionally to the population variations; (iv) changes in urban form influence macroscopic interaction ranges (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.

\subsection{Formalization}

We consider $N$ urban areas, represented at the macroscopic scale by their population $P_j(t)$ at time $t$, and at the mesoscopic scale by a population grid $p_{kl}^{(j)}(t)$.

The model runs for a total number $t_f$ of time steps, and we will assume that $\Delta t = 1$ for the sake of simplicity (the formulas can be generalized for arbitrary values of the time step, for example when running on real data with irregular time sampling).

The system is initialized with synthetic data with a parameter $\alpha_0$ for the initial hierarchy, $P_0 (0)$ for the initial population of the largest city, in a square world of size $w$ (reference unit for the decay parameter).

At each time step:

\begin{enumerate}
	\item Aggregated population are evolved according to

	\begin{equation}
		P_i(t+1) = P_i(t) \left(1 + \Delta t \cdot \left(g_i + \frac{w_i}{N} \cdot \sum_j \frac{V_{ij}}{<V_{ij}>} \right) \right)
	\end{equation}

	where the gravity interaction potential is given by

	\begin{equation}
		V_{ij} = \left(\frac{P_i P_j}{(\sum_k p_k)^2}\right)^{\gamma_G} \cdot \exp \left(- \frac{d_{ij}}{d_i} \right)
	\end{equation}

	and we write the population variations

	\begin{equation}
		\Delta P_i (t) = P_i (t + 1) - P_i (t)
	\end{equation}

	\item Mesoscopic parameters are modified following the evolution of population such that 
		\begin{itemize}
			\item the mesoscopic growth rate is adjusted to the population growth uniformaly over the time interval	
			    $N_G^{(i)} (t + 1) = \Delta P_i / t_m$
			\item The sprawl parameter evolves according to a fixed multiplier and the relative population increase following 
			\begin{equation}
				\beta_i (t+1) = \beta_i (t) \cdot \left(1 + \delta\beta \cdot \frac{\Delta P_i (t)}{\max_k  \Delta P_k (t)}\right)
			\end{equation}
			where the multiplier parameter $\delta\beta$ allows testing different scenarios: a negative value corresponds to transit-oriented development while a positive value corresponds to an uncontrolled sprawl
			\item The aggregation parameter evolves in a similar way but as a function of accessibility increase
			\begin{equation}
				\alpha_i (t+1) = \alpha_i (t) \cdot \left(1 + \delta\alpha \cdot \frac{\Delta Z_i (t)}{\max_k  \Delta Z_k (t)}\right)
			\end{equation}
			where the multiplier parameter $\delta\alpha$ allows switching between a metropolization scenario (more aggregation) and an uniformization scenario (less aggregation), and accessibility is given by
			\begin{equation}
				Z_i = \sum_j \frac{P_j}{\sum_k P_k} \cdot \exp( - d_{ij} / d_i)
			\end{equation}
			\item Change in the level of sprawl depends on the population pressure only, while aggregation depends on accessibility since it is linked to metropolization processes
			\item \textit{Note: the linear scale for these two parameters may not be relevant depending on the distribution of increments ?} $\rightarrow$ to be tested
		\end{itemize}

	\item Mesoscopic grids are evolved by the updated parameters, and $t_m$ time steps, following the aggregation-diffusion model, with $n_d$ unchanged. Slight differences in the end (due to rounding in computing the number of steps) is corrected by adjusting the macroscopic increments by the effective mesoscopic increments (which are assumed to be more precise).

	\item Macroscopic parameters are updated: for the sake of simplicity, only interaction decays are updated, assuming that urban form pattern play a role in the global insertion of the city. More precisely, we compute gravity flows within the area, and aggregate their value as an economic activity with a squared negative externality interpreted as a congestion with a cost $\lambda$ following

	\begin{equation}
		U_i = \sum_{kl} \left( \frac{P_k P_l}{P^2} \cdot \frac{1}{d_{kl}} - \lambda \left(\frac{P_k P_l}{P^2} \cdot \frac{1}{d_{kl}}\right)^2 \right)
	\end{equation}

	We do not add gravity parameter nor hierarchy parameter for the sake of simplicity. This utility $U_i$ is used to update the interaction decays following

	\begin{equation}
		d_i (t+1) = d_i (t) \left( 1 + \delta d \cdot \frac{U_i}{\max_k \left|U_k\right|} \right)
	\end{equation}
	where the multiplier parameter $\delta d$ allows controlling for the influence of local performance on global insertion.
\end{enumerate}

\subsection{Parameters}

The Table~\ref{tab:parameters} summarizes

\begin{table}
	\caption{Summary of model parameters\label{tab:parameters}}
	\centering
	\begin{tabular}{|c|c|c|c|}
	\hline
		Type & Parameter & Process & Range \\\hline
		\multirow{4}{*}{Macro} & $g_i = g_0$ & Endogenous growth & \\
		& $w_i = w_G$ & Interactions weight & \\
		& $\gamma_i = \gamma_G$ & Interactions hierarchy & \\
		& $d_i$ & Interactions decay & \\ \hline
		\multirow{4}{*}{Meso} & $\alpha_i$ & Aggregation & \\
		& $\beta_i$ & Diffusion & \\
		& $t_m$ & Urban growth speed & \\
		& $n_d$ & Diffusion & \\ \hline
		\multirow{3}{*}{Multiscale} & $\delta\alpha$ & Downward feedback & \\
		& $\delta\beta$ & Downward feedback & \\
		& $\delta d$ & Upward feedback & \\\hline
	\end{tabular}
\end{table}

\section{Results}

The model is applied on synthetic systems of cities typical of a continental range (500km, hierarchy around 1, 20 cities), with initial local population grid configurations as monocentric. Parameter space is explored with the OpenMOLE model exploration software \cite{reuillon2013openmole}, eased by the implementation of the model in scala \cite{model}. First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario (negative feedback of population growth on diffusion) to a ``sprawl'' scenario (positive feedback). Similarly, mesoscopic urban form indicators are significantly influenced by the coupling process.

\section{Discussion}

\subsection{Multi-modeling and concurrent processes}

This model is only a first structural sketch with very restrictive assumption, in particular regarding the downward and upward feedbacks on submodel parameters. There may be no link between urban form and global insertion, or it may be due to other processes, be expressed as an other functional form. An important stage before shifting to robust knowledge will consist in (i) reviewing and making a typology of such potential processes across scales; (ii) including most in a multi-modeling fashion to compare possible concurrent mechanisms.

\subsection{Developments}

Further work will consist in more targeted simulation experiments, including specific exploration algorithms such as diversity search for model regimes \cite{reuillon2013openmole}, to test the model as a proof-of-concept of models for policies. Such a model can also be calibrated on real city systems and urban form trajectories, to extrapolate coupling parameters that would be difficult to obtain otherwise. Our contribution is thus a first step towards multi-scalar simulation models for systems of cities.

We also did not include explicitly transportation networks in this model.

\bibliographystyle{apalike}
\bibliography{biblio}

\end{document}

### 2019-09-12 Models/urbangrowth/src/main/scala/urbangrowth/models/multiscale/MultiscaleModel.scala

@@ -162,6 +162,7 @@ object MultiscaleModel {
 
     assert(mesoStates.map{_.populationGrid.flatten.sum>0}.filter(_==false).size==0,"existing null meso pop grids")
 
+    // note: congestedFlow function in spatialdata computes \sum (flow - lambda flow^2 )
     val utilities = mesoStates.map{s => GridMorphology.congestedFlows(s.populationGrid.map{_.toArray}.toArray,congestionCost)}
 
     assert(utilities.filter(_.isNaN).size==0,"Nan in utilities : "+utilities)

### 2019-09-19 Docs/Papers/Multiscale/paper.tex
multiscal_targeted_summary

@@
 
+\title{Multiscalar modeling of urban growth}
+\date{}
+
 \begin{document}
+
+\maketitle
+
+\begin{abstract}
+	Urban evolution processes occur at different scales, with intricate interactions between levels and relatively distinct type of processes.
+To what extent actual urban dynamics include an actual strong coupling between scales, in the sense of both top-down and bottom-up feedbacks, remains an open issue with important practical application for the management of urban growth.
+We introduce in this paper a multi-scalar simulation model of urban growth, coupling an interaction model at the macroscopic scale with morphogenesis models for the evolution of urban form at the scale of metropolitan areas.
+\end{abstract}
 
@@
 
-The Table~\ref{tab:parameters} summarizes
+The Table~\ref{tab:parameters} summarizes model parameters.
 
@@
 
+\subsection{Indicators}
+
+Model behavior is characterized using the following indicators:
+
 \section{Results}
+
+\subsection{Implementation}
 
@@
 Parameter space is explored with the OpenMOLE model exploration software \cite{reuillon2013openmole}, eased by the implementation of the model in scala \cite{model}.
+
+\subsection{Grid exploration}
+
+% 20190429_193158_MULTISCALE_GRID_GRID.csv
+
 First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario (negative feedback of population growth on diffusion) to a ``sprawl'' scenario (positive feedback).
 Similarly, mesoscopic urban form indicators are significantly influenced by the coupling process.
+
+% 20190506_135221_MULTISCALE_TARGETEDGRID_GRID
+
+We do a targeted experiment to look at the influence of the macroscopic interaction decay and the mesoscopic evolution speed, under stylized scenarios for the downward feedback.
+
+We run 50 replications for each parameter value.
+
+Congestion cost for upward feedback is fixed to a rather strong effect.
+% note : we should control here compare with / without feedback ?
+%  -> first thing to check : with/ without the model
 

### 2019-09-19 Models/Analysis/multiscale.R

@@ -4,14 +4,16 @@ setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/urbangrowth/openmole'))
 library(dplyr)
 library(ggplot2)
 
-source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+#source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
 
 
-resPrefix = '20190429_193158_MULTISCALE_GRID_GRID'
+#resPrefix = '20190429_193158_MULTISCALE_GRID_GRID'
+resPrefix = '20190506_135221_MULTISCALE_TARGETEDGRID_GRID'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Multiscale/',resPrefix,'/');dir.create(resdir)
 
-res <- as.tbl(read.csv(paste0('exploration/',resPrefix,'.csv')))
+#res <- as.tbl(read.csv(paste0('exploration/',resPrefix,'.csv')))
+res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'.csv'),stringsAsFactors = F))
 
 params = c("macroGrowthRate","macroInteractionDecay","macroInteractionGamma",
            "mesoAlpha","mesoBeta","mesoTimeSteps",
@@ -39,29 +41,41 @@ mesoindics=paste0("delta",mesoindics)
 sres = res %>% group_by(
   macroGrowthRate,macroInteractionDecay,macroInteractionGamma,mesoAlpha,mesoBeta,mesoTimeSteps,
   macroMesoAlphaUpdateMax,macroMesoBetaUpdateMax,mesoMacroCongestionCost,mesoMacroDecayUpdateMax
-) %>% summarise(deltaHierarchymacroAccessibilities=mean(deltaHierarchymacroAccessibilities),
-deltaHierarchymacroClosenesses=mean(deltaHierarchymacroClosenesses),
-deltaHierarchymacroPopulations=mean(deltaHierarchymacroPopulations),
-deltamesoSlopes=mean(deltamesoSlopes),deltamesoMorans=mean(deltamesoMorans),
[…]

### 2019-09-26 Models/Multiscale/multiscale.R
extract_data

@@ -4,17 +4,19 @@ setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/urbangrowth/openmole'))
 library(dplyr)
 library(ggplot2)
 
-#source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
 
 
 #resPrefix = '20190429_193158_MULTISCALE_GRID_GRID'
 #resPrefix = '20190506_135221_MULTISCALE_TARGETEDGRID_GRID'
 resPrefix = '20190919_161009_MULTISCALE_TARGETEDGRID_GRID'
+#resPrefix = '20190919_141607_MULTISCALE_TARGETEDGRID_GRID'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Multiscale/',resPrefix,'/');dir.create(resdir)
 
 #res <- as.tbl(read.csv(paste0('exploration/',resPrefix,'.csv')))
-res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'.csv'),stringsAsFactors = F))
+#res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'.csv'),stringsAsFactors = F))
+res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'_sample.csv'),stringsAsFactors = F))
 
 params = c("macroGrowthRate","macroInteractionDecay","macroInteractionGamma",
            "mesoAlpha","mesoBeta","mesoTimeSteps",
@@ -29,6 +31,18 @@ mesoindics = c("mesoSlopes","mesoMorans","mesoEntropy","mesoDistances")
 ncities = 20
 tsteps = 20
 
+## test
+#macroindic="macroPopulations"
+macroindic="macroAccessibilities"
+#macroindic="macroClosenesses"
+i=1
+P0 = unlist(res[i,paste0(macroindic,0:(ncities-1))])
+Pf = unlist(res[i,paste0(macroindic,(ncities*tsteps - 1):(ncities*(tsteps + 1) - 1))])
+plot(log(1:length(Pf)),sort(log(Pf),decreasing = T))
[…]

### 2019-09-26 Models/Multiscale/summary.R
.

@@ -13,8 +13,8 @@ source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
 
 #resPrefix = '20190429_193158_MULTISCALE_GRID_GRID'
 #resPrefix = '20190506_135221_MULTISCALE_TARGETEDGRID_GRID'
-#resPrefix = '20190919_161009_MULTISCALE_TARGETEDGRID_GRID'
-resPrefix = '20190919_141607_MULTISCALE_TARGETEDGRID_GRID'
+resPrefix = '20190919_161009_MULTISCALE_TARGETEDGRID_GRID'
+#resPrefix = '20190919_141607_MULTISCALE_TARGETEDGRID_GRID'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Multiscale/',resPrefix,'/');dir.create(resdir)
 
 #res <- as.tbl(read.csv(paste0('exploration/',resPrefix,'.csv')))

### 2019-09-26 Models/Multiscale/multiscale.R
filter_timeseries_in_MultiscaleResult

@@ -12,11 +12,13 @@ source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
 #resPrefix = '20190506_135221_MULTISCALE_TARGETEDGRID_GRID'
 resPrefix = '20190919_161009_MULTISCALE_TARGETEDGRID_GRID'
 #resPrefix = '20190919_141607_MULTISCALE_TARGETEDGRID_GRID'
+#resPrefix = '20190926_114834_MULTISCALE_LHS_TEST'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Multiscale/',resPrefix,'/');dir.create(resdir)
 
 #res <- as.tbl(read.csv(paste0('exploration/',resPrefix,'.csv')))
 #res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'.csv'),stringsAsFactors = F))
-res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'_sample.csv'),stringsAsFactors = F))
+#res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'_sample.csv'),stringsAsFactors = F))
+res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'_filtered.csv'),stringsAsFactors = F))
 
 params = c("macroGrowthRate","macroInteractionDecay","macroInteractionGamma",
            "mesoAlpha","mesoBeta","mesoTimeSteps",
@@ -33,31 +35,60 @@ tsteps = 20
 
 ## test
 #macroindic="macroPopulations"
-macroindic="macroAccessibilities"
+#macroindic="macroAccessibilities"
 #macroindic="macroClosenesses"
-i=1
-P0 = unlist(res[i,paste0(macroindic,0:(ncities-1))])
-Pf = unlist(res[i,paste0(macroindic,(ncities*tsteps - 1):(ncities*(tsteps + 1) - 1))])
-plot(log(1:length(Pf)),sort(log(Pf),decreasing = T))
-plot(log(1:length(P0)),sort(log(P0),decreasing = T),col='red')
-hierarchy(Pf) - hierarchy(P0)
+#i=10
+#P0 = unlist(res[i,paste0(macroindic,0:(ncities-1))])
[…]

### 2019-09-26 Models/Multiscale/summary.R

@@ -13,8 +13,8 @@ source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
 
 #resPrefix = '20190429_193158_MULTISCALE_GRID_GRID'
 #resPrefix = '20190506_135221_MULTISCALE_TARGETEDGRID_GRID'
-resPrefix = '20190919_161009_MULTISCALE_TARGETEDGRID_GRID'
-#resPrefix = '20190919_141607_MULTISCALE_TARGETEDGRID_GRID'
+#resPrefix = '20190919_161009_MULTISCALE_TARGETEDGRID_GRID'
+resPrefix = '20190919_141607_MULTISCALE_TARGETEDGRID_GRID'
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Multiscale/',resPrefix,'/');dir.create(resdir)
 
 #res <- as.tbl(read.csv(paste0('exploration/',resPrefix,'.csv')))

### 2019-09-26 Models/urbangrowth/openmole/copyRes.py
prepare_for_proper_submodule

@@ -1,31 +1,0 @@
-import sys,os,shutil
-
-# copy latest generation for each run in a folder
-
-SOURCE=sys.argv[1]
-TARGET=sys.argv[2]
-MULTIPLE=False
-if len(sys.argv) > 3:
-    MULTIPLE=True
-
-def getLatestGen(dir):
-    files = [s for s in os.listdir(dir) if not s.startswith('Calib')]
-    #print(files)
-    print(sorted(map(lambda s:int(s.split('population')[1].split('.')[0]),files),reverse=True)[0])
-    return(str(sorted(map(lambda s : int(s.split('population')[1].split('.')[0]),files),reverse=True)[0]))
-
-
-sourcedirs = os.listdir(SOURCE)
-targetdirs = os.listdir(TARGET)
-
-for dir in sourcedirs :
-    if not os.path.isdir(TARGET+'/'+dir):
-        os.mkdir(TARGET+'/'+dir)
-    gen = getLatestGen(SOURCE+'/'+dir)
-    gens = [gen]
-    if MULTIPLE :
-        #gens = range((int(gen) - 1000),int(gen),100)
-        gens = [1,10,100,500]+list(range(1000,int(gen)+1,1000))
-    for gen in gens :
-        if not os.path.isfile(TARGET+'/'+dir+'/population'+str(gen)+'.csv') and os.path.isfile(SOURCE+'/'+dir+'/population'+str(gen)+'.csv'):
-            shutil.copy(SOURCE+'/'+dir+'/population'+str(gen)+'.csv',TARGET+'/'+dir+'/population'+str(gen)+'.csv')

### 2019-09-26 Models/urbangrowth/openmole/runCalib.sh

@@ -1,1 +1,0 @@
-cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 120G --logger-level ALL;mkdir calibration/`ls -t calib | head -n 1`; cp Calib"$1$2".oms calibration/`ls -t calib | head -n 1`"}'|sh

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/Run.scala

@@ -1,19 +1,0 @@
-
-package urbangrowth
-
-import models.marius.RunMarius
-import models.coevolution.RunCoevolution
-
-object Run extends App {
-
-  assert(args.length > 0,"Provide at least a model to run")
-
-  val modelName = args(0)
-
-  modelName match {
-    case "Marius" => RunMarius.run(args)
-    case "Coevolution" => RunCoevolution.run(args)
-    case _ => assert(1==0,"Unknown model")
-  }
-
-}

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/indicators/Indicators.scala

@@ -1,45 +1,0 @@
-
-package urbangrowth.indicators
-
-import Jama.Matrix
-
-
-object Indicators {
-
-
-  /**
-    * MSE on logs of populations
-    *
-    * @param m
-    * @param target
-    * @return
-    */
-  def mselog(m: Matrix,target: Matrix): Double = {
-    val logres = if (m.getColumnDimension==target.getColumnDimension&m.getRowDimension==target.getRowDimension)
-      new Matrix(m.getArray().map { _.map { d => Math.log(d) } })
-    else new Matrix(target.getRowDimension,target.getColumnDimension,0.0)
-    val logreal = new Matrix(target.getArray().map { _.map { d => Math.log(d) } })
-    val sqdiff = logres.minus(logreal).arrayTimes(logres.minus(logreal))
-    return sqdiff.getArray().flatten.sum
-  }
-
-  /**
-    * Log of MSE
-    *
-    * @param m
-    * @param target
-    * @return
-    */
-  def logmse(m: Matrix,target: Matrix): Double = {
-    //println(m.getColumnDimension)
-    //println(m.getRowDimension)
-    val sqdiff = if(m.getColumnDimension==target.getColumnDimension&m.getRowDimension==target.getRowDimension)
-      m.minus(target).arrayTimes(m.minus(target)) else {
-      val zeros = new Matrix(target.getRowDimension,target.getColumnDimension,0.0)
-        zeros.minus(target).arrayTimes(zeros.minus(target))
-    }
-    return Math.log(sqdiff.getArray().flatten.sum)
-  }
-
-
-}

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/coevolution/Coevolution.scala

@@ -1,222 +1,0 @@
-
-package urbangrowth.models.coevolution
-
-import java.io.File
-
-import Jama.Matrix
-
-import urbangrowth.models.MacroModel
-import urbangrowth.indicators._
-import urbangrowth.utils.io.FileUtils
-import urbangrowth.utils.math.MatrixUtils
-
-import urbangrowth._
-
-case class Coevolution(
-                        /**
-                          * Target pop matrix
-                          *   rows : cities ; cols : time
-                          */
-                        populationMatrix: Matrix,
-
-                        /**
-                          * Distance matrices (target in time)
-                          */
-                        distancesMatrices: Array[Matrix],
-
-                        /**
-                          * Network feedback distance matrix
-                          */
-                        feedbackDistancesMatrix: Matrix,
-
-                        /**
-                          * Dates
-                          */
-                        dates: Array[Double],
-                        growthRate: Double,
-                        gravityWeight: Double,
-                        gravityGamma: Double,
-                        gravityDecay: Double,
-                        feedbackWeight: Double,
-                        feedbackGamma: Double,
-                        feedbackDecay: Double
-                      ) extends MacroModel {
-
-  override def run(): Result = Coevolution.run(model = this)
-
-  override def toString: String = "Coevolution model with parameters"+
-   "\n\tgrowthRate = "+growthRate+"\n\tgravityWeight = "+gravityWeight+
-  "\n\tgravityGamma = "+gravityGamma+"\n\tgravityDecay = "+gravityDecay+
-  "\n\tfeedbackWeight = "+feedbackWeight+"\n\tfeedbackGamma = "+feedbackGamma+
-  "\n\tfeedbackDecay = "+feedbackDecay
-
-}
-
-
-
-object Coevolution {
-
-  //var populationMatrix: Matrix = null
-  //var distancesMatrix: Matrix = null
-  //var feedbackDistancesMatrix: Matrix = null
-  //var dates: Array[Double] = null
-  // setup matrices
[…]

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/coevolution/RunCoevolution.scala

@@ -1,26 +1,0 @@
-package urbangrowth.models.coevolution
-
-import java.io.File
-
-import urbangrowth.indicators._
-import urbangrowth.utils.io.FileUtils
-
-object RunCoevolution {
-
-  def run(args: Array[String]): Unit = {
-    assert(args.length>=3,"Coevol model needs at least 1 parameters")
-    val system = args(1)
-
-    val res: Result = args.length match {
-      case 3 =>  Coevolution(new File("data/processed/"+system+"_pops.csv"), null,null, new File("data/processed/"+system+"_dates.csv"),
-        args(2).toDouble,0.0, 1.0, 1.0, 0.0, 1.0, 1.0).run()
-      case 6 =>  Coevolution(new File("data/processed/"+system+"_pops.csv"), null,null, new File("data/processed/"+system+"_dates.csv"),
-        args(2).toDouble,args(3).toDouble, args(4).toDouble,args(5).toDouble, 0.0, 1.0, 1.0).run()
-    }
-
-    FileUtils.exportCSV(new File("res/popsim.csv"),res.simulatedPopulation)
-    FileUtils.exportCSV(new File("res/poptarget.csv"),res.simulatedPopulation)
-
-  }
-
-}

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/innovation/Innovation.scala

@@ -1,269 +1,0 @@
-
-package urbangrowth.models.innovation
-
-import Jama.Matrix
-import java.io.File
-
-import scala.collection.mutable.ArrayBuffer
-import scala.util.Random
-import urbangrowth.models.MacroModel
-import urbangrowth.utils.io.FileUtils
-import urbangrowth.indicators.Result
-import urbangrowth.utils.math.MatrixUtils
-
-
-
-case class Innovation(
-                       /**
-                         * Real population matrix
-                         */
-                       populationMatrix: Matrix,
-
-                       /**
-                         * Distance matrix
-                         */
-                      distanceMatrix: Matrix,
-
-                       /**
-                         * Dates
-                         */
-                       dates: Array[Double],
-
-                       /**
-                         * Model has its own rng
-                         */
-                       rng : Random,
-
-                       /**
-                         * Random seed
-                         */
-                       seed : Int,
-
-                       /**
-                         * Gibrat growth rate
-                         */
-                       growthRate: Double,
-
-                       /**
-                         * weight of innovation induced growth rate
-                         */
-                       innovationWeight: Double,
-
-                       /**
-                         * Decay of gravity interaction
-                         */
-                       gravityDecay: Double,
-
-                       /**
-                         * Decay of innovation diffusion
-                         */
-                       innovationDecay: Double,
-
-                       /**
-                         * Utility of the first innovation
-                         */
-                        innovationUtility: Double,
-
-                       /**
-                         * Growth of the innovation utility (default to 1.12)
-                         */
-                       innovationUtilityGrowth: Double,
-
-                       /**
-                         * Proportion of early adopters (defaults to 1%)
-                         */
-                       earlyAdoptersRate: Double,
-
[…]

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/Bonus.scala

@@ -1,51 +1,0 @@
-
-package urbangrowth.models.marius
-
-import DenseMatrix.Cell
-
-/**
-  * Bonus component which could be injected in the model
-  */
-trait Bonus { marius: Marius =>
-  /** Factor adjusting bonus value */
-  def bonusMultiplier: Double
-
-  /**
-    * Compute the bonuses given the diversity and the volume of transactions.
-    *
-    * @param t the transactions
-    * @return a sequence of bonuses indexed by city indexes
-    */
-  override def bonuses(t: Transacted): Seq[Double] = {
-    /** Measure of the exchange partners diversities */
-    def diversities = {
-      def transactedWith(transacted: Seq[Cell]) =
-        transacted.filter { case Cell(_, v) => v > 0 }.map { case Cell(to, _) => to }
-
-      (t.transacted.lines zip t.transposedTransacted.lines) map {
-        case (from, to) =>
-          (transactedWith(from).toSet union transactedWith(to).toSet).size / t.nbCities.toDouble
-      }
-    }
-
-    /** Volume of incoming transactions */
-    def importVolumes =
-      for {
-        (demand, i) <- t.demands.zipWithIndex
-      } yield t.transactedToSum(i)
-
-    /** Volume of outgoing transactions */
-    def exportVolumes =
-      for {
-        (supply, i) <- t.supplies.zipWithIndex
-      } yield t.transactedFromSum(i)
-
-    (importVolumes zip exportVolumes zip diversities) map {
-      case ((importVolume, exportVolume), diversity) =>
-        bonusMultiplier * (importVolume + exportVolume) * diversity
-    }
-  }
-
-}
-
-

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/BonusFixedCostModel.scala

@@ -1,45 +1,0 @@
-
-package urbangrowth.models.marius
-
-import java.io.File
-
-
-/** Model with fixed costs and bonuses */
-case class BonusFixedCostModel(
-                           modelDistanceMatrix: DistanceMatrix,
-                           modelConfiguration: MariusConfiguration,
-                           economicMultiplier: Double,
-                           sizeEffectOnSupply: Double,
-                           sizeEffectOnDemand: Double,
-                           distanceDecay: Double,
-                           wealthToPopulationExponent: Double,
-                           populationToWealthExponent: Double,
-                           bonusMultiplier: Double,
-                           fixedCost: Double) extends Marius with Bonus with FixedCost //with DefaultValues
-{
-  override def distanceMatrix: DistanceMatrix = modelDistanceMatrix
-
-  override def configuration: MariusConfiguration = modelConfiguration
-
-  override def maxStep: Int = modelConfiguration.dates.length - 1
-}
-
-object BonusFixedCostModel {
-  def apply(distanceMatrixFile: File,
-            confFile: File,
-            datesFile: File,
-            economicMultiplier: Double,
-            sizeEffectOnSupply: Double,
-            sizeEffectOnDemand: Double,
-            distanceDecay: Double,
-            wealthToPopulationExponent: Double,
-            populationToWealthExponent: Double,
-            bonusMultiplier: Double,
-            fixedCost: Double): BonusFixedCostModel =
-    BonusFixedCostModel(MariusFile.distances(distanceMatrixFile),MariusFile.configFromFile(confFile,datesFile) ,economicMultiplier,sizeEffectOnSupply,
-      sizeEffectOnDemand, distanceDecay, wealthToPopulationExponent, populationToWealthExponent,bonusMultiplier,fixedCost)
-}
-
-
-
-

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/DefaultValues.scala

@@ -1,9 +1,0 @@
-
-package urbangrowth.models.marius
-
-/** Default values for some parameters of the models */
-trait DefaultValues {
-  /** Simulate from 1959 to 1989 */
-  def maxStep = 30
-}
-

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/MariusConfiguration.scala

@@ -1,24 +1,0 @@
-
-package urbangrowth.models.marius
-
-import Jama.Matrix
-
-case class MariusConfiguration(
-                                dates: Seq[Int],
-                                realPopulations: Matrix,
-                                positions: Seq[Position],
-                                columnsBeforeDates: Int,
-                                arokatos: Seq[String],
-                                names: Seq[String],
-                                latitudes: Seq[Double],
-                                longitudes: Seq[Double],
-                                initialPopulations: Seq[Double],
-                                hydrocarbonDistribution: List[Boolean],
-                                regions: Iterator[String],
-                                regionCapitals: Iterator[Boolean],
-                                nationalCapitals: Iterator[Boolean],
-                                nations: Iterator[String]
-                              )
-
-
-

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/RunMarius.scala

@@ -1,27 +1,0 @@
-
-package urbangrowth.models.marius
-
-import java.io.File
-
-import urbangrowth.indicators._
-import urbangrowth.utils.io.FileUtils
-
-object RunMarius {
-
-  def run(args: Array[String]): Unit = {
-    assert(args.length>=6,"Marius model needs at least 4 parameters")
-    val system = args(1)
-
-    val res: Result = args.length match {
-      case 6 =>  SimpleModel(new File("data/processed/"+system+"_dist.csv"),new File("data/processed/"+system+"_pops.csv"),new File("data/processed/"+system+"_dates.csv"),
-        args(2).toDouble,args(3).toDouble, args(4).toDouble,args(5).toDouble, 1.0,1.0).run()
-      case 8 =>  SimpleModel(new File("data/processed/"+system+"_dist.csv"),new File("data/processed/"+system+"_pops.csv"),new File("data/processed/"+system+"_dates.csv"),
-        args(2).toDouble,args(3).toDouble, args(4).toDouble,args(5).toDouble, args(6).toDouble,args(7).toDouble).run()
-    }
-
-    FileUtils.exportCSV(new File("res/popsim.csv"),res.simulatedPopulation)
-    FileUtils.exportCSV(new File("res/poptarget.csv"),res.simulatedPopulation)
-
-  }
-
-}

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/marius/SimpleModel.scala

@@ -1,76 +1,0 @@
-
-package urbangrowth.models.marius
-
-import java.io.File
-
-import Jama.Matrix
-import urbangrowth.indicators._
-
-import scala.collection.mutable.ArrayBuffer
-
-import urbangrowth.models.MacroModel
-
-/** Simple model with only core mechanisms */
-case class SimpleModel(
-                      modelDistanceMatrix: DistanceMatrix,
-                      modelConfiguration: MariusConfiguration,
-                      economicMultiplier: Double,
-                      sizeEffectOnSupply: Double,
-                      sizeEffectOnDemand: Double,
-                      distanceDecay: Double,
-                      wealthToPopulationExponent: Double,
-                      populationToWealthExponent: Double) extends MacroModel with Marius //with DefaultValues
- {
-
-  override def distanceMatrix: DistanceMatrix = modelDistanceMatrix
-
-  override def configuration: MariusConfiguration = modelConfiguration
-
-   override def maxStep: Int = modelConfiguration.dates.length - 1
-
-   override def toString: String = "Simple Marius Model with parameters\n\teconomicMultiplier = "+economicMultiplier+"\n\tsizeEffectOnSupply = "+sizeEffectOnSupply+
-   "\n\tsizeEffectOnDemand = "+sizeEffectOnDemand+"\n\tdistanceDecay = "+distanceDecay+"\n\twealthToPopulationExponent = "+wealthToPopulationExponent+
-   "\n\tpopulationToWealthExponent = "+populationToWealthExponent
-
-  def run(): Result = SimpleModel.run(this)
-
-}
-
-
-object SimpleModel {
-
-  def apply(
-             distanceMatrixFile: File,
-             confFile: File,
-             datesFile: File,
-             economicMultiplier: Double,
-             sizeEffectOnSupply: Double,
-             sizeEffectOnDemand: Double,
-             distanceDecay: Double,
-             wealthToPopulationExponent: Double,
-             populationToWealthExponent: Double
-           ): SimpleModel =
-    SimpleModel(MariusFile.distances(distanceMatrixFile),MariusFile.configFromFile(confFile,datesFile) ,economicMultiplier,sizeEffectOnSupply,
-      sizeEffectOnDemand, distanceDecay, wealthToPopulationExponent, populationToWealthExponent)
-
-
-  def run(model: SimpleModel): Result = {
-    println("Running "+model.toString)
[…]

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/multiscale/MultiscaleModel.scala

@@ -1,201 +1,0 @@
-package urbangrowth.models.multiscale
-
-import scala.util.Random
-import org.openmole.spatialdata.grid.measures.GridMorphology
-import urbangrowth._
-import urbangrowth.models.multiscale
-
-case class MultiscaleModel(
-                          timeSteps: Int,
-                            // macro parameters
-                            macroNcities: Int,
-                            macroInitialHierarchy: Double,
-                            macroInitialMaxPop: Double,
-                            macroRange : Double,
-
-                            macroGrowthRate: Double,
-                            macroInteractionDecay: Double,
-                            macroInteractionWeight: Double,
-                            macroInteractionGamma: Double,
-
-                          // meso parameters
-                            mesoGridSize: Int,
-                            mesoCenterDensity: Double,
-
-                            mesoAlpha: Double,
-                            mesoBeta: Double,
-                            mesoNdiff: Int,
-                            mesoTimeSteps: Int,
-
-                            // coupling parameters
-                            macroMesoBetaUpdateMax: Double, // in share of beta
-                            macroMesoAlphaUpdateMax: Double,
-                            mesoMacroCongestionCost: Double,
-                            mesoMacroDecayUpdateMax: Double
-                          ) {
-
-  /**
-    * initial state
-    * @param rng
-    * @return
-    */
-  def initialState(implicit rng: Random): MultiscaleState = {
-    val macrostate = InteractionMacroState.initialSyntheticState(macroNcities,macroInitialHierarchy,macroInitialMaxPop,macroRange,macroGrowthRate,macroInteractionDecay,macroInteractionWeight,macroInteractionGamma)
-    val mesoStates = macrostate.populations.map{pi => {
-      val initState = ReactionDiffusionMesoState.initialSyntheticState(mesoGridSize, mesoCenterDensity, math.sqrt(pi / (2 * math.Pi * mesoCenterDensity)), mesoAlpha,
-        mesoBeta, mesoNdiff, 0.0, mesoTimeSteps)
[…]

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/models/multiscale/MultiscaleState.scala

@@ -1,71 +1,0 @@
-package urbangrowth.models.multiscale
-
-import urbangrowth.models._
-
-case class MultiscaleState(
-                          time: Int,
-                          macroState: InteractionMacroState,
-                          mesoStates: Vector[ReactionDiffusionMesoState]
-                          )
-
-
-/**
-  * computed indicators for a multiscale simu
-  * @param rawStates
-  * @param macroPopulations
-  * @param macroClosenesses
-  * @param macroAccessibilities
-  * @param mesoMorans
-  * @param mesoDistances
-  * @param mesoEntropy
-  * @param mesoSlopes
-  * @param mesoSlopeRsquared
-  * @param mesoCongestedFlows
-  */
-case class MultiScaleResult(
-                           rawStates: Vector[MultiscaleState],
-                           // all flattened indics
-                           macroPopulations: Vector[Double],
-                           macroClosenesses: Vector[Double],
-                           macroAccessibilities: Vector[Double],
-                           mesoMorans: Vector[Double],
-                           mesoDistances: Vector[Double],
-                           mesoEntropy: Vector[Double],
-                           mesoSlopes: Vector[Double],
-                           mesoSlopeRsquared: Vector[Double],
-                           mesoCongestedFlows: Vector[Double],
-                           mesoMissingPopulations: Vector[Double],
-                           fullTrajectories: Boolean
-                           ) {
-  def asArrayTuple =  (macroPopulations.toArray,macroClosenesses.toArray,macroAccessibilities.toArray,
-    mesoMorans.toArray,mesoDistances.toArray,mesoEntropy.toArray,mesoSlopes.toArray,mesoSlopeRsquared.toArray,
-    mesoCongestedFlows.toArray,mesoMissingPopulations.toArray
-  )
-}
-
-object MultiScaleResult {
-
-  def apply(rawStates: Vector[MultiscaleState],fulltrajs: Boolean): MultiScaleResult = {
-    val states: Vector[MultiscaleState] = if (fulltrajs) rawStates else rawStates.take(1)++rawStates.takeRight(1)
-    val morphologies: Vector[(Double,Double,Double,Double,Double)] = states.flatMap{_.mesoStates.map{_.morphology}}
-    val macroIndics: Vector[(Vector[Double],Vector[Double],Vector[Double])] = states.map{_.macroState.indicators}
-    MultiScaleResult(
[…]

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/package.scala

@@ -1,26 +1,0 @@
-package object urbangrowth {
-
-
-
-  type Populations = Array[Double]
-  type Distances = Array[Array[Double]]
-
-
-
-  sealed trait Logger
-
-  case object Printer extends Logger
-  case object DummyLogger extends Logger
-
-  implicit val logger = Printer
-
-  def log(msg: => String)(implicit logger: Logger) = {
-    logger match {
-      case Printer => println(msg)
-      case DummyLogger =>
-    }
-  }
-
-
-}
-

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/test/Test.scala

@@ -1,126 +1,0 @@
-
-package urbangrowth.test
-
-import java.io.File
-
-import Jama.Matrix
-
-import urbangrowth._
-import urbangrowth.indicators.Indicators
-import urbangrowth.models.innovation.Innovation
-import urbangrowth.models.multiscale._
-
-import scala.collection.mutable.ArrayBuffer
-import scala.util.Random
-
-
-object Test extends App {
-
-  TestModels.testMultiScale()
-  //TestModels.testInnovation()
-  //TestModels.testMarius()
-  //TestModels.testIntGib()
-
-}
-
-
-
-object TestModels {
-
-  implicit val rng = new Random
-
-
-  def testMultiScale(): Unit = {
-
-    val n = 20
-
-    val model = MultiscaleModel(timeSteps = 20,macroNcities = n,macroInitialHierarchy = 1.0,macroInitialMaxPop = 100000,macroRange=500,
-      macroGrowthRate = 0.01,macroInteractionDecay = 200.0,macroInteractionWeight=0.001,macroInteractionGamma=1.5,
-      mesoGridSize=50,mesoCenterDensity=1000,mesoAlpha = 1.1,mesoBeta=0.05,mesoNdiff = 1,mesoTimeSteps = 5,
-      macroMesoAlphaUpdateMax = 0.1,macroMesoBetaUpdateMax = 0.01,mesoMacroCongestionCost = 1.0,mesoMacroDecayUpdateMax = 10.0
-    )
-
-    val res = model.modelRun
-
-    //println(res)
-    //println(res.mesoMorans.sliding(n,n).mkString("\n"))
-    // deltapops <-> NaNs ?
-    log(res.macroPopulations.sliding(n,n).mkString("\n"))
-    log(res.mesoMorans.sliding(n,n).mkString("\n"))
-    //log(res.mesoCongestedFlows.sliding(n,n).mkString("\n"))
-  }
-
-
-  def testInnovation(): Unit = {
-    //val pop = new File("data/processed/FR_pops.csv")
-    val pop = new File("data/processed/CN_pops.csv")
-    //val dists = new File("data/processed/FR_dist.csv")
-    val dists = new File("data/processed/CN_dist.csv")
-    //val dates = new File("data/processed/FR_dates.csv")
-    val dates = new File("data/processed/CN_dates.csv")
-    val rng = new scala.util.Random
-    //val model = Innovation(pop,dists,dates,rng.nextInt,0.02,0.0001,200.0,50.0)
[…]

### 2019-09-26 Models/urbangrowth/src/main/scala/urbangrowth/utils/math/MatrixUtils.scala

@@ -1,83 +1,0 @@
-
-package urbangrowth.utils.math
-
-import Jama.Matrix
-
-import urbangrowth._
-
-import scala.util.Random
-
-
-object MatrixUtils {
-
-
-  def rowMatrix(d:Array[Double]): Matrix = {
-    new Matrix(d,1)
-  }
-
-  def colMatrix(d:Array[Double]): Matrix = {
-    new Matrix(d,d.length)
-  }
-
-
-
-  def diag(m: Matrix): Matrix = {
-    val n = m.getRowDimension()
-    val res = Matrix.identity(n, n)
-    for (i <- 0 to n - 1) {
-      res.set(i, i, m.get(i, 0))
-    }
-    return res
-  }
-
-  def diag(d : Array[Double]): Matrix = {
-    val res = new Matrix(d.length,d.length,0.0)
-    for(i <- 0 to d.length - 1){res.set(i,i,d(i))}
-    res
-  }
-
-  def setDiag(m: Matrix, s: Double): Unit = {
-    val n = m.getRowDimension()
-    for (i <- 0 to n - 1) {
-      m.set(i, i, s)
-    }
-  }
-
-  def repRow(d: Array[Double]): Matrix = diag(d).times(new Matrix(d.size,d.size,1.0))
-
-  def repCol(d: Array[Double]): Matrix = (new Matrix(d.size,d.size,1.0)).times(diag(d))
-
-
-  // implicit conversions to and from matrices ?
-  //class MatrixDecorator
-
-
-  /**
-    * random distance matrix
-    * @param n
-    * @param worldSize
-    * @param rng
-    * @return
-    */
-  def randomDistanceMatrix(n: Int,worldSize: Double)(implicit rng: Random): Vector[Vector[Double]] = {
-    val xcoords = Vector.fill(n)(rng.nextDouble()*worldSize)
-    val ycoords = Vector.fill(n)(rng.nextDouble()*worldSize)
-
-    val flatres = for {
-      x1 <- xcoords.zip(ycoords)
-      x2 <- xcoords.zip(ycoords)
-    } yield math.sqrt(math.pow((x1._1-x2._1),2)+math.pow(x1._2-x2._2,2))
-
-    log("dmat flat : "+flatres.size)
-
-    //flatres.sliding(n).toVector
-    flatres.sliding(n,n).toVector
-  }
-
-
-
-
-
-
-}
-

### 2019-09-26 Docs/Communications/2019/CCS2019/Abstract/ccs2019.tex
skeleton_slides

--- 2019-04-30 Docs/Communications/2019/CCS2019/ccs2019.tex
@@
 \affil[2]{\footnotesize CASA, UCL, London, UK}
-\affil[3]{\footnotesize UMR CNRS 8504 Géographie-cités, Paris, France}
+\affil[3]{\footnotesize UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
 \date{\vspace{-5ex}}
@@
 \noindent
+
+% keywords
+% Systems of cities
+%Multi-scalar model
+%Spatial interaction
+%Reaction-diffusion
 
@@
 More precisely, urban areas viewed as a population grid are embedded into the macroscopic interaction model.
-To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved at a given speed conditionally to the population variation; (iv) changes in urban form influence macroscopic interaction range (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.
+To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved with the reaction-diffusion models at a given speed conditionally to the population variations; (iv) changes in urban form influence macroscopic interaction ranges (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.
 
@@
 Parameter space is explored with the OpenMOLE model exploration software \cite{reuillon2013openmole}, eased by the implementation of the model in scala \cite{model}.
-First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario
+First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario (negative feedback of population growth on diffusion) to a ``sprawl'' scenario (positive feedback).
+Similarly, mesoscopic urban form indicators are significantly influenced by the coupling process.
+
+Further work will consist in more targeted simulation experiments, including specific exploration algorithms such as diversity search for model regimes \cite{reuillon2013openmole}, to test the model as a proof-of-concept of models for policies.
+Such a model can also be calibrated on real city systems and urban form trajectories, to extrapolate coupling parameters that would be difficult to obtain otherwise.
+Our contribution is thus a first step towards multi-scalar simulation models for systems of cities.
 
@@
 
-\footnotesize
+%\footnotesize
 

### 2019-09-26 Docs/Communications/2019/CCS2019/Communication/slides.tex
skeleton_slides

\input{header_slides.tex}

\begin{document}

\title[Multiscale urban growth]{A multi-scalar model for system of cities}
\author[Raimbault]{J.~Raimbault$^{1,2,3\ast}$\\\medskip
$^{\ast}$\texttt{j.raimbault@ucl.ac.uk}
}

\institute[UCL]{$^{1}$CASA, UCL\\
$^{2}$UPS CNRS 3611 Complex Systems Institute Paris\\
$^{3}$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}

\date[1st October 2019]{CCS 2019\\
Urban Complexity\\
October 1st 2019
}

\frame{\maketitle}

% keywords Systems of cities Multi-scalar model Spatial interaction Reaction-diffusion

\section{Introduction}

\sframe{Modeling urban growth}{

%The modeling of urban growth is a crucial issue for the design of sustainable territorial policies, through the understanding of past urbanization processes and the forecasting of future urban trajectories.

}

\sframe{Growth models at the mesoscopic scale}{

%Several models have been proposed at different scales and integrating different dimensions of urban systems, such as land-use transport interaction models \cite{wegener2004land} 
%At the scale of a metropolitan area, Land-use Transport Interaction models \cite{wegener2004land} are for example a privileged tool to anticipate the answer of spatial distributions of activities (mostly residential location and economic activities) to an evolution of the accessibility landscape permitted by new transportation infrastructures. At the same scale, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}.

}

\sframe{Interaction models for systems of cities}{

% or systems of cities models \cite{pumain2017urban}.
% At the smaller scale of the system of cities, macroscopic models of urban growth have focused on reproducing the distribution of city sizes, either through economic processes as e.g. \cite{gabaix1999zipf}, or from a geographical point of view focusing on interactions between cities \cite{favaro2011gibrat}.

}

\sframe{Towards multi-scalar models}{

%While multi-scalar models are recognized as crucial for the study of such systems \cite{Rozenblat2018}, they remain in practice unexplored.
%Territorial dynamics, and more particularly urban dynamics, have according to \cite{pumain1997pour} an intrinsic multi-scalar nature, with successive autonomous levels of emergence from individual microscopic agents to the mesoscopic scale of the city and the macroscopic scale of the system of cities. Furthermore, the need for sustainable territorial policies would imply the construction of multi-scalar models to take into account issues associated to each relevant scale \cite{Rozenblat2018}.

}

\sframe{Strong coupling and multi-scalarity}{

% on model coupling: example diagram Mike flooding

}

\sframe{Research objective}{

%This contribution contributes to that open question by introducing a multi-scale model of urban growth which focuses on the spatial structure of processes rather than on their multi-dimensionality. Therefore, we take into account only population variables, but both at the macroscopic scale of the system of cities in the legacy of \cite{pumain2017urban} and at the mesoscopic scale of the metropolitan area with an urban morphogenesis model. The coupling of these scales is a crucial novel feature of our model. We describe in the following stylized facts justifying the approach, describe the model, and summarize preliminary results from its exploration and calibration.
%This contribution introduces a parsimonious multi-scalar model for systems of cities, based on simple dimensions (mainly populations) with stylized processes, but yielding an effective strong coupling between the metropolitan mesoscopic scale and the macroscopic scale of the system of cities.

}

\section{Model description}

\sframe{Model Rationale}{

%The model couples the spatial interaction model of \cite{raimbault2018indirect} for the macro scale with the reaction-diffusion model for urban form studied by \cite{raimbault2018calibration}. More precisely, urban areas viewed as a population grid are embedded into the macroscopic interaction model. To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved with the reaction-diffusion models at a given speed conditionally to the population variations; (iv) changes in urban form influence macroscopic interaction ranges (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.

}

\sframe{Model formalization}{

}

\section{Results}

\sframe{Implementation}{

% Parameter space is explored with the OpenMOLE model exploration software \cite{reuillon2013openmole}, eased by the implementation of the model in scala \cite{model}.

% + pub openmole

}

\sframe{Synthetic configuration}{

% The model is applied on synthetic systems of cities typical of a continental range (500km, hierarchy around 1, 20 cities), with initial local population grid configurations as monocentric.

}

\sframe{Statistical consistency}{

% one factor sampling

}

\sframe{Impact of policy parameters}{

% First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario (negative feedback of population growth on diffusion) to a ``sprawl'' scenario (positive feedback). Similarly, mesoscopic urban form indicators are significantly influenced by the coupling process.
% ! abstract is wrong : say it !

}

\section{Discussion}

\sframe{Discussion}{

%Further work will consist in more targeted simulation experiments, including specific exploration algorithms such as diversity search for model regimes \cite{reuillon2013openmole}, to test the model as a proof-of-concept of models for policies. Such a model can also be calibrated on real city systems and urban form trajectories, to extrapolate coupling parameters that would be difficult to obtain otherwise.

}

\sframe{Conclusion}{

% Our contribution is thus a first step towards multi-scalar simulation models for systems of cities.

$\rightarrow$

$\rightarrow$

\bigskip
\bigskip

\textbf{Open repository for model and results at}

\texttt{https://github.com/JusteRaimbault/UrbanGrowth}

\textbf{Simulation data at}

\texttt{}% dataverse

\bigskip

\textbf{Acknowledgments}: thanks to the \textit{European Grid Infrastructure} for access to the infrastructure.

}

\sframe{Reserve slides}{

\centering

\Large

\textbf{Reserve Slides}

}

%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{biblio}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2019-09-29 Docs/Communications/2019/CCS2019/Communication/slides.tex
slides_v0

@@
 
+% Q : integrate macro indics ? bof, not used the more elaborated ones
+
+% striking image intro
+
+% history of Singapore: emerged within a multi-scale context - and then local growth (strongly shaped by its interactions at many scales)
+
 }
@@
 At the same scale, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}.
+
+\begin{columns}
+	\begin{column}{0.5\linewidth}
+
+		% includegraphics
+
+		% Nedum model ?
+
+		\includegraphics[width=\linewidth]{figures/nedum.png}
+
+		\medskip
+
+		\textit{Land-use transport models}
+
+		\cite{viguie2014downscaling}
+
+	\end{column}
+	\begin{column}{0.5\linewidth}
+
+		\begin{center}
+	\includegraphics[width=0.85\textwidth]{figures/intro_RBD_lattice.png}
+	\end{center}
+
+	%\footnotesize
+\textit{Hybrid urban morphogenesis model}
+
+	\cite{raimbault2014hybrid}
+
+ %Raimbault, J., Banos, A., \& Doursat, R.
+(2014, June).
+A Hybrid Network/Grid Model of Urban Morphogenesis and Optimization.
+In 4th International Conference on Complex Systems and Applications (pp.
+51-60).
+
+	\end{column}
+\end{columns}
 
@@
 
+\begin{center}
+\includegraphics[width=0.55\textwidth]{figures/simpop1.png}
+\includegraphics[width=0.42\textwidth]{figures/setup_synth_1_tick100.png}
+\end{center}
+
+\medskip
+
+\textit{The series of Simpop models: from Simpop1 \cite{sanders1997simpop} to SimpopNet \cite{schmitt2014modelisation}}
+
 }
@@
 
+% images scales Denise
+
+\begin{center}
+\includegraphics[height=0.65\textheight]{figures/evoltheory_scales}
+\end{center}
+
+\textit{Scales of urban systems of systems described by the evolutionary urban theory} \cite{pumain2018evolutionary}
+
+%\textit{Systems of cities as co-evolutive systems in which interactions are crucial}
+%\cite{pumain1997pour}
+%\cite{pumain2008socio}
+%\cite{pumain2018evolutionary}
+
 }
@@
 % on model coupling: example diagram Mike flooding
+
+\justify
+
+$\rightarrow$ Weak inter-scale coupling, such as progressive resolution for land-use model, does not considers emergence and autonomous scales
+
+\bigskip
+
+$\rightarrow$ An integrated model, or strongly coupled, would be \textit{a new model extending the two coupled models in the sense that it includes them in some parameter settings or limit conditions}
+
+\bigskip
+
+\textbf{\textit{Urban multi-scalar complexity must be capture by a strongly coupled model}}
 
@@
 
+$\rightarrow$ no strongly coupled multi-scalar model in the literature
+
+\medskip
+
+$\rightarrow$ need to however consider ``simple'' models to be able to understand their behavior and extract knowledge from them
+
+\bigskip
+\bigskip
+
+\textbf{Research objective: }
+
+\medskip
+
+\textit{Investigate a strong coupling of a simple urban system interaction network at the macroscopic scale with an urban morphogenesis model at the mesoscopic scale}
+
 }
@@
 
-\sframe{Model Rationale}{
+\sframe{Model rationale}{
+
+\centering
+
+\includegraphics{figures/multiscale_morph.pdf}
+
+}
+
+\sframe{Model rationale}{
 
@@
 
+\textbf{Model scales and objects: }
+
+\begin{itemize}
+	\item Cities as agents at the macroscopic scale
+	\item Cities as population density grids at the mesoscopic scale
+\end{itemize}
+
+\medskip
+
+\textbf{Processes: }
+
+\begin{itemize}
+	\item Interaction flows between cities and endogenous growth
+	\item Urban sprawl and aggregation within urban areas
+	\item \textbf{Downward feedback} Adaptation of urban form parameters to population and accessibility growth (``policy response'')
+	\item \textbf{Upward feedback} Adaptation of city interaction range depending on internal performance (congested flows)
+\end{itemize}
+
 }
@@
 \sframe{Model formalization}{
+
+At each time step:
+
+\medskip
+
+\begin{enumerate}
+	\item Macroscopic population are evolved with the interaction model
+	\item Population and accessibility differences modify the mesoscopic parameter of diffusion (sprawl) and aggregation (metropolization) \textbf{\textit{(downward feedback)}}
+	\item Urban forms are evolved for each city at the mesoscopic scale to match population increases
+	\item City interaction ranges are modified according to the internal performance of each city (congested flows) \textbf{\textit{(upward feedback)}}
+\end{enumerate}
+
+}
+
+\sframe{Macroscopic model}{
+
+% rq : does the macro mmodel captures many rank clock inversions ? kind of given variations of the rank correlation (cite macrocoevol)
+
+% + detail indicators here
+
+\justify
+
+\textit{System of cities interaction model including network evolution; production of multiple co-evolution regimes and calibration for France.}
+
+\cite{raimbault2018indirect}
+
+\medskip
+
+\begin{center}
+\includegraphics[width=0.72\textwidth]{figures/macrocoevol_en.png}
+\end{center}
+
+%\nocite{raimbault2019modeling}
+
+%\bigskip
+%\tiny
+
+%Raimbault, J. (2018). Indirect evidence of network effects in a system of cities. Environment and Planning B: Urban Analytics and City Science, 2399808318774335.
+
+%\smallskip
+
+%Raimbault, J. (2019). Modeling the co-evolution of cities and networks. In Niel, Z., Rozenblat, C., eds. \textit{Handbook of Cities and Networks}, Edwar Elgar Publishing, \textit{in press}.
+
+\footnotesize
+\textit{Indicators at the macroscopic scale: distributions of population, accessibilities, centralities (summarized by average, hierarchy, entropy)}
+
+}
+
+\sframe{Mesoscopic model}{
+
+% + detail urban form indicators
+
+\textit{Aggregation-diffusion model at the mescoscopic scale, covering most existing urban forms in Europe} \cite{raimbault2018calibration}
+
+\medskip
+
+\begin{center}
+\includegraphics[height=0.6\textheight]{figures/density_Fig2}
+\end{center}
+
+%\footnotesize\textit{Examples of generated territorial shapes}
+
+\footnotesize 
+\textit{Indicators at the mesoscopic scale: urban form captured by Moran index, average distance, hierarchy, entropy}
+
+}
+
+\sframe{Model parameters}{
+
+\centering
+
+		\begin{tabular}{|c|c|c|c|}
+	\hline %\tiny
+		Type & Parameter & Process & Range \\\hline
+		\multirow{4}{*}{Macro} & $g_i \textrm{ = } g_0$ & Endogenous growth & $\left[0 ; 0.05 \right]$ \\
+		& $w_i \textrm{ = } w_G$ & Interactions weight & $\left[0 ; 0.01 \right]$  \\
+		& $\gamma_i \textrm{ = } \gamma_G$ & Interactions hierarchy & $\left[0 ; 5 \right]$  \\
+		& $d_i$ & Interactions decay & $\left[0 ; 1000 \right]$  \\ \hline
+		\multirow{4}{*}{Meso} & $\alpha_i$ & Aggregation & $\left[0 ; 5 \right]$  \\
+		& $\beta_i$ & Diffusion & $\left[0 ; 0.1 \right]$  \\
+		& $t_m$ & Urban growth speed & $\{1 ; 20 \}$  \\
+		& $n_d$ & Diffusion steps &  $\left[1 ; 5 \right]$ \\ \hline
+		\multirow{4}{*}{Multiscale} & $\delta\alpha$ & Metropolization (downward) & $\left[-0.2 ; 0.2 \right]$  \\
+		& $\delta\beta$ & Sprawl (downward) & $\left[-0.2 ; 0.2 \right]$ \\
+		& $\delta d$ & Efficiency (upward) &$\left[-0.2 ; 0.2 \right]$ \\
+		& $\lambda$ & Congestion cost (upward) & $\left[0 ; 10 \right]$ \\\hline
+	\end{tabular}
 
@@
 
+%\footnotesize
+\justify
+
+\textit{Performance constraints: simulate $N$ mesoscopic morphogenesis models in parallel (macroscopic interactions are efficient as based on matrices)}
+
+\medskip
+
+$\rightarrow$ model implemented in \texttt{scala} and integrated within a broader library (including implementations of \cite{raimbault2018indirect} \cite{raimbault2018calibration} \cite{favaro2011gibrat} \cite{cottineau2015modular})
+
+\bigskip
+
+\textit{Large number of parameters and output indicators}
+
+\medskip
+
+$\rightarrow$ integration into the OpenMOLE model exploration open source software \cite{reuillon2013openmole}
+
+\begin{center}
+\includegraphics[width=0.55\textwidth]{figures/openmole.png}
+\end{center}
+
+(X year of computation for the results presented here)
+
 }
@@
 % The model is applied on synthetic systems of cities typical of a continental range (500km, hierarchy around 1, 20 cities), with initial local population grid configurations as monocentric.
+
+Model applied on synthetic systems of cities:
+
+\begin{itemize}
+	\item random positions and rank-size hierarchy ($\alpha \textrm{ = } 1.0$ and $P_0 \textrm{ = } 100,000$)
+	\item countrywide urban system scale: 500km and 20 cities
+	\item initial population grids as monocentric (grid of size 50 and center cell density 1000 units)
+	\item simulated for 20 macroscopic time steps (order of magnitude of half a century)
+\end{itemize}
 
@@
 
+\textit{For all indicators, median sharpe ratios computing for a parameter point across repetitions are all larger than 1.6}
+%(min around 1 for 3 indicators, above 5 for 2 and around 0.5 for 2)}
+
+\medskip
+
+\centering
+
+\includegraphics[width=0.8\textwidth]{figures/allindics_hist.png}
+
 }
 
-\sframe{Impact of policy parameters}{
+\sframe{Model exploration}{
 
@@
 
+\begin{center}
+\includegraphics[width=\textwidth]{figures/onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
+\end{center}
+
+\footnotesize
+\textit{U-shape behavior of both macroscopic and mesoscopic indicators as a function of $\delta d$}
+
 }
+
+\sframe{Model exploration}{
+
+% what is interesting: that macro -> meso -> macro has indeed a role (or the other way around)
+
+\begin{center}
+\includegraphics[height=0.8\textheight]{figures/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
+\end{center}
+
+\footnotesize
+\textit{Non-trivial influence of coupled feedbacks on the different scales}
+
+}
+
+\sframe{Impact of policy parameters}{
+
+\begin{center}
+\includegraphics[height=0.7\textheight]{figures/deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+\end{center}
+
+\textit{Mesoscopic centralization appears at a $\delta \beta$ critical value for low $\delta \alpha$; influenced by upward feedback}
+
+}
+
+\sframe{Impact of policy parameters}{
+
+\begin{center}
+\includegraphics[height=0.7\textheight]{figures/deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+\end{center}
+
+\textit{Mesoscopic hierarchy has a U-shape of $\delta \beta$ in negative $\delta \alpha$, but has a plateau for positive values} 
+%$\rightarrow$ sprawl is needed in a sense
+
+}
+
+% TODO 
+%\sframe{Calibration}{
+%}
 
@@
 Such a model can also be calibrated on real city systems and urban form trajectories, to extrapolate coupling parameters that would be difficult to obtain otherwise.
+\justify
+
+\textbf{Theoretical and practical implications}
+
+\medskip
+
+$\rightarrow$ model effectively captures an interaction between downward and upward feedback: weak emergence \cite{bedau2002downward}
+
+\medskip
+
+$\rightarrow$ coupling ``simple three parameters models'' yield a complicated and complex simulation model: necessity of complexity and simulation models to understand urban complexity? 
+% 'suggest in a way failure of reductionist empistemology (in the sense of a la Barthelemy) - at least for grabbing the multi-scalar nature of systems' (seems a tautology though
+
+\medskip
+
+$\rightarrow$ progressive integration towards models for policy?
+
+\bigskip
+
+\textbf{Developments}
+
+$\rightarrow$ diversity search algorithm to find e.g. regimes with the strongest effect of feedback
+
+\medskip
+
+$\rightarrow$ parametrization on real systems; possibly calibration
+
+(see \cite{raimbault2019ilus} forthcoming presentation at ILUS)
 
@@
 
-$\rightarrow$
+$\rightarrow$ A first step towards \textit{strongly} multi-scalar models to capture urban complexity towards policy models
 
-$\rightarrow$
+\medskip
+
+$\rightarrow$ Towards integrative models and theories for urban systems \cite{raimbault2019methods}
 
@@
 
-\textbf{Open repository for model and results at}
+\footnotesize
 
-\texttt{https://github.com/JusteRaimbault/UrbanGrowth}
+\textbf{Open repositories for}
 
-\textbf{Simulation data at}
-
-\texttt{}% dataverse
+\begin{itemize}
+ \item Model: \texttt{https://github.com/JusteRaimbault/UrbanGrowth-model}
+ \item Project: \texttt{https://github.com/JusteRaimbault/UrbanGrowth}
+ \item Simulation data: \texttt{}% dataverse
+\end{itemize}
 
@@
 
+\sframe{Mesoscopic model}{
+
+$\rightarrow$ Grid world with cell populations $\left(P_i\left(t\right)\right)_{1\leq i\leq N^2}$.
+
+\bigskip
+
+$\rightarrow$ At each time step:
+
+\begin{enumerate}
+\item Population growth with exogenous rate $N_G$, attributed independently to a cell following a preferential attachment of strength $\alpha$
+%\begin{equation}
+%\Pb{P_i(t+1)=P_i(t)+1|P(t+1)=P(t)+1}=\frac{(P_i(t)/P(t))^{\alpha}}{\sum(P_i(t)/P(t))^{\alpha}}
+%\end{equation}
+%The attribution being uniformly drawn if all population are equal to 0.
+\item Population is diffused $n_d$ times with strength $\beta$
+\end{enumerate}
+
+\bigskip
+
+$\rightarrow$ Stopping criterion: fixed maximal population $P_m$.
+
+%To avoid bord effects such as reflecting diffusion waves, border cells diffuse their due proportion outside of the world, implying that the total population at time $t$ is strictly smaller than $N_G\cdot t$.
+
+\bigskip
+
+$\rightarrow$ Output measured by morphological indicators: Moran index, average distance, rank-size hierarchy, entropy.
+
+}
+
+\sframe{Model formalization}{
+
+\textit{Macroscopic interactions}
+
+\medskip
+
+\begin{equation}
+		P_i\left( t \textrm{ + } 1 \right) \textrm{ = } P_i\left(t\right) \left(1 \textrm{ + } \Delta t \cdot \left(g_i \textrm{ + } \frac{w_i}{N} \cdot \sum_j \frac{V_{ij}}{<V_{ij}>} \right) \right)
+	\end{equation}
+
+where the gravity interaction potential is given by
+
+\begin{equation}
+		V_{ij} \textrm{ = } \left(\frac{P_i P_j}{(\sum_k p_k)^2}\right)^{\gamma_G} \cdot \exp \left(- \frac{d_{ij}}{d_i} \right)
+\end{equation}
+
+\medskip
+
+Accessibility given by
+
+\begin{equation}
+	Z_i \textrm{ = } \sum_j \frac{P_j}{\sum_k P_k} \cdot \exp( - d_{ij} / d_i)
+\end{equation}
+
+}
+
+\sframe{Model formalization}{
+
+\begin{itemize}
+	\item Mesoscopic growth rate
+			    $N_G^{\left(i\right)} \left(t \textrm{ + } 1\right) = \Delta P_i / t_m$
+	\item Sprawl parameter (population pressure)
+		\begin{equation}
+			\beta_i \left(t \textrm{ + } 1\right) \textrm{ = } \beta_i \left(t\right) \cdot \left(1 \textrm{ + } \delta\beta \cdot \frac{\Delta P_i (t)}{\max_k  \Delta P_k \left(t\right)}\right)
+		\end{equation}
+	\item Aggregation parameter (metropolization process)
+	    \begin{equation}
+			\alpha_i \left(t \textrm{ + } 1\right) \textrm{ = } \alpha_i \left(t\right) \cdot \left(1 \textrm{ + } \delta\alpha \cdot \frac{\Delta Z_i (t)}{\max_k  \Delta Z_k \left(t\right)}\right)
+		\end{equation}
+
+\end{itemize}
+
+}
+
+\sframe{Model formalization}{
+
+Congested flows within the mesoscopic zones:
+
+\begin{equation}
+		U_i \textrm{ = } \sum_{kl} \left( \frac{P_k P_l}{P^2} \cdot \frac{1}{d_{kl}} - \lambda \left(\frac{P_k P_l}{P^2} \cdot \frac{1}{d_{kl}}\right)^2 \right)
+\end{equation}
+
+\bigskip
+
+Update the macroscopic interaction distance accordingly
+
+\begin{equation}
+	d_i \left(t \textrm{ + } 1\right) \textrm{ = } d_i \left(t\right) \left( 1 \textrm{ + } \delta d \cdot \frac{U_i}{\max_k \left|U_k\right|} \right)
+\end{equation}
+
+}
+
+\sframe{Morphological indicators}{
+
+\begin{enumerate}
+\item Rank-size slope $\gamma$, given by $\ln \left( P_{\tilde{i}}/P_0\right) \sim k + \gamma\cdot \ln \left(\tilde{i}/i_0\right)$ where $\tilde{i}$ are the indexes of the distribution sorted in decreasing order.
+\item Entropy of the distribution:
+\begin{equation}
+\mathcal{E} = \sum_{i=1}^{M}\frac{P_i}{P}\cdot \ln{\frac{P_i}{P}}
+\end{equation}
+$\mathcal{E}=0$ means that all the population is in one cell whereas $\mathcal{E}=0$ means that the population is uniformly distributed.
+\item Spatial-autocorrelation given by Moran index, with simple spatial weights given by $w_{ij} = 1/d_{ij}$
+\[
+I = M \cdot \frac{\sum_{i\neq j} w_{ij} \left(P_i - \bar{P}\right)\cdot\left(P_j - \bar{P}\right)}{\sum_{i\neq j} w_{ij} \sum_{i}{\left( P_i - \bar{P}\right)}^2}
+\]
+\item Mean distance between individuals
+\[
+\bar{d} = \frac{1}{d_M}\cdot \sum_{i<j} \frac{P_i P_j}{P^2} \cdot d_{ij}
+\]
+where $d_M$ is a normalisation constant
+\end{enumerate}
+
+}
+
 %%%%%%%%%%%%%%%%%%%%%

### 2019-09-29 Models/Multiscale/multiscale.R

@@ -3,6 +3,7 @@ setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/urbangrowth/openmole'))
 
 library(dplyr)
 library(ggplot2)
+library(reshape2)
 
 source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
@@ -10,18 +11,23 @@ source(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/Analysis/functions.R'))
 
 #resPrefix = '20190429_193158_MULTISCALE_GRID_GRID'
 #resPrefix = '20190506_135221_MULTISCALE_TARGETEDGRID_GRID'
-resPrefix = '20190919_161009_MULTISCALE_TARGETEDGRID_GRID'
+#resPrefix = '20190919_161009_MULTISCALE_TARGETEDGRID_GRID';filtered=T;fulltrajs=T;oneFactor=F
 #resPrefix = '20190919_141607_MULTISCALE_TARGETEDGRID_GRID'
 #resPrefix = '20190926_114834_MULTISCALE_LHS_TEST'
+#resPrefix = '20190926_173110_MULTISCALE_ONEFACTOR_GRID';filtered=F;fulltrajs=F;oneFactor=T
+resPrefix = '20190927_131754_MULTISCALE_TARGETEDGRID_GRID';filtered=F;fulltrajs=F;oneFactor=F
+
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/Multiscale/',resPrefix,'/');dir.create(resdir)
 
-#res <- as.tbl(read.csv(paste0('exploration/',resPrefix,'.csv')))
-#res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'.csv'),stringsAsFactors = F))
-#res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'_sample.csv'),stringsAsFactors = F))
-res <- as.tbl(read.csv(paste0(resdir,'data/',resPrefix,'_filtered.csv'),stringsAsFactors = F))
-
-params = c("macroGrowthRate","macroInteractionDecay","macroInteractionGamma",
-           "mesoAlpha","mesoBeta","mesoTimeSteps",
+if(!filtered){
[…]

### 2019-10-01 Docs/Communications/2019/CCS2019/Communication/slides.tex
calib_results

@@
 
-$\rightarrow$ model implemented in \texttt{scala} and integrated within a broader library (including implementations of \cite{raimbault2018indirect} \cite{raimbault2018calibration} \cite{favaro2011gibrat} \cite{cottineau2015modular})
+$\rightarrow$ model implemented in \texttt{scala} and integrated within a broader library (including implementations of \cite{raimbault2018indirect} \cite{raimbault2018calibration}
+
+\cite{favaro2011gibrat} \cite{cottineau2015modular})
 
@@
 \begin{center}
-\includegraphics[width=0.55\textwidth]{figures/openmole.png}
+\includegraphics[height=0.13\textheight]{figures/iconOM.png}
+\includegraphics[height=0.13\textheight]{figures/openmole.png}
 \end{center}
 
-(X year of computation for the results presented here)
+%(X year of computation for the results presented here)
+
+\footnotesize
+\textit{Enables seamlessly (i) model embedding; (ii) access to HPC resources; (iii) exploration and optimization algorithms}
+
+\medskip
+
+\textbf{Come to the satellite \textit{New Methods and Epistemologies to Explore Simulation Models} tomorrow afternoon in LHN-TR+05}
 
@@
 Model applied on synthetic systems of cities:
+
+\medskip
 
@@
 
-\textit{For all indicators, median sharpe ratios computing for a parameter point across repetitions are all larger than 1.6}
 %(min around 1 for 3 indicators, above 5 for 2 and around 0.5 for 2)}
@@
 
-\centering
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/allindics_hist.png}
+\end{center}
 
-\includegraphics[width=0.8\textwidth]{figures/allindics_hist.png}
+\textit{For all indicators, median sharpe ratios computing for a parameter point across repetitions are all larger than 1.6}
 
@@
 \begin{center}
-\includegraphics[height=0.8\textheight]{figures/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
+\includegraphics[height=0.75\textheight]{figures/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
 \end{center}
 
-\footnotesize
+%\footnotesize
 \textit{Non-trivial influence of coupled feedbacks on the different scales}
@@
 
-% TODO 
-%\sframe{Calibration}{
-%}
+\sframe{Multiscale optimization}{
+
+\begin{center}
+	\includegraphics[width=0.8\textheight]{figures/pareto_macroHierarchy-mesoHierarchy.png}
+\end{center}
+
+\textit{Pareto front obtained with Genetic Algorithm optimization for two contradictory objective of macrosopic and mesoscopic hierarchies}
+
+}
 

### 2019-10-01 Docs/Communications/2019/CCS2019/Communication/slides.tex
slides_just_in_time

@@
 
-\sframe{Modeling urban growth}{
+\sframe{Urban growth across time and scales}{
 
@@
 % history of Singapore: emerged within a multi-scale context - and then local growth (strongly shaped by its interactions at many scales)
+
+\centering
+
+\includegraphics[width=0.48\textwidth]{figures/ConceptPlan1971.jpg}\hspace{0.1cm}
+\includegraphics[width=0.48\textwidth]{figures/ConceptPlan1991.jpg}\\
+\includegraphics[width=0.48\textwidth]{figures/2001-CONCEPTPLAN.jpg}\hspace{0.1cm}
+\includegraphics[width=0.48\textwidth]{figures/conceptplan2011.jpeg}
 
@@
 
+$\rightarrow$ Distinct processes and objects within each different scale
+
+\bigskip
+
 $\rightarrow$ Weak inter-scale coupling, such as progressive resolution for land-use model, does not considers emergence and autonomous scales
@@
 
-\textbf{\textit{Urban multi-scalar complexity must be capture by a strongly coupled model}}
+\textbf{\textit{Urban multi-scalar complexity must be captured by a strongly coupled model}}
 
@@
 
-$\rightarrow$ no strongly coupled multi-scalar model in the literature
+$\rightarrow$ no strongly coupled multi-scalar model in the literature (few examples such as \cite{murcio2015urban} but without distinct ontologies)
 
@@
  \item Project: \texttt{https://github.com/JusteRaimbault/UrbanGrowth}
- \item Simulation data: \texttt{}% dataverse
+ \item Simulation data: \texttt{https://doi.org/10.7910/DVN/IRHMQK}% dataverse
 \end{itemize}

### 2019-11-20 Models/MesoCalibration/runRange.sh
.

@@ -1,1 +1,1 @@
-cat $1 | awk '{print "rm Calib.oms; echo \"val areaind="$1"\" >> Calib.oms; cat MesoCalibration.oms >> Calib.oms; openmole -p density_2.12-1.0.jar --script Calib.oms --password-file omlpsswd"}'| sh
+cat $1 | awk '{print "rm Calib.oms; echo \"val areaind="$1"\" >> Calib.oms; cat MesoCalibration.oms >> Calib.oms; openmole -p density_2.12-1.0.jar --mem 30G --script Calib.oms --password-file omlpsswd"}'| sh

### 2019-11-20 Docs/Communications/2019/ILUS2019/Communication/ilus2019.tex

\input{header_slides}

\begin{document}

\title[Models of urban morphogenesis]{Models of urban morphogenesis to link urban form and function}
\author[J. Raimbault]{J.~Raimbault$^{1,2,3\ast}$\\\medskip
$^{\ast}$\texttt{j.raimbault@ucl.ac.uk}
}

\institute[UCL]{$^{1}$CASA, UCL\\
$^{2}$UPS CNRS 3611 Complex Systems Institute Paris\\
$^{3}$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}

\date[15th November 2019]{TQG Debates 2019\\
3.1: Fractals and Multi-fractals\\
November 15th 2019
}

\frame{\maketitle}

% \textbf{Keywords: }\textit{Urban growth; Reaction-diffusion; World urban areas; Model calibration}

%  Understanding drivers of urban growth and land-use change at the metropolitan scale often requires to isolate and estimate stylized processes of urban evolution. Several land-use change and urban growth models for example based on cellular automata have been introduced in the literature, but their lack of generality and need for rich data may become a shortcoming. We propose here to apply and calibrate a simple model of urban growth based on reaction-diffusion processes for population density, providing a rather general insight into world urban areas from simple data. This four-parameter model introduced by \cite{raimbault2018calibration} combines aggregation forces (positive externalities) with dispersal forces (negative externalities) to capture dynamics of urban form. The model was previously only simulated on synthetic initially empty spaces and statically calibrated by comparing the projection of final simulated configurations in a space of urban morphology indicators (Moran index, entropy, average distance, hierarchy) to real measures computed on a fixed-size moving window for Europe. We significantly extend these result by (i) applying the model on the 1000 largest urban areas of the Global Human Settlements database \citep{Florczyk2019ghs}; (ii) proceeding to a calibration on three successive time windows (1975-1990, 1990-2000 and 2000-2015) what provides dynamical values for parameters; (iii) using genetic optimization algorithms to minimise the distance on morphological indicators for each area. The computational cost being high, results are obtained by using the OpenMOLE model exploration software \citep{reuillon2013openmole}, which provides in a transparent way the optimization algorithms and the distribution of calibrations on a computation grid. We obtain therein comparable values for aggregation, diffusion and growth speed parameters on all these areas for the three periods given above. These values can be used for short terms projections, but also high-level policies by situating urban areas in the model phase diagrams and for example comparing them to more or less close phase transition parameter values. We also illustrate possible policy applications by relating estimated parameters to the EDGAR greenhouse gases emission database. This work thus shows how simple models can be useful for global comparisons, which are necessary to contextualize territorial policies.

\section{Introduction}

\sframe{Complex processes of Urban Morphogenesis}{

\centering

% center of Paris
%\includegraphics[width=0.88\textwidth]{figures/intro_paname}
%\includegraphics[width=0.9\textwidth]{figures/intro_bp}

\footnotesize\textit{Source: Geoportail}
}

\sframe{Defining Morphogenesis}{

% precise notions and defs

\justify

\textit{Proposition of an interdisciplinary definition}

\bigskip

\textbf{Meta-epistemological framework of imbricated notions:}

Self-organization $\supsetneq$ Morphogenesis $\supsetneq$ Autopoiesis $\supsetneq$ Life

\bigskip

\textbf{Properties:}

\begin{itemize}
\item Architecture links form and function
\item Emergence strength~\cite{bedau2002downward} increases with notion depth, as bifurcations~\cite{thom1974stabilite}
\end{itemize}

\bigskip

\textbf{Definition of Morphogenesis :} \textit{Emergence of the form and the function in a strongly coupled manner, producing an emergent architecture \cite{doursat2012morphogenetic}}

}

\sframe{Which models for Urban Morphogenesis ?}{

% why modeling, exemple of rbd
% -> champ ; difficultés ("verrous". rq : j'aime pas la metaphore du verrou, trop cadrant et suppose que deja construit et qu'il suffit d'ouvrir. demarche epistemo de co-evol des connaissances, ni inductif ni deductif. il faut contruire. horizons d'exploration plus appropriés. rejoint l'idee de faire rentrer dans "cadre analytique" : cases prédéfinies, alors qu'il s'agit au contraire de définir ces cases. -> relire hdr arnaud

\justify

\begin{columns}
\column{0.35\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/intro_RBD_lattice}\\
\footnotesize\textit{Example: a basic hybrid model based on elementary processes for density and network \cite{raimbault2014hybrid}}
\column{0.6\textwidth}
\justify

%\vspace{-1cm}

$\rightarrow$\textit{At the crossroad between Urban Simulation and Artificial Life, few models try to integrate and explain the link between Urban Form and Function}

\medskip

$\rightarrow$\textit{Importance of parcimonious, stylized models: modeling as a tool to understand processes}

\end{columns}

\bigskip

\textbf{Research Objective : } Explore simple models to capture morphogenesis based on abstract representation of urban processes; test their ability to reproduce existing urban systems.

}

\section{Reaction-diffusion model}

\sframe{A simple Reaction-diffusion model}{

% comment Arnaud : reaction-diffusion ?

% model rationale and processes

\justify

$\rightarrow$ Crucial role of the interplay between concentration forces and dispersion forces~\cite{fujita1996economics} in keeping Urban Systems at the border of chaos

\bigskip

$\rightarrow$ Potentiality of aggregation mechanisms (such as Simon model) to produce power laws \cite{dodds2017simon}

\bigskip

$\rightarrow$ Link with Reaction-diffusion approaches in Morphogenesis~\cite{turing1952chemical}

\bigskip

$\rightarrow$ Extension of a DLA-type model introduced by \cite{batty1991generating}, with simple abstract processes of population aggregation and diffusion

\bigskip
\bigskip

{\tiny

Raimbault, J. (2018). Calibration of a density-based model of urban morphogenesis. PloS one, 13(9), e0203516.
}

}

\sframe{Model Formalization}{

% model formalization and indicators

$\rightarrow$ Grid world with cell populations $\left(P_i\left(t\right)\right)_{1\leq i\leq N^2}$.

\bigskip

$\rightarrow$ At each time step:

\begin{enumerate}
\item Population growth with exogenous rate $N_G$, attributed independently to a cell following a preferential attachment of strength $\alpha$
%\begin{equation}
%\Pb{P_i(t+1)=P_i(t)+1|P(t+1)=P(t)+1}=\frac{(P_i(t)/P(t))^{\alpha}}{\sum(P_i(t)/P(t))^{\alpha}}
%\end{equation}
%The attribution being uniformly drawn if all population are equal to 0.
\item Population is diffused $n_d$ times with strength $\beta$
\end{enumerate}

\bigskip

$\rightarrow$ Stopping criterion: fixed maximal population $P_m$.

%To avoid bord effects such as reflecting diffusion waves, border cells diffuse their due proportion outside of the world, implying that the total population at time $t$ is strictly smaller than $N_G\cdot t$.

\bigskip

$\rightarrow$ Output measured by morphological indicators: Moran index, average distance, rank-size hierarchy, entropy.

}

\sframe{Generating Population Distributions}{

% examples : fig 2 of paper

\centering

\includegraphics[height=0.8\textheight]{figures/density_Fig2}

\footnotesize\textit{Examples of generated territorial shapes}

}

\sframe{Model behavior}{

\centering

\includegraphics[width=0.9\textwidth]{figures/density_Fig3}

\footnotesize\textit{Phase transitions of indicators unveiled by exploration of the parameter space (80000 parameter points, 10 repetitions each)}

}

\sframe{Path-dependence and frozen accidents}{

\centering

\includegraphics[width=0.8\textwidth]{figures/density_Fig4}

\footnotesize\textit{Illustration of path-dependence in a simplified one-dimensional version of the model: cell trajectories in time for 9 independent repetitions from the same initial configuration.}

}

\sframe{Empirical Data for Calibration}{

% comments Arnaud : Quel lien avec les slides avant et après ? "Real" : le terme me paraît discutable : données empiriques plutôt ?
%  Décrire les données car ici tu passes directement aux indicateurs

\begin{columns}
\column{0.6\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/density_indics_morpho_discrquantiles}
\column{0.3\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/density_cluster_pca_k5_morpho}\\
\includegraphics[width=\textwidth]{figures/density_cluster_map_k5_morpho}
\end{columns}

\justify

\footnotesize\textit{Computation of morphological indicators on population density data for Europe (shown here on France), morphological classification.}

}

\sframe{Model Calibration}{

\includegraphics[width=0.9\textwidth]{figures/density_synth}

\footnotesize\textit{Brute force calibration by exploring the parameter space. Reproduction of most existing configuration in the morphological sense (here in principal plan).}

}

\sframe{Model Targeted Exploration}{

\centering

\includegraphics[width=0.8\textwidth]{figures/density_Fig6}

\footnotesize\textit{Potentialities of targeted model explorations: here feasible space using Pattern Space Exploration algorithm \cite{10.1371/journal.pone.0138212}.}

}

\section{Co-evolution model}

\sframe{Including more complex processes ?}{

% transition : representation of territories

\textit{Which ontology to include more complex functional properties ?}

\medskip

$\rightarrow$ Territorial systems as the strong coupling between territories and (potential and realized) networks \cite{dupuy1987vers}.

\medskip

$\rightarrow$ Networks convey functional notions of centralities and accessibility, among others; have furthermore proper topological properties.

}

\section{Discussion}

\sframe{Discussion}{

\justify

\vspace{-1cm}

\textbf{Implications}

$\rightarrow$ This rather simple model reproduces most of existing urban forms in Europe for both population distribution and road network : which intrinsic dimension to the urban system and its morphological aspect ?

$\rightarrow$ Ability to reproduce static correlations and a variety of dynamical lagged correlation regimes suggests that the model captures some of the processes of co-evolution

% implications for morphogenesis ?

\bigskip

\textbf{Developments}

$\rightarrow$ Towards a dynamical calibration ? Need of dynamical data

$\rightarrow$ Investigate the link between spatial non-stationarity and non-ergodicity through simulation by the model

$\rightarrow$ Compare network generation in a ``fair'' way (correcting for additional parameters, open question for models of simulation)

}

\sframe{Towards policy applications}{

% more data

\justify

\textbf{More realistic models?}

\medskip

$\rightarrow$ Introducing more concrete ontologies, economic processes \\
\cite{bonin2014modelisation}, qualitative differentiation \cite{bonin2012modele}, governance processes \cite{le2015modeling}

\medskip

$\rightarrow$ Possible bridges with Land-use change models/Land-use Transport models \cite{wegener2004land}, with systems of cities models\\
 \cite{pumain2017urban}

\medskip

\textbf{More data-driven models?}

\medskip

$\rightarrow$ Work in progress: calibration of the reaction-diffusion model on world urban areas with the Global Human Settlements Layer database

\medskip

$\rightarrow$ Link with sustainability indicators: GHG emissions, economics, etc. \cite{raimbault2019multi}

\medskip

$\rightarrow$ Study models on hybrid synthetic data \cite{raimbault2018space}: systematic conclusions for policies

}

\sframe{Conclusion}{

\justify

$\rightarrow$ A novel model of urban morphogenesis at the mesoscopic scale systematically explored: \textbf{need for more coupling and comparison of models.}

\medskip

$\rightarrow$ At the macro scale of the system of cities? \textbf{Need for multi-scale models.}

\medskip

$\rightarrow$ With more refined urban characteristics and other dimensions ? \textbf{Need for more interdisciplinarity.}

\bigskip
\bigskip
\bigskip

\footnotesize{ - Code, data and results available at\\ \texttt{https://github.com/JusteRaimbault/CityNetwork}\\
- Acknowledgments: Thanks to the \textit{European Grid Infrastructure} and its \textit{National Grid Initiatives} (\textit{France-Grilles} in particular) to give the technical support and the infrastructure.

}

}

\section*{Reserve slides}

\sframe{Reserve slides}{

\centering

\Large

\textbf{Reserve Slides}

}

\sframe{Model classification : PDE}{

% derived PDE

The one-dimensional model verifies the PDE :

\begin{equation}\label{eq:pde}
\begin{split}
\delta t \cdot \frac{\partial p}{\partial t} \textrm{ = } \frac{N_G \cdot p^{\alpha}}{P_{\alpha}(t)} \textrm{ + } \frac{\alpha \beta \left(\alpha - 1\right) \delta x^2}{2}\cdot \frac{N_G \cdot p^{\alpha-2}}{P_{\alpha}\left(t\right)} \cdot \left(\frac{\partial p}{\partial x}\right)^2\\
\textrm{ + } \frac{\beta \delta x^2}{2} \cdot \frac{\partial^2 p}{\partial x^2} \cdot\left[ 1 \textrm{ + } \alpha \frac{N_G p^{\alpha - 1}}{P_{\alpha(t)}} \right]
\end{split}
\end{equation}

}

\sframe{Stationary behavior of 1D model}{
\centering
\includegraphics[width=\textwidth]{figures/density_stationary}
}

\sframe{Stationary behavior of 1D model}{
\centering
\includegraphics[width=0.48\textwidth]{figures/density_pmax_alpha}
\includegraphics[width=0.48\textwidth]{figures/density_pmax_logbeta}

}

\sframe{Morphological indicators}{

\begin{enumerate}
\item Rank-size slope $\gamma$, given by $\ln \left( P_{\tilde{i}}/P_0\right) \sim k \textrm{ + } \gamma\cdot \ln \left(\tilde{i}/i_0\right)$ where $\tilde{i}$ are the indexes of the distribution sorted in decreasing order.
\item Entropy of the distribution:
\begin{equation}
\mathcal{E} \textrm{ = } \sum_{i=1}^{M}\frac{P_i}{P}\cdot \ln{\frac{P_i}{P}}
\end{equation}
$\mathcal{E}\textrm{ = }0$ means that all the population is in one cell whereas $\mathcal{E}\textrm{ = }0$ means that the population is uniformly distributed.
\item Spatial-autocorrelation given by Moran index, with simple spatial weights given by $w_{ij} \textrm{ = } 1/d_{ij}$
\[
I \textrm{ = } M \cdot \frac{\sum_{i\neq j} w_{ij} \left(P_i - \bar{P}\right)\cdot\left(P_j - \bar{P}\right)}{\sum_{i\neq j} w_{ij} \sum_{i}{\left( P_i - \bar{P}\right)}^2}
\]
\item Mean distance between individuals
\[
\bar{d} \textrm{ = } \frac{1}{d_M}\cdot \sum_{i<j} \frac{P_i P_j}{P^2} \cdot d_{ij}
\]
where $d_M$ is a normalisation constant
\end{enumerate}

}

\sframe{Model behavior : Convergence}{

Large number of repetitions show good convergence properties

% hist examples

\includegraphics[width=0.5\textwidth]{figures/density_hist_moran}
\includegraphics[width=0.5\textwidth]{figures/density_hist_slope}

}

\sframe{Model behavior}{

\includegraphics[width=0.5\textwidth]{figures/density_pc_colalpha}
\includegraphics[width=0.5\textwidth]{figures/density_pc_colbeta}

}

\sframe{Empirical indicators computation}{

$\rightarrow$ Eurostat population density raster (100m, simplified at 500m resolution)

\medskip

$\rightarrow$ Overlapping (10km offset) squares of 50km side : equivalent to smoothing, removes window shape effect. Not very sensitive to window size (tested with 30km and 100km)

\medskip

$\rightarrow$ Indicators computed using Fast Fourier Transform Convolution

\medskip

$\rightarrow$ Classification using repeated k-means ; number of clusters taken at transition in clustering coefficient.

}

\sframe{Model calibration: all indicators}{

\centering
\includegraphics[width=0.65\textwidth]{figures/density_scatter}

}

%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{biblio}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\sframe{What is Morphogenesis ? Examples}{

% illustrations : ants, geomorphology, neurons, self-assembly ; ARBOTRON ; paper nature aile avion
% all from netlogo library ? would be nice illustration of generative nature

% remark : do not put classical biological example to show how it has percolated to other fields
%\vspace{-0.3cm}
%\includegraphics[width=\textwidth,height=0.82\textheight]{figures/intro_examples}

%\justify

%\vspace{-0.5cm}

%\footnotesize\textit{Sources (in order by column). Ants, Erosion, Game of Life: NetLogo Library; Arbotron \cite{jun2005formation}; Industrial design \cite{Aage:2017aa}; Swarm chemistry \cite{sayama2007decentralized}}
% sources : ants netlogo ; erosion netlogo ; arbotron 
%  inge : 
%}

\end{document}

### 2019-12-01 Models/MesoCalibration/runRangeYear.sh
1-41

@@ -1,1 +1,1 @@
-cat $1 | awk '{print "rm Calib.oms; echo \"val (areaind,endyear) = ("$1","$2")\" >> Calib.oms; cat MesoCalibrationYear.oms >> Calib.oms; /home/juste/openmole-9.2/openmole -p density_2.12-1.0.jar --script Calib.oms --password-file omlpsswd"}'| sh
+cat $1 | awk -F"," '{print "rm Calib.oms;echo "$2" ; echo \"val (areaind,endyear) = ("$1","$2")\" >> Calib.oms; cat MesoCalibrationYear.oms >> Calib.oms; /home/juste/openmole-9.2/openmole -p density_2.12-1.0.jar --script Calib.oms --password-file omlpsswd"}' | sh

### 2019-12-01 Models/MesoCalibration/calibration.R
run-500-byyears

@@ -14,7 +14,13 @@ latestgen <- function(dir){max(as.integer(sapply(strsplit(sapply(strsplit(list.f
 resdir = 'calibration/MESOCALIB_CALIB_LOCAL10_204_1990_20191119_051844/'
 res <- as.tbl(read.csv(paste0(resdir,'population',latestgen(resdir),'.csv')))
 
+g=ggplot(res,aes(x=popfit,y=indicsfit,color=alpha))
+g+geom_point()
+
 g=ggplot(res[res$evolution.samples>2,],aes(x=popfit,y=indicsfit,color=alpha))
+g+geom_point()
+
+g=ggplot(res,aes(x=popfit,y=indicsfit,color=beta))
 g+geom_point()
 
 g=ggplot(res[res$evolution.samples>2,],aes(x=popfit,y=indicsfit,color=beta))

### 2019-12-02 Models/MesoCalibration/analysis.R
.

@@ -8,6 +8,7 @@ library(rgdal)
 library(raster)
 library(reshape2)
 library(ggplot2)
+library(maps)
 
 source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 
@@ -16,8 +17,13 @@ source('morphology.R')
 
 ucdbsf <- st_as_sf(loadUCDBData(paste0(Sys.getenv('CS_HOME'),'/Data/JRC_EC/GHS/GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0'),'GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0'))
 
+resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/GHS/Morphology/')
+
 years = c('1975','1990','2000','2015')
 indics=c('totalPop','maxPop','minPop','moran','avgDist','entropy','alpha','alphaRSquared')
+indicnames = list('totalPop'='Total population','maxPop'='Maximal population','minPop'='Minimal population',
+                  'moran'='Spatial autocorrelation','avgDist'='Average distance','entropy'='Entropy',
+                  'alpha'='Hierarchy','alphaRSquared'='Fierarchy fit')
 
 if(!file.exists('configs/morphologies.csv')){
   # real morphologies
@@ -35,8 +41,13 @@ if(!file.exists('configs/morphologies.csv')){
   write.table(morphos,file='configs/morphologies.csv',col.names = T,row.names = F,sep = ";")
 }
 
+morphos <- read.csv(file='configs/morphologies.csv',header = T,sep = ";")
+
+
 # filter bad fit slope
 #morphos = morphos[morphos$alphaRSquared1975>0.5,]
+# better with outliers
+morphos = morphos[morphos$alpha1975>-4&morphos$alpha1990>-4&morphos$alpha2000>-4&morphos$alpha2015>-4,]
 
 timedf <- data.frame()
 for(year in years){
@@ -49,7 +60,8 @@ classdata = data.frame(id = 1:nrow(morphos))
 for(yearind in 2:length(years)){
   classdata=cbind(classdata,morphos[,paste0(classifindics,years[yearind])] - morphos[,paste0(classifindics,years[yearind-1])])
 }
-
+# normalize for clustering !
+for(j in 2:ncol(classdata)){classdata[,j] = (classdata[,j] - min(classdata[,j]))/(max(classdata[,j])-min(classdata[,j]))}
 
[…]

### 2019-12-02 Models/MesoCalibration/calibration.R

@@ -4,19 +4,32 @@ setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/MesoCalibration'))
 library(dplyr)
 library(ggplot2)
 
+morphos <- read.csv(file='configs/morphologies.csv',header = T,sep = ";")
 
-latestgen <- function(dir){max(as.integer(sapply(strsplit(sapply(strsplit(list.files(dir,pattern=".csv"),"population"),function(s){s[2]}),".csv"),function(s){s[1]})))}
+
+latestgen <- function(dir){
+  if(is.null(dir)){return(NULL)}
+  else{return(max(as.integer(sapply(strsplit(sapply(strsplit(list.files(dir,pattern=".csv"),"population"),function(s){s[2]}),".csv"),function(s){s[1]}))))}
+}
 latestdir <- function(row,year){
   dirs = list.dirs('calibration')
-  m = matrix(unlist(sapply(strsplit(sapply(strsplit(dirs[2:length(dirs)],split = '/'),function(r){r[2]}),split='_'),
-         function(r){if(as.numeric(r[4])==row&as.numeric(r[5])==year){return(list(r=r,sort=paste0(r[6],r[7])))}}
-  )),ncol=8,byrow=T)
+  rows = unlist(sapply(strsplit(sapply(strsplit(dirs[2:length(dirs)],split = '/'),function(r){r[2]}),split='_'),
+                       function(r){if(as.numeric(r[4])==row&as.numeric(r[5])==year){return(list(r=r,sort=paste0(r[6],r[7])))}}
+  ))
+  if(is.null(rows)){return(NULL)}
+  m = matrix(rows,ncol=8,byrow=T)
   latest = order(m[,8],decreasing = T)[1]
   return(paste0('calibration/',paste(m[latest,1:7],collapse='_')))
 }
 # latestgen(latestdir(row,year))
 
-for(row in 1:985){for(year in c(1990,2000,2015)){show(latestgen(latestdir(row,year)))}}
+allres = data.frame()
+for(row in 1:985){for(year in c(1990,2000,2015)){
+  dir = latestdir(row,year);gen=latestgen(dir)
+  show(paste0(row,'-',year,':',gen))
+  currentdata = read.csv(file=paste0(dir,'/population',gen,'.csv'))
[…]

### 2019-12-02 Models/MesoCalibration/dataPreparation.R

@@ -205,7 +205,13 @@ write.table(
   row.names = F,col.names = F,sep=",",file='runYears509-985'
 )
 
+inds = 204:500
+years = c(1990,2000,2015)
 
+write.table(
+  data.frame(ind = c(t(matrix(rep(inds,length(years)),ncol=length(years)))),year = rep(years,length(inds))),
+  row.names = F,col.names = F,sep=",",file='runYears204-500'
+)
 
 
 

### 2019-12-03 Models/MesoCalibration/runRangeYear.sh
200-250

@@ -1,1 +1,1 @@
-cat $1 | awk -F"," '{print "rm Calib.oms;echo "$2" ; echo \"val (areaind,endyear) = ("$1","$2")\" >> Calib.oms; cat MesoCalibrationYear.oms >> Calib.oms; openmole -p density_2.12-1.0.jar --script Calib.oms --password-file omlpsswd"}' | sh
+cat $1 | awk -F"," '{print "rm Calib.oms;echo "$2" ; echo \"val (areaind,endyear) = ("$1","$2")\" >> Calib.oms; cat MesoCalibrationYear.oms >> Calib.oms; /home/juste/openmole-9.2/openmole -p density_2.12-1.0.jar --script Calib.oms --password-file omlpsswd"}' | sh

### 2019-12-04 Models/MesoCalibration/analysis.R
empirical_reuslts

@@ -8,7 +8,10 @@ library(rgdal)
 library(raster)
 library(reshape2)
 library(ggplot2)
+library(scales)
+library(sf)
 library(maps)
+library(mapproj)
 
 source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 
@@ -125,43 +128,185 @@ areasmorph$areaid = inds
 
 areasmorph = left_join(areasmorph,morphos,by=c('areaid'='areaid'))
 
-map<- function(var){
-  WorldData <- map_data('world') %>% filter(region != "Antarctica") %>% fortify
-  g=ggplot() + geom_map(data = WorldData, map = WorldData,aes(group = group, map_id=region),
-             fill = "white", colour = "#7f7f7f", size=0.5) + 
-    geom_point(data=areasmorph,aes(x=GCPNT_LON,y=GCPNT_LAT,color=moran2015))+
+countries <- st_read(paste0(Sys.getenv('CS_HOME'),'/Data/Countries/'),'countries')
+
+map<- function(data,var,sizevar,filename,discrete=FALSE,legendtitle=NULL,legendsizetitle=NULL,xlim=c(-130,150),ylim=c(-50, 60)){
+  #WorldData <- map_data('world') %>% filter(region != "Antarctica") %>% fortify
+  #sizes = log10(data[[sizevar]]);sizes = (sizes - min(sizes,na.rm = T)) / (max(sizes,na.rm = T) - min(sizes,na.rm = T))
+  #data[['sizes']]=sizes
+  g=ggplot()+
+    #geom_map(data = WorldData, map = WorldData,aes(group = group, map_id=region),fill = "white", colour = "#7f7f7f", size=0.1) + 
+    geom_sf(data=countries,fill = "white", colour = "#7f7f7f", size=0.1)+
+    geom_point(data=data,aes_string(x='GCPNT_LON',y='GCPNT_LAT',color=var,size=sizevar),alpha=0.6)+
+    scale_size_area(name=ifelse(is.null(legendsizetitle),sizevar,legendsizetitle))+
     #geom_map(data = areasmorph, map=WorldData,
     #         aes(fill=moran2015),#, map_id=region),
     #         colour="#7f7f7f", size=0.5) +
-    coord_map("rectangular", lat0=0, xlim=c(-180,180), ylim=c(-60, 90)) +
[…]

### 2019-12-04 Docs/Communications/2019/ILUS2019/Communication/ilus2019.tex
.
maps_est/phase-diags

@@
 
-\title[Models of urban morphogenesis]{Models of urban morphogenesis to link urban form and function}
+\title[]{Worldwide estimation of parameters for a simple reaction-diffusion model of urban growth}
 \author[J. Raimbault]{J.~Raimbault$^{1,2,3\ast}$\\\medskip
@@
 
-\date[15th November 2019]{TQG Debates 2019\\
+\date[5th December 2019]{International Land-Use Symposium 2019\\
 3.1: Fractals and Multi-fractals\\
-November 15th 2019
+5th December 2019
 }
@@
 % \textbf{Keywords: }\textit{Urban growth; Reaction-diffusion; World urban areas; Model calibration}
-
-%  Understanding drivers of urban growth and land-use change at the metropolitan scale often requires to isolate and estimate stylized processes of urban evolution.
-Several land-use change and urban growth models for example based on cellular automata have been introduced in the literature, but their lack of generality and need for rich data may become a shortcoming.
-We propose here to apply and calibrate a simple model of urban growth based on reaction-diffusion processes for population density, providing a rather general insight into world urban areas from simple data.
-This four-parameter model introduced by \cite{raimbault2018calibration} combines aggregation forces (positive externalities) with dispersal forces (negative externalities) to capture dynamics of urban form.
-The model was previously only simulated on synthetic initially empty spaces and statically calibrated by comparing the projection of final simulated configurations in a space of urban morphology indicators (Moran index, entropy, average distance, hierarchy) to real measures computed on a fixed-size moving window for Europe.
-We significantly extend these result by (i) applying the model on the 1000 largest urban areas of the Global Human Settlements database \citep{Florczyk2019ghs}; (ii) proceeding to a calibration on three successive time windows (1975-1990, 1990-2000 and 2000-2015) what provides dynamical values for parameters; (iii) using genetic optimization algorithms to minimise the distance on morphological indicators for each area.
-The computational cost being high, results are obtained by using the OpenMOLE model exploration software \citep{reuillon2013openmole}, which provides in a transparent way the optimization algorithms and the distribution of calibrations on a computation grid.
-We obtain therein comparable values for aggregation, diffusion and growth speed parameters on all these areas for the three periods given above.
-These values can be used for short terms projections, but also high-level policies by situating urban areas in the model phase diagrams and for example comparing them to more or less close phase transition parameter values.
-We also illustrate possible policy applications by relating estimated parameters to the EDGAR greenhouse gases emission database.
-This work thus shows how simple models can be useful for global comparisons, which are necessary to contextualize territorial policies.
 
@@
 
-% center of Paris
-%\includegraphics[width=0.88\textwidth]{figures/intro_paname}
-%\includegraphics[width=0.9\textwidth]{figures/intro_bp}
+\includegraphics[width=0.88\textwidth]{figures/intro_paname}
 
@@
 
-\sframe{Defining Morphogenesis}{
+\sframe{Complex processes of Urban Morphogenesis}{
 
-% precise notions and defs
+\centering
 
-\justify
+\includegraphics[width=0.9\textwidth]{figures/intro_bp}
 
-\textit{Proposition of an interdisciplinary definition}
+\footnotesize\textit{Source: Geoportail}
+}
 
-\bigskip
+\sframe{Modeling urban growth and land-use change}{
 
-\textbf{Meta-epistemological framework of imbricated notions:}
+%  Understanding drivers of urban growth and land-use change at the metropolitan scale often requires to isolate and estimate stylized processes of urban evolution.
+Several land-use change and urban growth models for example based on cellular automata have been introduced in the literature, but their lack of generality and need for rich data may become a shortcoming.
 
-Self-organization $\supsetneq$ Morphogenesis $\supsetneq$ Autopoiesis $\supsetneq$ Life
+\cite{engelen2008validating} calibration/validation + map comparison
 
-\bigskip
+\cite{zhou2018gini} urban form / emissions
 
-\textbf{Properties:}
+$\rightarrow$\textit{At the crossroad between Urban Simulation and Artificial Life, few models try to integrate and explain the link between Urban Form and Function}
 
-\begin{itemize}
-\item Architecture links form and function
-\item Emergence strength~\cite{bedau2002downward} increases with notion depth, as bifurcations~\cite{thom1974stabilite}
-\end{itemize}
-
-\bigskip
-
-\textbf{Definition of Morphogenesis :} \textit{Emergence of the form and the function in a strongly coupled manner, producing an emergent architecture \cite{doursat2012morphogenetic}}
+$\rightarrow$\textit{Importance of parcimonious, stylized models: modeling as a tool to understand processes}
 
@@
 
-\sframe{Which models for Urban Morphogenesis ?}{
+\sframe{Contribution}{
 
-% why modeling, exemple of rbd
-% -> champ ; difficultés ("verrous".
-rq : j'aime pas la metaphore du verrou, trop cadrant et suppose que deja construit et qu'il suffit d'ouvrir.
-demarche epistemo de co-evol des connaissances, ni inductif ni deductif.
-il faut contruire.
-horizons d'exploration plus appropriés.
-rejoint l'idee de faire rentrer dans "cadre analytique" : cases prédéfinies, alors qu'il s'agit au contraire de définir ces cases.
--> relire hdr arnaud
+%  We propose here to apply and calibrate a simple model of urban growth based on reaction-diffusion processes for population density, providing a rather general insight into world urban areas from simple data.
 
-\justify
+% We significantly extend these result by (i) applying the model on the 1000 largest urban areas of the Global Human Settlements database \citep{Florczyk2019ghs}; (ii) proceeding to a calibration on three successive time windows (1975-1990, 1990-2000 and 2000-2015) what provides dynamical values for parameters; (iii) using genetic optimization algorithms to minimise the distance on morphological indicators for each area.
 
-\begin{columns}
-\column{0.35\textwidth}
-\centering
-\includegraphics[width=\textwidth]{figures/intro_RBD_lattice}\\
-\footnotesize\textit{Example: a basic hybrid model based on elementary processes for density and network \cite{raimbault2014hybrid}}
-\column{0.6\textwidth}
-\justify
-
-%\vspace{-1cm}
-
-$\rightarrow$\textit{At the crossroad between Urban Simulation and Artificial Life, few models try to integrate and explain the link between Urban Form and Function}
-
-\medskip
-
-$\rightarrow$\textit{Importance of parcimonious, stylized models: modeling as a tool to understand processes}
-
-\end{columns}
-
-\bigskip
-
-\textbf{Research Objective : } Explore simple models to capture morphogenesis based on abstract representation of urban processes; test their ability to reproduce existing urban systems.
+Explore simple models to capture morphogenesis based on abstract representation of urban processes; test their ability to reproduce existing urban systems.
 
@@
 
-% comment Arnaud : reaction-diffusion ?
-
-% model rationale and processes
+% This four-parameter model introduced by \cite{raimbault2018calibration} combines aggregation forces (positive externalities) with dispersal forces (negative externalities) to capture dynamics of urban form.
+The model was previously only simulated on synthetic initially empty spaces and statically calibrated by comparing the projection of final simulated configurations in a space of urban morphology indicators (Moran index, entropy, average distance, hierarchy) to real measures computed on a fixed-size moving window for Europe.
 
@@
 
-\sframe{Model Formalization}{
-
-% model formalization and indicators
+\sframe{Model formalization}{
 
@@
 \item Population growth with exogenous rate $N_G$, attributed independently to a cell following a preferential attachment of strength $\alpha$
-%\begin{equation}
-%\Pb{P_i(t+1)=P_i(t)+1|P(t+1)=P(t)+1}=\frac{(P_i(t)/P(t))^{\alpha}}{\sum(P_i(t)/P(t))^{\alpha}}
-%\end{equation}
-%The attribution being uniformly drawn if all population are equal to 0.
 \item Population is diffused $n_d$ times with strength $\beta$
@@
 
-\sframe{Generating Population Distributions}{
+\sframe{Measuring urban morphology}{
 
-% examples : fig 2 of paper
-
-\centering
-
-\includegraphics[height=0.8\textheight]{figures/density_Fig2}
-
-\footnotesize\textit{Examples of generated territorial shapes}
+\begin{enumerate}
+\item Rank-size slope $\gamma$, given by $\ln \left( P_{\tilde{i}}/P_0\right) \sim k \textrm{ + } \gamma\cdot \ln \left(\tilde{i}/i_0\right)$ where $\tilde{i}$ are the indexes of the distribution sorted in decreasing order.
+\item Entropy of the distribution:
+\begin{equation}
+\mathcal{E} \textrm{ = } \sum_{i=1}^{M}\frac{P_i}{P}\cdot \ln{\frac{P_i}{P}}
+\end{equation}
+$\mathcal{E}\textrm{ = }0$ means that all the population is in one cell whereas $\mathcal{E}\textrm{ = }0$ means that the population is uniformly distributed.
+\item Spatial-autocorrelation given by Moran index, with simple spatial weights given by $w_{ij} \textrm{ = } 1/d_{ij}$
+\[
+I \textrm{ = } M \cdot \frac{\sum_{i\neq j} w_{ij} \left(P_i - \bar{P}\right)\cdot\left(P_j - \bar{P}\right)}{\sum_{i\neq j} w_{ij} \sum_{i}{\left( P_i - \bar{P}\right)}^2}
+\]
+\item Mean distance between individuals
+\[
+\bar{d} \textrm{ = } \frac{1}{d_M}\cdot \sum_{i<j} \frac{P_i P_j}{P^2} \cdot d_{ij}
+\]
+where $d_M$ is a normalisation constant
+\end{enumerate}
 
@@
 
-\sframe{Model behavior}{
+\sframe{Model behavior (synthetic data))}{
 
@@
 
-\sframe{Path-dependence and frozen accidents}{
-
-\centering
-
-\includegraphics[width=0.8\textwidth]{figures/density_Fig4}
-
-\footnotesize\textit{Illustration of path-dependence in a simplified one-dimensional version of the model: cell trajectories in time for 9 independent repetitions from the same initial configuration.}
-
-}
-
-\sframe{Empirical Data for Calibration}{
-
-% comments Arnaud : Quel lien avec les slides avant et après ? "Real" : le terme me paraît discutable : données empiriques plutôt ?
-%  Décrire les données car ici tu passes directement aux indicateurs
-
-\begin{columns}
-\column{0.6\textwidth}
-\centering
-\includegraphics[width=\textwidth]{figures/density_indics_morpho_discrquantiles}
-\column{0.3\textwidth}
-\centering
-\includegraphics[width=\textwidth]{figures/density_cluster_pca_k5_morpho}\\
-\includegraphics[width=\textwidth]{figures/density_cluster_map_k5_morpho}
-\end{columns}
-
-\justify
-
-\footnotesize\textit{Computation of morphological indicators on population density data for Europe (shown here on France), morphological classification.}
-
-}
-
-\sframe{Model Calibration}{
+\sframe{Model application on a spatial grid}{
 
@@
 
-\sframe{Model Targeted Exploration}{
+\section{Empirical analysis}
+
+\sframe{GHSL database}{
+	% brief description of GHSL; illustration
+
+% + empirical approach: 1000 areas, window 0.25%
+
+$\rightarrow$ Application on worldwide urban areas using the GHSL population grid
+
+\medskip
+
+$\rightarrow$ Dynamical calibration on three successive time windows (1975-1990, 1990-2000, 2000-2015)
+
+\medskip
+
+$\rightarrow$ Extraction of neighborhood (spatial span times 1.5) of 1000 larger urban areas (2015 population)
+
+\medskip
+
+$\rightarrow$ Computation of morphological indicators on corresponding population grids
+
+}
+
+\sframe{Global indicators distribution: Moran}{
 
@@
 
-\includegraphics[width=0.8\textwidth]{figures/density_Fig6}
+\includegraphics[width=\linewidth]{figures/mapindic_moran_2015.png}
 
-\footnotesize\textit{Potentialities of targeted model explorations: here feasible space using Pattern Space Exploration algorithm \cite{10.1371/journal.pone.0138212}.}
+\medskip
+
+\small
+
+\textit{Regional patterns for Moran index}
 
@@
 
-\section{Co-evolution model}
+\sframe{Global indicators distribution: average distance}{
 
-\sframe{Including more complex processes ?}{
+\centering
 
-% transition : representation of territories
-
-\textit{Which ontology to include more complex functional properties ?}
+\includegraphics[width=\linewidth]{figures/mapindic_avgDist_2015.png}
 
@@
 
-$\rightarrow$ Territorial systems as the strong coupling between territories and (potential and realized) networks \cite{dupuy1987vers}.
+\small
+
+\textit{Complementarity of average distance indicator}
+
+}
+
+\sframe{Clustering morphological trajectories}{
+
+\includegraphics[width=0.55\textwidth]{figures/distributions_cluster-year.png}
+\includegraphics[width=0.4\textwidth]{figures/PCA_trajectories.png}
 
@@
 
-$\rightarrow$ Networks convey functional notions of centralities and accessibility, among others; have furthermore proper topological properties.
+\small
+
+\textit{Disjoint distributions and temporal trajectories: model application is relevant}
+
+}
+
+\sframe{Distribution of clusters}{
+
+\centering
+
+\includegraphics[width=\linewidth]{figures/mapindic_cluster.png}
+
+}
+
+\sframe{Correlations}{
+
+\centering
+
+\includegraphics[width=0.48\linewidth]{figures/correlations_deltavars-morpho_2015.png}
+\includegraphics[width=0.48\linewidth]{figures/correlations_vars-morpho_2015.png}
+
+}
+
+\section{Model calibration}
+
+\sframe{Calibration method}{
+ % setup, error
+}
+
+\sframe{Implementation}{
+
+% openmole slide
+
+% The computational cost being high, results are obtained by using the OpenMOLE model exploration software \citep{reuillon2013openmole}, which provides in a transparent way the optimization algorithms and the distribution of calibrations on a computation grid.
+
+%\footnotesize
+\justify
+
+\textit{Performance constraints: simulate $N$ mesoscopic morphogenesis models in parallel (macroscopic interactions are efficient as based on matrices)}
+
+\medskip
+
+$\rightarrow$ model implemented in \texttt{scala} and integrated within a broader library (including implementations of \cite{raimbault2018indirect} \cite{raimbault2018calibration}
+
+\cite{favaro2011gibrat} \cite{cottineau2015modular})
+
+\bigskip
+
+\textit{Large number of parameters and output indicators}
+
+\medskip
+
+$\rightarrow$ integration into the OpenMOLE model exploration open source software \cite{reuillon2013openmole}
+
+\begin{center}
+\includegraphics[height=0.13\textheight]{figures/iconOM.png}
+\includegraphics[height=0.13\textheight]{figures/openmole.png}
+\end{center}
+
+%(X year of computation for the results presented here)
+
+\footnotesize
+\textit{Enables seamlessly (i) model embedding; (ii) access to HPC resources; (iii) exploration and optimization algorithms}
+
+\medskip
+
+\textbf{Come to the satellite \textit{New Methods and Epistemologies to Explore Simulation Models} tomorrow afternoon in LHN-TR+05}
+
+}
+
+\sframe{Examples of results}{
+
+% pareto fronts for two different areas
+
+\centering
+
+\includegraphics[width=0.49\textwidth]{figures/pareto_indics-relpop_1_coloralpha.png}
+\includegraphics[width=0.49\textwidth]{figures/pareto_indics-relpop_100_coloralpha.png}
+
+}
+
+\sframe{Adjusted parameters}{
+
+% maps : alpha, beta, tstep
+
+% We obtain therein comparable values for aggregation, diffusion and growth speed parameters on all these areas for the three periods given above.
+
+ \centering
+
+\includegraphics[height=0.49\textheight]{figures/mapparam_meanalpha_2015.png}
+\includegraphics[height=0.49\textheight]{figures/mapparam_meanbeta_2015.png}
+
+}
+
+\sframe{Trajectories in phase diagrams}{
+
+% These values can be used for short terms projections, but also high-level policies by situating urban areas in the model phase diagrams and for example comparing them to more or less close phase transition parameter values.
+
+}
+
+\sframe{Correlations}{
+
+% We also illustrate possible policy applications by relating estimated parameters to the EDGAR greenhouse gases emission database.
+
+\centering
+
+\includegraphics[width=0.49\textwidth]{figures/deltavar-params_2000.png}
+\includegraphics[width=0.49\textwidth]{figures/deltavar-params_2015.png}
 
@@
 \sframe{Discussion}{
+
+%  This work thus shows how simple models can be useful for global comparisons, which are necessary to contextualize territorial policies.
+
+% \textit{Which ontology to include more complex functional properties ?}
+%$\rightarrow$ Territorial systems as the strong coupling between territories and (potential and realized) networks \cite{dupuy1987vers}.
+%$\rightarrow$ Networks convey functional notions of centralities and accessibility, among others; have furthermore proper topological properties.
 
@@
 
+\sframe{Generating Population Distributions}{
+
+% examples : fig 2 of paper
+
+\centering
+
+\includegraphics[height=0.8\textheight]{figures/density_Fig2}
+
+\footnotesize\textit{Examples of generated territorial shapes}
+
+}
+
+\sframe{Path-dependence and frozen accidents}{
+
+\centering
+
+\includegraphics[width=0.8\textwidth]{figures/density_Fig4}
+
+\footnotesize\textit{Illustration of path-dependence in a simplified one-dimensional version of the model: cell trajectories in time for 9 independent repetitions from the same initial configuration.}
+
+}
+
+\sframe{Empirical Data (moving window)}{
+
+\begin{columns}
+\column{0.6\textwidth}
+\centering
+\includegraphics[width=\textwidth]{figures/density_indics_morpho_discrquantiles}
+\column{0.3\textwidth}
+\centering
+\includegraphics[width=\textwidth]{figures/density_cluster_pca_k5_morpho}\\
+\includegraphics[width=\textwidth]{figures/density_cluster_map_k5_morpho}
+\end{columns}
+
+\justify
+
+\footnotesize\textit{Computation of morphological indicators on population density data for Europe (shown here on France), morphological classification.}
+
+}
+
 \sframe{Model classification : PDE}{
@@
 \includegraphics[width=0.48\textwidth]{figures/density_pmax_logbeta}
-
-}
-
-\sframe{Morphological indicators}{
-
-\begin{enumerate}
-\item Rank-size slope $\gamma$, given by $\ln \left( P_{\tilde{i}}/P_0\right) \sim k \textrm{ + } \gamma\cdot \ln \left(\tilde{i}/i_0\right)$ where $\tilde{i}$ are the indexes of the distribution sorted in decreasing order.
-\item Entropy of the distribution:
-\begin{equation}
-\mathcal{E} \textrm{ = } \sum_{i=1}^{M}\frac{P_i}{P}\cdot \ln{\frac{P_i}{P}}
-\end{equation}
-$\mathcal{E}\textrm{ = }0$ means that all the population is in one cell whereas $\mathcal{E}\textrm{ = }0$ means that the population is uniformly distributed.
-\item Spatial-autocorrelation given by Moran index, with simple spatial weights given by $w_{ij} \textrm{ = } 1/d_{ij}$
-\[
-I \textrm{ = } M \cdot \frac{\sum_{i\neq j} w_{ij} \left(P_i - \bar{P}\right)\cdot\left(P_j - \bar{P}\right)}{\sum_{i\neq j} w_{ij} \sum_{i}{\left( P_i - \bar{P}\right)}^2}
-\]
-\item Mean distance between individuals
-\[
-\bar{d} \textrm{ = } \frac{1}{d_M}\cdot \sum_{i<j} \frac{P_i P_j}{P^2} \cdot d_{ij}
-\]
-where $d_M$ is a normalisation constant
-\end{enumerate}
 
@@
 
+\sframe{Model Targeted Exploration}{
+
+\centering
+
+\includegraphics[width=0.8\textwidth]{figures/density_Fig6}
+
+\footnotesize\textit{Potentialities of targeted model explorations: here feasible space using Pattern Space Exploration algorithm \cite{10.1371/journal.pone.0138212}.}
+
+}
+
+\sframe{Defining Morphogenesis}{
+
+% precise notions and defs
+
+\justify
+
+\textit{Proposition of an interdisciplinary definition}
+
+\bigskip
+
+\textbf{Meta-epistemological framework of imbricated notions:}
+
+Self-organization $\supsetneq$ Morphogenesis $\supsetneq$ Autopoiesis $\supsetneq$ Life
+
+\bigskip
+
+\textbf{Properties:}
+
+\begin{itemize}
+\item Architecture links form and function
+\item Emergence strength~\cite{bedau2002downward} increases with notion depth, as bifurcations~\cite{thom1974stabilite}
+\end{itemize}
+
+\bigskip
+
+\textbf{Definition of Morphogenesis :} \textit{Emergence of the form and the function in a strongly coupled manner, producing an emergent architecture \cite{doursat2012morphogenetic}}
+
+}
+
 %%%%%%%%%%%%%%%%%%%%%

### 2019-12-04 Models/MesoCalibration/analysis.R

@@ -8,17 +8,13 @@ library(rgdal)
 library(raster)
 library(reshape2)
 library(ggplot2)
-library(scales)
-library(sf)
-library(maps)
-library(mapproj)
 
 source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 
 source('functions.R')
 source('morphology.R')
 
-ucdbsf <- st_as_sf(loadUCDBData(paste0(Sys.getenv('CS_HOME'),'/Data/JRC_EC/GHS/GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0'),'GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0'))
+ucdbsf <- loadUCDBData(paste0(Sys.getenv('CS_HOME'),'/Data/JRC_EC/GHS/GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0'),'GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0')
 
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/GHS/Morphology/')
 
@@ -46,20 +42,27 @@ if(!file.exists('configs/morphologies.csv')){
 
 morphos <- read.csv(file='configs/morphologies.csv',header = T,sep = ";")
 
+inds = c(1:946,948:1000) # 
+#morphosall = data.frame(matrix(unlist(res),nrow=length(res),byrow = T))
+#names(morphosall)<-names(res[[1]])
+#morphos <- read.csv(file='configs/morphologies.csv',header = T,sep = ";")
+#rownames(morphos)<-morphos$areaid
+areasmorph = ucdbsf[inds,]
+areasmorph$areaid = inds
+
+areasmorph = left_join(areasmorph,morphos,by=c('areaid'='areaid'))
+
+
 
 # filter bad fit slope
 #morphos = morphos[morphos$alphaRSquared1975>0.5,]
 # better with outliers
 morphos = morphos[morphos$alpha1975>-4&morphos$alpha1990>-4&morphos$alpha2000>-4&morphos$alpha2015>-4,]
 
-timedf <- data.frame()
-for(year in years){
-  currentd = morphos[,paste0(indics,year)];names(currentd)<-indics
-  timedf = rbind(timedf,cbind(area = 1:nrow(currentd),currentd,year=rep(year,nrow(currentd))))
-}
[…]

### 2019-12-04 Models/MesoCalibration/calibration.R

@@ -4,7 +4,17 @@ setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/MesoCalibration'))
 library(dplyr)
 library(ggplot2)
 
+source('functions.R')
+
+resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/MesoCalibration/');
+dir.create(resdir);
+
+
+ucdbsf <- loadUCDBData(paste0(Sys.getenv('CS_HOME'),'/Data/JRC_EC/GHS/GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0'),'GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0')
+
 morphos <- read.csv(file='configs/morphologies.csv',header = T,sep = ";")
+inds = c(1:946,948:1000);areasmorph = ucdbsf[inds,];areasmorph$areaid = inds
+areasmorph = as.data.frame(left_join(areasmorph,morphos,by=c('areaid'='areaid')))
 
 
 latestgen <- function(dir){
@@ -23,49 +33,145 @@ latestdir <- function(row,year){
 }
 # latestgen(latestdir(row,year))
 
-allres = data.frame()
-for(row in 1:985){for(year in c(1990,2000,2015)){
+# missing: 1-1975; 204-1990; 275-2000; 309-2015; 336-1990; 385-2000; 414-2015
+allrestmp = data.frame()
+for(row in 415:nrow(morphos)) {for(year in c(1990,2000,2015)){
   dir = latestdir(row,year);gen=latestgen(dir)
   show(paste0(row,'-',year,':',gen))
-  currentdata = read.csv(file=paste0(dir,'/population',gen,'.csv'))
-  allres = rbind(allres,cbind(areaid = rep(morphos[row,"areaid"],nrow(currentdata)),year = rep(year,nrow(currentdata)),currentdata))
+  if(!is.null(gen)){
+    currentdata = read.csv(file=paste0(dir,'/population',gen,'.csv'))
+    allrestmp = rbind(allrestmp,cbind(areaid = rep(morphos[row,"areaid"],nrow(currentdata)),year = rep(year,nrow(currentdata)),currentdata))
+  }
 }}
 
+
+timedf = as_temporal_df(areasmorph,indics)
+allres <- left_join(allrestmp,timedf[timedf$year>1975,],by=c('areaid'='areaid','year'='year')) # %>%
[…]

### 2019-12-04 Models/MesoCalibration/dataPreparation.R

@@ -199,11 +199,7 @@ write.table(
 
 inds = 509:985
 years = c(1990,2000,2015)
-
-write.table(
-  data.frame(ind = c(t(matrix(rep(inds,length(years)),ncol=length(years)))),year = rep(years,length(inds))),
-  row.names = F,col.names = F,sep=",",file='runYears509-985'
-)
+write.table(data.frame(ind = c(t(matrix(rep(inds,length(years)),ncol=length(years)))),year = rep(years,length(inds))),row.names = F,col.names = F,sep=",",file='runYears509-985')
 
 inds = 204:500
 years = c(1990,2000,2015)
@@ -212,6 +208,11 @@ write.table(
   data.frame(ind = c(t(matrix(rep(inds,length(years)),ncol=length(years)))),year = rep(years,length(inds))),
   row.names = F,col.names = F,sep=",",file='runYears204-500'
 )
+
+inds = 592:985
+years = c(1990,2000,2015)
+write.table(data.frame(ind = c(t(matrix(rep(inds,length(years)),ncol=length(years)))),year = rep(years,length(inds))),row.names = F,col.names = F,sep=",",file='runYears592-985')
+
 
 
 

### 2019-12-04 Models/MesoCalibration/functions.R

@@ -1,6 +1,97 @@
 
 
 
+library(scales)
+library(sf)
+#library(maps)
+#library(mapproj)
+
+
+as_temporal_df <- function(data,indics){
+  currenttimedf <- data.frame()
+  for(year in years){
+    currentd = data[,paste0(indics,year)];names(currentd)<-indics
+    currenttimedf = rbind(currenttimedf,cbind(areaid = data$areaid,currentd,year=rep(year,nrow(currentd))))
+  }
+  return(currenttimedf)
+}
+
+getParetoFront<-function(d,c1,c2){
+  inds=c()
+  for(i in 1:nrow(d)){
+    if(length(which(d[i,c1]>d[,c1]&d[i,c2]>d[,c2]))==0){inds=append(inds,i)}
+  }
+  return(d[inds,])
+}
+
+
+library(corrplot)
+
+correlations<-function(currentdata,corvars,corvarsnames,years,mode,fileprefix,classifindics=NULL){
+  
+  for(year in years){
+    rho = matrix(0,length(corvars),length(corvars));colnames(rho)<-corvarsnames;rownames(rho)<-corvarsnames
+    rhomin = rho;rhomax = rho
+    for(i in 1:length(corvars)){
+      for(j in 1:length(corvars)){
+        if(mode=="wide"){
+        ivar= paste0(corvars[i],year);jvar=paste0(corvars[j],year)
+        if(!corvars[i]%in%classifindics){ivar= paste0(corvars[i],substr(year,3,4))}
+        if(!corvars[j]%in%classifindics){jvar=paste0(corvars[j],substr(year,3,4))}
+        x=unlist(currentdata[,ivar]);y=unlist(currentdata[,jvar])
+        }
+        if(mode=="long"){
+          x=unlist(currentdata[currentdata$year==year,corvars[i]])
+          y=unlist(currentdata[currentdata$year==year,corvars[j]])
+        }
+        rhotest = cor.test(x,y)
+        rho[i,j]=rhotest$estimate;rhomin[i,j]=rhotest$conf.int[1];rhomax[i,j]=rhotest$conf.int[2]
+      }
+    }
+    png(filename = paste0(fileprefix,'_',year,'.png'),width = 25,height = 20,units='cm',res = 300)
+    corrplot(rho,lowCI.mat = rhomin, uppCI.mat = rhomax,type = 'upper',title = year,bg='lightgrey',
[…]

### 2019-12-05 Models/MesoCalibration/calibration.R
ext_phasediag

@@ -121,12 +121,14 @@ for(param in params){
 ####
 ## Correlations
 
-years = c(2000,2015)
+years = c(1990,2000,2015)
 
 vars=c('P','B','E','G','DP','DB','DE','DG')
 currenttimedf <- data.frame()
 for(year in years){
-  currentd = areasmorph[,paste0(vars,substr(year,3,4))];names(currentd)<-vars
+  if(year>1990){currentvars = vars}else{currentvars = vars[c(-4,-8)]}
+  currentd = areasmorph[,paste0(currentvars,substr(year,3,4))];names(currentd)<-currentvars
+  if(year==1990){currentd$G=rep(NA,nrow(currentd));currentd$DG=rep(NA,nrow(currentd))}
   currenttimedf = rbind(currenttimedf,cbind(areaid = areasmorph$areaid,currentd,year=rep(year,nrow(currentd))))
 }
 
@@ -146,15 +148,20 @@ correlations(cordata,corvars,corvarsnames,years,fileprefix = paste0(resdir,'Cali
 ####
 ## Trajectories in phase diag
 
+areanames = sapply(strsplit(areasmorph$UC_NM_LST,";"),function(r){r[1]})
+
 # load phase diags: (alpha,beta) stratified by tf
 
-phasediag <- as.tbl(read.csv('exploration/20191204_182426_EXPLORATION_LOCAL4.csv'))
+phasediag <- as.tbl(read.csv('exploration/20191204_225527_EXPLORATION_LOCAL4.csv'))
 sphasediag <- phasediag %>% group_by(alpha,beta,tsteps) %>% summarize(
   avgDistance=mean(avgDistance),
   entropy=mean(entropy),
   moran=mean(moran),
   slope=mean(slope)
 )
+
+# restrict phase diag to beta < 0.1 (boundary for calibration)
+sphasediag = sphasediag[sphasediag$beta<=0.1&sphasediag$alpha<=2,]
 
 indics = c("avgDistance","entropy","moran","slope")
 #indicnames =list("avgDistance","entropy","moran","slope")
@@ -163,14 +170,28 @@ tsteps= unique(sphasediag$tsteps);dists=sapply(cordata$meantsteps,function(t){mi
 for(indic in indics){
   for(tstep in tsteps){
     currentest = cordata[abs(cordata$meantsteps-tstep)==dists&cordata$meanalpha<max(sphasediag$alpha),]
-  ggplot()+
[…]

### 2019-12-05 Models/MesoCalibration/dataPreparation.R

@@ -209,9 +209,9 @@ write.table(
   row.names = F,col.names = F,sep=",",file='runYears204-500'
 )
 
-inds = 592:985
+inds = 593:985
 years = c(1990,2000,2015)
-write.table(data.frame(ind = c(t(matrix(rep(inds,length(years)),ncol=length(years)))),year = rep(years,length(inds))),row.names = F,col.names = F,sep=",",file='runYears592-985')
+write.table(data.frame(ind = c(t(matrix(rep(inds,length(years)),ncol=length(years)))),year = rep(years,length(inds))),row.names = F,col.names = F,sep=",",file='runYears593-985')
 
 
 

### 2019-12-05 Models/MesoCalibration/calibration.R
.

@@ -35,7 +35,7 @@ latestdir <- function(row,year){
 
 # missing: 1-1975; 204-1990; 275-2000; 309-2015; 336-1990; 385-2000; 414-2015
 allrestmp = data.frame()
-for(row in 415:nrow(morphos)) {for(year in c(1990,2000,2015)){
+for(row in 1:nrow(morphos)) {for(year in c(1990,2000,2015)){
   dir = latestdir(row,year);gen=latestgen(dir)
   show(paste0(row,'-',year,':',gen))
   if(!is.null(gen)){
@@ -44,6 +44,8 @@ for(row in 415:nrow(morphos)) {for(year in c(1990,2000,2015)){
   }
 }}
 
+years = c(1975,1990,2000,2015)
+indics=c('totalPop','maxPop','minPop','moran','avgDist','entropy','alpha','alphaRSquared')
 
 timedf = as_temporal_df(areasmorph,indics)
 allres <- left_join(allrestmp,timedf[timedf$year>1975,],by=c('areaid'='areaid','year'='year')) # %>%
@@ -107,6 +109,7 @@ for(param in params){
   for(year in years){
   currentdata=areasmorphest[areasmorphest$year==year,]
   if(param=='meanalpha'){currentdata=currentdata[currentdata[,'meanalpha']<3,]}
+  if(param=='meanbeta'){currentdata=currentdata[!is.na(currentdata[,'meanbeta']),]}
   map(
     currentdata,
     param,
@@ -133,6 +136,8 @@ for(year in years){
 }
 
 cordata <- left_join(sres,currenttimedf,by=c("areaid"="areaid","year"="year"))
+
+years=c(2000,2015)
 
 corvars = c('P','B','E','G','meanalpha','meanbeta','meantsteps')
 corvarsnames = c('Pop','Built','Emissions','GDP','alpha','beta','tsteps')

### 2019-12-05 Models/MesoCalibration/functions.R

@@ -7,7 +7,7 @@ library(sf)
 #library(mapproj)
 
 
-as_temporal_df <- function(data,indics){
+as_temporal_df <- function(data,indics,years){
   currenttimedf <- data.frame()
   for(year in years){
     currentd = data[,paste0(indics,year)];names(currentd)<-indics

### 2020-02-15 Models/urbangrowth/openmole/runCalib.sh
.

@@ -1,1 +1,1 @@
-cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 120G --logger-level ALL;mkdir calibration/`ls -t calib | head -n 1`; cp Calib"$1$2".oms calibration/`ls -t calib | head -n 1`"}'|sh
+cat $1| awk -F"," '{print "echo '\''val countrycode = \""$2"\"'\'' > Calib"$1$2".oms ; tail -n +2 Calibration"$1".oms >> Calib"$1$2".oms ; openmole --workspace /home/juste/.openmole --script Calib"$1$2".oms --password-file omlpsswd --mem 120G --logger-level ALL;mkdir calibration/`ls -t calib | head -n 1`; cp Calib"$1$2".oms calibration/`ls -t calib | head -n 1`"}'|sh

### 2020-02-15 Docs/Communications/2019/ECTQG2019/ectqg2019_urbgrowth.tex

\documentclass[11pt]{article}
\input{header.tex}
\begin{document}

\title{\vspace{-1.5cm}An evolutionary theory for the spatial dynamics of urban systems worldwide
\\\medskip
\textit{ECTQG 2019}
}
\author{\noun{J. Raimbault}$^{1,2,3,\ast}$, \noun{E. Denis}$^3$ and \noun{D.Pumain}$^3$\medskip\\
$^1$ CASA, UCL\\
$^2$ UPS CNRS 3611 ISC-PIF\\
$^3$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
\medskip\\
$^{\ast}$\texttt{juste.raimbault@polytechnique.edu}
}
\date{}

\maketitle

\justify

\pagenumbering{gobble}

\textbf{Keywords: }\textit{Urban Systems; Dynamic models; Evolutionary Theory}

\medskip

Analyzing the spatial dynamics of complex urban systems clearly deserves an evolutionary frame. Following the methodology and results already obtained in the GeoDiverCity project \citep{pumain2015multilevel,cura2017old,pumain2017urban}, including USA, Europe and BRICS countries, we complete them with new datasets at world scale and other types of models. These models conceived for explaining city size and urban growth distributions establish a correspondence between urban trajectories when observed at the level of cities and systems of cities. We test the validity and representativeness of several models of complex urban systems and their variations across regions of the world at different spatial scales \citep{raimbault2018calibration,raimbault2018indirect}. The originality of the approach is in considering spatial interaction and evolutionary path dependence as major features in the general behavior of urban entities. We investigate models of urban growth at different scales and on different urban systems: a model of urban morphogenesis at the metropolitan scale, which we calibrate dynamically, using the diachronic population grid on largest urban clusters, and interaction models for systems of cities at the macroscopic scale on main systems of cities across the world. We also suggest research directions towards the coupling of these models into a multi-scale model of urban growth.

Complex systems’ dynamics is in principle unpredictable, but contextualizing it regarding demographic, income and resource components may help in minimizing the forecasting errors. We use among others a new unique source correlating population and build-up footprint at world scale: the Global Human Settlement built-up areas (GHS-BU). Already explored statistically for comparing urban sprawl trends in the countries of the world by \cite{denis2019population}, the dataset is available at different dates between 1975 and 2015. In 2015 the source delineates precisely some 13 000 urban agglomerations between 50000 and tens million inhabitants in the world. These data help in further empirical testing to the hypotheses of the evolutionary theory of urban systems and partially revising them.

%%%%%%%%%%%%%%%%%%%%
%% Biblio
%%%%%%%%%%%%%%%%%%%%
%\tiny

%\begin{multicols}{2}

%\setstretch{0.3}
%\setlength{\parskip}{-0.4em}

\footnotesize

\bibliographystyle{apalike}
\bibliography{biblio}
%\end{multicols}

\end{document}

### 2020-02-15 Docs/Communications/2019/ILUS2019/Abstract/ilus2019.tex

\documentclass[11pt]{article}
\input{header.tex}
\begin{document}

\title{\vspace{-1cm}Worldwide estimation of parameters for a simple reaction-diffusion model of urban growth
\\\medskip
\textit{ILUS 2019}
}
\author{\noun{J. Raimbault}$^{1,2,3,\ast}$\medskip\\
$^1$ CASA, UCL\\
$^2$ UPS CNRS 3611 ISC-PIF\\
$^3$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
\medskip\\
$^{\ast}$\texttt{juste.raimbault@polytechnique.edu}
}
\date{}

\maketitle

\justify

\pagenumbering{gobble}

\textbf{Keywords: }\textit{Urban growth; Reaction-diffusion; World urban areas; Model calibration}

% doi:10.1177/2399808319843534 Singularity cities - alternative model

\medskip

Understanding drivers of urban growth and land-use change at the metropolitan scale often requires to isolate and estimate stylized processes of urban evolution. Several land-use change and urban growth models for example based on cellular automata have been introduced in the literature, but their lack of generality and need for rich data may become a shortcoming. We propose here to apply and calibrate a simple model of urban growth based on reaction-diffusion processes for population density, providing a rather general insight into world urban areas from simple data. This four-parameter model introduced by \cite{raimbault2018calibration} combines aggregation forces (positive externalities) with dispersal forces (negative externalities) to capture dynamics of urban form. The model was previously only simulated on synthetic initially empty spaces and statically calibrated by comparing the projection of final simulated configurations in a space of urban morphology indicators (Moran index, entropy, average distance, hierarchy) to real measures computed on a fixed-size moving window for Europe. We significantly extend these result by (i) applying the model on the 1000 largest urban areas of the Global Human Settlements database \citep{Florczyk2019ghs}; (ii) proceeding to a calibration on three successive time windows (1975-1990, 1990-2000 and 2000-2015) what provides dynamical values for parameters; (iii) using genetic optimization algorithms to minimise the distance on morphological indicators for each area. The computational cost being high, results are obtained by using the OpenMOLE model exploration software \citep{reuillon2013openmole}, which provides in a transparent way the optimization algorithms and the distribution of calibrations on a computation grid. We obtain therein comparable values for aggregation, diffusion and growth speed parameters on all these areas for the three periods given above. These values can be used for short terms projections, but also high-level policies by situating urban areas in the model phase diagrams and for example comparing them to more or less close phase transition parameter values. We also illustrate possible policy applications by relating estimated parameters to the EDGAR greenhouse gases emission database. This work thus shows how simple models can be useful for global comparisons, which are necessary to contextualize territorial policies.

%%%%%%%%%%%%%%%%%%%%
%% Biblio
%%%%%%%%%%%%%%%%%%%%
%\tiny

%\begin{multicols}{2}

%\setstretch{0.3}
%\setlength{\parskip}{-0.4em}

\footnotesize

\bibliographystyle{apalike}
\bibliography{biblio}
%\end{multicols}

\end{document}

### 2020-02-15 Docs/Communications/2019/ILUS2019/Communication/ilus2019.tex
ext_phasediag

@@
 \date[5th December 2019]{International Land-Use Symposium 2019\\
-3.1: Fractals and Multi-fractals\\
+Session Drivers, Mechanisms, Tools\\
 5th December 2019
@@
 
-\cite{engelen2008validating} calibration/validation + map comparison
+$\rightarrow$ Cellular automata models or models of land-use change generally highly data-driven
 
-\cite{zhou2018gini} urban form / emissions
+\medskip
 
-$\rightarrow$\textit{At the crossroad between Urban Simulation and Artificial Life, few models try to integrate and explain the link between Urban Form and Function}
+$\rightarrow$ May be case-specific or conclusions difficult to generalize (see Sleuth for example \cite{clarke2007decade})
 
-$\rightarrow$\textit{Importance of parcimonious, stylized models: modeling as a tool to understand processes}
+\medskip
+
+$\rightarrow$ Issues to calibrate and validate these models \cite{engelen2008validating}
+
+%calibration/validation + map comparison
+
+%\cite{zhou2018gini} urban form / emissions
+
+\bigskip
+
+\textbf{Urban morphogenesis models approach:}
+
+\smallskip
+
+$\rightarrow$\textit{At the crossroad between Urban Simulation and Artificial Life, stylized models of urban growth processes}
+
+\medskip
+
+$\rightarrow$\textit{Importance of parsimony: generative modeling as a tool to understand processes}
 
@@
 
-Explore simple models to capture morphogenesis based on abstract representation of urban processes; test their ability to reproduce existing urban systems.
+$\rightarrow$ Explore a simple model to capture morphogenesis based on abstract representation of urban processes; test their ability to reproduce existing urban systems.
+
+\bigskip
+
+$\rightarrow$ The simple reaction-diffusion model of urban growth is calibrated worldwide using GHSL, and in time.
+
+\bigskip
+
+$\rightarrow$ Globally comparable insights into processes of concentration and dispersion.
 
@@
 
-{\tiny
-
-Raimbault, J. (2018). Calibration of a density-based model of urban morphogenesis. PloS one, 13(9), e0203516.
-}
+\textit{Raimbault, J. (2018). Calibration of a density-based model of urban morphogenesis. PloS one, 13(9), e0203516.}
 
@@
 
-\centering
+\begin{center}
 
 \includegraphics[width=0.9\textwidth]{figures/density_Fig3}
+
+\end{center}
 
@@
 
-\sframe{Correlations}{
+\sframe{Correlations with urban area properties}{
 
@@
 
-\includegraphics[width=0.48\linewidth]{figures/correlations_deltavars-morpho_2015.png}
-\includegraphics[width=0.48\linewidth]{figures/correlations_vars-morpho_2015.png}
+\includegraphics[trim={4cm 0 0 0},width=0.48\linewidth]{figures/correlations_deltavars-morpho_2015.png}
+\includegraphics[trim={4cm 0 0 0},width=0.48\linewidth]{figures/correlations_vars-morpho_2015.png}
 
@@
  % setup, error
+
+\begin{itemize}
+	\item For a time period $\left[t_0,t_f\right]$, model initialized with population at $t_0$
+	\item Growth rate $N_G$ is computed by constraining with actual population growth: $N_G \textrm{ = } \left(P_i\left(t_f\right) - P_i \left(t_0\right)\right) / T$ where $T$ is a calibration parameter
+	\item Fitness functions: squared relative population error and squared relative morphological indicator error
+	\item Biobjective genetic algorithm (NSGA2) run for each areas ($\sim 600$ areas large enough) and each time windows (1000 generations, population of 100), on parameters $\alpha, \beta, T$
+\end{itemize}
+
 }
@@
 
-\textit{Performance constraints: simulate $N$ mesoscopic morphogenesis models in parallel (macroscopic interactions are efficient as based on matrices)}
+\textit{Performance constraints: large number of calibration algorithms to be run}
 
@@
 
-$\rightarrow$ model implemented in \texttt{scala} and integrated within a broader library (including implementations of \cite{raimbault2018indirect} \cite{raimbault2018calibration}
+$\rightarrow$ model implemented in \texttt{scala} and integrated within the \texttt{spatialdata} library for spatial sensitivity analysis (including implementations of
 
-\cite{favaro2011gibrat} \cite{cottineau2015modular})
+\cite{raimbault2018indirect} \cite{raimbault2018calibration} \cite{favaro2011gibrat} \cite{cottineau2015modular})
 
-\bigskip
-
-\textit{Large number of parameters and output indicators}
+See \url{https://github.com/openmole/spatialdata}
 
@@
 
-\textbf{Come to the satellite \textit{New Methods and Epistemologies to Explore Simulation Models} tomorrow afternoon in LHN-TR+05}
+%\textbf{Come to the satellite \textit{New Methods and Epistemologies to Explore Simulation Models} tomorrow afternoon in LHN-TR+05}
+
+\textbf{Apply for the OpenMOLE summer school before January 15th}
+
+\url{https://exmodelo.org/}
 
@@
 
-\centering
+\begin{center}
 
@@
 \includegraphics[width=0.49\textwidth]{figures/pareto_indics-relpop_100_coloralpha.png}
+
+\end{center}
+
+\textit{Different performance on indicator fit and forms of Pareto fronts}
 
@@
 
-\includegraphics[height=0.49\textheight]{figures/mapparam_meanalpha_2015.png}
-\includegraphics[height=0.49\textheight]{figures/mapparam_meanbeta_2015.png}
+\includegraphics[height=0.43\textheight]{figures/mapparam_meanalpha_2015.png}
+
+\includegraphics[height=0.43\textheight]{figures/mapparam_meanbeta_2015.png}
 
@@
 
+\begin{center}
+
+\includegraphics[height=0.49\textheight]{figures/phasediag_largestcities_moran_tsteps50.png}
+\includegraphics[height=0.49\textheight]{figures/phasediag_largestcities_slope_tsteps50.png}
+
+\end{center}
+
+\textit{Urban areas trajectories can be projected into model phase diagrams}
+
 }
@@
 \includegraphics[width=0.49\textwidth]{figures/deltavar-params_2015.png}
+
+\end{center}
+
+\textit{Slight significant correlations with growth rates of GHG emissions: more aggregation correlates with decreasing emissions e.g.}
 
@@
 
-$\rightarrow$ This rather simple model reproduces most of existing urban forms in Europe for both population distribution and road network : which intrinsic dimension to the urban system and its morphological aspect ?
+\smallskip
 
-$\rightarrow$ Ability to reproduce static correlations and a variety of dynamical lagged correlation regimes suggests that the model captures some of the processes of co-evolution
+$\rightarrow$ This rather simple model can effectively be calibrated on worldwide urban areas and provide comparable growth parameters
 
-% implications for morphogenesis ?
+\medskip
+
+$\rightarrow$ Link between urban form, urban dynamics and sustainability properties
 
@@
 
-$\rightarrow$ Towards a dynamical calibration ? Need of dynamical data
+\smallskip
+
+$\rightarrow$ Benchmark with other simple models (e.g. \cite{li2019singularity} presented yesterday)
+
+\medskip
+
+$\rightarrow$ Coupling with macro-scale model for system of cities \cite{raimbault:halshs-02351722}
+
+\medskip
 
@@
 
-$\rightarrow$ Compare network generation in a ``fair'' way (correcting for additional parameters, open question for models of simulation)
+%$\rightarrow$ Compare network generation in a ``fair'' way (correcting for additional parameters, open question for models of simulation)
 
@@
 
+Within the same stream of modeling (complementary to more classical approaches to land-use change):
+
+\medskip
+
 \textbf{More realistic models?}
 
-\medskip
+\smallskip
 
@@
 
-\medskip
+\smallskip
 
-$\rightarrow$ Work in progress: calibration of the reaction-diffusion model on world urban areas with the Global Human Settlements Layer database
-
-\medskip
-
-$\rightarrow$ Link with sustainability indicators: GHG emissions, economics, etc. \cite{raimbault2019multi}
+$\rightarrow$ More systematic link with sustainability indicators: GHG emissions, economics, etc. \cite{raimbault2019multi}
 
@@
 
-$\rightarrow$ A novel model of urban morphogenesis at the mesoscopic scale systematically explored: \textbf{need for more coupling and comparison of models.}
+$\rightarrow$ A simple model of urban morphogenesis at the mesoscopic scale systematically calibrated: \textbf{need for more benchmarking and comparison of models.}
 
-\medskip
+\smallskip
 
@@
 \bigskip
+
+\textbf{Use and contribute to OpenMOLE: } \url{https://github.com/openmole}
+
 \bigskip
 
-\footnotesize{ - Code, data and results available at\\ \texttt{https://github.com/JusteRaimbault/CityNetwork}\\
+\textbf{Apply to the summer school: } \url{https://exmodelo.org/}
+
+\bigskip
+\bigskip
+
+\footnotesize{ - Code, data and results available at \texttt{https://github.com/JusteRaimbault/UrbanGrowth}\\
 - Acknowledgments: Thanks to the \textit{European Grid Infrastructure} and its \textit{National Grid Initiatives} (\textit{France-Grilles} in particular) to give the technical support and the infrastructure.

### 2020-02-15 Docs/Communications/ignored/XTerM2019/extendedAbstract.tex

@@
 
+We propose the following specification for the strong coupling. At a time step of the macro level $\Delta t_{\mu}$:
+\begin{itemize}
+	\item Macro level interaction model is computed and population variations $\Delta P_i$ are obtained
+	\item Meso level sprawl is simulated in the interval for each area, given this increase.
+Parameters are fixed by a baseline value but modified as a function of $\Delta P_i$ following simple rules % TODO
+	\item The resulting urban form is used to compute an endogenous activity (still with very simple stylized rules) which modifies the macro parameters for each area (these depend then on $i$: $d_G^{(i)},g_0^{(i)}$
+\end{itemize}
+
 \begin{figure}[htp]

### 2020-03-13 Docs/Papers/CityGrowth/abstract.tex
citygrwoth_abstract

\documentclass[11pt]{article}

\begin{document}

\title{City growth regularities}

The growth of cities, when represented by a single quantity which is generally population, present some striking regularities. These are at the first order independent from geography but also city size itself, as postulated by Gibrat's law proposing random growth with an uniform expected growth rate. This chapter reviews this model and associated random growth models such as Simon's model, considering them in an unified framework. We then describe simulation models proposed to explain deviations to Gibrat's law, including interaction models for systems of cities. A second part of this chapter is dedicated to the empirical test of growth regularities across main systems of cities worldwide. We finally test the sensitivity of these to the definition of cities, using the Global Human Settlement layer database. These patterns have implications for sustainable territorial planning at large scales, for example by suggesting leverages to reduce inequalities within systems of cities.

\cite{gabaix2004evolution}
\cite{batty2008size}
\cite{mitzenmacher2004brief}
\cite{gabaix1999zipf}

\bibliographystyle{apalike}
\bibliography{biblio}

\end{document}

### 2020-05-20 Docs/Papers/EvolutionaryTheory/arxiv/v1/paper.tex
paper_evolth_submit

\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm,bbm}
\usepackage{graphicx,transparent,eso-pic}
\usepackage[document]{ragged2e}
%\usepackage{natbib}
\usepackage{multicol}
\usepackage{url}
\usepackage[margin=2cm]{geometry}
\usepackage{booktabs}

\newcommand{\noun}[1]{\textsc{#1}}

\begin{document}

\title{Empowering Urban Governance through Urban Science: Multi-scale Dynamics of Urban Systems Worldwide}

\author{\noun{Juste Raimbault}$^{1,2,3,\ast}$, \noun{Eric Denis}$^3$ and \noun{Denise Pumain}$^3$\medskip\\
$^1$ CASA, University College London\\
$^2$ UPS CNRS 3611 ISC-PIF\\
$^3$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\medskip\\
$^{\ast}$\texttt{juste.raimbault@polytechnique.edu}
}
\date{}

\maketitle

\justify

\begin{abstract}
The current science of cities can provide a useful foundation for future urban policies, provided that these proposals have been validated by correct observations of the diversity of situations in the world. However, international comparisons of the evolution of cities often produce uncertain results because national territorial frameworks are not always in strict correspondence with the dynamics of urban systems. We propose to provide various compositions of systems of cities to better take into account the dynamic networking of cities that go beyond regional and national territorial boundaries. Different models conceived for explaining city size and urban growth distributions enable to establish a correspondence between urban trajectories when observed at the level of cities and systems of cities. We test the validity and representativeness of several dynamic models of complex urban systems and their variations across regions of the world, at the macroscopic scale of systems of cities. The originality of the approach is in considering spatial interaction and evolutionary path dependence as major features in the general behavior of urban entities. The models studied include diverse and complementary processes, such as economic exchanges, diffusion of innovations and physical network flows. Complex systems' dynamics is in principle unpredictable, but contextualizing it regarding demographic, income and resource components may help in minimizing the forecasting errors. We use among others a new unique source correlating population and built-up footprint at world scale: the Global Human Settlement built-up areas (GHS-BU). Following the methodology and results already obtained in the GeoDiverCity project~\cite{pumain2015multilevel,cura2017old,pumain2017urban}, including USA, Europe and BRICS countries, we complete them with this new dataset at world scale and different models. This research helps in further empirical testing to the hypotheses of the evolutionary theory of urban systems and partially revising them. We also suggest research directions towards the coupling of these models into a multi-scale model of urban growth.\medskip\\
	\textbf{Keywords: } System of cities; Urban dynamics; Co-evolution regimes; Geodiversity
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

The urbanization process has profoundly transformed the distribution and organization of human societies on the surface of the earth since the emergence of the first cities some 10,000 years ago. Physically, it corresponds to a concentration of populations in densely populated built-up agglomerations, whose dimension, very unequal, expand over no less than four orders of magnitude (from $10^3$ to $10^7$ in number of inhabitants). Economically, urbanization translates into accumulations of capital, knowledge and wealth and the multiplication of networks that accompany creation and technological innovation. Socially, urbanization promotes the diversification and increasing sophistication of forms of institutional and political organization, including an intensification and refinement of the social division of labor. Culturally, urbanization, often held to be synonymous with ``civilization'', catalyzes the evolution of urban practices and collective representations through social mixing, hybridization and education, opening perspectives towards the possibility of a better knowledge, diversity and greater acceptance of the Other. All these trends seem to represent progress for the future of humanity, which is already reflected in synthetic indicators such as average income, the index of human development, or life expectancy in good health, despite a sharp increase in social inequalities since the 1980s and the neo-liberal shift in urban governance that accompanied the debt write-offs of cities and states. However, we know today that this evolution also marks the irruption of human activity in energy balances and the functioning of the terrestrial ecological system, which leads to propose a new stratum, called the Anthropocene, in geological time scales. The cities, which have accompanied and maintained the current overconsumption of the planet's resources, are part of a modification of human relationships with the Earth during this evolution that they seem threatened, not only in their future development, but sometimes even for their survival, by the scarcity of available materials and energy sources and the disasters that could be caused by major climate change initiated since at least two centuries. Is the success of urbanization responsible for that evolution and should the Anthropocene rather be called an Urbanocene?

Actually these cities, which have often shown great resilience in past centuries, may be a solution to the problem they could be a symptom of. Their organization, partially directed and partially spontaneous, in spatially distributed, hierarchical and complementary systems of cities for the exploitation, the control, but also the service, the maintenance and the adaptation of the territories, is undoubtedly a very important asset for a proper execution of the "ecological transition" being implemented in all parts of the world. This transition can succeed if it intelligently uses the evolutionary properties of city systems, ensuring both the "top down" dissemination of international regulations and the "bottom up" circulation of the many local initiatives in favor of technical and technological processes as well as organizations that will ensure respect for biological, cultural and geographical diversity, but also a more equal distribution of resources and wealth to reduce predatory relations to nature. Diversity in all these forms seems indeed essential to the continuation of the human and urban adventure. However, it is still necessary to develop a better understanding of this multi-scale urban dynamics before providing recommendations to planners and practitioners.  Geographers have collaborated with other specialists to build relevant knowledge of systems of cities, anchored in the knowledge and comparison of diverse world region’s contexts observed over many historical periods.

We illustrate in this paper how large sources of urban data and dynamic models when powerfully and safely handled with intense computing can help to identify the diversity of co-evolution regimes that have to be disentangled for being able to propose efficient solutions to the urban problems. More precisely, we tackle the question of how to understand the properties and dynamics of large systems of cities including complementary processes driving urban growth, using new sources of comprehensive urban data which are the Global Human Settlement Layer database~\cite{florczyk2019ghsl} and the Geodivercity database~\cite{pumain2015multilevel}. Our contribution relies on the following points: (i) we provide a theoretical framework to interpret evolutionary urban dynamics at the scale of systems of cities, building on the evolutionary urban theory proposed by~\cite{pumain2018evolutionary} and on the concept of co-evolution within urban systems defined by~\cite{raimbault2018caracterisation}; (ii) we study the empirical properties of large urban systems including patterns of urban growth and scaling properties, including different definitions of urban systems; (iii) we apply and calibrate simulation models for urban dynamics on 6 of these large systems worldwide, comparing very different processes including spatial interactions, transportation infrastructures, economic exchanges, and innovation diffusion, yielding for each urban system plausible underlying  mechanisms  driving their dynamic and providing potential policy insights.

The rest of this paper is organized as follows: in the next section, we develop the theoretical framework of the evolutionary urban theory and the underlying urban growth models based on spatial interactions; a third section complements this theoretical background by developing the concept of co-evolution within systems of cities. We then study empirical properties of large urban systems worldwide, and calibrate dynamical models of urban growth on these systems. We finally discuss future developments and possible implications for the sustainability of urban systems planning and management.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Geographical models of urban growth within systems of cities}

The growth of cities is often interpreted according to the decisions taken locally by a multitude of very diverse stakeholders, for instance decisions about urban policies, locational strategies from firms, or motivations for residential migrations. These actions from urban stakeholders appear at a first sight as the direct causes of urban growth, at this micro-level of individual decisions. Obviously the growth or decline of a city is resulting from aggregating all these processes, and also from possible radical changes in environmental conditions.  But because of until recently a lack of data at this level of observation, most models of urban growth were developed at meso-geographical level for subsets of urban units. As cities mostly grow demographically and expand spatially by aggregating new populations and activities from their center toward peripheries according to monocentric or polycentric patterns, the growth should be computed for evolving urban agglomerations or functional urban areas that are properly delineated at each date of the period under observation. The comparability of many results about urban growth and distribution of city sizes is too often hampered because authors did not apply this geographical principle when defining and delineating properly the urban units they consider (for instance: \cite{eeckhout2004gibrat} or \cite{xu2014discontinuities}, see \cite{cottineau2017metazipf} for a full review). Systematic investigations were made recently after developing original harmonized data bases on thousands of urban agglomerations over decades and even centuries in the GeoDiverCity project~\cite{pumain2015multilevel,cura2017old,pumain2017urban}, including USA, Europe and BRICS countries.

Such consistent statistical observations on thousands of cities and over centuries, enabled at first to confirm the conclusion that emerged in pioneer works on the growth process within integrated systems of cities (i.e. well connected cities obeying a unified system of political, cultural and economic rules that currently define a ``territory''): each city has a probability of growing similar to other cities belonging to the same territorial system. This was characterized as a `` distributed growth'' process that can be observed on the long run with many local and temporal fluctuations in any system of cities all over the world \cite{gibrat1931ingalits,robson1973view,pumain1982dynamique}: Gibrat’s model of urban growth constitutes a first good approximation of the distribution of urban growth within a system of cities. Gibrat's ``law of proportional effect'' means that growth rates are equiprobable whatever city size and are not correlated with previous growth rates.

This rather good fit already provides a double explanatory gain: it explains the remarkable persistence of urban spatial patterns and hierarchies over very long periods of time that may exhibit such meta-stability with very little changes over centuries. And moreover it provides an ``explanation'' for the statistical shape of urban sizes distribution (that is lognormal, close to Zipf’s law or other types of skewed distributions), according to a stochastic process that was also anticipated by Herbert Simon in 1955. Conversely, despite attempts at developing an economic interpretation of the genesis of urban hierarchies through firm choices or preferential attachment~\cite{gabaix1999zipf}. P. Krugman still considers it as ``a mystery'' regarding its explanation within economic theory~\cite{krugman1996confronting}.

Many observed distributions of city sizes (actually: settlement sizes including hamlets, villages, towns and SMAs) are close to lognormal distributions (evidence from \cite{robson1973view,pumain1982dynamique,eeckhout2004gibrat,decker2007global} and Gibrat's growth model mathematically leads on the very long run to a lognormal distribution of city sizes, but some of the hypothesis of Gibrat's growth model are sometimes partially rejected through statistical testing: more or less high correlations may be found between growth rates and city size (most of the time positive correlation), and positive correlation between successive growth rates also may be observed for some time periods.

That is why complementary models have been proposed that offer better fits when adding the possible effects of spatial interactions to Gibrat's model (which is a stochastic model where urban agglomerations are represented as independent units, which of course cannot be a relevant geographical model for epistemological reasons). \cite{favaro2007croissance} and \cite{favaro2011gibrat} developed a geographical model of urban growth including: a recurring emergence of clustered new innovations that create growth cycles (following a Schumpeterian process); a spatial diffusion of innovations occurring through a dynamic spatial interaction model \cite{wilson1971family}; a spatial diffusion of innovation according to a hierarchical process \cite{hagerstrand1952propagation}. The growth of a city depends on its share of labour force in each innovation cycle that induces a scaling parameter larger than one for the urban activities that participate in the innovation wave of the moment.

Analytically, the model can be expressed in a form that is very close to Gibrat’s model:

\begin{equation}
    \log{P_{i,t}} = \alpha + \log{P_{i,t-1}} + G_i + u_{i,t}
\end{equation}

where $G_i$ holds for the ``bias'' noticed in estimating Gibrat's model by Ordinary Least Squares (linked with spatial interaction processes, see \cite{favaro2011gibrat}). The noises $u_{i,t}$ are stochastic variables in time and space with zero averages, which allow capturing random fluctuations. They correspond to the random distribution of growth rate in the initial Gibrat's model. The endogenous growth rate $\alpha$ and the growth induced by interactions are deterministic given a realization of populations at the previous step.

This model has the advantage to replace a generic statistical model of growing independent entities (Gibrat’s urban growth model) by \emph{a model of spatially and temporally interdependent entities} (i.e. the geographical concept of ``system of cities'' or ``settlement system''). It reproduces the observations on differential scaling parameters for urban activities according to their age in innovation cycles~\cite{pumain2006evolutionary}. It also makes explicit the multilevel dynamics of interurban competition for capturing innovation, which may itself generate new innovation through interurban emulation, within an evolutionary perspective.

Moreover, this model enables us to interpret variants that represent the path dependence effects occurring through territorial differentiation. For instance, in new urban systems, as in USA, there is a spatial filling process that occur through spatial waves of urban growth (urban Frontier) corresponding to the diffusion of economic cycles \cite{bretagnolle2007formes}, whereas in mature urban systems, as in France, the innovations diffusion reaches cities that are not spatially regularly arranged but already experienced other growth periods according to distinct cycles of urban specialization~\cite{paulus2004coevolution,finance2016villes}.

Other types of models based on interactions have been introduced within the framework of the evolutionary urban theory~\cite{pumain2006evolutionary}. A deterministic version of the previous Favaro-Pumain model, based on economic exchanges between cities, was introduced by \cite{cottineau2014evolution}. This model called Marius was originally designed for the Former Soviet Union but applies to any system of cities, as only population trajectories are required to be parametrized to estimate and evaluate the model. Scaling laws are used to extrapolate economic activities of cities based on population, and economic exchanges are then simulated. The model is more precisely a \emph{family of models}, since different processes can be taken into account in a multi-modeling way (for example, local impact of energy resources, switch between top-down planned economic structure and bottom-up interactions only). \cite{cottineau2015growing} formulated this methodological framework as an ``evaluation-based incremental modeling method'', which allows testing of concurrent hypotheses to explain trajectories of systems of cities.

This type of models was also applied to the subject of transportation networks within urban systems. The SimpopNet model introduced by \cite{schmitt2014modelisation} captures entangled trajectories of city populations and network links between these cities in a stylized setting. It was shown by \cite{raimbault2020unveiling} that it effectively reproduces co-evolutionary dynamics (in a precise sense we develop below), which is a central theoretical feature in evolutionary urban theory. More recently, \cite{raimbault2020indirect} proposed a simplified version of the Favaro-Pumain model, in which the deviation due to interactions includes physical network effects, through direct spatial interaction models, but also indirect feedback of network flows on city growth. The model was calibrated on real data for France over a long time span (1830-2000). It was extended to a co-evolutionary model in \cite{raimbault2018modeling}, also calibrated on French data with railway network accessibility, for which the capacity to capture a broad range of co-evolutionary regimes was demonstrated.

These different models have the particularity to enter in the same theoretical frame of the evolutionary urban theory, of being backed-up by empirical studies and datasets constructed during the GeoDiverCity ERC project \cite{pumain2015multilevel}, and to simulate urban systems population trajectories on long time scales and large spatial scales (macroscopic scale of a region, country or continent). They are a fundamental component of our evidence-based approach to urban systems, and will be applied and compared in a benchmark below.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Systems of cities and co-evolution regimes}

We first need to introduce another theoretical component justifying the use and comparison of these different models within the evolutionary urban theory. Evolutionary economics~\cite{nelson2009evolutionary} have a wide experience in transferring the concept of evolution from its initial context in biology \cite{durham1991coevolution}. This requires to be able to identify within the studied systems several components, which are transmission processes, transformation processes and differentiation between populations of agents emerging from these processes. Artificial Life literally studies such artificial systems, extending biological systems to ``life as it could be'' \cite{langton1997artificial,bedau2003artificial}. Although the core component are not always explicit, and these concepts are sometimes relaxed to long-time structural dynamics of systems components, evolutionary concepts have been successfully applied in economics as mentioned above, but also in social sciences with the study of cultural evolution~\cite{mesoudi2017pursuing} or the evolution of social organizations \cite{volberda2003co}. The interplay between social components and biological components can even be considered \cite{bull2000meme}.

In that context, co-evolution originally corresponds to evolutionary changes in two species in strong mutual influence \cite{janzen1980coevolution}. The idea of diffuse co-evolution, taking into account the environment in which co-evolution takes places, including a broad network of other species, was proposed as a refinement of this concept \cite{strauss2005toward}. To effectively designate co-evolutionary processes, or more generally strongly entangled dynamics, it was also developed in economic geography \cite{schamp2010notion}. Thematic applications include the study of economic clusters \cite{ter2011co}, of technological change \cite{colletis2010co}, or environmental economics \cite{kallis2007coevolution}.

In geography, it was particularly developed by the evolutionary urban theory~\cite{pumain1997pour}, which in practice consists in a dynamical non-equilibrium approach to urban systems as complex adaptive systems, in which interactions between components are central \cite{paulus2004coevolution,schmitt2014modelisation}. In that context, building on the definition introduced by~\cite{raimbault2018co,raimbault2018caracterisation}, we propose a multi-level definition of co-evolution which is consistent with the evolutionary urban theory~\cite{pumain2018evolutionary}. First of all, transformation processes of territorial components should induce evolutionary dynamics at different scales. It is not clear what would be ``an urban genome'', but empirical evidence of innovation diffusion and cultural evolution (including social progress and changes in governance), but also physical transformations or physical flows between cities, suggest that these dynamics can be interpreted as evolutionary in a loose sense. Then, co-evolution can occur at a microscopic level between particular artifacts or agents (coupled dynamics), at a mesoscopic scale of a population of entities (in the initial biological and statistical sense, i.e. that characteristics of two populations of entities mutually influence themselves in their evolutionary trajectories), and finally at a macroscopic scale in the usual sense used in geography as strongly entangled dynamics at the system level.

This definition implies the existence of co-evolution niches in the sense of \cite{holland2012signals}, which can be understood as a system of boundaries and corresponding subsystems in which co-evolution takes place. This pertains particularly well for territorial systems, for which the niches will be consistent territorial ensembles, imbricated at different scales. The cities at a regional level are for example a first level of niche, embedded into the urban system at a larger territorial scale as the national, continental and even global one. Within a given niche, the co-evolving populations of entities will be in a specific co-evolution regime, concept introduced by \cite{raimbault2017identification} in the case of transportation networks and territories, to describe a given causal network between the variables considered. For example, population of cities and their centrality in the transportation network may be in a circular causal relationship (taken as a weak Granger causality), or in a unidirectional relation, or even in a triangle relationship with a third variable such as accessibility. Each case is a particular co-evolutionary structure - even if strictly speaking there is no co-evolution when there is no circularity, the term of co-evolution regime can be used in a broad way to describe this state of dynamical relationships.

This directly implies that (i) urban systems worldwide have each their own co-evolutionary trajectory, and thus their own driving processes; (ii) the way they could merge into a global urban system implies possibly other types of processes (emergence of mega-city regions e.g.) and is not incompatible with the persistence of local regional urban systems. A quantitative investigation of the second point is still out of the scope of this preliminary work, as it would imply more complicated and possibly multi-scale models, but also more accurate global data on a longer time span. We will however provide a first experiment below by testing models on a new global database. The first point is consistent with the idea of comparing different models with underlying processes of a different nature to try to reproduce the trajectories of urban systems, and this for different systems across the world. We will also provide results for such experiments below. The test of such dynamical models is an indirect way to dig into the diversity of co-evolution regimes, as data is generally missing for a direct investigation. The model fitting the best a given urban system and corresponding extrapolated parameters (for example interaction distance and hierarchy for a spatial interaction model) provide information on the underlying co-evolution dynamics, and can even in some case be associated with regimes directly identified in synthetic settings (such as the ones found by \cite{raimbault2018modeling}, but this elaborated investigation also remains out of the scope of this paper).

Building on this theoretical background, we will now investigate a new global database of urban areas and some stylized facts that can be extracted from it, and then test and compare dynamical models for urban systems.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A new source of data for comparing urban trajectories worldwide}

We test here the generality and robustness of stylised facts on systems of cities when using a new source of comparable urban data provided by the Global Human Settlement layer dataset (GHS\footnote{Source: GHSL (Global Human Settlement Layer) : GEO Human Planet Initiative (European Commission)}). The dataset was already explored statistically for comparing trends of urban sprawl in the countries of the world by \cite{denis2020population}, while \cite{melchiorri2019principles} computing a land use efficiency index measured a global trend towards densification with a diversity of urban trajectories according to regions of the world. The data delineate comparable morphological urban agglomerations by detecting built up area from satellite images at 40 m resolution, and combining it with a population layer generated at 250m resolution using local data at municipal and district levels provided by censuses on a regular 1 km\textsuperscript{2} grid \cite{dijkstra2014harmonised}.

The dataset is available at different dates between 1975 and 2015. In 2015 the source delineated precisely some 13,000 urban agglomerations larger than 50,000 inhabitants in the world. For each urban agglomeration figures on built-up surface and population are provided in 1975, 1990, 2000 and 2015. It also includes measurements of GDP, green areas and pollution levels from 1990 to 2015.

The short period of time considered is the major limitation of the source, that can be mitigated because of the multiple case studies that are now available from comparative analysis of urban trajectories over much longer time spans, such as \cite{robson1973view} for England and Wales 1801-1960; \cite{pumain1982dynamique} for France 1831-1975; \cite{bretagnolle1999systemes} for France 1831-1990; \cite{bretagnolle2007formes} for the US 1800-2000; \cite{swerts2013systemes} for India 1901-2011 and China 1982-2010; \cite{rozenblat2018international} on many large regions in the world and various time-spans.

\cite{dijkstra2018applying} have compared the GHSL data with the statistics provided by the World Urbanization Prospects \cite{desa2018world} and found a rather good fit with the data on individual cities larger than 300,000 inhabitants. We confirm that the GSHL source provides statistics that are roughly compatible with the results that were obtained in different countries of the world using our dedicated harmonized data bases (Table~\ref{tab:populations}). The total urban populations of each country or region should be smaller in GHSL because in GeoDiverCity we considered urban agglomerations larger than 10,000 inhabitants, but this is not always the case, especially for India and China whose urban populations could be overrepresented in GHSL. However, computing the slope of rank size distribution brings comparable results about the different structures of urban hierarchies, with less unequal distributions of city sizes in countries that developed earlier their systems of cities \cite{pumain2015multilevel}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{\textbf{Representativeness of GSHL source compared to harmonized data bases from GeoDiverCity.} Population is given in Millions. All indicators are given for the GHSL database, except for the population from the Geodivercity database for comparison (third column). The Rank-size exponent is estimated with a standard OLS on logarithms. Standard deviations of the rank-size exponent are all smaller than 0.02, and the adjusted R-squared larger than 0.97. FSU = Former Soviet Union.\label{tab:populations}}
\centering
\begin{tabular}{cccccc}
    \toprule
        System & Pop (M) & Pop geodiv. & Cities & Rank-size  \\
        \midrule
         Europe & 188 & 291 & 693 & $0.94$  \\
         China & 567 & 481 & 1850 & $0.91$  \\
         Brazil & 112 & 161 & 349 & $0.99$  \\
         India & 703 & 427 & 3248 & $0.78$  \\
         South Africa & 25 & 25 & 77 & $1.05$ \\
         US & 153 & 324 & 287 & $1.16$  \\
         FSU & 120 & 174 & 450 & $0.92$ \\
         \bottomrule
\end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In order to roughly check the global reliability of the data set we computed the now classical indices describing urban hierarchies with the slope of Zipf's rank-size rule. This can be done for a diversity of ways of grouping cities, i.e. dividing the world in consistent systems of cities. Delineating proper systems of cities at the macro-level of inquiry is an even more delicate exercise than delineating urban entities at meso-geographical level. Theoretically one should consider subsystems that have more internal than external interactions with other cities, which raise difficult issues of determining which interactions to consider and over which period of time. Even if it was possible to identify and measure well adapted data, in most cases we would find that the largest cities of any system of cities have a much larger scope of interactions than the smaller towns \cite{bretagnolle2010simulating}. We shall try to elaborate more on this particular problem in a next paper, thus in this first attempt here we choose to experiment on different types of groupings, using the seven national or regional urban systems for which we had developed alternative databases in the GeoDiverCity programme \cite{pumain2015multilevel} and comparing with two other ways of grouping countries: a first one according to the main five continents of the world as many studies on flows of air flight passengers or even networks of branches owned by multinational firms demonstrated that they often encompass clusters of stronger internal interactions and have discontinuities between them; a second one is made according to major economic trade zones that are also supposed to be subsets of denser inter-urban exchanges.

A first simple statistical analysis confirms results on stylised facts that were already observed on many systems of cities. We show in Figure~\ref{fig:fig1} the rank-size plot for all 5 continents, at three different dates. We find as already known that the urban hierarchy roughly remains constant in time. Numerical values for exponents in this estimation are provided in Table~\ref{tab:tab2} for the year 2015. We also retrieve the fact that more mature urban systems (Europe) have a exponent closer to one, while recently booming urban systems such as Africa have an exponent far from one (0.78 in 2015) – such a result may also be linked with the absence of smaller towns in GHSL. Asia has an intermediate value of 0.87, what would be consistent with the fact of mixing subsystems which have a very long history (Japan and China e.g.) but also recently underwent drastic demographic transitions, with other subsystems whose development is more rapid and recent (South-East Asia). The primacy indices are also consistent with what could be expected: note that it is larger for Europe when Moscow is included (it would be close to one if taking the EU only for example, Paris and London being approximately the same size).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig1.png}
\caption{Evolution of rank-size distributions by continents 1990-2000-2015.\label{fig:fig1}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{Rank-size by continents in 2015. Source: GHSL, European Commission.\label{tab:tab2}}
\centering
\begin{tabular}{cccccc}
\toprule
System & Population (M) & NB of cities & P1/P2 (primacy) & Slope of rank-size & R2 \\
\midrule
Europa & 288 & 1067 & 1.45 & $0.93 \pm 0.003$ & 0.991 \\
America & 547 & 1521 & 1.02 & $1.02 \pm 0.002$ & 0.996 \\
Asia & 2143 & 7737 & 1.12 & $0.87 \pm 0.0004$ & 0.998 \\
Africa & 585 & 2876 & 1.7 & $0.78 \pm 0.0008$ & 0.997 \\
Oceania & 19 & 86 & 1.08 & $0.91 \pm 0.027$ & 0.926\\
\bottomrule
\end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We also study the rank-size properties by grouping the countries into trade areas, which are supposed to capture subsystems with a higher level of interurban interactions. We show the temporal evolution of rank-size plot in Figure~\ref{fig:fig2}, and corresponding statistics in 2015 in Table~\ref{tab:tab3}. Although the trade areas considered highly overlap with continents (in particular for EEA with Europe and ASEAN with Asia), the exponents obtained are different from the previous ones, and closer to one. According to \cite{arcaute2020scaling}, varying the definition of cities yield varying Zipf exponents, with no endogenous privileged definition. We can indeed expect the same when varying the system of cities considered, as \cite{corral2020truncated} show that power-law are not valid anymore when considering the tail of the distribution, i.e. changing the number of cities included in the estimation. If Zipf’s law was a pure product of ergodic stochastic processes without any interactions between cities, randomly sampling a subset of a given system should yield roughly the same exponent (at least with large sample size and with an OLS estimator). Although the sampling is not random here (a more elaborated statistical analysis remains for future work), the discrepancy suggests the opposite of the previous case and therefore that urban systems are highly non-ergodic and that interactions between cities are crucial. We suggest that the exact same phenomenon occurs for urban scaling laws, since they are in the same way strongly dependent to system definition \cite{arcaute2015constructing,cottineau2017diverse}, and that the claims of universality by the mainstream scaling literature are at least not compatible with this empirical evidence, at worse inaccurate in terms of underlying dynamical processes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig2.png}
\caption{Evolution of rank-size distributions by trade zones 1990-2000-2015.\label{fig:fig2}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{Rank size for trade zones in 2015. Source: GHSL, European Commission. [ASEAN =Association of Southeast Asian Nations, 10 countries; MERCOSUR= Southern Common Market, 4 countries; COMESA= Common Market for Eastern and Southern Africa, 21 countries; EEA= European Economic Area, 31 countries]\label{tab:tab3}}
\centering
\begin{tabular}{cccccc}
\toprule
System & Population (M) & NB of cities & P1/P2 (primacy) & Slope of rank-size & R2 \\
\midrule
ASEAN & 293 & 874 & 1.67 & $0.92 \pm 0.003$ & 0.993 \\
MERCOSUR & 220 & 657 & 1.37 & $1.00 \pm 0.0016$ & 0.998 \\
COMESA & 252 & 1367 & 3.39 & $0.72 \pm 0.0014$ & 0.995 \\
EEA & 194 & 720 & 1.01 & $0.94 \pm 0.0026$ & 0.994 \\ 
\bottomrule
\end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We also proceed to a simple statistical analysis of other variables in the database. We confirm the basic assumption of the Gibrat's model that population and population growth are uncorrelated. The correlation coefficients on Figure~\ref{fig:fig3} are above 0.7 between built-up area and population, as well as GDP and pollutant emissions. Cities are places of concentration of human activities with their desirable and less desirable outputs. However, there are wide inequalities in urban densities according to the continents (Asian cities are more than twice as dense as European cities that are ten times denser than North American and Australian ones, cf \cite{bertaud2003spatial}). Even if the progression of built-up area between 2000 and 2015 is positively correlated with the one of population, with a 0.63 coefficient, there are discrepancies between regions of the world in the evolution towards rather compact or more spread urbanization (Figure~\ref{fig:fig4}). The map shows clearly that sub-Saharan regions in Africa and south and southeastern Asia are expanding more rapidly the urbanized surfaces than their urban population – although being as well regions of rapid demographic urban growth. The variations of other variables, i.e. CO2 emissions and GDP, are totally uncorrelated with the variations of population, what means that their recent dynamics are independent.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig3.png}
\caption{Correlations between GHSL indicators (GDP, Emissions, Built-up surfaces, Population  and their evolution 2000-2015 (Delta)).\label{fig:fig3}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig4.png}
\caption{Comparing spatial and demographic urban expansion 2000-2015. Source: GHSL, European Commission.\label{fig:fig4}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We also compute the exponents of the data on GDP and emissions as power functions of urban sizes, under the hypothesis that they would exhibit scaling laws. Actually, the relationship between urban indicators in GHSL are not linear but follow scaling laws with exponents above 1, indicating a larger production and pollution potential when considering cities of larger size. This can be observed for subsets of cities we had previously observed in the GeoDiverCity programme in Table~\ref{tab:tab4}. The adjustments are in several cases of a low quality (for India for example), suggesting the relative validity of scaling laws. We did not proceed to additional tests to verify if these relations are effectively power laws (see \cite{broido2019scale} for the case of scale-free networks), but the chances are high for not having power-laws in a strict statistical sense. The distributions are still loosely fat-tailed and the exponents provide useful information. They lie close to one in the case of built-up area as expected, supra-linear but close to one for GDP also as expected, and supra-linear with a high exponent for emissions. The last relation on emissions was less expected, since according to \cite{ribeiro2019effects} the literature has very heterogeneous conclusions on emissions scaling, and the results vary highly depending on city definitions. There is however a high supra-linear scaling when considering urban areas, which is consistent with our findings. We obtain similar order of magnitudes (maximal variations of 0.2) when considering other geographical areas as above (continents and trade areas), confirming a certain robustness in the qualitative interpretations that can be done from these estimations. Note that this database holds promises for future more robust and useful scaling studies, as it has a worldwide coverage and the systems considered can be varied at ease, and includes fundamental variables related to urban issues (many scaling studies focusing on anecdotal variables such as the number of cinemas).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{Scaling exponents for large urban systems in 2015. We provide for each system the OLS estimates on logarithms of each variable as a function of population, with standard errors and adjusted R squared. FSU= Former Soviet Union. Source: GHSL, European Commission.\label{tab:tab4}}
\centering
\begin{tabular}{cccccc}
    \toprule
    System & Built-up area & GDP & Emissions \\
    \midrule
Europe & $0.93 \pm 0.016$ (0.83) & $1.15 \pm 0.019$ (0.83) & $1.50 \pm 0.038$ (0.69) \\
China & $1.06 \pm 0.019$ (0.62)  & $1.14 \pm 0.011$ (0.85)  & $1.84 \pm 0.037$ (0.57) \\
Brazil & $0.98 \pm 0.025$ (0.81)  & $1.10 \pm 0.055 $ (0.54) & $1.71 \pm 0.053$ (0.75) \\
India & $1.34 \pm 0.031$ (0.36) & $1.25 \pm 0.022$ (0.50) & $1.54 \pm 0.031$ (0.42) \\
S. Africa & $1.18 \pm 0.090$ (0.69) & $1.08 \pm 0.028$ (0.95) & $1.56 \pm 0.087$  (0.81) \\
US & $0.97 \pm 0.015$ (0.92) & $1.04 \pm 0.069$ (0.99) &$ 1.34 \pm 0.03$ (0.84) \\
FSU & $0.97 \pm 0.035$ (0.63) & $1.17 \pm 0.041$ (0.65) & $1.95 \pm 0.088$ (0.52)\\
\bottomrule
\end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Comparing dynamic models of urban growth}

We now turn to the test of different dynamical models to simulate the observed population trajectories of cities. As detailed above, this should allow to compare the influence of different possible drivers of urban growth, and possibly investigate the diversity of co-evolution regimes. We apply four different families of models, on the GHSL database between 1990 and 2015 and on the GeoDiverCity database. The first one is the stochastic model of Gibrat that can be used as a benchmark because it has the smallest number of hypotheses, without any interactions between cities. The second model inspired from \cite{favaro2007croissance} introduces innovation waves as impulses of urban growth with hierarchical diffusion, as detailed in section 2. A third model family is the Marius model, introduced by \cite{cottineau2014evolution}, which focuses on economic interactions between cities. The last model family is a spatial interaction model taking into account physical networks, studied by \cite{raimbault2020indirect}. The underlying processes within each of these model families are potential partial explanations for urban growth, and each belong to very different classes of processes.

All these models can be formulated within a common framework which has been described by \cite{raimbault2020indirect}. First of all, we consider a deterministic version of the Gibrat's model, for which the extensions with interactions will capture the covariance structure between population trajectories. In the Gibrat's model, formulated as $P(t+1) = R(t) \cdot P(t)$ where $P$ and $R$ are independent random variables, we have $\mathbb{E}\left[P\right](t+1) = \mathbb{E}\left[R\right](t) \cdot \mathbb{E}\left[P\right](t)$. Furthermore, city trajectories are assumed independent, i.e. $\textrm{Cov}\left[P_i,P_j\right] = 0$ for any cities $i \neq j$. Writing $\mathbb{E}\left[P_i\right](t) = \mu_i(t)$, we generalize the deterministic formulation above to a non-linear one by taking $\mu_i(t+1) = f(\mu_i(t))$. The specification of the transition function or algorithm between these average populations will fully determine the model. The corresponding Gibrat's model (named ``gibrat'' on Figures~\ref{fig:fig5} to \ref{fig:fig9}) has one single parameter which is the average endogenous growth rate. Note that in this deterministic version, there is no additional parameter for the variance (or other moments depending on the distribution chosen) of growth rates.

The network interaction model proposed by \cite{raimbault2020indirect} includes linearly a Gibrat's component of fixed endogenous growth rates, a spatial interaction component given by the average interaction potential with all other cities for each city (with Euclidian or network distance, see \cite{raimbault2020indirect} for details) and a second order term of network flow feedback that we do not include here for simplicity. We consider two versions of the model, one (named ``intgib'' for ``Interaction Gibrat'' on Figures~\ref{fig:fig5} to \ref{fig:fig9}) with Euclidian distance between cities to determine interaction potential, the other (named ``intgibphysical'' for ``Interaction Gibrat Physical'' on Figures~\ref{fig:fig5} to \ref{fig:fig9}) with a physical network distance computed as shortest paths with a terrain slope impedance derived from a global Digital Elevation Model. Both models have the same four parameters, namely endogenous growth rate, weight of interactions, hierarchy of interactions, and geographical range of interactions.

The Favaro-Pumain model for the diffusion of innovation \cite{favaro2011gibrat} considers population of cities and additional variables representing an adoption rate of a given innovation. To evolve populations, (i) innovations are diffused in the network of cities following a spatial interaction model and with an intensity depending on the utility of the innovation; (ii) population are updated following another spatial interaction model, interaction potential being driven by the innovative characters of cities; (iii) we introduce exogenously a new innovation with an increased utility if a certain global adoption threshold is reached for the previous innovation, at a fixed initial penetration rate and in a city chosen with a probability calculated according to a scaling law of population. A first simplified version of this model (named ``innovation'' on Figures~\ref{fig:fig5} to \ref{fig:fig9}) has default parameter values from \cite{favaro2011gibrat} and four free parameters which are the endogenous growth rate, the weight of interactions, interaction range for innovation diffusion and interaction range for population growth. The full version (named ``innovationext'' for ``Innovation extended'' on Figures~\ref{fig:fig5} to \ref{fig:fig9}) has nine parameters, with additional parameters being the initial utility of the first innovation, the fixed growth rate of innovation utilities, the initial penetration rate, the adoption threshold for a new innovation, and the hierarchy exponent to determine innovative cities.

Finally, the Marius model family based on economic exchanges \cite{cottineau2014evolution} implements the following processes. Cities are attributed an initial wealth following a scaling law of populations. At each time step, (i) supply and demand are updated for each city as superlinear functions of populations; (ii) cities exchange goods according to a spatial interaction potential and their supply and demand, and wealth are updated accordingly; (iii) population are updated such that population difference follows a scaling law of wealth difference with a given economic multiplier and exponent. A restricted Marius model (named ``mariusrestr'' for ``Marius restricted'' on Figures~\ref{fig:fig5} to \ref{fig:fig9}) has four parameters, namely economic multiplier, supply and demand exponents, and the interaction distance. The full model (named ``marius'') has six parameters, adding the exponent for the initial wealth and the exponent for the population update.

Note that besides the benchmark Gibrat’s model, we have a version of each model with four parameters, which makes them directly comparable in terms of adjustment performance. Indeed, taking into account over fitting in such simulation models remains an open question as \cite{raimbault2020indirect} puts it, and a fair model comparison is ensured with the same number of parameters. We will however consider all model versions in the comparison and study absolute performance of models whatever their number of degrees of freedom. The models are implemented in Scala within a single library, and integrated into the OpenMOLE model exploration platform for exploration and calibration~\cite{reuillon2013openmole}. Source code and results are available on the repository of the project at \url{https://github.com/JusteRaimbault/UrbanGrowth}.

We show in Figure~\ref{fig:fig5} how different dynamic models succeed in simulating the population trajectories of all cities for BRICS countries, Europe and the United States. Here each model is adjusted on the GeoDiverCity dataset for each system of cities, for time spans covering 1960 to 2010 (precise dates depend on each system). Evaluating models on this dataset rather than on the GHSL database seems more relevant for these systems of cities, as data was specifically tailored for comparability and following a consistent geographical definition of urban areas with more refined estimation of their populations \cite{pumain2015multilevel}. The model calibration procedure is the following. Cities are initialized with observed population at the first date in the dataset. Given a model and associated parameter values, populations are then simulated for each date in the dataset according to the model. The fitness is evaluated with two complementary indicators: (i) the logarithm of the total mean square error between observed and simulated populations, taken in time and for all cities; (ii) the mean square error on logarithms of observed and simulated populations taken in time and for all cities. These two indicators are complementary, because of the hierarchical nature of urban systems: considering only the mean square error will favor the adjustment on very large cities only, while considering logarithms of population will put a higher emphasis on the role of medium-sized and small cities, which must not be neglected when considering an urban system \cite{aveline2018pathways,denis2017subaltern}. A multi-objective calibration algorithm (more precisely the NSGA2 algorithm), implemented in the OpenMOLE platform with a specific design to be distributed on a computation grid \cite{schmitt2015half}, was run with these two objectives for each city system and each model. The algorithm is stopped after a fixed number of steps, for which convergence was always obtained (in the sense of negligible variation in the final populations obtained).

The curves of different color in Figure~\ref{fig:fig5} are representing the Pareto fronts, i.e. the points forming an optimization compromise between the two objectives, for each system of cities (subpanels) and each model (colors). The curves represent Pareto fronts obtained with the final generation of the genetic algorithm, corresponding to 200 simulations of each model (one point corresponds to one simulation). On the whole, models with innovation and interaction are performing better for reconstructing the trajectories than the simplest stochastic model, but the patterns of fitness are very diverse. In some cases such as South Africa, Russia and Europe, the full Marius model is clearly dominating all other models, since its Pareto front performs better than the Pareto fronts of other models regarding the two objectives. In these cases, economic exchanges are thus better candidates than the other processes considered to explain urban growth. These are the most mature systems among the ones under study, what would suggest a correspondence between the age of the system and the fact that its dynamics are driven by economic exchanges (furthermore whatever the political and economic system in place, since Europe and Russia before and after the fall of USSR are covered). In some cases such as China, several models are in close competition, while in the remaining cases, different models are complementary to obtain the effective Pareto front.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig5.png}
\caption{\textbf{Test of six dynamic models simulating urban trajectories on systems of cities at national scale.} \textit{(Legend common to figures 5 to 9)}: two measurements of the fitness of each dynamic model (log of sum of deviations between observed and simulated populations on y axis [ensuring a better fit of the top of urban hierarchies] and sum of deviations between the logarithms of observed and simulated population of each city [ensuring a better fit for medium and smaller towns]) are related on the Cartesian graph exhibiting Pareto fronts that can be compared for the seven dynamic models. The calibration of exploration models were done using genetic algorithms and computation grid with the OpenMOLE platform.\label{fig:fig5}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We give more details on these cases where different models appear to be complementary. Figure~\ref{fig:fig6} and Figure~\ref{fig:fig7} represent contrasted patterns of the Pareto fronts in the case of India and Brazil. In the case of India, for the models shown here Gibrat's model provide the worse predictions of urban trajectories, whatever their position in the urban hierarchy. Obviously introducing interactions, and especially physical interactions, is a necessary and major improvement for simulating the development of Indian cities of all sizes. The interaction model and its physical version (blue and light green) are the best but very close to the full innovation diffusion Favaro-pumain model (dark green). In that case, the two processes are equivalent candidates to explain urban growth. Note that we show here a zoomed region of the previous plots in Figure~\ref{fig:fig5}, and that in that case the Marius economic model is outside the plot range, even dominated by the one-parameter Gibrat’s model. Several explanatory factors for this dominating role of spatial interactions in the Indian system of cities could be proposed, such as a history of a stratified urbanization due to successive colonial extraction periods, but also the capture of a long distance commerce which spans beyond the boundaries of India or its current insertion within globalized value chains.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig6.png}
\caption{Testing urban dynamic models on the Indian case.\label{fig:fig6}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For Brazil (Figure~\ref{fig:fig7}) the pattern is more complicated because multiple factors differentiate urban trajectories. The Pareto front is formed by the full Favaro-Pumain model (\textit{innovationext}, dark green). The innovation model is always the best whatever the city size considered,  what could be consistent with the historical and geographical context of Brazil: innovations spilling out of the newly founded Brazilia should necessarily have impacted surroundings medium-sized towns. However, we also obtain the physical interaction model (\textit{intgibphysical}, light blue) within the zoomed windows as a second Pareto front. This can be interpreted as also a good explanatory power for spatial interactions taking into account the physical terrain. In that sense, the uneven topography of Brazil must play a role in shaping transportation networks and urban interactions, which is consistent with inequalities in extraction potential, in particular for agriculture.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig7.png}
\caption{Testing urban dynamic models on the Brazilian case.\label{fig:fig7}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The case of Chinese cities is interesting, not only because of the peculiarities of the urbanization process in this rapidly urbanizing country with a still very old system of cities since a half century \cite{wu2020emerging}, but also because the complication of local sources of territorial information about urban populations \cite{swerts2013systemes,swerts2017data}. What emerges from Figure~\ref{fig:fig8} that represents the fitness of dynamic models is a striking quasi-equivalence between all models considered, except for the Marius model (purple) slightly dominating the other Pareto fronts when medium-sized cities are privileged. Otherwise, all models are very close to the Gibrat's model (red). Such an ``anomaly'' when compared to the simulations made on all other regions of the world can be interpreted to be produced by the systematic character of the Chinese urban planning aiming at developing urban areas in a parallel way all over the Chinese urban regions. Models introducing spatial interaction produce patterns expected when geographical bottom-up processes of interaction are considered only, but they cannot anticipate highly top-down planned urban development actions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig8.png}
\caption{Testing urban dynamic models on China.\label{fig:fig8}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We have also applied these models to the full set of cities in the world as documented from the GSHL database (Figure~\ref{fig:fig9}). Only such a broad coverage database, despite its potential bias, can allow applying such interaction models at this global scale, and this application thus shows the complementarity of data sources and the potentialities of GHSL. Although we can expect a larger diversity of urban trajectories at that scale because of the much contrasted demographic behavior in regions where the urban transition is achieved and those where it is still going on, this does not seem to hamper the performance of the models. Gibrat's remains the most approximate way of predicting urban trajectories, but what emerges at that world scale is a stronger differentiation between the ability to simulate the top or the bottom of urban hierarchies. Largest cities are rather well approximated with the innovation diffusion model (green), while smaller cities are better adjusted with the Marius economic model. This would be consistent with metropolization processes implying that large global cities interact more between themselves than with their hinterland, at least for highly innovative and value-added activities. Meanwhile, on a regional scale for medium-sized cities (these interacting in smaller ranges than large cities because of the size term in the spatial interaction model), economic exchanges are driving urban growth.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Fig9.png}
\caption{Test of six dynamic models simulating urban trajectories on the world system of cities (GHSL data source).\label{fig:fig9}}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To what extent such a worldwide estimation is relevant, compared to spatial locally stationary estimations on fixed (as done before) or on variable spatial extents, remains to be investigated. We however show how such models can be applied and compared worldwide, opening research avenues towards systematic benchmarks of dynamical models for systems of cities all over the world. Furthermore, we showed the complementarity of the different models, since none was dominating the others in all cases, confirming the diversity of co-evolution niches suggested before.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Towards multi-scale models}

While several applications and developments of the studies described previously will be necessary to strengthen our knowledge of evolutive urban systems, the development of new kinds of models will also be a crucial feature towards sustainable territorial governance. Following \cite{pumain2008socio}, urban systems imply diverse processes at different scales, with upward and downward feedbacks between these. The typical scales are first the intra-urban scale (microscopic scale), implying urban stakeholders, built environment, physical artifacts; second the metropolitan scale, or mesoscopic scale, for which the important processes are the location of population, economic activities and amenities, and for which abstractions done for example in Land-use Transport (LUTI) models are good approximations to grasp urban dynamics \cite{wegener2004land}; and third the scale of the system of cities, or the macroscopic scale, for which spatial interactions and the processes described in the model we used are appropriate abstractions. This view considerably extends and details the formulation of \cite{berry1964cities} of cities as systems of systems. In that context, \cite{rozenblat2018conclusion} point out that a multiscalar methodology is necessary for territorial policies, for example to be able to articulate global issues with local issues in the least contradictory way possible, with multiple and generally conflicting objectives. \cite{raimbault2019multi} illustrates this by empirically studying the economic and ecological performance of European mega-urban regions with a multi-level perspective through endogenous definitions of these. Furthermore, the recent transition from relatively modular regional urban systems to a globally interconnected urban system, and possibly associated new settlement patterns such as polycentric mega-urban regions \cite{le2017etalement}, implies a need for new models to possibly understand how this transition occurs, or simulate urban dynamics after the coalescence of regional urban systems.

An example of such multi-scalar models can be given between the mesoscopic and macroscopic scale. The macroscopic dynamics of a metropolitan area will strongly drive its internal development, for example because of population and economic flows coming from outside. This implies a downward feedback from the macroscopic urban dynamics to the mesoscopic dynamics. Reciprocally, the internal organization of a metropolitan area will influence its dynamics and insertion in the system at the macroscopic scale (whatever the dimension, for example economic performance), through positive and negative externalities such as congestion. This consists of upward feedback. A model strongly coupling these two scales and including explicitly these two feedbacks, coupling a reaction-diffusion model of urban growth at the mesoscopic scale \cite{raimbault2018calibration} with the interaction model of \cite{raimbault2020indirect} at the macroscopic scale, has been proposed by \cite{raimbault2019multiscalar}. Model exploration on synthetic systems unveils non-trivial non-linear effects from including the feedbacks, and for example intermediate optimal ranges for policy parameters influencing the level of sprawl (transit-oriented development) or the level of local aggregation (metropolization). The development of such models and their calibration on real dataset such as the GHSL dataset we used here, can become precious tools for evidence-based urban policies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

The essential point of evolutionary theory is to take into consideration the spatio-temporal dimension of the urban realm The aim is to link the development of cities to the many and diverse interrelations that make cities, since their emergence, entities that are not isolated, but on the contrary interdependent in their evolution, to the point of constituting ``systems of cities''. These systems are social adapters (in the sense that they carry and induce social change), complex, multi-scalar and open. The dynamics of these systems of cities, although they must always be placed in a context of time and space, include regularities that make it partly comparable and predictable, from one system to another and for certain scales of time. It is the micro-geographical level interactions, formed by the multiple interventions of a large number of stakeholders that produce the "behaviors" of cities and city systems at meso- and macro-geographical scales, because of the complex reflexive feedbacks introduced by the practices of so diverse stakeholders. It is important for these people and institutions to be informed of such knowledge about urban dynamics, to take advantage of this collective territorial intelligence and to make the most of the important adaptations required by the ecological and social tensions of our time. For complex dynamics such as the one observed in urban systems, policy interventions are always difficult as they can yield unexpected and even contrary effects. The compromise between an external top-down shock and endogenous bottom-up measures is also a subtle aspect to be determined. Understanding urban dynamics and co-evolution regimes in their diversity brings indirect elements of answer to these policy issues.

What modelling of urban dynamics brings in the discussion is at first a solid confirmation about the robustness of stylized facts that are integrated within the evolutive theory of urban systems. While also confirming the complementarity of processes and models it underlines the importance of the historical/political/geographical context which produces numerous effects of path-dependency.

Many questions remain open, as how to link urban scaling and dynamic models, how to define endogenously consistent urban systems, and how to develop data-driven multiscale models. However, a few messages can be conveyed towards citizens and practitioners: (i) there is statistical predictability of city growth and size on short time periods; (ii) largest metropolises are not “monstruopolises” but “normal” products of the urbanization process in their particular territorial conditions; iii) proactive adaptive strategies are necessary (through imitation, or anticipation and risk) for maintaining every city updating within a context that remain too often conceived as a rivalry or competition but that should more and more evolve towards emulation, according to a concept of co-opetition. We can have confidence in the future of cities because of the well observed robustness and sustainability of systems of cities, and the wide variations in their organization and evolution remind us about the fact that there are no norms nor any optimum in the territorial, social and cultural design of cities. Their diversity demonstrates the viability of different ways of being urban, and perhaps is a guarantee of the sustainability of the systems that they construct at the world scale.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\authorcontributions{Conceptualization, Juste Raimbault, Eric Denis and Denise Pumain; Data curation, Juste Raimbault, Eric Denis and Denise Pumain; Methodology, Juste Raimbault, Eric Denis and Denise Pumain; Software, Juste Raimbault, Eric Denis and Denise Pumain; Writing – original draft, Juste Raimbault, Eric Denis and Denise Pumain; Writing – review \& editing, Juste Raimbault, Eric Denis and Denise Pumain.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\funding{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgments}

Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure (http://www.egi.eu). We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.

%\bibliographystyle{unsrt}
%\bibliography{biblio.bib}

\begin{thebibliography}{10}

\bibitem{pumain2015multilevel}
D.~Pumain, E.~Swerts, C.~Cottineau, C.~Vacchiani-Marcuzzo, A.~Ignazzi,
  A.~Bretagnolle, F.~Delisle, R.~Cura, L.~Lizzi, and S.~Baffi.
\newblock Multi-level comparison of large urban systems.
\newblock {\em Cybergeo}, 706, 2015.

\bibitem{cura2017old}
Robin Cura, Cl{\'e}mentine Cottineau, Elfie Swerts, Cosmo Antonio~Ignazzi, Anne
  Bretagnolle, Celine Vacchiani-Marcuzzo, and Denise Pumain.
\newblock The old and the new: qualifying city systems in the world with
  classical models and new data.
\newblock {\em Geographical Analysis}, 49(4):363--386, 2017.

\bibitem{pumain2017urban}
Denise Pumain and Romain Reuillon.
\newblock {\em Urban dynamics and simulation models}.
\newblock Springer, 2017.

\bibitem{florczyk2019ghsl}
AJ~Florczyk, C~Corbane, D~Ehrlich, S~Freire, T~Kemper, L~Maffenini,
  M~Melchiorri, M~Pesaresi, P~Politis, M~Schiavina, et~al.
\newblock Ghsl data package 2019.
\newblock {\em European Commission Joint Research Center}, 29788, 2019.

\bibitem{pumain2018evolutionary}
Denise Pumain.
\newblock An evolutionary theory of urban systems.
\newblock In {\em International and transnational perspectives on urban
  systems}, pages 3--18. Springer, 2018.

\bibitem{raimbault2018caracterisation}
Juste Raimbault.
\newblock {\em Caract{\'e}risation et mod{\'e}lisation de la co-{\'e}volution
  des r{\'e}seaux de transport et des territoires}.
\newblock PhD thesis, Universit{\'e} Paris 7 Denis Diderot, 2018.

\bibitem{eeckhout2004gibrat}
Jan Eeckhout.
\newblock Gibrat's law for (all) cities.
\newblock {\em American Economic Review}, 94(5):1429--1451, 2004.

\bibitem{xu2014discontinuities}
Zengwang Xu and Robert Harriss.
\newblock Discontinuities in the evolution of the city system in texas from
  1850 to 2010.
\newblock {\em Computers, environment and urban systems}, 43:14--24, 2014.

\bibitem{cottineau2017metazipf}
Cl{\'e}mentine Cottineau.
\newblock Metazipf. a dynamic meta-analysis of city size distributions.
\newblock {\em PloS one}, 12(8), 2017.

\bibitem{gibrat1931ingalits}
Robert Gibrat.
\newblock Les ingalit{\'e}s {\'e}conomiques.
\newblock {\em Libraire du Recueil Sierey, Paris}, 235, 1931.

\bibitem{robson1973view}
BT~Robson.
\newblock A view on the urban sense.
\newblock {\em Studies in human geography}, pages 203--241, 1973.

\bibitem{pumain1982dynamique}
Denise Pumain.
\newblock {\em La dynamique des villes}.
\newblock Economica, 1982.

\bibitem{gabaix1999zipf}
Xavier Gabaix.
\newblock Zipf's law for cities: an explanation.
\newblock {\em The Quarterly journal of economics}, 114(3):739--767, 1999.

\bibitem{krugman1996confronting}
Paul Krugman.
\newblock Confronting the mystery of urban hierarchy.
\newblock {\em Journal of the Japanese and International economies},
  10(4):399--418, 1996.

\bibitem{decker2007global}
Ethan~H Decker, Andrew~J Kerkhoff, and Melanie~E Moses.
\newblock Global patterns of city size distributions and their fundamental
  drivers.
\newblock {\em PLoS One}, 2(9), 2007.

\bibitem{favaro2007croissance}
Jean-Marc Favaro.
\newblock {\em Croissance urbaine et cycles d'innovation dans les syst{\`e}mes
  de villes: une mod{\'e}lisation par les interactions spatiales}.
\newblock PhD thesis, Paris 1, 2007.

\bibitem{favaro2011gibrat}
Jean-Marc Favaro and Denise Pumain.
\newblock Gibrat revisited: An urban growth model incorporating spatial
  interaction and innovation cycles.
\newblock {\em Geographical Analysis}, 43(3):261--286, 2011.

\bibitem{wilson1971family}
Alan~Geoffrey Wilson.
\newblock A family of spatial interaction models, and associated developments.
\newblock {\em Environment and Planning A}, 3(1):1--32, 1971.

\bibitem{hagerstrand1952propagation}
Torsten Hagerstrand.
\newblock {\em The propagation of innovation waves}.
\newblock 1952.

\bibitem{pumain2006evolutionary}
Denise Pumain, Fabien Paulus, C{\'e}line Vacchiani-Marcuzzo, and Jos{\'e} Lobo.
\newblock An evolutionary theory for interpreting urban scaling laws.
\newblock {\em Cybergeo: European Journal of Geography}, 2006.

\bibitem{bretagnolle2007formes}
Anne Bretagnolle, Denise Pumain, and C{\'e}line Vacchiani-Marcuzzo.
\newblock Les formes des syst{\`e}mes de villes dans le monde, 2007.

\bibitem{paulus2004coevolution}
Fabien Paulus.
\newblock {\em Co{\'e}volution dans les syst{\`e}mes de villes: croissance et
  sp{\'e}cialisation des aires urbaines fran{\c{c}}aises de 1950 {\`a} 2000}.
\newblock PhD thesis, Universit{\'e} Panth{\'e}on-Sorbonne-Paris I, 2004.

\bibitem{finance2016villes}
Olivier Finance.
\newblock {\em Les villes fran{\c{c}}aises investies par les firmes
  transnationales {\'e}trang{\`e}res: des r{\'e}seaux d'entreprises aux
  {\'e}tablissements localis{\'e}s}.
\newblock PhD thesis, Paris 1, 2016.

\bibitem{cottineau2014evolution}
Cl{\'e}mentine Cottineau.
\newblock {\em L'{\'e}volution des villes dans l'espace post-sovi{\'e}tique.
  Observation et mod{\'e}lisations.}
\newblock PhD thesis, 2014.

\bibitem{cottineau2015growing}
Cl{\'e}mentine Cottineau, Paul Chapron, and Romain Reuillon.
\newblock Growing models from the bottom up. an evaluation-based incremental
  modelling method (ebimm) applied to the simulation of systems of cities.
\newblock {\em Journal of Artificial Societies and Social Simulation}, 18(4):9,
  2015.

\bibitem{schmitt2014modelisation}
Clara Schmitt.
\newblock {\em Mod{\'e}lisation de la dynamique des syst{\`e}mes de peuplement:
  de SimpopLocal {\`a} SimpopNet}.
\newblock PhD thesis, Universit{\'e} Panth{\'e}on-Sorbonne-Paris I, 2014.

\bibitem{raimbault2020unveiling}
Juste Raimbault.
\newblock Unveiling co-evolutionary patterns in systems of cities: a systematic
  exploration of the simpopnet model.
\newblock In {\em Theories and Models of Urbanization}, pages 261--278.
  Springer, 2020.

\bibitem{raimbault2020indirect}
Juste Raimbault.
\newblock Indirect evidence of network effects in a system of cities.
\newblock {\em Environment and Planning B: Urban Analytics and City Science},
  47(1):138--155, 2020.

\bibitem{raimbault2018modeling}
Juste Raimbault.
\newblock Modeling the co-evolution of cities and networks.
\newblock {\em arXiv preprint arXiv:1804.09430}, 2018.

\bibitem{nelson2009evolutionary}
Richard~R Nelson.
\newblock {\em An evolutionary theory of economic change}.
\newblock harvard university press, 2009.

\bibitem{durham1991coevolution}
William~H Durham.
\newblock {\em Coevolution: Genes, culture, and human diversity}.
\newblock Stanford University Press, 1991.

\bibitem{langton1997artificial}
Christopher~G Langton.
\newblock {\em Artificial life: An overview}.
\newblock Mit Press, 1997.

\bibitem{bedau2003artificial}
Mark~A Bedau.
\newblock Artificial life: organization, adaptation and complexity from the
  bottom up.
\newblock {\em Trends in cognitive sciences}, 7(11):505--512, 2003.

\bibitem{mesoudi2017pursuing}
Alex Mesoudi.
\newblock Pursuing darwin’s curious parallel: Prospects for a science of
  cultural evolution.
\newblock {\em Proceedings of the National Academy of Sciences},
  114(30):7853--7860, 2017.

\bibitem{volberda2003co}
Henk~W Volberda and Arie~Y Lewin.
\newblock Co-evolutionary dynamics within and between firms: From evolution to
  co-evolution.
\newblock {\em Journal of management studies}, 40(8):2111--2136, 2003.

\bibitem{bull2000meme}
Larry Bull, Owen Holland, and Susan Blackmore.
\newblock On meme--gene coevolution.
\newblock {\em Artificial life}, 6(3):227--235, 2000.

\bibitem{janzen1980coevolution}
Daniel~H Janzen.
\newblock When is it coevolution?
\newblock 1980.

\bibitem{strauss2005toward}
Sharon~Y Strauss, Heather Sahli, and Jeffrey~K Conner.
\newblock Toward a more trait-centered approach to diffuse (co) evolution.
\newblock {\em New Phytologist}, 165(1):81--90, 2005.

\bibitem{schamp2010notion}
Eike~W Schamp.
\newblock On the notion of co-evolution in economic geography.
\newblock Technical report, Edward Elgar Publishing, 2010.

\bibitem{ter2011co}
Anne~LJ Ter~Wal and Ron Boschma.
\newblock Co-evolution of firms, industries and networks in space.
\newblock {\em Regional studies}, 45(7):919--933, 2011.

\bibitem{colletis2010co}
Gabriel Colletis.
\newblock Co-{\'e}volution des territoires et de la technologie: une
  perspective institutionnaliste.
\newblock {\em Revue dEconomie Regionale Urbaine}, (2):235--249, 2010.

\bibitem{kallis2007coevolution}
Giorgos Kallis.
\newblock When is it coevolution?
\newblock {\em Ecological Economics}, 62(1):1--6, 2007.

\bibitem{pumain1997pour}
Denise Pumain.
\newblock Pour une th{\'e}orie {\'e}volutive des villes.
\newblock {\em L'Espace g{\'e}ographique}, pages 119--134, 1997.

\bibitem{raimbault2018co}
Juste Raimbault.
\newblock Co-evolution and morphogenetic systems.
\newblock {\em arXiv preprint arXiv:1803.11457}, 2018.

\bibitem{holland2012signals}
John~H Holland.
\newblock {\em Signals and boundaries: Building blocks for complex adaptive
  systems}.
\newblock Mit Press, 2012.

\bibitem{raimbault2017identification}
Juste Raimbault.
\newblock Identification de causalit{\'e}s dans des donn{\'e}es
  spatio-temporelles.
\newblock 2017.

\bibitem{denis2020population}
E~Denis.
\newblock Population, land, wealth and the global urban sprawl. drivers of
  urban built-up expansion across the world from 1990 to 2015.
\newblock In {\em Theories and Models of Urbanization. Springer Lecture Notes
  in Morphogenesis}, pages 235--258. 2020.

\bibitem{melchiorri2019principles}
Michele Melchiorri, Martino Pesaresi, Aneta~J Florczyk, Christina Corbane, and
  Thomas Kemper.
\newblock Principles and applications of the global human settlement layer as
  baseline for the land use efficiency indicator—sdg 11.3. 1.
\newblock {\em ISPRS International Journal of Geo-Information}, 8(2):96, 2019.

\bibitem{dijkstra2014harmonised}
Lewis Dijkstra and Hugo Poelman.
\newblock A harmonised definition of cities and rural areas: the new degree of
  urbanisation.
\newblock {\em WP}, 1:2014, 2014.

\bibitem{bretagnolle1999systemes}
Anne Bretagnolle.
\newblock {\em Les syst{\`e}mes de villes dans l'espace-temps: effets de
  l'accroissement des vitesses de d{\'e}placement sur la taille et l'espacement
  des villes}.
\newblock PhD thesis, Paris 1, 1999.

\bibitem{swerts2013systemes}
Elfie Swerts.
\newblock {\em Les Syst{\`e}mes de villes en Inde et en Chine}.
\newblock PhD thesis, 2013.

\bibitem{rozenblat2018international}
C{\'e}line Rozenblat, Denise Pumain, and Elkin Velasquez.
\newblock {\em International and Transnational Perspectives on Urban Systems}.
\newblock Springer, 2018.

\bibitem{dijkstra2018applying}
Lewis Dijkstra, Aneta Florczyk, Sergio Freire, Thomas Kemper, Martino Pesaresi,
  and M~Schiavina.
\newblock Applying the degree of urbanisation to the globe: A new harmonised
  definition reveals a different picture of global urbanisation.
\newblock In {\em Proceedings of the 16th IAOS Conference: Better Statistics
  for Better Lives, Paris, France}, pages 19--20, 2018.

\bibitem{desa2018world}
UN~DESA.
\newblock World urbanization prospects: 2018.
\newblock {\em United Nations Department for Economic and Social Affairs},
  2018.

\bibitem{bretagnolle2010simulating}
Anne Bretagnolle and Denise Pumain.
\newblock Simulating urban networks through multiscalar space-time dynamics:
  Europe and the united states, 17th-20th centuries.
\newblock {\em Urban Studies}, 47(13):2819--2839, 2010.

\bibitem{arcaute2020scaling}
Elsa Arcaute and Erez Hatna.
\newblock Scaling laws: insights and limitations.
\newblock In {\em Theories and models of urbanization}, pages 45--66. Springer,
  2020.

\bibitem{corral2020truncated}
{\'A}lvaro Corral, Frederic Udina, and Elsa Arcaute.
\newblock Truncated lognormal distributions and scaling in the size of
  naturally defined population clusters.
\newblock {\em Physical Review E}, 101(4):042312, 2020.

\bibitem{arcaute2015constructing}
Elsa Arcaute, Erez Hatna, Peter Ferguson, Hyejin Youn, Anders Johansson, and
  Michael Batty.
\newblock Constructing cities, deconstructing scaling laws.
\newblock {\em Journal of The Royal Society Interface}, 12(102):20140745, 2015.

\bibitem{cottineau2017diverse}
Cl{\'e}mentine Cottineau, Erez Hatna, Elsa Arcaute, and Michael Batty.
\newblock Diverse cities or the systematic paradox of urban scaling laws.
\newblock {\em Computers, environment and urban systems}, 63:80--94, 2017.

\bibitem{bertaud2003spatial}
Alain Bertaud and Stephen Malpezzi.
\newblock The spatial distribution of population in 48 world cities:
  Implications for economies in transition.
\newblock {\em Center for urban land economics research, University of
  Wisconsin}, 32(1):54--55, 2003.

\bibitem{broido2019scale}
Anna~D Broido and Aaron Clauset.
\newblock Scale-free networks are rare.
\newblock {\em Nature communications}, 10(1):1--10, 2019.

\bibitem{ribeiro2019effects}
Haroldo~V Ribeiro, Diego Rybski, and J{\"u}rgen~P Kropp.
\newblock Effects of changing population or density on urban carbon dioxide
  emissions.
\newblock {\em Nature communications}, 10(1):1--9, 2019.

\bibitem{reuillon2013openmole}
Romain Reuillon, Mathieu Leclaire, and Sebastien Rey-Coyrehourcq.
\newblock Openmole, a workflow engine specifically tailored for the distributed
  exploration of simulation models.
\newblock {\em Future Generation Computer Systems}, 29(8):1981--1990, 2013.

\bibitem{aveline2018pathways}
Natacha Aveline-Dubach.
\newblock Pathways of sustainable development across china: the cases of
  hangzhou, datong and zhuhai, 2018.

\bibitem{denis2017subaltern}
Eric Denis and Marie-H{\'e}l{\`e}ne Z{\'e}rah.
\newblock {\em Subaltern urbanisation in India: An introduction to the dynamics
  of ordinary towns}.
\newblock Springer, 2017.

\bibitem{schmitt2015half}
Clara Schmitt, S{\'e}bastien Rey-Coyrehourcq, Romain Reuillon, and Denise
  Pumain.
\newblock Half a billion simulations: Evolutionary algorithms and distributed
  computing for calibrating the simpoplocal geographical model.
\newblock {\em Environment and Planning B: Planning and Design},
  42(2):300--315, 2015.

\bibitem{wu2020emerging}
Fulong Wu.
\newblock Emerging cities and urban theories: a chinese perspective.
\newblock In {\em Theories and models of urbanization}, pages 171--182.
  Springer, 2020.

\bibitem{swerts2017data}
Elfie Swerts.
\newblock A data base on chinese urbanization: Chinacities.
\newblock {\em Cybergeo: European Journal of Geography}, 2017.

\bibitem{pumain2008socio}
Denise Pumain.
\newblock The socio-spatial dynamics of systems of cities and innovation
  processes: a multi-level model.
\newblock In {\em The Dynamics of Complex Urban Systems}, pages 373--389.
  Springer, 2008.

\bibitem{wegener2004land}
Michael Wegener and Franz F{\"u}rst.
\newblock Land-use transport interaction: state of the art.
\newblock {\em Available at SSRN 1434678}, 2004.

\bibitem{berry1964cities}
Brian~JL Berry.
\newblock Cities as systems within systems of cities.
\newblock {\em Papers in regional science}, 13(1):147--163, 1964.

\bibitem{rozenblat2018conclusion}
Celine Rozenblat and Denise Pumain.
\newblock Conclusion: Toward a methodology for multi-scalar urban system
  policies.
\newblock {\em International and Transnational Perspectives on Urban Systems},
  page 385, 2018.

\bibitem{raimbault2019multi}
Juste Raimbault.
\newblock Multi-dimensional urban network percolation.
\newblock {\em JIMIS}, 5:5, 2019.

\bibitem{le2017etalement}
F~Le~N{\'e}chet.
\newblock De l’{\'e}talement urbain aux r{\'e}gions m{\'e}tropolitaines
  polycentriques: formes de fonctionnement et formes de gouvernance.
\newblock {\em Peupler la terre-De la pr{\'e}histoire {\`a} l’{\`e}re des
  m{\'e}tropoles. Presses Universitaires Francois Rabelais, Tours}, 2017.

\bibitem{raimbault2018calibration}
Juste Raimbault.
\newblock Calibration of a density-based model of urban morphogenesis.
\newblock {\em PloS one}, 13(9), 2018.

\bibitem{raimbault2019multiscalar}
Juste Raimbault.
\newblock A multi-scalar model for systems of cities.
\newblock In {\em Conference on Complex Systems, Singapore}. 2019.

\end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2020-08-24 Docs/Papers/MesoBenchmark/paper/paper.tex
skeleton meso benchmark paper

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{A comparison of simple models for urban morphogenesis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Juste Raimbault\textsuperscript{1,2,3,*}
\\
\bigskip
\textbf{1} Center for Advanced Spatial Analysis, University College London, London, United Kingdom
\\
\textbf{2} UPS CNRS 3611 ISC-PIF, CNRS, Paris, France
\\
\textbf{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Universit{\'e} Paris Diderot, Paris, France
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* juste.raimbault@polytechnique.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The spatial distribution of population and activities within urban areas, or urban form at the mesoscopic scale, is the outcome of multiple antagonist processes. We propose in this paper to benchmark different models of urban morphogenesis, to systematically compare the urban forms they can produce. Different types of approaches are included, such as a reaction-diffusion model, a gravity-based model, and correlated percolation. Applying a diversity search algorithm, we estimate the feasible space of each model within a space of urban form indicators, in comparison of empirical values for worldwide urban areas. We find a complementarity of the different types of processes, advocating for a plurality of urban models.

\linenumbers

\section*{Introduction}

% urban form and sustainability - urban form at the mesoscopic scale

% broad overview of approaches: CA, LUTI. why study simple models (include studied models)

\cite{10.1371/journal.pone.0203516}

\cite{li2019singularity}

% urban multi-modeling/multidimensionality: importance of plurality of views

% research question / contributions

% organization

\section*{Materials and methods}

%\subsection*{Processes}
% kind of systematic review of possibly involved processes?

\subsection*{Urban morphogenesis models}

\subsubsection*{Gravity-based model}

\cite{li2019singularity}

% modified: exponent for spatial interaction model; multiple seeds (polycentricity)

\subsubsection*{Reaction-diffusion}

\cite{10.1371/journal.pone.0203516}
% modify: max density?

\subsubsection*{Correlated percolation}

\cite{makse1998modeling}

The method to generate a spatial field exhibiting long range correlations was introduced for problems in physics by \cite{makse1996method}.

\subsubsection*{Kernel mixtures}

Finally, to provide some kind of null model to understand the advantages of each approach compared to a simple description of population distribution, we also include urban forms generated as kernel mixtures.
\cite{anas1998urban}

% null model: white noise? (for robustness of indicators)

\subsection*{Measures of urban form}

\subsection*{Empirical data}

\section*{Results}

\subsection*{Behavior of models}

\subsection*{Feasible morphological spaces}

\section*{Discussion}

\section*{Conclusion}

%\section*{Supporting information}

\section*{Acknowledgments}

\nolinenumbers

\bibliography{biblio}

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
%

\end{document}

%\begin{eqnarray}
%\label{eq:schemeP}
%	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
%\end{eqnarray}

% Place figure captions after the first paragraph in which they are cited.
%\begin{figure}[!h]
%\caption{{\bf Bold the figure title.}
%Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
%\label{fig1}
%\end{figure}

% Place tables after the first paragraph in which they are cited.
%\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
%\centering
%\caption{
%{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
%\begin{tabular}{|l+l|l|l|l|l|l|l|}
%\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
%$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
%$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
%$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
%\end{tabular}
%\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
%\end{flushleft}
%\label{table1}
%\end{adjustwidth}
%\end{table}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
%\paragraph*{S1 Fig.}
%\label{S1_Fig}
%{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).

%\paragraph*{S2 Fig.}
%\label{S2_Fig}
%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

%\paragraph*{S1 File.}
%\label{S1_File}
%{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

%\paragraph*{S1 Video.}
%\label{S1_Video}
%{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

%\paragraph*{S1 Appendix.}
%\label{S1_Appendix}
%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

%\paragraph*{S1 Table.}
%\label{S1_Table}
%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

### 2020-08-27 Docs/Papers/MesoBenchmark/paper/paper.tex
pse_reactiondiff

@@
 
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+%\documentclass[10pt,letterpaper]{article}
+%\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\documentclass[10pt]{article}
+%\usepackage[margin=2cm]{geometry}
 
@@
 % Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
+%\raggedright
+%\setlength{\parindent}{0.5cm}
+%\textwidth 5.25in 
+%\textheight 8.75in
 
@@
 %\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
+%\pagestyle{fancy}
+%\fancyhf{}
+%%\setlength{\headheight}{27.023pt}
+%%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+%\rfoot{\thepage/\pageref{LastPage}}
+%\renewcommand{\headrulewidth}{0pt}
+%\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+%\fancyheadoffset[L]{2.25in}
+%\fancyfootoffset[L]{2.25in}
+%\lfoot{\today}
 
@@
 
-\begin{document}
-\vspace*{0.2in}
+%\begin{document}
+%\vspace*{0.2in}
 
 % Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{A comparison of simple models for urban morphogenesis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+%\begin{flushleft}
+%\begin{center}
+%{\Large
+%\textbf\newline{A comparison of simple models for urban morphogenesis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+%}
+%\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+%\\
+\title{A comparison of simple models for urban morphogenesis}
+\date{}
+
+\author{
+Juste Raimbault\textsuperscript{1,2,3,*}
+\bigskip\\
+\textsuperscript{1} Center for Advanced Spatial Analysis, University College London
+\\
+\textsuperscript{2} UPS CNRS 3611 ISC-PIF
+\\
+\textsuperscript{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
+\bigskip\\
+* juste.raimbault@polytechnique.edu
 }
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Juste Raimbault\textsuperscript{1,2,3,*}
-\\
-\bigskip
-\textbf{1} Center for Advanced Spatial Analysis, University College London, London, United Kingdom
-\\
-\textbf{2} UPS CNRS 3611 ISC-PIF, CNRS, Paris, France
-\\
-\textbf{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Universit{\'e} Paris Diderot, Paris, France
-\\
-\bigskip
 
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
+\begin{document}
 
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-%\ddag These authors also contributed equally to this work.
+\maketitle
 
-% Current address notes
-%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-%\dag Deceased
-
-% Group/Consortium Author Note
-%\textpilcrow Membership list can be found in the Acknowledgments section.
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* juste.raimbault@polytechnique.edu
-
-\end{flushleft}
+%\end{flushleft}
+%\end{center}
 % Please keep the abstract below 300 words
-\section*{Abstract}
+%\section*{Abstract}
+\vspace{1cm}
+\begin{abstract}
 The spatial distribution of population and activities within urban areas, or urban form at the mesoscopic scale, is the outcome of multiple antagonist processes.
@@
 We find a complementarity of the different types of processes, advocating for a plurality of urban models.
+\end{abstract}
 
-\linenumbers
+%\linenumbers
 
@@
 
+Understanding the dynamics of cities is an increasing issue for sustainability, since the proportion of the world population expected to live in cities will grow to a large majority in the next decades, and that cities combine both positive and negative externalities on most aspects.
+Their complexity implies that quantitative and qualitative predictions are not relevant, but planners can \emph{invent future cities} \cite{batty2018inventing}, what requires though a knowledge of key urban processes which can be acted upon.
+In that context, the growth of \emph{urban form} in its different definition and scales, is essential \cite{williams2000achieving}.
+Considering urban form at a mesoscopic scale, i.e.
+roughly the scale of urban areas, it can be understood as the spatial distribution of activities.
+More particularly the distribution of population density has a strong impact on commuting, energy consumption and emissions \cite{le2012urban}.
+Being able to link microscopic processes with the growth of different types of urban form is thus important for a long term planning of sustainable urban systems.
+
 % broad overview of approaches: CA, LUTI. why study simple models (include studied models)
+
+Urban modeling at the mesoscopic scale is the subject of diverse approaches and disciplines.
+Intra-urban urban economic models, building on works such as the Alonso-Mills-Muth model or the Fujita-Ogawa model
+Transportation and Urban Planning also have a long history in urban dynamics models, including Land-use transport interaction models \cite{wegener2004land}
+Spatial interaction models can also be used in a similar manner \cite{milton2019accelerating}
+At the interface of physics, artificial life and quantitative geography, a few approaches propose simple models to explain the growth of urban form. Although such models
 
@@
 
-% urban multi-modeling/multidimensionality: importance of plurality of views
+% urban multi-modeling/multidimensionality: importance of plurality of views: why benchmark
 
@@
 Finally, to provide some kind of null model to understand the advantages of each approach compared to a simple description of population distribution, we also include urban forms generated as kernel mixtures.
-\cite{anas1998urban}
+We consider in particular exponential mixtures \cite{anas1998urban}, written as
 
@@
 
-\subsection*{Empirical data}
+% already computed for dynamical calib for world urban areas -> use same measures. todo future work: benchmark measures world areas: landscape eco, etc.
+
+Quantitative measures of urban form
 
@@
 
-\subsection*{Behavior of models}
+\subsection*{Implementation}
+
+The models are implemented in \texttt{scala} and integrated into the \texttt{spatialdata} library for spatial sensitivity analysis \cite{raimbault2020scala}.
+The library is bundled as an OpenMOLE plugin for the numerical experiments.
+OpenMOLE is an open source software for model exploration and validation \cite{reuillon2013openmole} combining model embedding with state-of-the-art exploration methods (including for example sensitivity analysis, design of experiments, calibration with genetic algorithms) and a transparent distribution of computations on high performance computing environments.
+In our case, we use its workflow system and an integrated algorithm to determine the feasible space of models.
+
+\subsection*{Empirical data}
+
+%%%%%%%%%%%%%%
+\begin{figure}[!h]
+	\includegraphics[width=\linewidth]{figures/Fig1.png}
+	\caption{{\bf .}\label{fig:fig1}}
+\end{figure}
+%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%
+\begin{figure}[!h]
+	\includegraphics[width=\linewidth]{figures/Fig2.png}
+	\caption{{\bf .}\label{fig:fig1}}
+\end{figure}
+%%%%%%%%%%%%%%
+
+%\subsection*{Behavior of models}
+% also saltelli?
 
@@
 \section*{Discussion}
+
+% results, implications
+
+% future dev
+
+% dynamical calib
+\cite{raimbault:halshs-02406539}
 

### 2020-08-27 Docs/Papers/MesoBenchmark/paper/figures.sh

@@ -1,0 +1,12 @@
+
+JPGQUALITY=50
+PDFRESOLUTION=200
+WIDTH=2000
+HORIZONTALPADDING=10
+VERTICALPADDING=10
+
+# Fig 1
+montage figuresraw/mapindic_alpha_2015.png figuresraw/mapindic_moran_2015.png -tile 1x2 -geometry +0+"$VERTICALPADDING" -resize "$WIDTH"x figures/Fig1.png
+
+# Fig 2
+montage figuresraw/mapindic_cluster.png figuresraw/distributions_cluster-year.png -tile 1x2 -geometry +0+"$VERTICALPADDING" -resize "$WIDTH"x figures/Fig2.png

### 2020-08-27 Models/MesoBenchmark/realData.R

@@ -1,0 +1,32 @@
+setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/MesoBenchmark'))
+
+library(ggplot2)
+library(rgdal)
+library(dplyr)
+
+source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+source('../MesoCalibration/functions.R')
+source('../MesoCalibration/morphology.R')
+
+ucdbsf <- loadUCDBData(paste0(Sys.getenv('CS_HOME'),'/Data/JRC_EC/GHS/GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0'),'GHS_STAT_UCDB2015MT_GLOBE_R2019A_V1_0')
+
+years = c('1975','1990','2000','2015')
+indics=c('totalPop','maxPop','minPop','moran','avgDist','entropy','alpha','alphaRSquared')
+indicnames = list('totalPop'='Total population','maxPop'='Maximal population','minPop'='Minimal population',
+                  'moran'='Spatial autocorrelation','avgDist'='Average distance','entropy'='Entropy',
+                  'alpha'='Hierarchy','alphaRSquared'='Fierarchy fit')
+
+# load morphologies: see ../MesoCalibration/dataPreparation.R and ../MesoCalibration/analysis.R
+morphos <- read.csv(file='../MesoCalibration/configs/morphologies.csv',header = T,sep = ";")
+inds = c(1:946,948:1000)
+areasmorph = ucdbsf[inds,]
+areasmorph$areaid = inds
+areasmorph = left_join(areasmorph,morphos,by=c('areaid'='areaid'))
+
+
+
+
+
+
+
+

### 2020-08-27 Models/MesoCalibration/analysis.R

@@ -59,7 +59,7 @@ areasmorph = left_join(areasmorph,morphos,by=c('areaid'='areaid'))
 # better with outliers
 morphos = morphos[morphos$alpha1975>-4&morphos$alpha1990>-4&morphos$alpha2000>-4&morphos$alpha2015>-4,]
 
-timedf <- as_temporal_df(morphos,indics)
+timedf <- as_temporal_df(morphos,indics,years)
 
 classifindics = c('moran','avgDist','entropy','alpha')
 classdata = data.frame(id=morphos$areaid)
@@ -138,7 +138,7 @@ for(indic in indics){for(year in years){
 
 
 # clusters map
-areasmorphfilt = areasmorph[!apply(areasmorph[,names(morphosall)],1,function(r){length(which(is.na(r)))>0})&areasmorph$alpha1975>-4&areasmorph$alpha1990>-4&areasmorph$alpha2000>-4&areasmorph$alpha2015>-4,]
+areasmorphfilt = areasmorph[!apply(areasmorph[,names(morphos)],1,function(r){length(which(is.na(r)))>0})&areasmorph$alpha1975>-4&areasmorph$alpha1990>-4&areasmorph$alpha2000>-4&areasmorph$alpha2015>-4,]
 for(classindic in names(classdata)){
   currentd = classdata[[classindic]];names(currentd)<-classdata$id
   areasmorphfilt[[paste0("class",classindic)]] = currentd[areasmorphfilt$areaid]
@@ -146,7 +146,7 @@ for(classindic in names(classdata)){
 areasmorphfilt$classcluster=as.character(areasmorphfilt$classcluster)
 
 map(
-  areasmorphfilt,
+  areasmorphfilt[!is.na(areasmorphfilt$classcluster),],
   'classcluster',
   "totalPop2015",
   paste0(resdir,'mapindic_cluster.png'),

### 2020-08-27 Models/MesoBenchmark/pseResults.R
pse

@@ -1,0 +1,4 @@
+
+
+
+# for hypervolume intersection: https://rdrr.io/cran/hypervolume/man/hypervolume_set.html

### 2020-08-28 Docs/Papers/MesoBenchmark/paper/paper.tex
.

@@
 Urban modeling at the mesoscopic scale is the subject of diverse approaches and disciplines.
-Intra-urban urban economic models, building on works such as the Alonso-Mills-Muth model or the Fujita-Ogawa model
-Transportation and Urban Planning also have a long history in urban dynamics models, including Land-use transport interaction models \cite{wegener2004land}
-Spatial interaction models can also be used in a similar manner \cite{milton2019accelerating}
-At the interface of physics, artificial life and quantitative geography, a few approaches propose simple models to explain the growth of urban form. Although such models
+Intra-urban urban economic models, building on classic works such as the Alonso-Mills-Muth model or the Fujita-Ogawa model, propose models linking land-use with land and building markets, which are spatially explicit to different degrees \cite{viguie2012trade}.
+Transportation and Urban Planning also have a long history in urban dynamics models, including Land-use transport interaction models \cite{wegener2004land}.
+Spatial interaction models can also be used in a similar manner to study urban dynamics and as a by-product urban form \cite{milton2019accelerating}.
+Cellular automata models of urban growth are also a privileged approach to study the growth of urban form from a data-driven perspective \cite{batty1997cellular}.
 
-\cite{10.1371/journal.pone.0203516}
-
-\cite{li2019singularity}
+At the interface of physics, artificial life and quantitative geography, a few approaches propose simple models to explain the growth of urban form, and generally rely on an unidimensional description of urban form, namely the distribution of population or of the built environment.
+In that context, the correlated percolation model introduced by \cite{makse1998modeling} was a precursor.
+Such models can rely on abstract physical processes but also on agent behavior, such as in the Sugarscape model which according to \cite{batty2007cities} can be considered as a model for human settlements.
+\cite{murcio2015urban} use migration between cities at multiple scales to simulate urban growth.
+Diffusion-limited aggregation (DLA) is an other approach transferred from physics to urban modeling \cite{batty1989urban} and has shown relevant to reproduce fractal urban structures and urban migration processes \cite{murcio2009colored}.
+\cite{murcio2013second} combines DLA with percolation to obtain more realistic urban forms.
+Closer to the idea of urban morphogenesis, \cite{10.1371/journal.pone.0203516} proposes a reaction-diffusion model to capture fundamental urban growth processes.
+\cite{li2019singularity} describes an urban growth model based on geographical processes, namely an aggregation of population driven by spatial interaction.
+All these works have in common to model urban growth in synthetic settings, at a mesoscopic scale, considering population distribution only, and in a stylized way.
+They furthermore consider diverse processes, remaining simple in their structure although they lead to the emergence of a complex behavior.
+We will in this paper focus on such models, referring to them as \emph{models of urban morphogenesis}.
 
@@
 
+Exhibiting models with a few number of parameters and processes is useful from an explanative viewpoint, when these can reproduce real world configurations.
+Having multiple concurrent models which include diverse, complementary or contradictory processes, is furthermore useful for the construction of integrated urban theories, since concurrent explanations can be benchmarked, compared and possibly integrated into multi-modeling approaches.
+This plurality in urban modeling is intrinsic to a literature with multiple disciplines focusing on a same object of study \cite{pumain2020conclusion}.
+
 % research question / contributions
 
+We propose thus in this paper to benchmark several simple models of urban morphogenesis, in order to understand the potentialities of some of these models to exhibit a complex behavior and reproduce existing urban forms, and compare them in a systematic way.
+More precisely, our contributions are as follows: (i) we integrate four different models (correlated percolation, reaction-diffusion, gravity and exponential mixture) into a single software framework; (ii) we compute measures of urban form for urban areas worldwide; (iii) we apply a novelty search algorithm to the models in order to determine their feasible morphological space, and compare these to real urban form values.
+This contributes to a general understanding of the complementarity of urban models, more particularly for urban morphogenesis at this scale.
+
 % organization
+
+The rest of this paper is organized as follows:
 

### 2020-08-29 Models/MesoBenchmark/pseResults.R
pse_corrperco

@@ -1,4 +1,25 @@
 
+setwd(paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Models/MesoBenchmark'))
+
+library(ggplot2)
+library(dplyr)
+library(GGally)
+
+
+files = list(
+  'expmixture' = '20200828_1515_PSE_EXPMIXTURE_res/population20000.csv',
+  'reactiondiffusion' = '20200828_1624_PSE_REACTIONDIFFUSION_res/population50000.csv'
+)
+
+indics = c('moran','distance','entropy','slope')
+
+res = data.frame()
+for(model in names(files)){
+  currentres = as.tbl(read.csv(paste0('openmole/pse/',files[[model]])))
+  res = rbind(res,cbind(currentres[,c(indics,'evolution.samples')],model=rep(model,nrow(currentres))))
+}
+
+ggpairs(res[res$slope>-4,indics],aes(color=model))
 
 
 # for hypervolume intersection: https://rdrr.io/cran/hypervolume/man/hypervolume_set.html

### 2020-08-29 Docs/Papers/MesoBenchmark/paper/paper.tex
pse_corrperco
pse_gravity

@@
 We propose thus in this paper to benchmark several simple models of urban morphogenesis, in order to understand the potentialities of some of these models to exhibit a complex behavior and reproduce existing urban forms, and compare them in a systematic way.
-More precisely, our contributions are as follows: (i) we integrate four different models (correlated percolation, reaction-diffusion, gravity and exponential mixture) into a single software framework; (ii) we compute measures of urban form for urban areas worldwide; (iii) we apply a novelty search algorithm to the models in order to determine their feasible morphological space, and compare these to real urban form values.
+More precisely, our contributions are the following: (i) we integrate four different models (correlated percolation, reaction-diffusion, gravity and exponential mixture) into a single software framework; (ii) we compute measures of urban form for urban areas worldwide; (iii) we apply a novelty search algorithm to the models in order to determine their feasible morphological space, and compare these to real urban form values.
 This contributes to a general understanding of the complementarity of urban models, more particularly for urban morphogenesis at this scale.
@@
 
-The rest of this paper is organized as follows:
+The rest of this paper is organized as follows: we first describe the models benchmarked and the quantitative measures used for urban form; we describe empirical values of urban morphology indicators for urban areas worldwide; these are then compared to the feasible space of each model obtained with a diversity search algorithm.
+We finally discuss the implications of these results for theories of urban morphogenesis and possible developments.
 
@@
 
+Following the so-called ``first law of geography'', entities in space have interaction patterns which can be described with spatial interaction models, including the gravity model.
 \cite{li2019singularity}
@@
 In our case, we use its workflow system and an integrated algorithm to determine the feasible space of models.
+
+% TODO examples for each model
 

### 2020-08-30 Docs/Papers/MesoBenchmark/paper/paper.tex
.

@@
 
+We study and compare four different models of urban morphogenesis.
+We consider a population grid of size $N = W\times H$ (not necessarily square), each cell being characterized by its population $P_i$ with $1\leq i \leq N$.
+
 \subsubsection*{Gravity-based model}
-
-Following the so-called ``first law of geography'', entities in space have interaction patterns which can be described with spatial interaction models, including the gravity model.
-\cite{li2019singularity}
 
@@
 
+Following the so-called ``first law of geography'', entities in space have interaction patterns which can be described with spatial interaction models \cite{fotheringham1989spatial}, including the gravity model.
+\cite{li2019singularity} proposed an urban growth model including this process within an iterative growth with population aggregation processes, extending a more simple model introduced by \cite{rybski2013distance}.
+We generalize this model by adding (i) a hierarchy parameter regarding population aggregation and (ii) seeding multiple initial sites to allow the emergence of polycentric urban forms.
+
+Formally, an initial grid is seeded with $P_0^{(G)}$ sites with population 1, randomly selected.
+Then, iteratively, one unit of population is added to each cell at each time step with a probability proportional to
+\begin{equation}
+p_i \propto \frac{\sum_{j\neq i} P_j^{\gamma_P^{(G)}} \cdot d_{ij}^{- \gamma_D^{(G)}}}{\sum d_{ij}^{- \gamma_D^{(G)}}}
+\end{equation}
+
+where probabilities are rescaled such that the cell with the larger value has a probability of $g^{(G)}$.
+This last parameter allows modifying the speed of growth.
+The model is stopped when a total population $P_M^{(G)}$ is reached.
+
 \subsubsection*{Reaction-diffusion}
 
-\cite{10.1371/journal.pone.0203516}
-% modify: max density?
+% modify: max density? - NO
+
+In his attempt to understand embryogenesis, Alan Turing proposed to use chemical partial differential equations (PDEs) to model morphogenesis, introducing the nowadays famous reaction-diffusion equations \cite{turing1990chemical}.
+In such systems, chemical substances react together and diffuse in space, leading to the emergence of complex geometrical patterns.
+The concept of morphogenesis has been since well used in urban studies \cite{raimbault2018co}, but very few models have actually implemented reaction-diffusion equations, \cite{bonin2014modelisation} being a notable exception.
+\cite{10.1371/journal.pone.0203516} proposes to capture the fundamental processes of agglomeration economies (positive externalities) leading to aggregation and of congestion (negative externalities) leading to sprawl, as an ``aggregation-diffusion'' model of urban morphogenesis.
+The model yields indeed in certain limits reaction-diffusion PDEs.
+Formally, starting from an empty grid, $N^{(R)}$ units of population are added at each time step, and attributed independently to cells with a probability proportional to $P_i^{\alpha^{(R)}}$ (probabilities are rescaled to obtain a probability distribution over all cells).
+Population is then diffused in space $d^{(R)}$ times with a strength $\beta^{(R)}$.
+The model is stopped when a maximum population $P_M^{(R)}$ is reached.
 
@@
 
-\cite{makse1998modeling}
+The first two models presented are iterative and can in theory be used dynamically.
+Other approaches, closer to procedural modeling \cite{parish2001procedural}, do not simulate the progressive growth of population.
+They can however capture processes at play in the growth of urban form.
+The correlated percolation model described by \cite{makse1998modeling} integrates for example clustering processes in cities.
+A method to generate a spatial field exhibiting long range correlations was introduced for problems in physics by \cite{makse1996method}.
+It is combined to a monocentric density profile in \cite{makse1998modeling} to produce urban forms.
+In practice, a correlated field $p_i$ is generated by (i) generating a random spatial field; (ii) compute its spatial Fourier transform; (iii) introduce a correlation by multiplying it with a spectral density function with a power-law exponent $\alpha^{(C)}$; (iv) retrieve a long-range correlated spatial field by taking the inverse Fourier transform.
+This field is combined to a density field $\rho_i$ to determine a binary value for the cell: it is populated if $p_i > \theta_i$ with $\rho_i = \int_{-\infty}^{\theta_i} d\mathbb{P}(p_i)$.
+We generalize the initial model by taking a polycentric density field with
 
-The method to generate a spatial field exhibiting long range correlations was introduced for problems in physics by \cite{makse1996method}.
+\begin{equation}
+\rho_i (\vec{x}) = \sum_{j=1}^{n^{(C)}} \exp\left(\lVert\vec{x} - \vec{x}_j\rVert / r_j\right)
+\label{eq:kernelmixture}
+\end{equation}
+
+where the kernel centers are chosen at random and kernel sizes $r_j$ are taken such that kernel populations follow a rank-size law of hierarchy $\beta^{(C)}$ and the largest kernel has a fixed size $r_0^{(C)}$.
 
@@
 Finally, to provide some kind of null model to understand the advantages of each approach compared to a simple description of population distribution, we also include urban forms generated as kernel mixtures.
-We consider in particular exponential mixtures \cite{anas1998urban}, written as
+We consider in particular exponential mixtures \cite{anas1998urban}, where population density is written as previously in Eq.~\ref{eq:kernelmixture}.
+Parameters for this model are the number of kernels $n^{(E)}$, the rank-size hierarchy $\alpha^{(E)}$, the size of the largest kernel $r_0^{E}$.
+Contrary to previous models in which total population had an influence by controlling the speed of growth, density can here be rescaled arbitrarily (morphological indicators used are not changed through rescaling, see below), and we set the maximal density for one kernel to one.
 
@@
 
-Quantitative measures of urban form
+Quantitative measures of urban form are multiple and depend on the scale considered \cite{zhang2005metrics}.
+\cite{raimbault2019generating} for example introduces measures for buildings at the district scale.
+The field of Landscape Ecology has its own metrics similar to urban form measures \cite{10.1371/journal.pone.0225734}.
+For the scale we consider and considering population distribution only, metrics of urban form have been proposed for example to quantify sprawl \cite{tsai2005quantifying}.
+These can be related to fractal approaches to urban form \cite{chen2011derivation}.
+The effective dimension when applied to real cities is reasonably low \cite{schwarz2010urban}, and a few complementary indicators can be used.
+We thus follow \cite{10.1371/journal.pone.0203516} and consider urban form measures which are: (i) Moran index to capture spatial autocorrelation and the existence of centers; (ii) average distance between individuals which captures a level of aggregation; (iii) distribution entropy (aspatial) to capture the uniformity of the distribution; and (iv) rank-size slope which captures the hierarchy of population distribution.
 
@@
 In our case, we use its workflow system and an integrated algorithm to determine the feasible space of models.
-
-% TODO examples for each model
 
@@
 
+\subsection*{Generated urban forms}
+
+% examples for each model
+
+%%%%%%%%%%%%%%
+\begin{figure}[!h]
+	\includegraphics[width=\linewidth]{figures/Fig3.png}
+	\caption{{\bf .}\label{fig:fig3}}
+\end{figure}
+%%%%%%%%%%%%%%
+
 \subsection*{Feasible morphological spaces}
+
+%%%%%%%%%%%%%%
+\begin{figure}[!h]
+	\includegraphics[width=\linewidth]{figures/Fig4.png}
+	\caption{{\bf .}\label{fig:fig4}}
+\end{figure}
+%%%%%%%%%%%%%%
 
@@
 
-\section*{Acknowledgments}
+%\section*{Acknowledgments}
 

### 2020-08-30 Docs/Papers/MesoBenchmark/paper/figures.sh

@@ -10,3 +10,9 @@ montage figuresraw/mapindic_alpha_2015.png figuresraw/mapindic_moran_2015.png -t
 
 # Fig 2
 montage figuresraw/mapindic_cluster.png figuresraw/distributions_cluster-year.png -tile 1x2 -geometry +0+"$VERTICALPADDING" -resize "$WIDTH"x figures/Fig2.png
+
+# Fig 3
+montage figuresraw/CorrelatedPerco_size200_gradient002_corr04_centers3_nonBinary.png figuresraw/Expmixture_size200_centers10_maxRadiusRate0-2_hierarchy1_seed-2489151330471457438.png figuresraw/Gravity_size200_g03_gravity2-5_hierarchy0-5_centers3_pop20000_seed7990907742188865088.png figuresraw/ReactionDiffusion_size200_alpha0-8_beta0-2_diffSteps1_growthRate100_population5000_seed-2225875956205852410.png -tile 2x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x -border 2 -bordercolor Black figures/Fig3.png
+
+# Fig 4
+cp figuresraw/scatter_all.png figures/Fig4.png

### 2020-08-30 Models/MesoBenchmark/pseResults.R

@@ -5,21 +5,78 @@ library(ggplot2)
 library(dplyr)
 library(GGally)
 
+source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
+
+resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/MesoBenchmark/')
 
 files = list(
-  'expmixture' = '20200828_1515_PSE_EXPMIXTURE_res/population20000.csv',
-  'reactiondiffusion' = '20200828_1624_PSE_REACTIONDIFFUSION_res/population50000.csv'
+  'expmixture' = 'openmole/pse/20200829_1514_PSE_EXPMIXTURE_res/population100000.csv',
+  'reactiondiffusion' = 'openmole/pse/20200828_1624_PSE_REACTIONDIFFUSION_res/population50000.csv',
+  'correlatedperco' = 'openmole/pse/20200829_1330_PSE_CORRELATEDPERCO_res/population100000.csv',
+  'gravity' = 'openmole/pse/20200829_1526_PSE_GRAVITY_res/population37000.csv'
 )
 
 indics = c('moran','distance','entropy','slope')
 
 res = data.frame()
 for(model in names(files)){
-  currentres = as.tbl(read.csv(paste0('openmole/pse/',files[[model]])))
+  currentres = as.tbl(read.csv(files[[model]]))
   res = rbind(res,cbind(currentres[,c(indics,'evolution.samples')],model=rep(model,nrow(currentres))))
 }
 
-ggpairs(res[res$slope>-4,indics],aes(color=model))
+
+ggsave(plot = ggpairs(res[res$slope>-4,],aes(color=model),columns = indics,
+                      lower = list(continuous = wrap("points", alpha = 0.6,size=0.2,shape='.')),
+                      diag = list(continuous = wrap("densityDiag", alpha = 0.4))
+                      )+stdtheme,filename = paste0(resdir,'scatter_models.png'),width = 40,height=30,units='cm')
 
 
+# with real points
+morphos <- as.tbl(read.csv(file='../MesoCalibration/configs/morphologies.csv',header = T,sep = ";"))
+real = morphos[,c("moran2015","avgDist2015","entropy2015","alpha2015")]
+names(real)<- indics
+
+res = rbind(res,cbind(real,evolution.samples=rep(0,nrow(real)),model=rep('real',nrow(real))))
+
[…]

### 2020-10-01 Docs/Papers/MesoBenchmark/arxiv/v1/paper.tex
submission_mesobenchmark

--- 2020-08-30 Docs/Papers/MesoBenchmark/paper/paper.tex
@@
 % Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
+%\bibliographystyle{plos2015}
 
@@
 \begin{figure}[!h]
-	\includegraphics[width=\linewidth]{figures/Fig1.png}
-	\caption{{\bf .}\label{fig:fig1}}
+	\includegraphics[width=\linewidth]{Fig1.png}
+	\caption{{\bf Maps of urban morphology indicators for worldwide urban areas.} (Top) Rank-size hierarchy; (Bottom) Moran spatial autocorrelation index.
+Area of circles gives population.\label{fig:fig1}}
 \end{figure}
@@
 \begin{figure}[!h]
-	\includegraphics[width=\linewidth]{figures/Fig2.png}
-	\caption{{\bf .}\label{fig:fig1}}
+	\includegraphics[width=\linewidth]{Fig2.png}
+	\caption{{\bf Urban form typology.
+Using an unsupervised k-means clustering algorithm, we find clusters among urban areas.} (Top) Map of cluster belonging; (Bottom) Statistical distribution of indicator values, within each cluster and in time for all dates in the GHSL dataset.\label{fig:fig2}}
 \end{figure}
@@
 
+We compare model simulation results to urban form measures computed on worldwide urban areas.
+We use therefore the Global Human Settlement Layer database (GHSL), which provides an exhaustive worldwide population raster with a 1km resolution \cite{melchiorri2018unveiling}.
+\cite{Raimbault_2020} has shown the relevance of using this database for worldwide simulation models.
+It is available at four dates from 1975 onwards, but as not all models are dynamical we use the most recent population configuration only (2015).
+The database provides a layer of urban areas, within which we extract for the 1000 largest areas in terms of population a covering window ($\pm 25\%$ of extent on each side) from the population raster (note that some windows may be overlapping, as in the case of Hong-Kong which is separated from the main cluster of the Pearl River Delta mega-city region).
+Morphological indicators are computed on these extracted areas.
+
+We show in Fig.~\ref{fig:fig1} maps of indicators.
+More particularly, we map the rank-size hierarchy and Moran spatial autocorrelation which have meaningful geographical variations.
+Rank-size hierarchy will tell if the metropolitan area is strongly dominated by one center or if it more balanced.
+We retrieve the fact that in Europe, Paris and London are known for such a strong monocentricity, compared to cities in Germany for example.
+Similarly, mega-city regions in East Asia (Pearl River delta, Yangtze River delta, Beijing-Tianjin) are more polycentric and thus balanced than Wuhan or Seoul.
+Regarding spatial autocorrelation, we also observe a strong variation in East Asia (Tokyo compared to Chinese megacities for example), and within India (agglomerations in the Gange plain making a cluster of non-correlated, thus highly sprawled areas).
+
+The areas can be clustered following a non-supervised approach.
+We proceed to a k-means clustering on normalized indicator values, and find $k=4$ clusters as meaningful regarding the derivative of within-cluster variance.
+A map of cluster belonging and cluster profiles are shown in Fig.~\ref{fig:fig2}.
+We retrieve the variation within East Asia (Tokyo, Seoul and Shanghai being each in a different cluster) but less in Europe.
+The main cluster (light blue) corresponds to strongly monocentric urban areas.
+We see in density distributions of indicator values that clusters have clearly different profiles, which correspond to different typologies of urban morphology that the models try to approximate.
+For example, cluster 3 (light blue) and 1 (red) have the same level of hierarchy, but the latter has a much higher autocorrelation and entropy, and corresponds thus to more polycentric configurations.
+
 %\subsection*{Behavior of models}
 % also saltelli?
+
+%%%%%%%%%%%%%%
+\begin{figure}[!h]
+	\includegraphics[width=\linewidth]{Fig3.png}
+	\caption{{\bf Examples of generated urban forms for a square world of width $W = 50$.} From Left to Right and Top to Bottom: (i) Correlated percolation model for $r_0^{(C)} = 50$, $\alpha^{(C)} = 0.4$, $n^{(C)} = 3$; (ii) Exponential mixture model for $n^{(E)} = 10$, $r_0^{(E)} = 40$, $\alpha^{(E)} = 1$; (iii) Gravity model for $g^{(G)} = 0.3$, $\gamma_D^{(G)} = 2.5$, $\gamma_P^{(G)} = 0.5$, $P_0^{(G)} = 3$, $P_M^{(G)} = 20000$; (iv) Reaction-diffusion model for $\alpha^{(R)} = 0.8$, $\beta^{(R)} = 0.2$, $d^{(R)} = 1$, $N^{(R)} = 100$, $P_M^{(R)} = 5000$.\label{fig:fig3}}
+\end{figure}
+%%%%%%%%%%%%%%
 
@@
 
+We show in Fig.~\ref{fig:fig3} examples of generated urban forms for each model included in this study.
+Visually, these urban form look rather different.
+To what extent they are statistically distant for the morphology indicators can only be determined by systematic experiments.
+The gravity and percolation results look similar, although at a slightly different scale.
+This particular configuration of the reaction-diffusion model corresponds more to a rural or peri-urban configuration, while the exponential mixture is fuzzy and would resemble a blurred polycentric urban configuration.
+
 %%%%%%%%%%%%%%
 \begin{figure}[!h]
-	\includegraphics[width=\linewidth]{figures/Fig3.png}
-	\caption{{\bf .}\label{fig:fig3}}
+	\includegraphics[width=\linewidth]{Fig4.png}
+	\caption{{\bf Feasible morphological space for each simulation model.} The lower part of the plot matrix gives a scatterplot between each indicator.
+We also plot real urban areas in purple.
+The diagonal gives statistical distribution, while the correlation matrix is given in the upper part, conditionally to each model.\label{fig:fig4}}
 \end{figure}
@@
 
+We now turn to the main experiment of this paper: using a diversity search algorithm to determine the full feasible morphological space for each model.
+Therefore, we use the Pattern Space Exploration (PSE) algorithm \cite{cherel2015beyond} embedded in OpenMOLE.
+Diversity search was introduced in the field of artificial life as genetic algorithms with the aim to maximize diversity of the population \cite{lehman2008exploiting}.
+In the case of the PSE algorithm, a novelty criteria leads the search towards new regions of the indicator space, and results are stored in a hitmap.
+When running long enough, convergence in terms of number of solutions found is generally reached, and one can consider the final population as the feasible space of the model.
+We run here the algorithm up to 100,000 generations for each model, with grid of step 0.05 in the indicator space.
+Convergence was reached separately for each model.
+
+We show in Fig~\ref{fig:fig4} a scatterplot of the final population in the morphological space.
+First of all, we observe that correlation are very different for each model (for example opposite value for gravity and reaction-diffusion between entropy and slope), confirming that the way urban form is produced is very different although the final form may be the same.
+Then, we observe that the point clouds slightly intersect but also have their own proper space that which no other model can reach.
+For example, in the entropy-slope space, the reaction-diffusion model is very flexible and covers a large part of space, while gravity is restricted to a small region and correlated percolation and exponential mixture exhibit much less variability with a strong correlation between the two dimensions.
+The models cover a similar region for the Moran-entropy space, but all fail to capture an important part of real points (in purple), corresponding to configurations with a high Moran index but a relatively low entropy.
+Models systematically associate a high moran with a high entropy.
+In other dimensions, real points are reasonably covered.
+Thus, we can conclude that (i) the different models are complementary in terms of urban forms produced and (ii) that a part of real configurations are approached by some of the models, but other can not be reproduced.
+
 %%%%%%%%%%%%%%
 \begin{figure}[!h]
-	\includegraphics[width=\linewidth]{figures/Fig4.png}
-	\caption{{\bf .}\label{fig:fig4}}
+	\begin{center}
+	\includegraphics[width=0.6\linewidth]{Fig5.png}
+	\end{center}
+	\caption{{\bf Overlap between point cloud hypervolumes.}\label{fig:fig5}}
 \end{figure}
 %%%%%%%%%%%%%%
+
+To better quantify how each model are similar and how they approach real configurations, we compute the hypervolume (in the four dimensional indicator space) corresponding to each point cloud and the intersections between these hypervolumes.
+We use the \texttt{hypervolume} \texttt{R} package \cite{hypervolume} with a gaussian kernel density estimation with adaptable bandwidth.
+Then, we compute for each couple of model the ratio between the intersection of hypervolumes and the second model.
+The non-symmetric relative overlap matrix is plotted in Fig.~\ref{fig:fig5} (diagonal was removed for a better readability).
+We find that the closer model are the correlated percolation and reaction diffusion, with close to 50\% overlap.
+Then the gravity point cloud is mostly contained by reaction-diffusion and correlated percolation, but not the other way around.
+Thus models produce similar points in the parameter space, but most of their output cloud is original.
+Finally, we find regarding the proximity to real configurations that the best model is the exponential mixture.
+At the price of producing blurry urban forms, the flexibility therein is the best to approach the best existing configurations.
+Then comes the reaction-diffusion model, and the least flexible to approach real values is the correlated percolation model.
 
@@
 
+We have shown that different simple urban morphogenesis models are strongly complementary in the morphological space, confirming that a complementarity of processes also leads to a complementarity in patterns produced.
+This result rejoins the results obtained by \cite{raimbault2018multi} in the case of transportation network generative models at a similar mesoscopic scale, the results of \cite{raimbault2019generating} for generating building configurations, and the results of \cite{raimbault2018systematic} in the case of dynamical models for systems of cities.
+Both found a complementarity of approach including different types of processes.
+We argue that this is further evidence of the multi-dimensionality of urban systems and for a necessity of a plurality of urban models to capture both diverse processes but also outcomes.
+
 % future dev
@@
 % dynamical calib
-\cite{raimbault:halshs-02406539}
+
+This work is a first step towards a systematic benchmark of simple urban morphogenesis models.
+Future work should include the search for an explanation of the unreachable real urban configuration, and possibly alternative models approaching these.
+Other urban form indicators should also be tested.
+Finally, dynamical calibration of models remains an open question, investigated in the case of the reaction-diffusion model by \cite{raimbault:halshs-02406539}.
+An issue is that some models such as the correlated percolation model are not dynamical and should thus be adapted to be calibrated between successive points in time.
 
 \section*{Conclusion}
+
+We have implemented and systematically compared four very different simple models for urban morphogenesis, including reaction-diffusion, correlated percolation, exponential mixture and a gravitational aggregation model.
+We applied a diversity search algorithm to obtain the feasible morphological space.
+The results confirm a complementarity between the models and the relevance of a plurality in urban modeling approaches.
 
@@
 
-\bibliography{biblio}
+\begin{thebibliography}{10}
+
+\bibitem{batty2018inventing}
+Batty M.
+\newblock Inventing future cities.
+\newblock MIT press; 2018.
+
+\bibitem{williams2000achieving}
+Williams K, Burton E, Jenks M.
+\newblock Achieving sustainable urban form: an introduction.
+\newblock Achieving sustainable urban form. 2000;2000:1--5.
+
+\bibitem{le2012urban}
+Le~N{\'e}chet F.
+\newblock Urban spatial structure, daily mobility and energy consumption: a
+  study of 34 european cities.
+\newblock Cybergeo: European Journal of Geography. 2012;.
+
+\bibitem{viguie2012trade}
+Vigui{\'e} V, Hallegatte S.
+\newblock Trade-offs and synergies in urban climate policies.
+\newblock Nature Climate Change. 2012;2(5):334--337.
+
+\bibitem{wegener2004land}
+Wegener M, F{\"u}rst F.
+\newblock Land-use transport interaction: state of the art.
+\newblock Available at SSRN 1434678. 2004;.
+
+\bibitem{milton2019accelerating}
+Milton R, Roumpani F.
+\newblock Accelerating Urban Modelling Algorithms with Artificial Intelligence.
+\newblock In: Proceedings of the 5th International Conference on Geographical
+  Information Systems Theory, Applications and Management. vol.~1. INSTICC;
+  2019. p. 105--116.
+
+\bibitem{batty1997cellular}
+Batty M.
+\newblock Cellular automata and urban form: a primer.
+\newblock Journal of the American Planning Association. 1997;63(2):266--274.
+
+\bibitem{makse1998modeling}
+Makse HA, Andrade JS, Batty M, Havlin S, Stanley HE, et~al.
+\newblock Modeling urban growth patterns with correlated percolation.
+\newblock Physical Review E. 1998;58(6):7054.
+
+\bibitem{batty2007cities}
+Batty M.
+\newblock Cities and complexity: understanding cities with cellular automata,
+  agent-based models, and fractals.
+\newblock The MIT press; 2007.
+
+\bibitem{murcio2015urban}
+Murcio R, Morphet R, Gershenson C, Batty M.
+\newblock Urban transfer entropy across scales.
+\newblock PLoS One. 2015;10(7):e0133780.
+
+\bibitem{batty1989urban}
+Batty M, Longley P, Fotheringham S.
+\newblock Urban growth and form: scaling, fractal geometry, and
+  diffusion-limited aggregation.
+\newblock Environment and planning A. 1989;21(11):1447--1472.
+
+\bibitem{murcio2009colored}
+Murcio R, Rodr{\'\i}guez-Romo S.
+\newblock Colored diffusion-limited aggregation for urban migration.
+\newblock Physica A: Statistical Mechanics and its Applications.
+  2009;388(13):2689--2698.
+
+\bibitem{murcio2013second}
+Murcio R, Sosa-Herrera A, Rodriguez-Romo S.
+\newblock Second-order metropolitan urban phase transitions.
+\newblock Chaos, Solitons \& Fractals. 2013;48:22--31.
+
+\bibitem{10.1371/journal.pone.0203516}
+Raimbault J.
+\newblock Calibration of a density-based model of urban morphogenesis.
+\newblock PLOS ONE. 2018;13(9):1--18.
+\newblock doi:{10.1371/journal.pone.0203516}.
+
+\bibitem{li2019singularity}
+Li Y, Rybski D, Kropp JP.
+\newblock Singularity cities.
+\newblock Environment and Planning B: Urban Analytics and City Science. 2019;
+  p. 2399808319843534.
+
+\bibitem{pumain2020conclusion}
+Pumain D, Raimbault J.
+\newblock Conclusion: Perspectives on urban theories.
+\newblock In: Theories and Models of Urbanization. Springer; 2020. p. 303--330.
+
+\bibitem{fotheringham1989spatial}
+Fotheringham AS, O'Kelly ME.
+\newblock Spatial interaction models: formulations and applications. vol.~1.
+\newblock Kluwer Academic Publishers Dordrecht; 1989.
+
+\bibitem{rybski2013distance}
+Rybski D, Ros AGC, Kropp JP.
+\newblock Distance-weighted city growth.
+\newblock Physical Review E. 2013;87(4):042114.
+
+\bibitem{turing1990chemical}
+Turing AM.
+\newblock The chemical basis of morphogenesis.
+\newblock Bulletin of mathematical biology. 1990;52(1-2):153--197.
+
+\bibitem{raimbault2018co}
+Raimbault J.
+\newblock Co-evolution and morphogenetic systems.
+\newblock arXiv preprint arXiv:180311457. 2018;.
+
+\bibitem{bonin2014modelisation}
+Bonin O, Hubert JP.
+\newblock Mod{\'e}lisation morphog{\'e}n{\'e}tique de moyen terme des villes:
+  une sch{\'e}matisation du mod{\`e}le th{\'e}orique de Ritchot et Desmarais
+  dans le cadre du mod{\`e}le standard de l'{\'e}conomie urbaine.
+\newblock Revue dEconomie Regionale Urbaine. 2014;(3):471--497.
+
+\bibitem{parish2001procedural}
+Parish YI, M{\"u}ller P.
+\newblock Procedural modeling of cities.
+\newblock In: Proceedings of the 28th annual conference on Computer graphics
+  and interactive techniques; 2001. p. 301--308.
+
+\bibitem{makse1996method}
+Makse HA, Havlin S, Schwartz M, Stanley HE.
+\newblock Method for generating long-range correlations for large systems.
+\newblock Physical Review E. 1996;53(5):5445.
+
+\bibitem{anas1998urban}
+Anas A, Arnott R, Small KA.
+\newblock Urban spatial structure.
+\newblock Journal of economic literature. 1998;36(3):1426--1464.
+
+\bibitem{zhang2005metrics}
+Zhang M, Kukadia N.
+\newblock Metrics of urban form and the modifiable areal unit problem.
+\newblock Transportation Research Record. 2005;1902(1):71--79.
+
+\bibitem{raimbault2019generating}
+Raimbault J, Perret J.
+\newblock Generating urban morphologies at large scales.
+\newblock In: Artificial Life Conference Proceedings. MIT Press; 2019. p.
+  179--186.
+
+\bibitem{10.1371/journal.pone.0225734}
+Bosch M.
+\newblock PyLandStats: An open-source Pythonic library to compute landscape
+  metrics.
+\newblock PLOS ONE. 2019;14(12):1--19.
+\newblock doi:{10.1371/journal.pone.0225734}.
+
+\bibitem{tsai2005quantifying}
+Tsai YH.
+\newblock Quantifying urban form: compactness versus' sprawl'.
+\newblock Urban studies. 2005;42(1):141--161.
+
+\bibitem{chen2011derivation}
+Chen Y.
+\newblock Derivation of the functional relations between fractal dimension of
+  and shape indices of urban form.
+\newblock Computers, Environment and Urban Systems. 2011;35(6):442--451.
+
+\bibitem{schwarz2010urban}
+Schwarz N.
+\newblock Urban form revisited—Selecting indicators for characterising
+  European cities.
+\newblock Landscape and urban planning. 2010;96(1):29--47.
+
+\bibitem{raimbault2020scala}
+Raimbault J, Perret J, Reuillon R.
+\newblock A scala library for spatial sensitivity analysis.
+\newblock arXiv preprint arXiv:200710667. 2020;.
+
+\bibitem{reuillon2013openmole}
+Reuillon R, Leclaire M, Rey-Coyrehourcq S.
+\newblock OpenMOLE, a workflow engine specifically tailored for the distributed
+  exploration of simulation models.
+\newblock Future Generation Computer Systems. 2013;29(8):1981--1990.
+
+\bibitem{melchiorri2018unveiling}
+Melchiorri M, Florczyk AJ, Freire S, Schiavina M, Pesaresi M, Kemper T.
+\newblock Unveiling 25 years of planetary urbanization with remote sensing:
+  Perspectives from the global human settlement layer.
+\newblock Remote Sensing. 2018;10(5):768.
+
+\bibitem{Raimbault_2020}
+Raimbault J, Denis E, Pumain D.
+\newblock Empowering Urban Governance through Urban Science: Multi-Scale
+  Dynamics of Urban Systems Worldwide.
+\newblock Sustainability. 2020;12(15):5954.
+\newblock doi:{10.3390/su12155954}.
+
+\bibitem{cherel2015beyond}
+Ch{\'e}rel G, Cottineau C, Reuillon R.
+\newblock Beyond corroboration: Strengthening model validation by looking for
+  unexpected patterns.
+\newblock PloS one. 2015;10(9):e0138212.
+
+\bibitem{lehman2008exploiting}
+Lehman J, Stanley KO.
+\newblock Exploiting open-endedness to solve problems through the search for
+  novelty.
+\newblock In: ALIFE; 2008. p. 329--336.
+
+\bibitem{hypervolume}
+Blonder B, with contributions~from David J~Harris. hypervolume: High
+  Dimensional Geometry and Set Operations Using Kernel Density Estimation,
+  Support Vector Machines, and Convex Hulls; 2019.
+\newblock Available from: \url{https://CRAN.R-project.org/package=hypervolume}.
+
+\bibitem{raimbault2018multi}
+Raimbault J.
+\newblock Multi-modeling the morphogenesis of transportation networks.
+\newblock In: Artificial Life Conference Proceedings. MIT Press; 2018. p.
+  382--383.
+
+\bibitem{raimbault2018systematic}
+Raimbault J.
+\newblock A systematic comparison of interaction models for systems of cities.
+\newblock In: Conference on Complex Systems 2018; 2018.
+
+\bibitem{raimbault:halshs-02406539}
+Raimbault J.
+\newblock {Worldwide estimation of parameters for a simple reaction-diffusion
+  model of urban growth}.
+\newblock In: {International Land-use Symposium 2019}. Paris, France;
+  2019.Available from:
+  \url{https://halshs.archives-ouvertes.fr/halshs-02406539}.
+
+\end{thebibliography}
 

### 2020-10-01 Docs/Papers/MesoBenchmark/paper/coverletter.tex
submission_mesobenchmark

%% start of file `template.tex'.
%% Copyright 2006-2013 Xavier Danaux (xdanaux@gmail.com).
%
% This work may be distributed and/or modified under the
% conditions of the LaTeX Project Public License version 1.3c,
% available at http://www.latex-project.org/lppl/.

\documentclass[11pt,a4paper,sans]{moderncv}        % possible options include font size ('10pt', '11pt' and '12pt'), paper size ('a4paper', 'letterpaper', 'a5paper', 'legalpaper', 'executivepaper' and 'landscape') and font family ('sans' and 'roman')

\usepackage[document]{ragged2e}
% pour justifier

% moderncv themes
\moderncvstyle{banking}                            % style options are 'casual' (default), 'classic', 'oldstyle' and 'banking'
\moderncvcolor{red}                                % color options 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'
\renewcommand{\familydefault}{\rmdefault}         % to set the default font; use '\sfdefault' for the default sans serif font, '\rmdefault' for the default roman one, or any tex font name
%\nopagenumbers{}                                  % uncomment to suppress automatic page numbering for CVs longer than one page

% character encoding
\usepackage[utf8]{inputenc}                       % if you are not using xelatex ou lualatex, replace by the encoding you are using
%\usepackage{CJKutf8}                              % if you need to use CJK to typeset your resume in Chinese, Japanese or Korean

% adjust the page margins
\usepackage[scale=0.75]{geometry}
%\setlength{\hintscolumnwidth}{3cm}                % if you want to change the width of the column with the dates
%\setlength{\makecvtitlenamewidth}{10cm}           % for the 'classic' style, if you want to force the width allocated to your name and avoid line breaks. be careful though, the length is normally calculated to avoid any overlap with your personal info; use this at your own typographical risks...

% personal data
%\name{Yoann}{Potiron}
%\title{Keio University}                               % optional, remove / comment the line if not wanted
%\address{Assistant Professor Faculty Business and Commerce}{2-15-45 Mita, Minato-ku, Tokyo, 108-8345}{Japan}% optional, remove / comment the line if not wanted; the "postcode city" and and "country" arguments can be omitted or provided empty
%\phone[mobile]{+1~(234)~567~890}                   % optional, remove / comment the line if not wanted
%\phone[fixed]{+81~(0)3~5418~6571}                    % optional, remove / comment the line if not wanted
%\phone[fax]{+3~(456)~789~012}                      % optional, remove / comment the line if not wanted
%\email{potiron@fbc.keio.ac.jp}  
%\extrainfo{yoann.potiron@gmail.com}
% optional, remove / comment the line if not wanted
%\homepage{http://www.fbc.keio.ac.jp/\char`\~ potiron}                         % optional, remove / comment the line if not wanted
%\extrainfo{yoann.potiron@gmail.com}                 % optional, remove / comment the line if not wanted
%\photo[64pt][0.4pt]{picture}                       % optional, remove / comment the line if not wanted; '64pt' is the height the picture must be resized to, 0.4pt is the thickness of the frame around it (put it to 0pt for no frame) and 'picture' is the name of the picture file
%\quote{Some quote}                                 % optional, remove / comment the line if not wanted

% to show numerical labels in the bibliography (default is to show no labels); only useful if you make citations in your resume
%\makeatletter
%\renewcommand*{\bibliographyitemlabel}{\@biblabel{\arabic{enumiv}}}
%\makeatother
%\renewcommand*{\bibliographyitemlabel}{[\arabic{enumiv}]}% CONSIDER REPLACING THE ABOVE BY THIS

% bibliography with mutiple entries
%\usepackage{multibib}
%\newcites{book,misc}{{Books},{Others}}
%----------------------------------------------------------------------------------
%            content
%----------------------------------------------------------------------------------
\firstname{}
\lastname{}
\begin{document}
%-----       letter       ---------------------------------------------------------

% must contain :
%Summarize the study’s contribution to the scientific literature
%Relate the study to previously published work
%Specify the type of article (for example, research article, systematic review, meta-analysis, clinical trial)
%Describe any prior interactions with PLOS regarding the submitted manuscript
%Suggest appropriate Academic Editors to handle your manuscript (see the full list of Academic Editors)
%List any opposed reviewers

% recipient data
\recipient{Editors PLoS ONE}{}
\date{September 4th, 2020}
\opening{Dear Editors,}
\closing{Yours faithfully,\\
Juste Raimbault\\
University College London
}
         % use an optional argument to use a string other than "Enclosure", or redefine \enclname
\makelettertitle

\justify
I am pleased to submit an original research article entitled ``A comparison of simple models for urban morphogenesis'' for consideration for publication in PLoS ONE. This article gives a novel quantitative insight into four urban morphogenesis models. It does not contain previously published work. I did not have any previous interaction with PLOS regarding this manuscript.

This manuscript has not been published and is not under consideration for publication elsewhere. A preprint was deposited on arxiv simultaneously to the submission.

I have no conflicts of interest to disclose. I do not have any opposed reviewer. This paper is submitted to the PloS ONE collection ``Cities as Complex Systems'' edited by M. Gonzalez (University of California, Berkeley) and D. Rybski (Potsdam Institute for Climate Impact Research).

Thank you for your consideration.
\justify

\makeletterclosing

\end{document}

%% end of file `template.tex'.

### 2020-10-01 Docs/Papers/MesoBenchmark/paper/paper.tex
submission_mesobenchmark

@@
 
-%\documentclass[10pt,letterpaper]{article}
-%\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-\documentclass[10pt]{article}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+%\documentclass[10pt]{article}
 %\usepackage[margin=2cm]{geometry}
@@
 % Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
+\usepackage{setspace} 
+\doublespacing
 
 % Text layout
-%\raggedright
-%\setlength{\parindent}{0.5cm}
-%\textwidth 5.25in 
-%\textheight 8.75in
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
 
@@
 \usepackage{epstopdf}
-%\pagestyle{myheadings}
-%\pagestyle{fancy}
-%\fancyhf{}
-%%\setlength{\headheight}{27.023pt}
-%%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-%\rfoot{\thepage/\pageref{LastPage}}
-%\renewcommand{\headrulewidth}{0pt}
-%\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-%\fancyheadoffset[L]{2.25in}
-%\fancyfootoffset[L]{2.25in}
-%\lfoot{\today}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
 
@@
 
-%\begin{document}
-%\vspace*{0.2in}
+\begin{document}
+\vspace*{0.2in}
 
 % Title must be 250 characters or less.
-%\begin{flushleft}
-%\begin{center}
-%{\Large
-%\textbf\newline{A comparison of simple models for urban morphogenesis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-%}
-%\newline
+\begin{flushleft}
+{\Large
+\textbf\newline{A comparison of simple models for urban morphogenesis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
 % Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-%\\
-\title{A comparison of simple models for urban morphogenesis}
-\date{}
-
-\author{
+\\
 Juste Raimbault\textsuperscript{1,2,3,*}
@@
 * juste.raimbault@polytechnique.edu
-}
 
-\begin{document}
+\end{flushleft}
 
-\maketitle
-
-%\end{flushleft}
-%\end{center}
 % Please keep the abstract below 300 words
-%\section*{Abstract}
+\section*{Abstract}
 \vspace{1cm}
-\begin{abstract}
 The spatial distribution of population and activities within urban areas, or urban form at the mesoscopic scale, is the outcome of multiple antagonist processes.
@@
 We find a complementarity of the different types of processes, advocating for a plurality of urban models.
-\end{abstract}
 
-%\linenumbers
+% urban morphogenesis; reaction-diffusion; correlated percolation; gravity model; novelty search
+
+\linenumbers
 
@@
 In our case, we use its workflow system and an integrated algorithm to determine the feasible space of models.
+Source code for models is available in the open source code of the \texttt{spatialdata} library at \url{https://github.com/openmole/spatialdata}.
+Source code for numerical experiments, simulation data and results used in this paper are available on an open git repository at \url{https://github.com/JusteRaimbault/UrbanGrowth}.
 
@@
 \begin{figure}[!h]
-	\includegraphics[width=\linewidth]{figures/Fig1.png}
-	\caption{{\bf .}\label{fig:fig1}}
+	%\includegraphics[width=\linewidth]{figures/Fig1.png}
+	\vspace{2cm}
+	\caption{{\bf Maps of urban morphology indicators for worldwide urban areas.} (Top) Rank-size hierarchy; (Bottom) Moran spatial autocorrelation index.
+Area of circles gives population.\label{fig:fig1}}
 \end{figure}
@@
 \begin{figure}[!h]
-	\includegraphics[width=\linewidth]{figures/Fig2.png}
-	\caption{{\bf .}\label{fig:fig1}}
+	%\includegraphics[width=\linewidth]{figures/Fig2.png}
+	\vspace{2cm}
+	\caption{{\bf Urban form typology.
+Using an unsupervised k-means clustering algorithm, we find clusters among urban areas.} (Top) Map of cluster belonging; (Bottom) Statistical distribution of indicator values, within each cluster and in time for all dates in the GHSL dataset.\label{fig:fig2}}
 \end{figure}
@@
 
+We compare model simulation results to urban form measures computed on worldwide urban areas.
+We use therefore the Global Human Settlement Layer database (GHSL), which provides an exhaustive worldwide population raster with a 1km resolution \cite{melchiorri2018unveiling}.
+\cite{Raimbault_2020} has shown the relevance of using this database for worldwide simulation models.
+It is available at four dates from 1975 onwards, but as not all models are dynamical we use the most recent population configuration only (2015).
+The database provides a layer of urban areas, within which we extract for the 1000 largest areas in terms of population a covering window ($\pm 25\%$ of extent on each side) from the population raster (note that some windows may be overlapping, as in the case of Hong-Kong which is separated from the main cluster of the Pearl River Delta mega-city region).
+Morphological indicators are computed on these extracted areas.
+
+We show in Fig.~\ref{fig:fig1} maps of indicators.
+More particularly, we map the rank-size hierarchy and Moran spatial autocorrelation which have meaningful geographical variations.
+Rank-size hierarchy will tell if the metropolitan area is strongly dominated by one center or if it more balanced.
+We retrieve the fact that in Europe, Paris and London are known for such a strong monocentricity, compared to cities in Germany for example.
+Similarly, mega-city regions in East Asia (Pearl River delta, Yangtze River delta, Beijing-Tianjin) are more polycentric and thus balanced than Wuhan or Seoul.
+Regarding spatial autocorrelation, we also observe a strong variation in East Asia (Tokyo compared to Chinese megacities for example), and within India (agglomerations in the Gange plain making a cluster of non-correlated, thus highly sprawled areas).
+
+The areas can be clustered following a non-supervised approach.
+We proceed to a k-means clustering on normalized indicator values, and find $k=4$ clusters as meaningful regarding the derivative of within-cluster variance.
+A map of cluster belonging and cluster profiles are shown in Fig.~\ref{fig:fig2}.
+We retrieve the variation within East Asia (Tokyo, Seoul and Shanghai being each in a different cluster) but less in Europe.
+The main cluster (light blue) corresponds to strongly monocentric urban areas.
+We see in density distributions of indicator values that clusters have clearly different profiles, which correspond to different typologies of urban morphology that the models try to approximate.
+For example, cluster 3 (light blue) and 1 (red) have the same level of hierarchy, but the latter has a much higher autocorrelation and entropy, and corresponds thus to more polycentric configurations.
+
 %\subsection*{Behavior of models}
 % also saltelli?
+
+%%%%%%%%%%%%%%
+\begin{figure}[!h]
+	%\includegraphics[width=\linewidth]{figures/Fig3.png}
+	\vspace{2cm}
+	\caption{{\bf Examples of generated urban forms for a square world of width $W = 50$.} From Left to Right and Top to Bottom: (i) Correlated percolation model for $r_0^{(C)} = 50$, $\alpha^{(C)} = 0.4$, $n^{(C)} = 3$; (ii) Exponential mixture model for $n^{(E)} = 10$, $r_0^{(E)} = 40$, $\alpha^{(E)} = 1$; (iii) Gravity model for $g^{(G)} = 0.3$, $\gamma_D^{(G)} = 2.5$, $\gamma_P^{(G)} = 0.5$, $P_0^{(G)} = 3$, $P_M^{(G)} = 20000$; (iv) Reaction-diffusion model for $\alpha^{(R)} = 0.8$, $\beta^{(R)} = 0.2$, $d^{(R)} = 1$, $N^{(R)} = 100$, $P_M^{(R)} = 5000$.\label{fig:fig3}}
+\end{figure}
+%%%%%%%%%%%%%%
 
@@
 
+We show in Fig.~\ref{fig:fig3} examples of generated urban forms for each model included in this study.
+Visually, these urban form look rather different.
+To what extent they are statistically distant for the morphology indicators can only be determined by systematic experiments.
+The gravity and percolation results look similar, although at a slightly different scale.
+This particular configuration of the reaction-diffusion model corresponds more to a rural or peri-urban configuration, while the exponential mixture is fuzzy and would resemble a blurred polycentric urban configuration.
+
 %%%%%%%%%%%%%%
 \begin{figure}[!h]
-	\includegraphics[width=\linewidth]{figures/Fig3.png}
-	\caption{{\bf .}\label{fig:fig3}}
+	%\includegraphics[width=\linewidth]{figures/Fig4.png}
+	\vspace{2cm}
+	\caption{{\bf Feasible morphological space for each simulation model.} The lower part of the plot matrix gives a scatterplot between each indicator.
+We also plot real urban areas in purple.
+The diagonal gives statistical distribution, while the correlation matrix is given in the upper part, conditionally to each model.\label{fig:fig4}}
 \end{figure}
@@
 
+We now turn to the main experiment of this paper: using a diversity search algorithm to determine the full feasible morphological space for each model.
+Therefore, we use the Pattern Space Exploration (PSE) algorithm \cite{cherel2015beyond} embedded in OpenMOLE.
+Diversity search was introduced in the field of artificial life as genetic algorithms with the aim to maximize diversity of the population \cite{lehman2008exploiting}.
+In the case of the PSE algorithm, a novelty criteria leads the search towards new regions of the indicator space, and results are stored in a hitmap.
+When running long enough, convergence in terms of number of solutions found is generally reached, and one can consider the final population as the feasible space of the model.
+We run here the algorithm up to 100,000 generations for each model, with grid of step 0.05 in the indicator space.
+Convergence was reached separately for each model.
+
+We show in Fig~\ref{fig:fig4} a scatterplot of the final population in the morphological space.
+First of all, we observe that correlation are very different for each model (for example opposite value for gravity and reaction-diffusion between entropy and slope), confirming that the way urban form is produced is very different although the final form may be the same.
+Then, we observe that the point clouds slightly intersect but also have their own proper space that which no other model can reach.
+For example, in the entropy-slope space, the reaction-diffusion model is very flexible and covers a large part of space, while gravity is restricted to a small region and correlated percolation and exponential mixture exhibit much less variability with a strong correlation between the two dimensions.
+The models cover a similar region for the Moran-entropy space, but all fail to capture an important part of real points (in purple), corresponding to configurations with a high Moran index but a relatively low entropy.
+Models systematically associate a high moran with a high entropy.
+In other dimensions, real points are reasonably covered.
+Thus, we can conclude that (i) the different models are complementary in terms of urban forms produced and (ii) that a part of real configurations are approached by some of the models, but other can not be reproduced.
+
 %%%%%%%%%%%%%%
 \begin{figure}[!h]
-	\includegraphics[width=\linewidth]{figures/Fig4.png}
-	\caption{{\bf .}\label{fig:fig4}}
+	%\begin{center}
+	%\includegraphics[width=0.6\linewidth]{figures/Fig5.png}
+	%\end{center}
+	\vspace{2cm}
+	\caption{{\bf Overlap between point cloud hypervolumes.}\label{fig:fig5}}
 \end{figure}
 %%%%%%%%%%%%%%
+
+To better quantify how each model are similar and how they approach real configurations, we compute the hypervolume (in the four dimensional indicator space) corresponding to each point cloud and the intersections between these hypervolumes.
+We use the \texttt{hypervolume} \texttt{R} package \cite{hypervolume} with a gaussian kernel density estimation with adaptable bandwidth.
+Then, we compute for each couple of model the ratio between the intersection of hypervolumes and the second model.
+The non-symmetric relative overlap matrix is plotted in Fig.~\ref{fig:fig5} (diagonal was removed for a better readability).
+We find that the closer model are the correlated percolation and reaction diffusion, with close to 50\% overlap.
+Then the gravity point cloud is mostly contained by reaction-diffusion and correlated percolation, but not the other way around.
+Thus models produce similar points in the parameter space, but most of their output cloud is original.
+Finally, we find regarding the proximity to real configurations that the best model is the exponential mixture.
+At the price of producing blurry urban forms, the flexibility therein is the best to approach the best existing configurations.
+Then comes the reaction-diffusion model, and the least flexible to approach real values is the correlated percolation model.
 

### 2020-10-01 Docs/Papers/MesoBenchmark/paper/figures.sh

@@ -1,7 +1,7 @@
 
 JPGQUALITY=50
 PDFRESOLUTION=200
-WIDTH=2000
+WIDTH=1000
 HORIZONTALPADDING=10
 VERTICALPADDING=10
 
@@ -16,3 +16,6 @@ montage figuresraw/CorrelatedPerco_size200_gradient002_corr04_centers3_nonBinary
 
 # Fig 4
 cp figuresraw/scatter_all.png figures/Fig4.png
+
+# Fig 5
+cp figuresraw/overlap.png figures/Fig5.png

### 2020-10-01 Models/MesoBenchmark/pseResults.R

@@ -49,6 +49,7 @@ ggsave(plot = ggpairs(res[res$slope>-4,],aes(color=model),columns = indics,
 # for hypervolume intersection: https://rdrr.io/cran/hypervolume/man/hypervolume_set.html
 
 library(hypervolume)
+library(reshape2)
 
 models = c(names(files),'real')
 res=res[res$slope>-4,]
@@ -67,16 +68,19 @@ for(model1 in 1:length(models)){
   for (model2 in 1:length(models)){
     if(model1!=model2){
       show(paste0(models[model1],' / ',models[model2]))
-      hvset = hypervolume_set(hvs[[models[model1]]],hvs[[models[model2]]],check.memory=FALSE)
+      hvset = hypervolume_set(hvs[[models[model1]]],hvs[[models[model2]]],check.memory=FALSE,num.points.max=100000)
       overlaps[model1,model2] = hvset@HVList$Intersection@Volume / hvs[[models[model2]]]@Volume
     }
   }
 }
 rownames(overlaps)<-models
 colnames(overlaps)<-models
+diag(overlaps)=NA
+save(overlaps,hvs,file='hvs.RData')
 
-melt(overlaps)
-
+g=ggplot(melt(overlaps),aes(x=Var1,y=Var2,fill=value))
+g+geom_raster()+xlab('')+ylab('')+scale_fill_continuous(name='Overlap')+stdtheme
+ggsave(file=paste0(resdir,'overlap.png'),width=30,height=26,units='cm')
 
 
 

### 2020-10-01 Docs/Papers/HandbookEntropy/figures.sh
handbook_entropy

@@ -1,0 +1,12 @@
+
+JPGQUALITY=50
+PDFRESOLUTION=200
+WIDTH=1500
+HEIGHT=1000
+HORIZONTALPADDING=10
+VERTICALPADDING=10
+
+convert figuresraw/innovation_leg.png -gravity center -extent 3500x figuresraw/tmpleg.png
+montage figuresraw/innovation_1.png figuresraw/innovation_2.png figuresraw/innovation_3.png figuresraw/innovation_4.png -tile 2x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x -border 2 -bordercolor Black figuresraw/tmp3.png
+montage figuresraw/tmp3.png figuresraw/tmpleg.png -tile 1x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x figures/Fig3.jpg
+rm figuresraw/tmp3.png figuresraw/tmpleg.png

### 2020-10-28 Docs/Papers/HandbookEntropy/arxiv/v1/chapter.tex
chapter_arxiv

\documentclass[10pt]{article}

\usepackage{graphicx,amsmath,amssymb,ragged2e,natbib}

\usepackage[margin=2cm]{geometry}

\title{Spatial dynamics of complex urban systems within an evolutionary theory frame}

\date{}

\author{
Juste Raimbault\textsuperscript{1,2,3*},
Denise Pumain\textsuperscript{3}\bigskip\\
\textsuperscript{1} Centre for Advanced Spatial Analysis, UCL, London, United Kingdom
\\
\textsuperscript{2} UPS CNRS 3611 ISC-PIF, Paris, France
\\
\textsuperscript{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France\bigskip\\
* \texttt{juste.raimbault@polytechnique.edu}
}

\begin{document}

\maketitle

\begin{abstract}
This chapter is about Complexity and Spatial Dynamics in Urban Systems. Strong inequalities in the size of cities and the apparent difficulty of limiting their growth raise practical issues for spatial planning. At a time when new constraints in terms of limited energy and raw material resources or possible catastrophic events such as pandemics are challenging further urban expansion, it is important to consolidate the theories from various scientific disciplines to estimate to what extent the urban dynamics can be modified. While briefly reviewing the contributions to urban theories provided by the new developments in complexity sciences, we first advocate for the soundness of urban theories. Second, we develop our original approach considering spatial interaction and evolutionary path dependence as major features in the general behavior of urban entities. Third, we test these principles grounded in an evolutionary theory of urban systems by experimenting four dynamic models of urban growth calibrated on harmonized empirical data sets with comparisons across the whole world.\medskip

	\textbf{Keywords:} spatial dynamics; complex systems; system of cities; evolutionary theory; urban growth; simulation
\end{abstract}

\justify

\section{Introduction}

The context of this chapter, namely ``Entropy, Complexity, and Spatial Dynamics: The Rebirth of Theory'' is challenging. The problem is twofold: first, it means demonstrating how the insights provided by concepts and models associated with the notion of complexity have contributed to theories on spatial dynamics; second, it questions whether this is a rebirth of the theoretical approach in this field, which would therefore presume its relative abandon before.

This chapter provides answers to these two questions about a particular dimension of spatial dynamics, that of cities and systems of cities. In response to the second question, and considering the progress of research on urban complexity over the last forty years or so, we identify an enrichment and consolidation of existing theories by the principles of complex systems rather than a true renaissance of urban theory. Indeed, contrary to what some scholars claim, the theories developed around the spatial dynamics of urbanization processes and the evolution of settlement systems are far from being obsolete \citep{brenner2014urban}. Neither are they limited in their fundamentals and application to only one part of the world \citep{robinson2016comparative}. \cite{scott2015nature} and \cite{wu2020emerging} rightly criticized both ``theories''. We may add that many scholars did not abandon the former urban theories but on contrary tested, revised and completed them during the last four decades \citep{pumain1997pour,pumain1998urban,pumain2003approche,pumain2020theories,batty2013new}. The major theoretical challenge of that period was three folds: first, to shift from idiographic postures to nomothetic ones; second, from static views toward dynamic and evolutionary; and third, to really transfer concepts and models from natural sciences to build a socially relevant knowledge. Elements of knowledge from ancient theories can consolidate when situated in their geo-historical contexts, when tests of these theoretical propositions are carried out on subsets of empirical data that are well identified in time and space, and when the means are provided to articulate the observations made at multiple levels of geographic scales.

Starting from such a spiral theoretical conception of the cumulativeness of knowledge \citep{pumain2005cumulativite}, we try here to demonstrate how it is possible to construct an evolutionary theory of cities and systems of cities. This theory is abductive in the sense that it is constructed by frequent to-and-fro movements between empirical observations, logical propositions and mathematical and computer models. This theory is not entirely new as it integrates elements of knowledge already well identified by specialists in urban issues, such as geographers, economists, historians and archaeologists, sociologists, town planners and architects. Its originality stems from progress in international comparisons made through using new harmonized databases, methods for validating the results of computer simulation models, and putting these results into the perspective of complex systems theories.

This chapter brings a step forward in the long process of building and testing an evolutionary theory of systems of cities. This theory is based on empirical observations and tested with dynamic models that are designed for simulating urban development at various spatial and temporal scales. We summarize here the methodology and results already obtained in the GeoDiverCity project \citep{pumain2015multilevel,cura2017old,pumain2017urban}, including USA, Europe and BRICS countries, and the work of \cite{raimbault2020empowering} completing them with new datasets at world scale and other types of models. These models are designed for explaining urban growth and city size distributions with an increasing deepening in the complexity of the implemented processes \cite{cottineau2015growing}. They are all conceived for exploring the correspondences between urban trajectories observed at the meso-level of individual cities and the structuring of systems of cities at macro geographical scale. The validity and representativeness of these four models of complex urban systems are tested and their variations across regions of the world within different spatial frames are studied \citep{raimbault2018calibration,raimbault2020indirect}. This work is made possible by using different data sources, namely the GeoDivercity database and a new data source that for the first time compares the evolution of cities globally over a period between 1970 and 2015, the GHSL database. Eric \cite{denis2020more} already explored the source statistically for comparing urban sprawl trends in the countries of the world. \cite{raimbault2020empowering} calibrate four different models of urban growth on both sets of data, across different regions of the world, and analyze their results and assess for each their possible contribution to the theory of urban complex systems. We recall here the structure of these models, in relation to the question of how to build integrated urban theories.

The rest of this chapter is organized as follows: we first develop the importance of constructing urban theories and how they relate to models; we then develop the details of urban dynamics models introduced in the context of the evolutionary urban theory; we finally discuss important open problems crucial in the development of such theories.

\section{For a theoretical rearming in urban science}

To clarify our position vis-à-vis urban theories, we propose a first discussion of the diversity of possible meanings attributed to this word among the disciplines that have been interested in explaining the evolution of cities. This discussion is based on a few recent publications, without claiming to be exhaustive within the limits of this chapter. The main question we address is to understand how the disciplines most formerly interested in the scientific object ``city'' have been able to revise or complete their theoretical statements using the contributions proposed by the natural sciences, which have been advancing interesting concepts and models for thinking about and studying complex systems for some forty years. To begin with, we can start with the definition of complexity as proposed by Aura \cite{reggiani2009complexity} for whom ``the term "complexity" embeds both the assemblage of different units in a system and their intertwined dynamics. In other words, the term "complexity" is strictly related to the concept of networks''. In this definition we would emphasize the physical and societal interactions that are creating and using these networks with a consequence of making the evolution of ``different units'' strongly interdependent on each other. That leads to a specific deepening in the concept of co-evolution for cities belonging to systems of cities \citep{paulus2004coevolution,raimbault2020unveiling}.

Two repeated observations prompted the development of a theory of the evolution of systems of cities. The first, empirical, is the long persistence of urban networks in large integrated territories, which retain, over time, the same spatial configurations of the relative size of cities. This is visible for example on maps representing the population of cities in proportional circles more than a century apart, in Europe, in India, or even in the United States since 1950. The second results of applying to the evolution of these urban population distribution a simple stochastic statistical model formalized by the French statistician Gibrat since 1931. This model states that the growth of urban populations is proportional to the initial size of cities during each short time interval, with fluctuations such as growth rates (amount reported to the population) are statistically independent from the size of the cities and from one time interval to the next. The interest of this model of spatially distributed urban growth in integrated systems of cities is to predict that the result of such a growth process is always a lognormal distribution. Thus, the model provides a first statistical explanation to the ``mystery'' of the Zipf's law still recently mentioned by the economist Paul \cite{krugman1996confronting}. A huge literature is dedicated to Zipf’s law and Pareto distributions that are observed in many natural and social systems, to the extent that these ``universal'' laws, together with fractal spatial patterns, have been sometimes considered as the ``signature'' of complexity in different systems. The underlying regularities in the urban spatial dynamics that these models capture have attracted many scientists from various disciplines to propose different theories of urban complexity.

\subsection{Which acceptation of ``theory'' for complex urban systems?}

A preprint posted on the Internet in January 2020 \citep{lobo2020urban} deserves particular attention at first sight for three reasons. First, it announces an ``integrated theory''. Second, it covers the full historical period ``from the first cities to sustainable metropolises''. Third, 35 names of prominent scholars signed it, which is rather rare among papers dealing with objects of social sciences.

On which propositions do they come all in agreement? A first one is about the necessary multidisciplinary character of ``urban science'', which we fully agree with as well. A second proposal is about acknowledging the fact that cities are the product of a long but relatively recent and socially driven historical evolution, which should lead to the conclusion that urban theories are part of social sciences dealing with historical objects. However, the paper acknowledges, ``a basic tension in current urban science between approaches from the social sciences and those from the natural sciences'' (p.3). The authors conclude nevertheless: ``the time is ripe for a targeted integration of the social and physical sciences of cities and urbanism'' (p.4).

The author’s argumentation about this point becomes ambiguous and even contradictory. When they enunciate what the urban theory should be, they suggest Darwinian biology as a model of science to imitate in focusing on processes rather than forms, which we can agree. But they add: ``A theoretically grounded science of cities should capture fundamental processes that lie at the core of all human spatial agglomerations, whether these are past or present, agrarian or industrial, in developed economies or developing countries'' (p. 11). They clearly tend to think of universal ``fundamental principles'' that would appear as the core of urban science, while the categories that were elaborated and coined by history or other disciplines of social sciences would be considered as part of a “context”. Should we not require that a “historically grounded” urban science would have processes from societal theories as ``fundamental principles''? In other words, the processes leading to the elaboration of concepts such as ``agrarian'' or ``industrial'' society or ``developing country'' may receive at least the same level of attention that the one of ``agglomeration'' if we want to understand urban evolution and properly integrate the disciplinary points of view.

That is why we are not sure that the ``scaling analytic framework'' is indeed the first thing to mention of what we have learned from ``urban science injected into the study of cities and urbanization'' – even if that may appear paradoxical speaking from a geographer’s point of view, always caring about scale! We can nevertheless subscribe to most of the authors’s 13 propositions enabling them to converge toward ``unified perspectives'' –although a few would deserve more discussion. As the authors, we are convinced that there is a need for urban theory and that is it necessarily grounded in history, properly tested with abundant and diverse data compared in space and time. Of course, we agree to consider cities as places of high density, frequent social interaction and mixed activities, knowing that their size is both a cause and consequence of their creativity, acknowledging the universal character of urban hierarchies and the specific role of innovations in their development. However, we could question the list of nine propositions that the authors consider as new insights provided by the new ``urban science''. The list of 23 questions to solve in future research is not so new either. We acknowledge that the sciences of complex systems that developed during the second half of 20th century can be granted for having challenged social sciences toward more interest in mathematical and computational formalisms and having provided a series of formalized tools and models and a common vocabulary that established a bridge between researchers in mathematics and physics and social sciences. Nevertheless, the urban theory cannot be reduced to its quantifiable dimensions only.

We may wonder if any ``integrated theory'' is yet imaginable because at a given moment in science, and in social sciences especially, there are a plurality of theories. Social sciences may deserve a specific definition of complexity that would depend on the number of different disciplines that are required for providing a satisfying explanation or interpretation of a particular object, by borrowing concepts and models from these disciplines, each having investigated a category of complex processes, to make the object intelligible at a particular granularity level of description. Thus it is likely that while we pretend currently to build an evolutionary theory of urban systems, through integrating as much as possible from accumulated empirical knowledge in geographical, archeological and historical tradition, using statistical and simulation models from the new ``urban science'', other fields are not yet satisfied with a theory of urban hierarchy, such as \cite{krugman1996confronting} for whom it remains a ``mystery'' that cannot be properly derived from the principles of economic theory only.

\subsection{Minimum requirement for a theory of urban complex systems}

When adopting a nomothetic attitude in urban research, we adopt a series of epistemological concepts and practices that are common with other sciences. The starting point is to work with \textit{empirical data that are properly defined before being properly measured}. Taking care for meaningful definitions of cities is not a trivial exercise, because it supposes to enter and to understand the criteria that societies chose for establishing a distinction between rural and urban settlements \citep{rozenblat2020extending}. It is obvious for any social scientist that such a distinction has a political origin, even if many state governments in the last two centuries decided to objectivize as much as possible their definition through elaborating statistics. For centuries being ``urban'' was part of a social status, an attribute conveyed to people rather than to places. Most social scientists are uneasy with attempts at qualifying the urban areas delineated from satellite images of built-up areas or street networks as ``natural cities'' \citep{jiang2015zipf}. Of course, cities belong to ``nature'' in using material resources and energy and hosting humans. They may even become greener if they succeed in managing the next ecological transition... However, social organizations are driving cities' generative processes, rather than natural features. There are traces of the history of cities as political constructions in the social representations opposing peasants and urban citizens, with sometimes practical and economic consequences on their way of life, if you think for instance of the former Chinese hukou system \citep{wu2020emerging}. There are traces as well in the decisions of naming ``urban'' different places such as India distinguishing ``Statutory towns'' and ``Census towns'' \citep{swerts2018diffuse}. Some countries periodically delineate the perimeters of urban areas to follow the spatial expansion of their cities, such as Germany or Russia, whereas others do not.

This did not prevent too many experimented and incidentally prominent scholars to capture “urban data” without any care about their quality, selection, claiming and striking results after crunching them with sophisticated statistical or mathematical models. Such an attitude is detrimental, for instance regarding the endless controversies about the models of city sizes distribution, for building a sound science that accumulates knowledge with reproducible means \citep{pumain2012theorie}. The resulting cacophony is sometimes disentangled in courageous articles, such as for instance \citep{cottineau2017metazipf,cottineau2020metametazipf} in a meta-review of hundreds of references, but certainly represents a waste of time and efforts that could be avoided. A recent paper seemed promising in announcing the observation of worldwide trends in urbanization from 1950 to 2030 using some 1857 cities larger than 300 000 inhabitants from 155 countries \citep{egidi2020long}. The paper offers surprising results such as a systematic ``inverse relationship'' between urban growth and city size in all regions ``up to the late 1990s''. At first sight, this could result from mixing countries with opposite urban growth trends, but the authors observe it as well in homogenous sub regions. However, after questioning the authors who kindly accepted to precise the origin of their data, it seems that the selected cities were delineated over time in a way not measuring the spatial expansion of the urban area. This could explain why the results seem in contradiction with those obtained from large parts of the world during the same period with harmonized urban databases \citep{cura2017old}. It also reminds about the necessity of starting with a sound conceptual definition of cities for measuring their attributes and controlling the possible effects of the data selection on the results. The urban definition may vary but should be in accordance with concepts and theory, i.e. with the research question and the selection of attributes that are measured for answering them \citep{rozenblat2020extending}.

To specify the fundamentals of our theory we retain a conceptual framework that relies on the long history of social sciences investigating cities and interpreting them \citep{pumain1996theoriser}. Thus even if we can integrate insights from complex systems science such as fractals and scaling laws, we do not agree that all fundamentals should stem from ``physics for society'' \citep{caldarelli2018physics}. While maintaining a strong attention to the spatial dimension, our concern with space is neither static nor geometric. It is geographical, and geo-historical. Space is crucial in the definition of cities and systems of cities, it is a social space whose properties are revised and updated with the changing technologies, and it is as well made of transformed natural space in the process of using earth resources and developing political territories. The dimensionality of cities is constrained by the accessible space for face-to-face daily activities, whereas transportation networks enabling less frequent encounters may generate and sustain interdependencies at much longer distances leading to identify systems of cities. At each observation level, those of individuals (persons or institutions), cities and systems of cities, new attributes and new concepts do correspond to emergent phenomena in complex systems that are produced from multiple interactions.

Although the recent development of networks and the accelerating circulation of information seem to blur the three-level conceptual definition of these geographical objects, we believe that this representation remains useful and relevant for guiding urban research and applications. An interesting contribution by C. \cite{roth2006reconstruction} discusses the emergentist view compared to the reductionist and suggests that different modes of access and not only bottom- up interactions could be part of the generative processes of the features constituting one level – which is the view point from statistical mechanics in envisaging self-organizing systems. This rejoins our theory for which urban interactions shaping cities and systems of cities are mostly but not exclusively conceived as emanating from one low level toward the higher level, in societal processes that take a long time, often longer than a human life. Over decades and sometimes centuries, most frequent interactions among urban stakeholders at micro level shape the emerging properties of a city at meso-geographical level, and inter-urban interactions generate the hierarchical organization and the functional differentiation of cities within the system of cities at macro-level. However, for instance, a location decision taken by an individual firm (at micro-level) can operate a definite change in the relative specialization of a particular city within the system of cities (at macro level), as well as many policy regulations decided at macro-level may directly change the situation of an individual firm or citizen at micro-level inside a city.

Sociologists such as Michel \cite{grossetti2020matiere} discuss the ontology of entities that scholars identify in social sciences. A science of cities is possible because cities are remarkably persistent entities, among all societal institutions entities of the meso level, in an intermediary position between individual agents and the territories that embed them. Through their interactions however, they build at a much higher level, comparable to the one of civilizations, spatial organizations that we name systems of cities and maintain their ``emerging'' properties over much longer durations. The rhythm of relative societal change within cities is much slower than the changes in human life and change in systems of cities is much slower than change in individual cities. Our theory of cities as complex social systems embedded within systems of cities and territories is \emph{evolutionary}. However, the historical evolution is different from biological, of course because its processes are social, meaning they are guided with human intentions (even if that causality is not always directly effective) and they proceed much more rapidly than changes in biology.

While trying to identify major processes guiding the evolution of cities within system of cities, we stressed the role of information and innovation \citep{lane2009complexity}. Innovation when defined as a socially accepted invention may represent a form of social organization or a belief or a practice as well as a technological device or a new service. We shall develop further in the models presented below how innovation and its more or less always hierarchical diffusion among cities within systems of cities appears as a major dynamic process for explaining the expansion of urbanization and the persistency of cities’ relative situation in urban systems.

\subsection{Defining integrated theories}

An issue worth mentioning before developing further the relation between theories and models is the definition of a theory itself, and furthermore of an ``integrated theory''. \citep{lobo2020urban} use this term several times without detailing its exact meaning - to what extent ambiguity can be harmful in this aspect? The nature, structure and function of scientific theories has always been one core subject in philosophy of science. According to \cite{winther2016structure}, three main views are currently accepted for the structure of scientific theories: (i) the syntactic view which sees theories as mathematical logical constructions; (ii) the semantic view, which considers a set-theoretic approach and understands theories as set of models, the theory being itself a model; and (iii) the pragmatic view, which includes more complex ways of knowledge production and components, including informal or implicit knowledge such as analogies, but also taking into account how knowledge is socially produced and organized. Following \cite{suppe2000understanding}, the semantic view was proposed to overcome limitations of the syntactic view (such as the fact that theories are not axiomatic systems). It was suggested by \citep{halvorson2012scientific} that the semantic view is also limited to be able to grasp the complexity of scientific knowledge production. One stream of ideas within the pragmatic view is the pragmatic view of models, focusing on the diversity and functions of models. In that context, the epistemological positioning of scientific perspectivism \citep{giere2010scientific} advocates for a plurality of scientific perspectives, of which models are media. According to \citep{callebaut2012scientific}, this approach is highly relevant to deal with complex and multi-scalar problems. In the context of urban theories, \citep{pumain2020conclusion} showed a plurality of theories and models in practice. We therefore suggest that a pragmatic view of theory is the most relevant to build integrated theories, which will have to couple multiple heterogeneous components. This aspect also comes into play when considering tools or methods used to support the theory. Although it is not clearly stated, it seemed implied by \citep{lobo2020urban} that an integrated theory will have to be ``rigorous'' and formulated mainly with mathematics. While theories in physics almost necessarily include mathematical formalism and derivations, this is not the case in social science, and knowledge can be communicated and formalised through other ways, as recalled by \cite{banos2013pour} as one of the crucial principles for modeling and simulation in social science.

Finally, the meaning of ``integrated'' is a rather open question. There are many terms to distinguish between different degrees of interaction between disciplines, from multidisciplinarity (combining approaches from different disciplines) to actual interdisciplinarity which implies that new knowledge is created beyond the disciplines \citep{huutoniemi2010analyzing}. There is also no clear definition of model coupling in the literature, nor of theory coupling. Model coupling can range from a ``loose coupling'' where outputs of one model are used as inputs of another \citep{clarke1998loose}, to stronger couplings where both model dynamics influence each other in time or even where additional coupling mechanisms are constructed and implemented, leading to new hybrid models \citep{mustafa2017coupling}. Furthermore, there may exist different types of integration: in the roadmap for the study of complex systems, \citep{chavalarias2009french} propose a two-dimensional reading of integration, through the construction of integrated disciplines spanning across one field at multiple scales (vertical integration), and the investigation of transversal fundamental questions that are pervasive to different types of complex systems (horizontal integration).

We postulate here that ``integration'' refers to such strength of model and theory coupling, but the question of its proper definition and quantification remains open. We will in the following build on possible interactions between models and theories, leading to the position that coupling models is a possible way to build integrated theories.

\subsection{How to make models and theory dialogue?}

If the theory is constructed from empirical observations, prioritized and simplified into stylized facts, the design of the model should enable to reproduce these stylized facts. Dynamic models start from an initial observed situation that is described in a simplified way by state variables (such as the population of cities, their income level or their employment structure). These models reconstruct the evolution of these quantities on the basis of mechanisms (rules and parameters), which represent the interactions (i.e. social processes) that are supposed to explain the transformations of the state variables over time \citep{pumain2013theoretical}.

For a model to be an interesting tool for testing a theory, a correspondence must be established between the way it is implemented in the model and the expected result of the simulation. Selecting the stylized facts to be reconstructed and the granularity of their description is a crucial choice for the success of the exercise. For example, it is rather pointless to give the model the sole objective of correctly simulating a distribution of city sizes, in the form of a lognormal or Pareto law. This form of distribution (described as an 'over-identified model' by B. \cite{robson1973urban}) can indeed emerge from a whole series of mathematical or computer mechanisms, some of which have little to do with the functioning of a system of cities. Without aiming for absolute realism but in order to get closer to it and make better use of modelling, it is necessary, for example, to go one-step further and ensure that the growth process produced by the model corresponds in detail to what is summarized in \cite{gibrat1931inegalites}'s statistical model \citep{modica2017methodological}.

All models described below include the well-established fact that the urban hierarchy receives a relevant statistical explanation from Gibrat's model of urban growth. However, this random mathematical model has at its core a hypothesis that makes it little credible to found a theory of cities, since it assumes that all cities (elements of the statistical distribution) are independent of each other, as well as the random events that lead to their growth. Such a hypothesis enters in conflict with the definition, the functioning of cities and the processes of their emergence and their evolution, which organizes them into systems of co-evolution, counteracting these simplifying assumptions a priori. In fact, the slight deviations noted in relation to the hypotheses of the model have led to the completion of other hypotheses that adapt this simple model, powerful but too general, to different historical and geographical contexts.

In the Simpop family of models, the reconstruction of cities' trajectories is based on their interactions using the multi-agents systems as computing technique for simulation. This technique was chosen because, when compared to mathematical systems of differential equations, they enable more flexible representation of the variety of spatial interactions that characterize inter-urban trade according to the functional specialization of cities. In brief, the so-called ``central functions'' providing services to the population interact with their market according to gravitational principles, while manufacturing and touristic cities select sometimes distant but specific places to interact with, and administrative functions deal in a systematic way with all other cities in their circumscription. More detailed precisions can be injected in the simulation, as for instance the political selection of some urban places for developing industries of national interest during the socialist regime in the former Soviet Union \citep{cottineau2014evolution}.

So, logically the theory is as much as possible embedded in the model, but how can the model contribute to the building of the theory? Of course, the usual recursive processes during model building may ensure that the major theoretical principles are correctly framed in the implemented model. Nevertheless, until recently there was too little confidence in the results obtained from simulation models when using multi-agents modelling techniques \citep{rey2015plateforme}. The validation procedures were considered as not sufficiently reliable for guaranteeing that the estimated parameters were the only values capable of generating the required features of the system’s evolution and that all the included rules were necessary to produce the simulated result. A real scientific breakthrough for SSH occurred when the combined use of genetic algorithms and distributed computing enabled to shift from the usual few hundred simulations of the same model toward several hundred millions \citep{schmitt2015half}. This experiment not only provides an almost full exploration of the parameter space but also ensures that all parameters leading to the expected result are both sufficient and necessary \citep{reuillon2015new,raimbault2019methods}. Although large-scale computing has always been anchored within quantitative geography practices \citep{rey2019calcul}, these results were particularly novel in the way models, methods and theory were intertwined.

This process of an iterative construction of theories and models corresponds to the latest developments of an ``abductive'' epistemology, which succeeded historically to deductive approaches in classical geography and purely inductive approaches \citep{varenne2018theories}. In that context, the relation between theories and models can be understood through the lens of knowledge domains \citep{raimbault2017applied}: scientific knowledge production is achieved through perspectives which rely fundamentally on the model as their main medium, but are also composed of other types of knowledge: theoretical (defined in this case in the pragmatic way), empirical (facts derived from data), data itself, methods, and tools implementing methods. This typology may not be exhausted but already captures the strong interaction between different components of knowledge, and how the progressive iteration between their contents, including models and theories, is central. We see here that a third dimension of integration could be proposed, as the integration between knowledge domains, accounting for how strong are the dependencies between models and theories for example. \citep{raimbault2020relating} has proposed a similar view as the core of an ``applied perspectivism'', an epistemological stance aimed at expliciting disciplinary positions through reflexivity and enhancing interdisciplinarity. In that context, a way forward to couple theories and build integrated theories would be through the coupling of models themselves. In that context, model validation methods are key, and these can be concretely diffused into diverse disciplines with interdisciplinary and reflexive knowledge dissemination practices \citep{leclaire2019retour}. The approach of applied perspectivism is based on a strong interaction between models and theories, and as we put it above, on the use of new model exploration and validation methods.

\section{Simulation models for systems of cities}

Following what we just developed, simulation models and their systematic exploration seem to be a powerful way to (re-)build urban theories. We propose now to detail the structure of a few simulation models for systems of cities, which were all introduced within the frame of the evolutionary urban theory. We describe in particular the processes taken into account by each model, and what the corresponding exploration results may imply for the theory.

The different models we describe below were benchmarked and applied to several systems of cities worldwide by \citep{raimbault2020empowering}. They rely on the same fundamental assumptions and share a common basic structure and formulation: (i) agents are cities, characterized a main state variable which is population; (ii) building on the Gibrat model, these add to endogenous growth additional processes which account for space and spatial interactions; (iii) they simulate not stochastic distributions of city sizes, but their average in time, and are thus specified on these averages only. The interaction between cities, which could be included in the covariance structure for a fully stochastic model, is in that case captured by the spatial interaction terms.

\subsection{A model developing urban hierarchy with transportation network}

One dimension of urban systems which is expected to significantly influence system trajectories is physical infrastructure, and more particularly transportation infrastructure \citep{dupuy1987vers}. This aspect is included in a model of spatial interaction taking into account physical transportation networks, studied by \citep{raimbault2020indirect}. It was extended into a co- evolution model between cities and abstract transportation networks by \citep{raimbault2020modeling}, and refined into a co-evolution model taking into account an explicit spatial structure of the network by \citep{raimbault2020hierarchy}. The population of cities $P_i$ are updated at each time $t$ following

\begin{equation}
	P_i (t + \Delta t) = P_i(t) \cdot \left[ 1 + \Delta t \cdot \left( r_0 + \hat{w_G} \cdot \sum_{i \neq j} \left( \frac{P_i(t)P_j(t)}{P(t)^2}\right)^{\gamma_G}  \cdot \exp \left( - \frac{d_{ij}}{d_G} \right)\right) \right]	
\end{equation}

with $\Delta t$ time step, $r_0$ fixed endogenous growth rate, $\hat{w_G}$ relative share of spatial interactions (the parameter is in practice renormalized by the average of interaction potentials), $P(t)$ total population, $\gamma_G$ hierarchy of spatial interactions, $d_G$ range of spatial interactions, and $d_{ij}$ geographical or network distance (yielding two different versions of the model). The initial model of (Raimbault, 2018b) includes an additional term accounting for centrality effects in the network, with feedback processes of flows going through a city on its population growth rate. This model is completed by a network evolution module to give the co-evolution model: network link effective speeds are updated at each time by

\begin{equation}
	d_l(t+ \Delta t) = d_l(t) \cdot \left[ 1 + \Delta t \cdot g_M \left( \frac{1 - \left(\frac{\varphi_l}{\varphi_0}\right)^{\gamma_N}}{1 + \left(\frac{\varphi_l}{\varphi_0}\right)^{\gamma_N}} \right) \right]	
\end{equation}

where $d_l(t)$ is the length of link $l$, $g_M$ a maximal speed increase parameter, $\varphi_l$ the flow within the link, $\varphi_0$ the threshold parameter (above which link speed increase) determined in practice through a quantile parameter $\varphi_0^{(q)}$, and $\gamma_N$ the hierarchy of the process. Flows within a physical network can be determined by any network assignment algorithm, given the spatial interaction flows obtained at the previous step.

The complete model workflow is synthesized in Fig.~\ref{fig:fig1}. This summarizes the superposition of the different processes linked to cities and the transportation network. The exploration and application of this model can bring new insights for theory building, including for example: (i) do network-mediated spatial interactions explain well urban trajectories? - it was shown by \citep{raimbault2020indirect} that in terms of fitting error for the French system of cities, the feedback of network flows actually improved model performance, implying that these network processes are good candidates as explanatory variables for urban dynamics; (ii) can co- evolution in the statistical and population sense be effectively reproduced between cities and transportation infrastructures? - \citep{raimbault2020modeling} exhibits a large variety of co-evolution regimes produced by the model; (iii) what are the common dynamics of urban and network hierarchies? - \citep{raimbault2020hierarchy} investigated this question and showed much richer dynamics than a simple self-reinforcement. These results suggest that next theoretical developments should not neglect infrastructure networks, the role of co-evolution, and to rethink concepts of hierarchy in systems of cities.

\begin{figure}
	\includegraphics[width=\linewidth]{Fig1.jpg}
	\caption{Schematic representation of the different processes included in the macroscopic co- evolution model for cities and transportation networks. Cities are characterized by their population (circle area) and are linked by a transportation network (dashed lines). City population grows endogenously (Gibrat model, first panel), through spatial interactions (second panel) and feedback of flows (third panel). Network links evolve according to the flow they carry (fourth panel).\label{fig:fig1}}	
\end{figure}

\subsection{A model explaining urban growth with economic and environmental processes}

Economic processes are the subject of large parts of the literature studying urban dynamics, for example in urban economics, economic geography, or regional science, but not only. The MARIUS model family developed by \citep{cottineau2014evolution} is based on economic exchanges between cities, within the frame of the evolutionary urban theory. Following \citep{cottineau2015modular}, the model family is composed of a baseline model capturing spatial interactions (the Gibrat mechanism is not included in this model which can however been seen as having the same basis as other models presented here, where the endogenous growth mechanism has been deactivated), and incremental mechanisms which are a growth bonus for cities with more interactions, a fixed cost of transactions, geographical effects such as the local presence of resources and policies with a territorial impact, and the possibility to modulate urbanisation speed with external temporal constraints. We describe here only the baseline model, reformulated by replacing the power spatial interaction function with an exponential, for the sake of simplicity. Given population of cities at the initial step $P_i(0)$, an abstract wealth state variables is constructing following a scaling law by taking $W_i(0)=P_i^{\alpha_W}$, where $\alpha_W$ is a model parameter. At each time step, supply and demand functions are computed as scaling laws of the current population for each city, potential flows are computed with spatial interactions, balance of exchanges and wealths are updated accordingly. This gives the following non-linear spatial interaction model to update city wealth:

\begin{multline}
	W_i (t + \Delta t) = W_i (t) + w_0 \cdot \sum_j \exp \left( - \frac{d_{ij}}{d_G} \right)	\cdot \left( \min \left[ \frac{P_i^{\alpha_S} P_j^{\alpha_D}}{\sum_k P_k^{\alpha_S} \exp(-d_{ij}/d_G)} ; \frac{P_i^{\alpha_D} P_j^{\alpha_S}}{\sum_k P_k^{\alpha_D} \exp(-d_{ij}/d_G)} \right]\right. \\ \left. - \min \left[ \frac{P_i^{\alpha_D} P_j^{\alpha_S}}{\sum_k P_k^{\alpha_S} \exp(-d_{ij}/d_G)} ; \frac{P_i^{\alpha_S} P_j^{\alpha_D}}{\sum_k P_k^{\alpha_D} \exp(-d_{ij}/d_G)} \right] \right)
\end{multline}

where $w_0$ is a normalization parameter, $d_G$ the spatial interaction range, $\alpha_S$ the supply scaling parameter, and $\alpha_D$ the demand supply scaling parameter. The population increment is then computed as a function of the scaled wealth increment, as

\begin{equation}
	P_i (t+\Delta t) = P_i (t) + \frac{1}{w_0} \left( W_i (t+\Delta t)^{\alpha_P} - W_i (t)^{\alpha_P} \right)	
\end{equation}

where $\alpha_P$ is the population scaling parameter. More detailed model explanation, exploration, and application to the Former Soviet Union can be found in \citep{cottineau2014evolution,cottineau2015growing,cottineau2015modular}.

Some insights for theory from this model family are for example detailed by \citep{cottineau2015modular}: (i) which mechanisms among the different economic processes included are better at explaining urban trajectories, when taking into account parsimony, i.e. considering only models with the baseline and one other mechanism? - different optimal model structure were found depending on the calibration structure, suggesting a strong break in stationarity due to the fall the Soviet Union; applying a similar approach to other systems of cities remains difficult as the model was tailored to that case, but would bring significant insight into the question of non-stationarity of urban trajectories; (ii) when considering all processes, which are the most important to explain city trajectories? - a surrogate statistical model applied on values of the calibration error as a function of parameters, for an optimal model population composed of a fixed number of individual for each model instance, gives the relative contribution of each mechanism to the fit improvement; (iii) what are the properties of cities on which the model does not perform well? - this provides feedback on the specific theoretical assumptions and mechanism survey done during the construction of the model family for this particular case study.

\subsection{A model linking urban growth with innovation as spatial diffusion process}

\subsubsection{The Favaro-Pumain model}

The spatial diffusion of innovations has been coined as a fundamental process driving urban dynamics \citep{hagerstrand1968innovation}. In economics, many approaches consider innovation as crucial, such as in endogenous growth theory \citep{aghion1998endogenous}. The theoretical and empirical study of innovation clusters and knowledge spillovers is also a considerable stream in economic literature \citep{audretsch2005does}. In the context of the evolutionary urban theory, the SimpopLocal model \citep{schmitt2014modelisation} proposes to link the diffusion of innovation with the emergence of cities in an agent-based approach. Based on similar ideas, but within a discrete dynamical system formulation which inspired the formulation of other models we gave here,  \cite{favaro2011gibrat} introduced an extension of the Gibrat model considering spatial interactions for the dynamics of populations and for the diffusion of innovations. We describe here the model as specified by \citep{raimbault2020model}. In this model, cities are characterized by their population $P_i(t)$ and by a level of adoption of a set of innovations $\delta_{c,i,t}$ corresponding to proportions such that $\sum_c \delta_{c,i,t} = 1$. Populations are updated following a spatial interaction model, where attractivity of cities is modulated by the local level of adoption relative to the global level, by

\begin{equation}
	P_i (t+\Delta t) = P_i(t) \cdot \left[1 + \Delta t \cdot \left(r_0 + \hat{w_I} \cdot \sum_{i\neq j} \frac{P_i(t) P_j(t)}{P(t)^2} \cdot \exp \left(- \frac{d_{ij}}{d_G} \right) \cdot \prod_c \delta_{c,i,t}^{\varphi_{c,t}} \right) \right]
\end{equation}

The parameter $\hat{w_I}$ is in practice normalized by the average of potentials. The global adoption level is defined by $\varphi_{c,t} = \sum_i \delta_{i,c,t} P_i / \sum_{i,c} \delta$. Parameters $r_0$ and $d_G$ are as previously the endogenous growth rate and spatial interaction range. Innovations are diffused between cities also with a spatial interaction model, according to

\begin{equation}
	\delta_{c,i,t+\Delta t} = \frac{\sum_j p_{c,j,t}^{1/u_c}\cdot \exp\left(-d_{ij}/d_I\right)}{\sum_c\sum_j p_{c,j,t}^{1/u_c}\cdot \exp\left(-d_{ij}/d_I\right)}	
\end{equation}

where $p_{c,i,t} = \delta_{c,i,t}\cdot \frac{P_i}{P}$ is the population share in city $i$ adopting innovation $c$, and $u_c$ is the innovation utility (innovations with a higher utility will diffuse faster). The last mechanism in a model step corresponds to the introduction of new innovations - we describe it in the next subsection which corresponds to a modified version of the model tailored to capture urban evolution.

\subsubsection{An explicit model of urban evolution}

The initial Favaro-Pumain model described by \citep{favaro2011gibrat} is an urban dynamics model, but not an urban evolution model in a sense that would extend biological and cultural evolution \citep{mesoudi2017pursuing}. To have evolution in a population, one must have fundamental processes of transmission and transformation, leading to differentiating sub- populations. Elementary units of transmission are genes in biology, while the concept of meme was introduced in cultural evolution. In what would consist elementary units and processes of an urban evolution in that sense remains an open question, but \citep{raimbault2020model} proposed a possible instantiation, extending the Favaro-Pumain model. Elementary units are innovation, and the urban genome corresponds to the innovation proportions $\delta_{c,i,t}$. Transmission processes correspond to the spatial diffusion of innovations and the evolution of city sizes due to spatial interactions, as a kind of spatial crossover. Transformation processes are achieved through local mutations of the genome, corresponding to random discoveries of new innovations. In the initial model, the stylized fact of creation-destruction was implemented by deterministically inserting a new innovation when a certain adoption threshold was attained, and innovation utility was multiplied by a fixed factor. In the urban evolution model, at each time step each city has a probability to innovate given by

\begin{equation}
	p = \beta \cdot \left( \frac{P_i(t)}{\max_k P_k(t)} \right)^{\alpha_I}	
\end{equation}

where $\beta$ is an intrinsic innovativity parameter, and the probability to innovate is proportional to a scaling function of size with exponent $\alpha_I$. The utility of a new innovation is drawn stochastically from a distribution with average the current empirical mean of existing innovation and with standard deviation a fixed parameter $\sigma_U$. The new innovation has a penetration share of $\delta_0$ in the new city, which genome is modified by rescaling the share of other innovations. A schematic summary of the innovation diffusion urban evolution model is shown in Fig.~\ref{fig:fig2}.

\begin{figure}
	\includegraphics[width=\linewidth]{Fig2.jpg}	
	\caption{Schematic representation of the different processes included in the innovation diffusion urban evolution model. Cities are characterized by their population (circle area) and their innovation shares or genome (color boxes). Innovations are diffused between cities with a spatial interaction model (second panel). City sizes are updated according to spatial interactions and attractivity linked to the share of the dominant innovation (third panel). Genomes are mutated with possibly the introduction of new innovations (fourth panel).\label{fig:fig2}}
\end{figure}

Insights for the construction of theory that can be drawn from this model are diverse. First, with the Favaro-Pumain model only, one can investigate to what extent hierarchical diffusion of innovation occurs and successive waves influence urban growth. When considering the urban evolution model, the possibility for the model to lead to the emergence of effective co- evolution niches (i.e. regions of space diverging significantly in terms of genome composition in time) is an important feature for the role of co-evolution in the urban theory. Results of the exploration by \citep{raimbault2020model} were preliminary and did not investigate specific indicators quantifying co-evolution, but found intermediate spatial interaction ranges leading to structures similar to innovation clusters. Finally, perspectives to parametrize this model with empirical innovation data (using patents as proxy for example) can be linked to feedback on the effective role of innovation diffusion in urban dynamics.

\section{Discussion}

We have previously discussed the importance of theory (whatever its meaning) in urban studies, and its relation to modeling. In particular, we recalled how models and theories can dialogue to strengthen conclusions on both sides, and advance towards the construction of integrated theories. We finally illustrated with concrete urban simulation models diverse ways theory can be instantiated into simulation models and how the exploration, validation and application of these models can inform back the theory. Many obstacles still remain in that context, and we discuss now two particular issues that in our opinion will require a particular attention to reach effectively integrated theories.

\subsection{Multi-modeling and concurrent hypotheses}

One significant difficulty, particularly linked to the integration of complementary/concurrent hypotheses or theories, is how to properly compare them in terms of relevance to the integrated knowledge built. To tackle it, we discuss above how \cite{cottineau2015modular} followed the procedure of multi-modeling, in the sense of a full model in which incremental mechanisms can be included or not. Combining specific technical tools such as automatic code generation and methodological tools with a dedicated niched genetic algorithm to calibrate a large set of these models, they were able to produce a population of optimal solution and evaluate statistically the contribution of diverse mechanisms to model fit, yielding therein a benchmark of mechanisms in terms of explanatory power. This approach is however possible only when all models can be integrated into a single framework and directly compared. In the case of \citep{raimbault2020empowering}, which benchmarked simple versions of the models we described above, such automatic generation and niched calibration could not be achieved, and models were independently calibrated on two objectives and the final Pareto fronts compared. This was still possible as the models have the same input and output data and a similar structure. How to couple and integrate models when they are highly heterogeneous, at different scales and on totally different objects? How to build multi-models of heterogeneous models, ideally automatically towards the idea of ``model crushing'' coined by \citep{openshaw1993modelling}? These questions remain rather open, although initiatives such as the model exploration platform OpenMOLE are built towards the facilitation of tackling such issues \citep{raimbault2019methods}. Another methodological difficulty in model comparison that should not be neglected is the fair comparison of models, i.e. taking into account overfitting and keeping parsimonious models. Generic information criteria for simulation models that would render a fair comparison possible as Akaike Information Criterion does for statistical models, also remains an open problem \citep{raimbault2020indirect}. These questions directly relate to the nature of integration, of model coupling, and the construction of multi-scale models.

\subsection{Between synthetic and real systems of cities}

One classical mantra in French geography is the tension between ``the general and the particular'' \citep{durand1991particulier}. This opposition can directly be found at the core of theories for systems of cities: what are the general and robust stylized facts that can be attributed to systems of cities in general, and what are the peculiar geo-historical conditions that lead to a particular trajectory. Beyond the myth of ``universality'' advocated by physicists which would imply universal laws derived from the same microscopic processes and which would assume some kind of ergodicity in the system \citep{pumain2012urban}, robust stylized facts such as Zipf and Gibrat laws are well identified in a first approximation. In that context and within the interaction between models and theories, being able to distinguish between intrinsic model dynamics, robust effects due to space that one observe on similar geographical structure, and effects due to the peculiar initial conditions and geo-historical conditions, is a difficult problem. \citep{raimbault2019space} introduced a novel methodological pipeline for sensitivity analysis of spatial models, which enables to test the sensitivity of model outcomes to spatial initial conditions. This requires generating spatial synthetic data that resembles real configurations, in this case population grids at the scale of urban areas. \citep{raimbault2020scala} develop generators at other scales (microscopic building configurations, macroscopic systems of cities, transportation networks). Such work is the basis towards a better understanding on how urban dynamics models behave on synthetic systems of cities compared to real case studies. This is essential for the construction of integrated theories, to understand their level of generality, and how they relate to specific geographical situations.

\section{Conclusion}

We have recalled here the meaning and importance of theories to study complex urban systems, and how a diversity of theories is needed to grasp the multi-dimensional nature of such systems. This diversity can be the basis of coupling between theories, and the construction of integrated theories. What such integration consists of and how to reach it remains in our opinion an open question, which is nevertheless partly answered by the use of simulation models, their coupling and a dynamical dialogue between models and theories. We illustrated these concepts with three urban dynamics models developed within the frame of an evolutionary theory of cities. Many obstacles remain on the path towards an evidence-based integrated knowledge of complex urban systems, but we believe elementary bricks already exist and methodological ways forward are progressively developed.

\begin{thebibliography}{}

\bibitem[Aghion et~al., 1998]{aghion1998endogenous}
Aghion, P., Ljungqvist, L., Howitt, P., Howitt, P.~W., Brant-Collett, M.,
  Garc{\'\i}a-Pe{\~n}alosa, C., et~al. (1998).
\newblock {\em Endogenous growth theory}.
\newblock MIT press.

\bibitem[Audretsch and Lehmann, 2005]{audretsch2005does}
Audretsch, D.~B. and Lehmann, E.~E. (2005).
\newblock Does the knowledge spillover theory of entrepreneurship hold for
  regions?
\newblock {\em Research policy}, 34(8):1191--1202.

\bibitem[Banos, 2013]{banos2013pour}
Banos, A. (2013).
\newblock {\em Pour des pratiques de mod{\'e}lisation et de simulation
  lib{\'e}r{\'e}es en g{\'e}ographie et SHS}.
\newblock PhD thesis.

\bibitem[Batty, 2013]{batty2013new}
Batty, M. (2013).
\newblock {\em The new science of cities}.
\newblock MIT press.

\bibitem[Brenner and Schmid, 2014]{brenner2014urban}
Brenner, N. and Schmid, C. (2014).
\newblock The ‘urban age’in question.
\newblock {\em International journal of urban and regional research},
  38(3):731--755.

\bibitem[Caldarelli et~al., 2018]{caldarelli2018physics}
Caldarelli, G., Wolf, S., and Moreno, Y. (2018).
\newblock Physics of humans, physics for society.
\newblock {\em Nature Physics}, 14(9):870--870.

\bibitem[Callebaut, 2012]{callebaut2012scientific}
Callebaut, W. (2012).
\newblock Scientific perspectivism: A philosopher of science’s response to
  the challenge of big data biology.
\newblock {\em Studies in History and Philosophy of Science Part C: Studies in
  History and Philosophy of Biological and Biomedical Sciences}, 43(1):69--80.

\bibitem[Chavalarias et~al., 2009]{chavalarias2009french}
Chavalarias, D., Bourgine, P., Perrier, E., Amblard, F., Arlabosse, F., Auger,
  P., Baillon, J.-B., Barreteau, O., Baudot, P., Bouchaud, E., et~al. (2009).
\newblock French roadmap for complex systems 2008-2009.

\bibitem[Clarke and Gaydos, 1998]{clarke1998loose}
Clarke, K.~C. and Gaydos, L.~J. (1998).
\newblock Loose-coupling a cellular automaton model and gis: long-term urban
  growth prediction for san francisco and washington/baltimore.
\newblock {\em International journal of geographical information science},
  12(7):699--714.

\bibitem[Cottineau, 2014]{cottineau2014evolution}
Cottineau, C. (2014).
\newblock {\em L'{\'e}volution des villes dans l'espace post-sovi{\'e}tique.
  Observation et mod{\'e}lisations.}
\newblock PhD thesis, Universit{\'e} Paris 1 Panth{\'e}on-Sorbonne.

\bibitem[Cottineau, 2017]{cottineau2017metazipf}
Cottineau, C. (2017).
\newblock Metazipf. a dynamic meta-analysis of city size distributions.
\newblock {\em PloS one}, 12(8):e0183919.

\bibitem[Cottineau, 2020]{cottineau2020metametazipf}
Cottineau, C. (2020).
\newblock Metametazipf. what do analyses of city size distributions have in
  common?
\newblock {\em arXiv preprint arXiv:2008.11473}.

\bibitem[Cottineau et~al., 2015a]{cottineau2015growing}
Cottineau, C., Chapron, P., and Reuillon, R. (2015a).
\newblock Growing models from the bottom up. an evaluation-based incremental
  modelling method (ebimm) applied to the simulation of systems of cities.
\newblock {\em Journal of Artificial Societies and Social Simulation}, 18(4):9.

\bibitem[Cottineau et~al., 2015b]{cottineau2015modular}
Cottineau, C., Reuillon, R., Chapron, P., Rey-Coyrehourcq, S., and Pumain, D.
  (2015b).
\newblock A modular modelling framework for hypotheses testing in the
  simulation of urbanisation.
\newblock {\em Systems}, 3(4):348--377.

\bibitem[Cura et~al., 2017]{cura2017old}
Cura, R., Cottineau, C., Swerts, E., Antonio~Ignazzi, C., Bretagnolle, A.,
  Vacchiani-Marcuzzo, C., and Pumain, D. (2017).
\newblock The old and the new: qualifying city systems in the world with
  classical models and new data.
\newblock {\em Geographical Analysis}, 49(4):363--386.

\bibitem[Denis, 2020]{denis2020more}
Denis, E. (2020).
\newblock More urban constructions for whom? drivers of urban built-up
  expansion across the world from 1990 to 2015.
\newblock In {\em Theories and Models of Urbanization}, pages 235--258.
  Springer.

\bibitem[Dupuy, 1987]{dupuy1987vers}
Dupuy, G. (1987).
\newblock Vers une th{\'e}orie territoriale des r{\'e}seaux: une application au
  transport urbain.
\newblock In {\em Annales de g{\'e}ographie}, pages 658--679. JSTOR.

\bibitem[Durand-Dast{\`e}s, 1991]{durand1991particulier}
Durand-Dast{\`e}s, F. (1991).
\newblock Le particulier et le g{\'e}n{\'e}ral en g{\'e}ographie.
\newblock In {\em Actes du sixi{\`e}me colloque de didactique de l'histoire, de
  la g{\'e}ographie et des sciences sociales}, volume~6, pages 209--219.
  Institut national de recherche p{\'e}dagogique.

\bibitem[Egidi et~al., 2020]{egidi2020long}
Egidi, G., Salvati, L., and Vinci, S. (2020).
\newblock The long way to tipperary: City size and worldwide urban population
  trends, 1950-2030.
\newblock {\em Sustainable Cities and Society}, page 102148.

\bibitem[Favaro and Pumain, 2011]{favaro2011gibrat}
Favaro, J.-M. and Pumain, D. (2011).
\newblock Gibrat revisited: An urban growth model incorporating spatial
  interaction and innovation cycles.
\newblock {\em Geographical Analysis}, 43(3):261--286.

\bibitem[Gibrat, 1931]{gibrat1931inegalites}
Gibrat, R. (1931).
\newblock {\em Les In{\'e}galit{\'e}s {\'E}conomiques, Librairie du Recueil
  Sirey, Paris}.

\bibitem[Giere, 2010]{giere2010scientific}
Giere, R.~N. (2010).
\newblock {\em Scientific perspectivism}.
\newblock University of Chicago Press.

\bibitem[Grossetti, 2020]{grossetti2020matiere}
Grossetti, M. (2020).
\newblock Mati{\`e}re sociale-introduction: {\`a} la recherche d'une ontologie
  robuste.

\bibitem[Hagerstrand et~al., 1968]{hagerstrand1968innovation}
Hagerstrand, T. et~al. (1968).
\newblock {\em Innovation diffusion as a spatial process.}
\newblock Chicago, USA: Univ. Chicago Press.

\bibitem[Halvorson, 2012]{halvorson2012scientific}
Halvorson, H. (2012).
\newblock What scientific theories could not be.
\newblock {\em Philosophy of science}, 79(2):183--206.

\bibitem[Huutoniemi et~al., 2010]{huutoniemi2010analyzing}
Huutoniemi, K., Klein, J.~T., Bruun, H., and Hukkinen, J. (2010).
\newblock Analyzing interdisciplinarity: Typology and indicators.
\newblock {\em Research policy}, 39(1):79--88.

\bibitem[Jiang et~al., 2015]{jiang2015zipf}
Jiang, B., Yin, J., and Liu, Q. (2015).
\newblock Zipf’s law for all the natural cities around the world.
\newblock {\em International Journal of Geographical Information Science},
  29(3):498--522.

\bibitem[Krugman, 1996]{krugman1996confronting}
Krugman, P. (1996).
\newblock Confronting the mystery of urban hierarchy.
\newblock {\em Journal of the Japanese and International economies},
  10(4):399--418.

\bibitem[Lane et~al., 2009]{lane2009complexity}
Lane, D., Pumain, D., van~der Leeuw, S.~E., and West, G. (2009).
\newblock {\em Complexity perspectives in innovation and social change},
  volume~7.
\newblock Springer Science \& Business Media.

\bibitem[Leclaire et~al., 2019]{leclaire2019retour}
Leclaire, M., Rey-Coyrehourcq, S., Charton, B., Arduin, H., Chapron, P.,
  Ch{\'e}rel, G., Delay, E., Gaston, B., Lavall{\'e}e, F., Passerat-Palmbach,
  J., et~al. (2019).
\newblock Retour d'exp{\'e}rience de l'{\'e}cole openmole" exmodelo",
  organis{\'e}e en partenariat avec le m{\'e}so-centre du criann.
\newblock In {\em Journ{\'e}es Calcul et Donn{\'e}es 2019}.

\bibitem[Lobo et~al., 2020]{lobo2020urban}
Lobo, J., Alberti, M., Allen-Dumas, M., Arcaute, E., Barthelemy, M.,
  Bojorquez~Tapia, L.~A., Brail, S., Bettencourt, L., Beukes, A., Chen, W.-Q.,
  et~al. (2020).
\newblock Urban science: Integrated theory from the first cities to sustainable
  metropolises.

\bibitem[Mesoudi, 2017]{mesoudi2017pursuing}
Mesoudi, A. (2017).
\newblock Pursuing darwin’s curious parallel: Prospects for a science of
  cultural evolution.
\newblock {\em Proceedings of the National Academy of Sciences},
  114(30):7853--7860.

\bibitem[Modica et~al., 2017]{modica2017methodological}
Modica, M., Reggiani, A., and Nijkamp, P. (2017).
\newblock Methodological advances in gibrat’s and zipf’s laws: a
  comparative empirical study on the evolution of urban systems.
\newblock In {\em Socioeconomic Environmental Policies and Evaluations in
  Regional Science}, pages 37--59. Springer.

\bibitem[Mustafa et~al., 2017]{mustafa2017coupling}
Mustafa, A., Cools, M., Saadi, I., and Teller, J. (2017).
\newblock Coupling agent-based, cellular automata and logistic regression into
  a hybrid urban expansion model (huem).
\newblock {\em Land Use Policy}, 69:529--540.

\bibitem[Openshaw, 1993]{openshaw1993modelling}
Openshaw, S. (1993).
\newblock Modelling spatial interaction using a neural net.
\newblock In {\em Geographic information systems, spatial modelling and policy
  evaluation}, pages 147--164. Springer.

\bibitem[Paulus, 2004]{paulus2004coevolution}
Paulus, F. (2004).
\newblock {\em Co{\'e}volution dans les syst{\`e}mes de villes: croissance et
  sp{\'e}cialisation des aires urbaines fran{\c{c}}aises de 1950 {\`a} 2000}.
\newblock PhD thesis, Universit{\'e} Panth{\'e}on-Sorbonne-Paris I.

\bibitem[Pumain, 1997]{pumain1997pour}
Pumain, D. (1997).
\newblock Pour une th{\'e}orie {\'e}volutive des villes.
\newblock {\em L'Espace g{\'e}ographique}, pages 119--134.

\bibitem[Pumain, 1998]{pumain1998urban}
Pumain, D. (1998).
\newblock Urban research and complexity.
\newblock In {\em The city and its sciences}, pages 323--361. Springer.

\bibitem[Pumain, 2003]{pumain2003approche}
Pumain, D. (2003).
\newblock Une approche de la complexit{\'e} en g{\'e}ographie.
\newblock {\em G{\'e}ocarrefour}, 78(1):25--31.

\bibitem[Pumain, 2005]{pumain2005cumulativite}
Pumain, D. (2005).
\newblock Cumulativit{\'e} des connaissances.
\newblock {\em Revue europ{\'e}enne des sciences sociales. European Journal of
  Social Sciences}, (XLIII-131):5--12.

\bibitem[Pumain, 2012]{pumain2012urban}
Pumain, D. (2012).
\newblock Urban systems dynamics, urban growth and scaling laws: The question
  of ergodicity.
\newblock In {\em Complexity theories of cities have come of age}, pages
  91--103. Springer.

\bibitem[Pumain, 2020]{pumain2020theories}
Pumain, D. (2020).
\newblock {\em Theories and Models of Urbanization}.
\newblock Springer.

\bibitem[Pumain et~al., 2012]{pumain2012theorie}
Pumain, D. et~al. (2012).
\newblock Une th{\'e}orie g{\'e}ographique pour la loi de zipf.
\newblock {\em R{\'e}gion et d{\'e}veloppement}, 36(31-54):125--150.

\bibitem[Pumain and Raimbault, 2020]{pumain2020conclusion}
Pumain, D. and Raimbault, J. (2020).
\newblock Conclusion: Perspectives on urban theories.
\newblock In {\em Theories and Models of Urbanization}, pages 303--330.
  Springer.

\bibitem[Pumain and Reuillon, 2017]{pumain2017urban}
Pumain, D. and Reuillon, R. (2017).
\newblock {\em Urban dynamics and simulation models}.
\newblock Springer.

\bibitem[Pumain and Robic, 1996]{pumain1996theoriser}
Pumain, D. and Robic, M. (1996).
\newblock Th{\'e}oriser la ville in derycke ph, huriot jm, pumain d.
\newblock {\em Penser la Ville. Th{\'e}ories et mod{\`e}les}.

\bibitem[Pumain and Sanders, 2013]{pumain2013theoretical}
Pumain, D. and Sanders, L. (2013).
\newblock Theoretical principles in interurban simulation models: a comparison.
\newblock {\em Environment and Planning A}, 45(9):2243--2260.

\bibitem[Pumain et~al., 2015]{pumain2015multilevel}
Pumain, D., Swerts, E., Cottineau, C., Vacchiani-Marcuzzo, C., Ignazzi, C.~A.,
  Bretagnolle, A., Delisle, F., Cura, R., Lizzi, L., and Baffi, S. (2015).
\newblock Multilevel comparison of large urban systems.
\newblock {\em Cybergeo: European Journal of Geography}.

\bibitem[Raimbault, 2017]{raimbault2017applied}
Raimbault, J. (2017).
\newblock An applied knowledge framework to study complex systems.
\newblock In {\em Complex Systems Design \& Management}, pages 31--45.

\bibitem[Raimbault, 2018]{raimbault2018calibration}
Raimbault, J. (2018).
\newblock Calibration of a density-based model of urban morphogenesis.
\newblock {\em PloS one}, 13(9):e0203516.

\bibitem[Raimbault, 2020a]{raimbault2020hierarchy}
Raimbault, J. (2020a).
\newblock Hierarchy and co-evolution processes in urban systems.
\newblock {\em arXiv preprint arXiv:2001.11989}.

\bibitem[Raimbault, 2020b]{raimbault2020indirect}
Raimbault, J. (2020b).
\newblock Indirect evidence of network effects in a system of cities.
\newblock {\em Environment and Planning B: Urban Analytics and City Science},
  47(1):138--155.

\bibitem[Raimbault, 2020c]{raimbault2020model}
Raimbault, J. (2020c).
\newblock A model of urban evolution based on innovation diffusion.
\newblock {\em Artificial Life Conference Proceedings}, (32):500--508.

\bibitem[Raimbault, 2020d]{raimbault2020modeling}
Raimbault, J. (2020d).
\newblock Modeling the co-evolution of cities and networks.
\newblock In {\em Handbook of Cities and Networks, Rozenblat C., Niel Z., eds.
  Edward Elgar Publishing. In press}.

\bibitem[Raimbault, 2020e]{raimbault2020relating}
Raimbault, J. (2020e).
\newblock Relating complexities for the reflexive study of complex systems.
\newblock In {\em Theories and models of urbanization}, pages 27--41. Springer.

\bibitem[Raimbault, 2020f]{raimbault2020unveiling}
Raimbault, J. (2020f).
\newblock Unveiling co-evolutionary patterns in systems of cities: a systematic
  exploration of the simpopnet model.
\newblock In {\em Theories and Models of Urbanization}, pages 261--278.
  Springer.

\bibitem[Raimbault et~al., 2019]{raimbault2019space}
Raimbault, J., Cottineau, C., Le~Texier, M., Le~Nechet, F., and Reuillon, R.
  (2019).
\newblock Space matters: Extending sensitivity analysis to initial spatial
  conditions in geosimulation models.
\newblock {\em Journal of Artificial Societies and Social Simulation}, 22(4).

\bibitem[Raimbault et~al., 2020a]{raimbault2020empowering}
Raimbault, J., Denis, E., and Pumain, D. (2020a).
\newblock Empowering urban governance through urban science: Multi-scale
  dynamics of urban systems worldwide.
\newblock {\em Sustainability}, 12(15):5954.

\bibitem[Raimbault et~al., 2020b]{raimbault2020scala}
Raimbault, J., Perret, J., and Reuillon, R. (2020b).
\newblock A scala library for spatial sensitivity analysis.
\newblock In {\em GISRUK 2020 Proceedings. Birkbeck, University of London}.

\bibitem[Raimbault and Pumain, 2019]{raimbault2019methods}
Raimbault, J. and Pumain, D. (2019).
\newblock Methods for exploring simulation models.
\newblock {\em Geographical Modeling: Cities and Territories}, 2:125--150.

\bibitem[Reggiani and Nijkamp, 2009]{reggiani2009complexity}
Reggiani, A. and Nijkamp, P. (2009).
\newblock {\em Complexity and spatial networks: in search of simplicity}.
\newblock Springer Science \& Business Media.

\bibitem[Reuillon et~al., 2015]{reuillon2015new}
Reuillon, R., Schmitt, C., De~Aldama, R., and Mouret, J.-B. (2015).
\newblock A new method to evaluate simulation models: the calibration profile
  (cp) algorithm.
\newblock {\em Journal of Artificial Societies and Social Simulation},
  18(1):12.

\bibitem[Rey-Coyrehourcq, 2015]{rey2015plateforme}
Rey-Coyrehourcq, S. (2015).
\newblock {\em Une plateforme int{\'e}gr{\'e}e pour la construction et
  l'{\'e}valuation de mod{\`e}les de simulation en g{\'e}ographie.(An
  integrated platform for building and evaluating model of simulation in
  geography).}
\newblock PhD thesis, Pantheon-Sorbonne University, Paris, France.

\bibitem[Rey-Coyrehourcq et~al., 2019]{rey2019calcul}
Rey-Coyrehourcq, S., Banos, A., and Raimbault, J. (2019).
\newblock Le calcul intensif en g{\'e}ographie: une tradition bien ancr{\'e}e.
\newblock In {\em JCAD Journ{\'e}es Calcul et Donn{\'e}es 2019}.

\bibitem[Robinson, 2016]{robinson2016comparative}
Robinson, J. (2016).
\newblock Comparative urbanism: new geographies and cultures of theorizing the
  urban.
\newblock {\em International Journal of Urban and Regional Research},
  40(1):187--199.

\bibitem[Robson, 1973]{robson1973urban}
Robson, B. (1973).
\newblock {\em Urban Growth: an approach}.
\newblock Methuen \& Co Ltd., London.

\bibitem[Roth, 2006]{roth2006reconstruction}
Roth, C. (2006).
\newblock Reconstruction failures: Questioning level design.
\newblock In {\em International Workshop on Epistemological Aspects of Computer
  Simulation in the Social Sciences}, pages 89--98. Springer.

\bibitem[Rozenblat, 2020]{rozenblat2020extending}
Rozenblat, C. (2020).
\newblock Extending the concept of city for delineating large urban regions
  (lur) for the cities of the world.
\newblock {\em Cybergeo: European Journal of Geography}.

\bibitem[Schmitt, 2014]{schmitt2014modelisation}
Schmitt, C. (2014).
\newblock {\em Mod{\'e}lisation de la dynamique des syst{\`e}mes de peuplement:
  de SimpopLocal {\`a} SimpopNet}.
\newblock PhD thesis, Universit{\'e} Panth{\'e}on-Sorbonne-Paris I.

\bibitem[Schmitt et~al., 2015]{schmitt2015half}
Schmitt, C., Rey-Coyrehourcq, S., Reuillon, R., and Pumain, D. (2015).
\newblock Half a billion simulations: Evolutionary algorithms and distributed
  computing for calibrating the simpoplocal geographical model.
\newblock {\em Environment and Planning B: Planning and Design},
  42(2):300--315.

\bibitem[Scott and Storper, 2015]{scott2015nature}
Scott, A.~J. and Storper, M. (2015).
\newblock The nature of cities: The scope and limits of urban theory.
\newblock {\em International journal of urban and regional research},
  39(1):1--15.

\bibitem[Suppe, 2000]{suppe2000understanding}
Suppe, F. (2000).
\newblock Understanding scientific theories: An assessment of developments,
  1969-1998.
\newblock {\em Philosophy of science}, 67:S102--S115.

\bibitem[Swerts et~al., 2018]{swerts2018diffuse}
Swerts, E., Denis, E., and Mukhopadhyay, P. (2018).
\newblock Diffuse urbanization and mega-urban regions in india: Between
  reluctant and restrictive urbanism?
\newblock In {\em International and Transnational Perspectives on Urban
  Systems}, pages 237--262. Springer.

\bibitem[Varenne, 2018]{varenne2018theories}
Varenne, F. (2018).
\newblock {\em Th{\'e}ories et mod{\`e}les en sciences humaines: le cas de la
  g{\'e}ographie}.
\newblock {\'E}ditions Mat{\'e}riologiques.

\bibitem[Winther, 2016]{winther2016structure}
Winther, R.~G. (2016).
\newblock The structure of scientific theories.
\newblock In {\em The Stanford Encyclopedia of Philosophy Edward N. Zalta
  (ed.)}.

\bibitem[Wu, 2020]{wu2020emerging}
Wu, F. (2020).
\newblock Emerging cities and urban theories: a chinese perspective.
\newblock In {\em Theories and models of urbanization}, pages 171--182.
  Springer.

\end{thebibliography}

\end{document}

### 2020-11-20 Docs/Papers/HandbookEntropy/chapter.tex

--- 2020-10-28 Docs/Papers/HandbookEntropy/arxiv/v1/chapter.tex
@@
-\documentclass[10pt]{article}
+\documentclass[10pt,letterpaper]{article}
 
@@
 \begin{figure}
-	\includegraphics[width=\linewidth]{Fig1.jpg}
+	\includegraphics[width=\linewidth]{figures/Fig1.jpg}
 	\caption{Schematic representation of the different processes included in the macroscopic co- evolution model for cities and transportation networks.
@@
 \begin{figure}
-	\includegraphics[width=\linewidth]{Fig2.jpg}	
+	\includegraphics[width=\linewidth]{figures/Fig2.jpg}	
 	\caption{Schematic representation of the different processes included in the innovation diffusion urban evolution model.
@@
 
-\begin{thebibliography}{}
-
-\bibitem[Aghion et~al., 1998]{aghion1998endogenous}
-Aghion, P., Ljungqvist, L., Howitt, P., Howitt, P.~W., Brant-Collett, M.,
-  Garc{\'\i}a-Pe{\~n}alosa, C., et~al. (1998).
-\newblock {\em Endogenous growth theory}.
-\newblock MIT press.
-
-\bibitem[Audretsch and Lehmann, 2005]{audretsch2005does}
-Audretsch, D.~B. and Lehmann, E.~E. (2005).
-\newblock Does the knowledge spillover theory of entrepreneurship hold for
-  regions?
-\newblock {\em Research policy}, 34(8):1191--1202.
-
-\bibitem[Banos, 2013]{banos2013pour}
-Banos, A. (2013).
-\newblock {\em Pour des pratiques de mod{\'e}lisation et de simulation
-  lib{\'e}r{\'e}es en g{\'e}ographie et SHS}.
-\newblock PhD thesis.
-
-\bibitem[Batty, 2013]{batty2013new}
-Batty, M. (2013).
-\newblock {\em The new science of cities}.
-\newblock MIT press.
-
-\bibitem[Brenner and Schmid, 2014]{brenner2014urban}
-Brenner, N. and Schmid, C. (2014).
-\newblock The ‘urban age’in question.
-\newblock {\em International journal of urban and regional research},
-  38(3):731--755.
-
-\bibitem[Caldarelli et~al., 2018]{caldarelli2018physics}
-Caldarelli, G., Wolf, S., and Moreno, Y. (2018).
-\newblock Physics of humans, physics for society.
-\newblock {\em Nature Physics}, 14(9):870--870.
-
-\bibitem[Callebaut, 2012]{callebaut2012scientific}
-Callebaut, W. (2012).
-\newblock Scientific perspectivism: A philosopher of science’s response to
-  the challenge of big data biology.
-\newblock {\em Studies in History and Philosophy of Science Part C: Studies in
-  History and Philosophy of Biological and Biomedical Sciences}, 43(1):69--80.
-
-\bibitem[Chavalarias et~al., 2009]{chavalarias2009french}
-Chavalarias, D., Bourgine, P., Perrier, E., Amblard, F., Arlabosse, F., Auger,
-  P., Baillon, J.-B., Barreteau, O., Baudot, P., Bouchaud, E., et~al. (2009).
-\newblock French roadmap for complex systems 2008-2009.
-
-\bibitem[Clarke and Gaydos, 1998]{clarke1998loose}
-Clarke, K.~C. and Gaydos, L.~J. (1998).
-\newblock Loose-coupling a cellular automaton model and gis: long-term urban
-  growth prediction for san francisco and washington/baltimore.
-\newblock {\em International journal of geographical information science},
-  12(7):699--714.
-
-\bibitem[Cottineau, 2014]{cottineau2014evolution}
-Cottineau, C. (2014).
-\newblock {\em L'{\'e}volution des villes dans l'espace post-sovi{\'e}tique.
-  Observation et mod{\'e}lisations.}
-\newblock PhD thesis, Universit{\'e} Paris 1 Panth{\'e}on-Sorbonne.
-
-\bibitem[Cottineau, 2017]{cottineau2017metazipf}
-Cottineau, C. (2017).
-\newblock Metazipf. a dynamic meta-analysis of city size distributions.
-\newblock {\em PloS one}, 12(8):e0183919.
-
-\bibitem[Cottineau, 2020]{cottineau2020metametazipf}
-Cottineau, C. (2020).
-\newblock Metametazipf. what do analyses of city size distributions have in
-  common?
-\newblock {\em arXiv preprint arXiv:2008.11473}.
-
-\bibitem[Cottineau et~al., 2015a]{cottineau2015growing}
-Cottineau, C., Chapron, P., and Reuillon, R. (2015a).
-\newblock Growing models from the bottom up. an evaluation-based incremental
-  modelling method (ebimm) applied to the simulation of systems of cities.
-\newblock {\em Journal of Artificial Societies and Social Simulation}, 18(4):9.
-
-\bibitem[Cottineau et~al., 2015b]{cottineau2015modular}
-Cottineau, C., Reuillon, R., Chapron, P., Rey-Coyrehourcq, S., and Pumain, D.
-  (2015b).
-\newblock A modular modelling framework for hypotheses testing in the
-  simulation of urbanisation.
-\newblock {\em Systems}, 3(4):348--377.
-
-\bibitem[Cura et~al., 2017]{cura2017old}
-Cura, R., Cottineau, C., Swerts, E., Antonio~Ignazzi, C., Bretagnolle, A.,
-  Vacchiani-Marcuzzo, C., and Pumain, D. (2017).
-\newblock The old and the new: qualifying city systems in the world with
-  classical models and new data.
-\newblock {\em Geographical Analysis}, 49(4):363--386.
-
-\bibitem[Denis, 2020]{denis2020more}
-Denis, E. (2020).
-\newblock More urban constructions for whom? drivers of urban built-up
-  expansion across the world from 1990 to 2015.
-\newblock In {\em Theories and Models of Urbanization}, pages 235--258.
-  Springer.
-
-\bibitem[Dupuy, 1987]{dupuy1987vers}
-Dupuy, G. (1987).
-\newblock Vers une th{\'e}orie territoriale des r{\'e}seaux: une application au
-  transport urbain.
-\newblock In {\em Annales de g{\'e}ographie}, pages 658--679. JSTOR.
-
-\bibitem[Durand-Dast{\`e}s, 1991]{durand1991particulier}
-Durand-Dast{\`e}s, F. (1991).
-\newblock Le particulier et le g{\'e}n{\'e}ral en g{\'e}ographie.
-\newblock In {\em Actes du sixi{\`e}me colloque de didactique de l'histoire, de
-  la g{\'e}ographie et des sciences sociales}, volume~6, pages 209--219.
-  Institut national de recherche p{\'e}dagogique.
-
-\bibitem[Egidi et~al., 2020]{egidi2020long}
-Egidi, G., Salvati, L., and Vinci, S. (2020).
-\newblock The long way to tipperary: City size and worldwide urban population
-  trends, 1950-2030.
-\newblock {\em Sustainable Cities and Society}, page 102148.
-
-\bibitem[Favaro and Pumain, 2011]{favaro2011gibrat}
-Favaro, J.-M. and Pumain, D. (2011).
-\newblock Gibrat revisited: An urban growth model incorporating spatial
-  interaction and innovation cycles.
-\newblock {\em Geographical Analysis}, 43(3):261--286.
-
-\bibitem[Gibrat, 1931]{gibrat1931inegalites}
-Gibrat, R. (1931).
-\newblock {\em Les In{\'e}galit{\'e}s {\'E}conomiques, Librairie du Recueil
-  Sirey, Paris}.
-
-\bibitem[Giere, 2010]{giere2010scientific}
-Giere, R.~N. (2010).
-\newblock {\em Scientific perspectivism}.
-\newblock University of Chicago Press.
-
-\bibitem[Grossetti, 2020]{grossetti2020matiere}
-Grossetti, M. (2020).
-\newblock Mati{\`e}re sociale-introduction: {\`a} la recherche d'une ontologie
-  robuste.
-
-\bibitem[Hagerstrand et~al., 1968]{hagerstrand1968innovation}
-Hagerstrand, T. et~al. (1968).
-\newblock {\em Innovation diffusion as a spatial process.}
-\newblock Chicago, USA: Univ. Chicago Press.
-
-\bibitem[Halvorson, 2012]{halvorson2012scientific}
-Halvorson, H. (2012).
-\newblock What scientific theories could not be.
-\newblock {\em Philosophy of science}, 79(2):183--206.
-
-\bibitem[Huutoniemi et~al., 2010]{huutoniemi2010analyzing}
-Huutoniemi, K., Klein, J.~T., Bruun, H., and Hukkinen, J. (2010).
-\newblock Analyzing interdisciplinarity: Typology and indicators.
-\newblock {\em Research policy}, 39(1):79--88.
-
-\bibitem[Jiang et~al., 2015]{jiang2015zipf}
-Jiang, B., Yin, J., and Liu, Q. (2015).
-\newblock Zipf’s law for all the natural cities around the world.
-\newblock {\em International Journal of Geographical Information Science},
-  29(3):498--522.
-
-\bibitem[Krugman, 1996]{krugman1996confronting}
-Krugman, P. (1996).
-\newblock Confronting the mystery of urban hierarchy.
-\newblock {\em Journal of the Japanese and International economies},
-  10(4):399--418.
-
-\bibitem[Lane et~al., 2009]{lane2009complexity}
-Lane, D., Pumain, D., van~der Leeuw, S.~E., and West, G. (2009).
-\newblock {\em Complexity perspectives in innovation and social change},
-  volume~7.
-\newblock Springer Science \& Business Media.
-
-\bibitem[Leclaire et~al., 2019]{leclaire2019retour}
-Leclaire, M., Rey-Coyrehourcq, S., Charton, B., Arduin, H., Chapron, P.,
-  Ch{\'e}rel, G., Delay, E., Gaston, B., Lavall{\'e}e, F., Passerat-Palmbach,
-  J., et~al. (2019).
-\newblock Retour d'exp{\'e}rience de l'{\'e}cole openmole" exmodelo",
-  organis{\'e}e en partenariat avec le m{\'e}so-centre du criann.
-\newblock In {\em Journ{\'e}es Calcul et Donn{\'e}es 2019}.
-
-\bibitem[Lobo et~al., 2020]{lobo2020urban}
-Lobo, J., Alberti, M., Allen-Dumas, M., Arcaute, E., Barthelemy, M.,
-  Bojorquez~Tapia, L.~A., Brail, S., Bettencourt, L., Beukes, A., Chen, W.-Q.,
-  et~al. (2020).
-\newblock Urban science: Integrated theory from the first cities to sustainable
-  metropolises.
-
-\bibitem[Mesoudi, 2017]{mesoudi2017pursuing}
-Mesoudi, A. (2017).
-\newblock Pursuing darwin’s curious parallel: Prospects for a science of
-  cultural evolution.
-\newblock {\em Proceedings of the National Academy of Sciences},
-  114(30):7853--7860.
-
-\bibitem[Modica et~al., 2017]{modica2017methodological}
-Modica, M., Reggiani, A., and Nijkamp, P. (2017).
-\newblock Methodological advances in gibrat’s and zipf’s laws: a
-  comparative empirical study on the evolution of urban systems.
-\newblock In {\em Socioeconomic Environmental Policies and Evaluations in
-  Regional Science}, pages 37--59. Springer.
-
-\bibitem[Mustafa et~al., 2017]{mustafa2017coupling}
-Mustafa, A., Cools, M., Saadi, I., and Teller, J. (2017).
-\newblock Coupling agent-based, cellular automata and logistic regression into
-  a hybrid urban expansion model (huem).
-\newblock {\em Land Use Policy}, 69:529--540.
-
-\bibitem[Openshaw, 1993]{openshaw1993modelling}
-Openshaw, S. (1993).
-\newblock Modelling spatial interaction using a neural net.
-\newblock In {\em Geographic information systems, spatial modelling and policy
-  evaluation}, pages 147--164. Springer.
-
-\bibitem[Paulus, 2004]{paulus2004coevolution}
-Paulus, F. (2004).
-\newblock {\em Co{\'e}volution dans les syst{\`e}mes de villes: croissance et
-  sp{\'e}cialisation des aires urbaines fran{\c{c}}aises de 1950 {\`a} 2000}.
-\newblock PhD thesis, Universit{\'e} Panth{\'e}on-Sorbonne-Paris I.
-
-\bibitem[Pumain, 1997]{pumain1997pour}
-Pumain, D. (1997).
-\newblock Pour une th{\'e}orie {\'e}volutive des villes.
-\newblock {\em L'Espace g{\'e}ographique}, pages 119--134.
-
-\bibitem[Pumain, 1998]{pumain1998urban}
-Pumain, D. (1998).
-\newblock Urban research and complexity.
-\newblock In {\em The city and its sciences}, pages 323--361. Springer.
-
-\bibitem[Pumain, 2003]{pumain2003approche}
-Pumain, D. (2003).
-\newblock Une approche de la complexit{\'e} en g{\'e}ographie.
-\newblock {\em G{\'e}ocarrefour}, 78(1):25--31.
-
-\bibitem[Pumain, 2005]{pumain2005cumulativite}
-Pumain, D. (2005).
-\newblock Cumulativit{\'e} des connaissances.
-\newblock {\em Revue europ{\'e}enne des sciences sociales. European Journal of
-  Social Sciences}, (XLIII-131):5--12.
-
-\bibitem[Pumain, 2012]{pumain2012urban}
-Pumain, D. (2012).
-\newblock Urban systems dynamics, urban growth and scaling laws: The question
-  of ergodicity.
-\newblock In {\em Complexity theories of cities have come of age}, pages
-  91--103. Springer.
-
-\bibitem[Pumain, 2020]{pumain2020theories}
-Pumain, D. (2020).
-\newblock {\em Theories and Models of Urbanization}.
-\newblock Springer.
-
-\bibitem[Pumain et~al., 2012]{pumain2012theorie}
-Pumain, D. et~al. (2012).
-\newblock Une th{\'e}orie g{\'e}ographique pour la loi de zipf.
-\newblock {\em R{\'e}gion et d{\'e}veloppement}, 36(31-54):125--150.
-
-\bibitem[Pumain and Raimbault, 2020]{pumain2020conclusion}
-Pumain, D. and Raimbault, J. (2020).
-\newblock Conclusion: Perspectives on urban theories.
-\newblock In {\em Theories and Models of Urbanization}, pages 303--330.
-  Springer.
-
-\bibitem[Pumain and Reuillon, 2017]{pumain2017urban}
-Pumain, D. and Reuillon, R. (2017).
-\newblock {\em Urban dynamics and simulation models}.
-\newblock Springer.
-
-\bibitem[Pumain and Robic, 1996]{pumain1996theoriser}
-Pumain, D. and Robic, M. (1996).
-\newblock Th{\'e}oriser la ville in derycke ph, huriot jm, pumain d.
-\newblock {\em Penser la Ville. Th{\'e}ories et mod{\`e}les}.
-
-\bibitem[Pumain and Sanders, 2013]{pumain2013theoretical}
-Pumain, D. and Sanders, L. (2013).
-\newblock Theoretical principles in interurban simulation models: a comparison.
-\newblock {\em Environment and Planning A}, 45(9):2243--2260.
-
-\bibitem[Pumain et~al., 2015]{pumain2015multilevel}
-Pumain, D., Swerts, E., Cottineau, C., Vacchiani-Marcuzzo, C., Ignazzi, C.~A.,
-  Bretagnolle, A., Delisle, F., Cura, R., Lizzi, L., and Baffi, S. (2015).
-\newblock Multilevel comparison of large urban systems.
-\newblock {\em Cybergeo: European Journal of Geography}.
-
-\bibitem[Raimbault, 2017]{raimbault2017applied}
-Raimbault, J. (2017).
-\newblock An applied knowledge framework to study complex systems.
-\newblock In {\em Complex Systems Design \& Management}, pages 31--45.
-
-\bibitem[Raimbault, 2018]{raimbault2018calibration}
-Raimbault, J. (2018).
-\newblock Calibration of a density-based model of urban morphogenesis.
-\newblock {\em PloS one}, 13(9):e0203516.
-
-\bibitem[Raimbault, 2020a]{raimbault2020hierarchy}
-Raimbault, J. (2020a).
-\newblock Hierarchy and co-evolution processes in urban systems.
-\newblock {\em arXiv preprint arXiv:2001.11989}.
-
-\bibitem[Raimbault, 2020b]{raimbault2020indirect}
-Raimbault, J. (2020b).
-\newblock Indirect evidence of network effects in a system of cities.
-\newblock {\em Environment and Planning B: Urban Analytics and City Science},
-  47(1):138--155.
-
-\bibitem[Raimbault, 2020c]{raimbault2020model}
-Raimbault, J. (2020c).
-\newblock A model of urban evolution based on innovation diffusion.
-\newblock {\em Artificial Life Conference Proceedings}, (32):500--508.
-
-\bibitem[Raimbault, 2020d]{raimbault2020modeling}
-Raimbault, J. (2020d).
-\newblock Modeling the co-evolution of cities and networks.
-\newblock In {\em Handbook of Cities and Networks, Rozenblat C., Niel Z., eds.
-  Edward Elgar Publishing. In press}.
-
-\bibitem[Raimbault, 2020e]{raimbault2020relating}
-Raimbault, J. (2020e).
-\newblock Relating complexities for the reflexive study of complex systems.
-\newblock In {\em Theories and models of urbanization}, pages 27--41. Springer.
-
-\bibitem[Raimbault, 2020f]{raimbault2020unveiling}
-Raimbault, J. (2020f).
-\newblock Unveiling co-evolutionary patterns in systems of cities: a systematic
-  exploration of the simpopnet model.
-\newblock In {\em Theories and Models of Urbanization}, pages 261--278.
-  Springer.
-
-\bibitem[Raimbault et~al., 2019]{raimbault2019space}
-Raimbault, J., Cottineau, C., Le~Texier, M., Le~Nechet, F., and Reuillon, R.
-  (2019).
-\newblock Space matters: Extending sensitivity analysis to initial spatial
-  conditions in geosimulation models.
-\newblock {\em Journal of Artificial Societies and Social Simulation}, 22(4).
-
-\bibitem[Raimbault et~al., 2020a]{raimbault2020empowering}
-Raimbault, J., Denis, E., and Pumain, D. (2020a).
-\newblock Empowering urban governance through urban science: Multi-scale
-  dynamics of urban systems worldwide.
-\newblock {\em Sustainability}, 12(15):5954.
-
-\bibitem[Raimbault et~al., 2020b]{raimbault2020scala}
-Raimbault, J., Perret, J., and Reuillon, R. (2020b).
-\newblock A scala library for spatial sensitivity analysis.
-\newblock In {\em GISRUK 2020 Proceedings. Birkbeck, University of London}.
-
-\bibitem[Raimbault and Pumain, 2019]{raimbault2019methods}
-Raimbault, J. and Pumain, D. (2019).
-\newblock Methods for exploring simulation models.
-\newblock {\em Geographical Modeling: Cities and Territories}, 2:125--150.
-
-\bibitem[Reggiani and Nijkamp, 2009]{reggiani2009complexity}
-Reggiani, A. and Nijkamp, P. (2009).
-\newblock {\em Complexity and spatial networks: in search of simplicity}.
-\newblock Springer Science \& Business Media.
-
-\bibitem[Reuillon et~al., 2015]{reuillon2015new}
-Reuillon, R., Schmitt, C., De~Aldama, R., and Mouret, J.-B. (2015).
-\newblock A new method to evaluate simulation models: the calibration profile
-  (cp) algorithm.
-\newblock {\em Journal of Artificial Societies and Social Simulation},
-  18(1):12.
-
-\bibitem[Rey-Coyrehourcq, 2015]{rey2015plateforme}
-Rey-Coyrehourcq, S. (2015).
-\newblock {\em Une plateforme int{\'e}gr{\'e}e pour la construction et
-  l'{\'e}valuation de mod{\`e}les de simulation en g{\'e}ographie.(An
-  integrated platform for building and evaluating model of simulation in
-  geography).}
-\newblock PhD thesis, Pantheon-Sorbonne University, Paris, France.
-
-\bibitem[Rey-Coyrehourcq et~al., 2019]{rey2019calcul}
-Rey-Coyrehourcq, S., Banos, A., and Raimbault, J. (2019).
-\newblock Le calcul intensif en g{\'e}ographie: une tradition bien ancr{\'e}e.
-\newblock In {\em JCAD Journ{\'e}es Calcul et Donn{\'e}es 2019}.
-
-\bibitem[Robinson, 2016]{robinson2016comparative}
-Robinson, J. (2016).
-\newblock Comparative urbanism: new geographies and cultures of theorizing the
-  urban.
-\newblock {\em International Journal of Urban and Regional Research},
-  40(1):187--199.
-
-\bibitem[Robson, 1973]{robson1973urban}
-Robson, B. (1973).
-\newblock {\em Urban Growth: an approach}.
-\newblock Methuen \& Co Ltd., London.
-
-\bibitem[Roth, 2006]{roth2006reconstruction}
-Roth, C. (2006).
-\newblock Reconstruction failures: Questioning level design.
-\newblock In {\em International Workshop on Epistemological Aspects of Computer
-  Simulation in the Social Sciences}, pages 89--98. Springer.
-
-\bibitem[Rozenblat, 2020]{rozenblat2020extending}
-Rozenblat, C. (2020).
-\newblock Extending the concept of city for delineating large urban regions
-  (lur) for the cities of the world.
-\newblock {\em Cybergeo: European Journal of Geography}.
-
-\bibitem[Schmitt, 2014]{schmitt2014modelisation}
-Schmitt, C. (2014).
-\newblock {\em Mod{\'e}lisation de la dynamique des syst{\`e}mes de peuplement:
-  de SimpopLocal {\`a} SimpopNet}.
-\newblock PhD thesis, Universit{\'e} Panth{\'e}on-Sorbonne-Paris I.
-
-\bibitem[Schmitt et~al., 2015]{schmitt2015half}
-Schmitt, C., Rey-Coyrehourcq, S., Reuillon, R., and Pumain, D. (2015).
-\newblock Half a billion simulations: Evolutionary algorithms and distributed
-  computing for calibrating the simpoplocal geographical model.
-\newblock {\em Environment and Planning B: Planning and Design},
-  42(2):300--315.
-
-\bibitem[Scott and Storper, 2015]{scott2015nature}
-Scott, A.~J. and Storper, M. (2015).
-\newblock The nature of cities: The scope and limits of urban theory.
-\newblock {\em International journal of urban and regional research},
-  39(1):1--15.
-
-\bibitem[Suppe, 2000]{suppe2000understanding}
-Suppe, F. (2000).
-\newblock Understanding scientific theories: An assessment of developments,
-  1969-1998.
-\newblock {\em Philosophy of science}, 67:S102--S115.
-
-\bibitem[Swerts et~al., 2018]{swerts2018diffuse}
-Swerts, E., Denis, E., and Mukhopadhyay, P. (2018).
-\newblock Diffuse urbanization and mega-urban regions in india: Between
-  reluctant and restrictive urbanism?
-\newblock In {\em International and Transnational Perspectives on Urban
-  Systems}, pages 237--262. Springer.
-
-\bibitem[Varenne, 2018]{varenne2018theories}
-Varenne, F. (2018).
-\newblock {\em Th{\'e}ories et mod{\`e}les en sciences humaines: le cas de la
-  g{\'e}ographie}.
-\newblock {\'E}ditions Mat{\'e}riologiques.
-
-\bibitem[Winther, 2016]{winther2016structure}
-Winther, R.~G. (2016).
-\newblock The structure of scientific theories.
-\newblock In {\em The Stanford Encyclopedia of Philosophy Edward N. Zalta
-  (ed.)}.
-
-\bibitem[Wu, 2020]{wu2020emerging}
-Wu, F. (2020).
-\newblock Emerging cities and urban theories: a chinese perspective.
-\newblock In {\em Theories and models of urbanization}, pages 171--182.
-  Springer.
-
-\end{thebibliography}
+\bibliographystyle{apalike}
+\bibliography{biblio}
 

### 2020-11-24 Models/MesoBenchmark/pseResults.R
replications

@@ -9,14 +9,54 @@ source(paste0(Sys.getenv('CS_HOME'),'/Organisation/Models/Utils/R/plots.R'))
 
 resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/MesoBenchmark/')
 
-files = list(
-  'expmixture' = 'openmole/pse/20200829_1514_PSE_EXPMIXTURE_res/population100000.csv',
-  'reactiondiffusion' = 'openmole/pse/20200828_1624_PSE_REACTIONDIFFUSION_res/population50000.csv',
-  'correlatedperco' = 'openmole/pse/20200829_1330_PSE_CORRELATEDPERCO_res/population100000.csv',
-  'gravity' = 'openmole/pse/20200829_1526_PSE_GRAVITY_res/population37000.csv'
+resdirs = list(
+  'expmixture' = 'openmole/pse/20200829_1514_PSE_EXPMIXTURE_res',
+  'reactiondiffusion' = 'openmole/pse/20200828_1624_PSE_REACTIONDIFFUSION_res',
+  'correlatedperco' = 'openmole/pse/20201121_1802_PSE_CORRELATEDPERCO_GRID/',
+  'gravity' = 'openmole/pse/20200829_1526_PSE_GRAVITY_res'
 )
 
+gen<-function(filename){as.integer(sapply(strsplit(sapply(strsplit(filename,"population"),function(s){s[2]}),".csv"),function(s){s[1]}))}
+
+latestgen <- function(dir){
+  if(is.null(dir)){return(NULL)}
+  else{return(max(sapply(list.files(dir,pattern=".csv"),gen)))}
+}
+
 indics = c('moran','distance','entropy','slope')
+
+#####
+# Convergence
+
+allres = data.frame()
+for(model in names(resdirs)){
+  currentdir = resdirs[[model]]
+  currentfiles = paste0(currentdir,'/',list.files(currentdir,pattern = '.csv'))
+  for(currentfile in currentfiles){
+    currentres = as.tbl(read.csv(currentfile));currentgen=gen(currentfile)
+    allres = rbind(allres,cbind(currentres[,c(indics,'evolution.samples')],model=rep(model,nrow(currentres)),generation=rep(currentgen,nrow(currentres))))
+  }
+}
+
+sres = allres %>% group_by(model,generation) %>% summarise(patterns = n())
+
+g=ggplot(sres[sres$model=='correlatedperco',],aes(x=generation,y=patterns,color=model,group=model))
[…]

### 2020-11-25 Models/MesoBenchmark/pseResults.R
myoml_results

@@ -12,9 +12,11 @@ resdir = paste0(Sys.getenv('CS_HOME'),'/UrbanGrowth/Results/MesoBenchmark/')
 resdirs = list(
   'expmixture' = 'openmole/pse/20200829_1514_PSE_EXPMIXTURE_res',
   'reactiondiffusion' = 'openmole/pse/20200828_1624_PSE_REACTIONDIFFUSION_res',
-  'correlatedperco' = 'openmole/pse/20201121_1802_PSE_CORRELATEDPERCO_GRID/',
+  'correlatedperco' = 'openmole/pse/20201124_1816_PSE_CORRELATEDPERCO_REPLICATIONS_GRID/',
   'gravity' = 'openmole/pse/20200829_1526_PSE_GRAVITY_res'
 )
+
+stochastic = FALSE
 
 gen<-function(filename){as.integer(sapply(strsplit(sapply(strsplit(filename,"population"),function(s){s[2]}),".csv"),function(s){s[1]}))}
 
@@ -34,7 +36,12 @@ for(model in names(resdirs)){
   currentfiles = paste0(currentdir,'/',list.files(currentdir,pattern = '.csv'))
   for(currentfile in currentfiles){
     currentres = as.tbl(read.csv(currentfile));currentgen=gen(currentfile)
-    allres = rbind(allres,cbind(currentres[,c(indics,'evolution.samples')],model=rep(model,nrow(currentres)),generation=rep(currentgen,nrow(currentres))))
+    # toggle for stochastic runs
+    if (stochastic){
+      allres = rbind(allres,cbind(currentres[,c(indics,'evolution.samples')],model=rep(model,nrow(currentres)),generation=rep(currentgen,nrow(currentres))))
+    }else{
+      allres = rbind(allres,cbind(currentres[,indics],model=rep(model,nrow(currentres)),generation=rep(currentgen,nrow(currentres))))
+    }
   }
 }
 
@@ -43,14 +50,15 @@ sres = allres %>% group_by(model,generation) %>% summarise(patterns = n())
 g=ggplot(sres[sres$model=='correlatedperco',],aes(x=generation,y=patterns,color=model,group=model))
 g+geom_point()+geom_line()
 
-sres = allres[allres$evolution.samples>=10,] %>% group_by(model,generation) %>% summarise(patterns = n())
-
-g=ggplot(sres[sres$model=='correlatedperco',],aes(x=generation,y=patterns,color=model,group=model))
-g+geom_point()+geom_line()
+if (stochastic){
[…]

### 2020-11-25 Models/MesoBenchmark/pseResults.R
readme_papers

@@ -77,7 +77,7 @@ for(model in names(resdirs)){
   }
 }
 
-#plot(res[res$model=='correlatedperco',indics])
+#plot(res[res$model=='correlatedperco'&res$slope>-4,indics])
 
 
 ggsave(plot = ggpairs(res[res$slope>-4,],aes(color=model),columns = indics,

### 2020-12-06 Docs/Papers/MesoBenchmark/paper/paper.tex
rerun
myoml_results
time_pse_expmixture

@@
 
+\cite{jun2020modling} entropy and urban growth
+
 \section*{Introduction}
@@
 Source code for numerical experiments, simulation data and results used in this paper are available on an open git repository at \url{https://github.com/JusteRaimbault/UrbanGrowth}.
+Full simulation data is available on the dataverse repository at \url{https://doi.org/10.7910/DVN/VVIEYO}.
 
@@
 Convergence was reached separately for each model.
+% equivalent in terms of Islands generation?
+% which spatialdata commit? 4ef6bc7c4e7887bb35138773ae87b158e6e9a19b
+% (replace plugin on myoml-27/08 16h52 + remove urbgrowth acc)
+% latest 36fb676ce20f1ca6586defb9f8a15ea2a366477b
+% (remove urbanevolution_2.12 29/04 15:03: pb compat) (remove saltelli exec - time ok)(remove zombiesbundle 31/08 9:55)
 
@@
 
+Correlations/PCA for each model and for all points? (plus analysis in link to params? not possible in stoch? (does replication strain params? Y!)
+
 To better quantify how each model are similar and how they approach real configurations, we compute the hypervolume (in the four dimensional indicator space) corresponding to each point cloud and the intersections between these hypervolumes.
@@
 \section*{Discussion}
+
+geodesign: uncertainty - compromises - no prediction models
+
+\cite{10.1371/journal.pone.0242479} emissions - density: policies when linked to models?
+
+\cite{d2019new}
 

### 2020-12-07 Docs/Seminars/2020/ISWEE2020/Communication/iswee2020.tex
slides_iswee

\input{header_slides.tex}

\begin{document}

\title[Dynamics of Urban Systems]{Empowering Urban Governance through Urban Science: Multi-Scale Dynamics of Urban Systems Worldwide}
\author[Raimbault, Denis and Pumain]{J.~Raimbault$^{1,2,3\ast}$, E. Denis$^3$ and D.Pumain$^3$\\
\texttt{j.raimbault@ucl.ac.uk}
}

\institute[CASA, UCL]{$^{1}$CASA, UCL\\
$^{2}$UPS CNRS 3611 Complex Systems Institute Paris\\
$^{3}$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}

\date[December 7th 2020]{ISWEE 2020\\
Session 2: Sustainable Cities and Society\\
December 7th 2020
}

\frame{\maketitle}

%

%Abstract:	Cities are facing many sustainability issues in the context of the current global interdependency characterized by an economic uncertainty coupled to climate changes, which challenge their local policies aiming to better conciliate reasonable growth with livable urban environment. The urban dynamic models developed by the so-called “urban science” can provide a useful foundation for more sustainable urban policies. It implies that their proposals have been validated by correct observations of the diversity of situations in the world. However, international comparisons of the evolution of cities often produce unclear results because national territorial frameworks are not always in strict correspondence with the dynamics of urban systems. We propose to provide various compositions of systems of cities in order to better take into account the dynamic networking of cities that go beyond regional and national territorial boundaries. Different models conceived for explaining city size and urban growth distributions enable the establishing of a correspondence between urban trajectories when observed at the level of cities and systems of cities. We test the validity and representativeness of several dynamic models of complex urban systems and their variations across regions of the world, at the macroscopic scale of systems of cities. The originality of the approach resides in the way it considers spatial interaction and evolutionary path dependence as major features in the general behavior of urban entities. The models studied include diverse and complementary processes, such as economic exchanges, diffusion of innovations, and physical network flows. Complex systems dynamics is in principle unpredictable, but contextualizing it regarding demographic, income, and resource components may help in minimizing the forecasting errors. We use, among others, a new unique source correlating population and built-up footprint at world scale: the Global Human Settlement built-up areas (GHS-BU). Following the methodology and results already obtained in the European GeoDiverCity project, including USA, Europe, and BRICS countries, we complete them with this new dataset at world scale and different models. This research helps in further empirical testing of the hypotheses of the evolutionary theory of urban systems and partially revising them. We also suggest research directions towards the coupling of these models into a multi-scale model of urban growth.

\section{Introduction}

\sframe{How to explain urban growth?}{

\begin{itemize}
	\item Apparent direct \textbf{causes} : intentions/actions from urban actors (policies, locational strategies from firms, residential migrations \ldots)
	\item But \textbf{statistical observation} (thousands of cities, over centuries) : each city has a probability of growing similar to other cities belonging to the same territorial system	
\end{itemize}

\medskip
$\rightarrow$ ``distributed growth'' on the long run with many local and temporal fluctuations

}

\sframe{Gibrat's model}{

``Proportional'' growth = growth rates are equiprobable for any city size and not correlated with previous rate

\medskip

Good fit $\rightarrow$ double explanatory gain:

\begin{itemize}
	\item Persistency of urban spatial patterns and hierarchies
	\item The statistical shape of urban sizes distribution (Zipf’s law or lognormal) as generated from growth process
\end{itemize}

\cite{gibrat1931inegalits} \cite{robson1973urban} \cite{pumain1982dynamique}

}

\sframe{Gibrat’s model}{
%
\textit{A satisfying proxy but some empirical contradiction}
%
\bigskip
%
\begin{enumerate}
\item The observed distributions of city sizes (actually: settlement sizes including hamlets, villages, towns and SMAs) are lognormal (evidence from \cite{robson1973urban}, \cite{pumain1982dynamique}, \cite{eeckhout2004gibrat}, \cite{decker2007global})
\item Gibrat’s growth model leads to a lognormal distribution of city sizes
\item But Gibrat’s growth model hypothesis are rejected (correlation between growth rates and city size, correlation between successive growth rates)
\end{enumerate}

}
%
%\sframe{Evolutionary urban theory}{
%
%\textit{Innovation diffusion and spatial integration of urban systems}
%
%\bigskip
%
%\begin{itemize}
%\item In new urban systems, as in USA, there is a spatial filling process that occur through waves of urban growth (urban Frontier) corresponding to the diffusion of economic cycles
%\item In mature urban systems, like in France, the innovations diffusion reaches cities that are not spatially regularly arranged but already experienced other growth periods according to distinct cycles of urban specialisation
%\end{itemize}
%
%}
%
%
\sframe{Evolutionary urban theory and scaling laws}{

An \textbf{Evolutionary Urban Theory} linking scaling laws to a geographical model of urban growth with spatial interaction and innovation cycles \cite{pumain1997pour} \cite{pumain2018evolutionary}

\bigskip

\begin{itemize}
\item \cite{pumain2006evolutionary} suggest to replace a generic statistical model of growing independent entities (Gibrat’s urban growth model) by a model of spatially and temporally interdependent entities (i.e. the geographical concept of ``system of cities'' or ``settlement system'')
\item It reproduces the observations on differential scaling parameters for urban activities according to their age in innovation cycles \cite{favaro2011gibrat}
\item It also makes explicit the multilevel dynamics of interurban competition for capturing innovation, which may itself generate new innovation through interurban emulation, within an evolutionary perspective
\end{itemize}

}

\sframe{Testing the evolutionary urban theory}{

\footnotesize

\textit{How are stylized facts on systems of cities robust and general ?}

\medskip

$\rightarrow$ empirical study with the new Global Human Settlement layer dataset

\medskip

\textit{How can dynamical models of urban systems be applied in the context of the evolutionary urban theory ?}

\medskip

$\rightarrow$ test of six dynamical models, based on geographical interactions between cities but different dimensions, on different systems of cities and worldwide

%\medskip

\begin{center}
\includegraphics[width=0.77\textwidth]{figures/paper.png}
\end{center}

}

\section{Data analysis}

\sframe{A new source of data on global urbanization}{

\begin{itemize}
	\item GHSL (Global Human Settlement Layer) : GEO Human Planet Initiative (European Commission)
	\item Built up area from satellite images 40 m + population data 250 m $\rightarrow$ 1 km2 grid
	\item 13 000 urban areas > 50 000 inhab.
	\item Surface, population in 1975, 1990, 2000, 2015
	\item GDP, Green surfaces, Pollutants 1990-2015
\end{itemize}

}

% number of areas per country
% summary(countrycount$count)
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
%   1.00    4.00   13.00   72.17   48.75 3248.00

%
%\sframe{Population distributions}{
%
%\begin{center}
%    \includegraphics[width=0.7\textwidth]{figures/population_distributions.png}
%\end{center}
%
%\textit{Temporal stability of the shape of global population distribution}
%
%}

\sframe{Urban systems summary}{

    % \cite{pumain2015multilevel} : number of cities / urban pop for each urban system
    % Summary table : number of cities, rank size slope, primacy index, total population

    % summary - 2015 - geodivercity urban systems

    \textit{Summary statistics in 2015 for urban systems \cite{pumain2015multilevel}}

    \smallskip

%    {\tiny
%    \begin{center}
%    \begin{tabular}{|c|c|c|c|c|c|}
%    \hline
%        System & Population & Cities & Primacy & Rank-size & R2  \\\hline
%         Europe & 188Mio & 693 & 1.01 & $0.94 \pm 0.0026$ & 0.994  \\
%         China & 567.2Mio & 1850 & 1.66 & $0.91 \pm 0.0011$ & 0.997  \\
%         Brazil & 111.7Mio & 349 & 1.95 & $0.99 \pm 0.0026$ & 0.998  \\
%         India & 703.1Mio & 3248 & 1.22 & $0.78 \pm 0.0009$ & 0.996  \\
%         South Africa & 25.3Mio & 77 & 1.85 & $1.05 \pm 0.020$ & 0.974  \\
%         US & 153Mio & 324 & 1.12 & $1.16 \pm 0.0054$ & 0.992  \\
%         FSU & 120.3Mio & 450 & 3.27 & $0.92 \pm 0.0062$ & 0.979  \\\hline
%    \end{tabular}
%    \end{center}
%    }
{%\tiny
    \begin{center}
    \begin{tabular}{|c|c|c|c|c|c|}
    \hline
        System & Pop (M) & Pop geodiv. & Cities & Rank-size  \\\hline
         Europe & 188 & 291 & 693 & $0.94$  \\
         China & 567 & 481 & 1850 & $0.91$  \\
         Brazil & 112 & 161 & 349 & $0.99$  \\
         India & 703 & 427 & 3248 & $0.78$  \\
         South Africa & 25 & 25 & 77 & $1.05$ \\
         US & 153 & 324 & 287 & $1.16$  \\
         FSU & 120 & 174 & 450 & $0.92$ \\\hline
    \end{tabular}
    \end{center}
    }

    %\smallskip

	%\begin{center}
%		\includegraphics[height=0.4\textheight]{figures/geodivercity-summary.png}
%	\end{center}

}

\sframe{Urban systems hierarchy}{

Reproducing results of \cite{pumain2015multilevel} for large urban systems

% Brazil, China, US, Europe, South Africa, URSS, India

\begin{center}
    \includegraphics[width=0.53\textwidth,height=0.5\textheight]{../../../../Communications/2019/ECTQG2019/Communication/figures/RankSize_Europe-China-Brazil-India-SouthAfrica-UnitedStates-FormerSovietUnion_years-90-00-15.png}
    \includegraphics[width=0.45\textwidth]{../../../../Communications/2019/ECTQG2019/Communication/figures/geodivercity-ranksize.png}
\end{center}

\medskip

$\rightarrow$ \textbf{Robustness of qualitative stylized facts to the database}

}

\sframe{Rank-size by continents or trade areas}{

\begin{center}
    \includegraphics[width=0.48\textwidth]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig1.png}
    \includegraphics[width=0.48\textwidth]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig2.png}
\end{center}

\medskip

$\rightarrow$ \textbf{Possibility to extend analysis to other consistent geographical ensembles}

}

\sframe{Correlations between urban indicators}{

% includes test of gibrat law : correlation pops / growth rates

\begin{center}
    %\includegraphics[width=0.46\textwidth]{figures/correlations_00.png}\hspace{0.2cm}
    \includegraphics[width=0.7\textwidth]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig3.png}
\end{center}

%\medskip

%$\rightarrow$ no apparent deviation to Gibrat's law with this dataset (importance of definition of cities in scaling laws \cite{cottineau2017diverse})

}

\sframe{Linking urban growth and built-up area growth}{

\begin{center}
\includegraphics[width=0.85\textwidth]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig4.png}
\end{center}

\footnotesize

\textit{Geographical structure in the relation between population growth and built-up area growth}

}

%
%\sframe{Scaling laws}{
%
%%reproduction / Generalizing existing studies
%
%\begin{center}
%%    \includegraphics[width=0.32\textwidth]{figures/Built-fitted_Europe-China-Brazil-India-SouthAfrica-UnitedStates-FormerSovietUnion_years-90-00-15.png}
%    \includegraphics[width=0.48\textwidth]{figures/GDP-fitted_Europe-China-Brazil-India-SouthAfrica-UnitedStates-FormerSovietUnion_years-90-00-15.png}
%    \includegraphics[width=0.48\textwidth]{figures/Emissions-fitted_Europe-China-Brazil-India-SouthAfrica-UnitedStates-FormerSovietUnion_years-90-00-15.png}
%\end{center}
%
%}

\sframe{Evolution of scaling exponents}{

% evolution of scaling exponents in time

\begin{center}
    %\includegraphics[width=0.32\textwidth,height=0.45\textheight]{figures/Built-evolution_Europe-China-Brazil-India-SouthAfrica-UnitedStates-FormerSovietUnion_years-90-00-15.png}
    \includegraphics[width=0.48\textwidth,height=0.45\textheight]{../../../../Communications/2019/ECTQG2019/Communication/figures/Emissions-evolution_Europe-China-Brazil-India-SouthAfrica-UnitedStates-FormerSovietUnion_years-90-00-15.png}
    \includegraphics[width=0.48\textwidth,height=0.45\textheight]{../../../../Communications/2019/ECTQG2019/Communication/figures/GDP-evolution_Europe-China-Brazil-India-SouthAfrica-UnitedStates-FormerSovietUnion_years-90-00-15.png}
\end{center}

\medskip

\textit{All indicators are stable in their confidence range}

}

\sframe{Summary of scaling exponents}{

\begin{center}
\small
    \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    System & Built-up area & GDP & Emissions \\\hline
Europe & $0.93 \pm 0.016$ (0.83) & $1.15 \pm 0.019$ (0.83) & $1.50 \pm 0.038$ (0.69) \\
China & $1.06 \pm 0.019$ (0.62)  & $1.14 \pm 0.011$ (0.85)  & $1.84 \pm 0.037$ (0.57) \\
Brazil & $0.98 \pm 0.025$ (0.81)  & $1.10 \pm 0.055 $ (0.54) & $1.71 \pm 0.053$ (0.75) \\
India & $1.34 \pm 0.031$ (0.36) & $1.25 \pm 0.022$ (0.50) & $1.54 \pm 0.031$ (0.42) \\
S. Africa & $1.18 \pm 0.090$ (0.69) & $1.08 \pm 0.028$ (0.95) & $1.56 \pm 0.087$  (0.81) \\
US & $0.97 \pm 0.015$ (0.92) & $1.04 \pm 0.069$ (0.99) &$ 1.34 \pm 0.03$ (0.84) \\
FSU & $0.97 \pm 0.035$ (0.63) & $1.17 \pm 0.041$ (0.65) & $1.95 \pm 0.088$ (0.52)\\\hline
\end{tabular}
\end{center}

\bigskip

$\rightarrow$ more general, more or less consistent study of scaling (``basic'' indicators but on consistent and global geographical areas)
% vs shitty on number of cinemas etc - we do not really care.

}

%\sframe{Fitting scaling laws}{

% \cite{finance2018absent}
% \cite{clauset2009power}
% \cite{leitao2016scaling}
% \cite{broido2019scale}
% \cite{smith2019biochemical}

% summary table of Goodness of fit for - classical scaling, with cutoff, as piecewise linear

%\textit{Alternative models for better fit of scaling laws}

% test alternative scaling models ?

%}

\section{Dynamical models}

\sframe{Dynamical models of urban growth}{

\textit{Testing interaction-based dynamical models for urban growth}

\bigskip

\begin{itemize}
	\item The Favaro-Pumain model for the diffusion of innovation \cite{favaro2011gibrat}
	\item The Marius model family based on economic exchanges \cite{cottineau2014evolution}
	\item An interaction model including physical transportation networks \cite{raimbault2018indirect}
\end{itemize}

}

\sframe{Description of models}{

% increasing complexity for the presentation of models

\begin{columns}
	\begin{column}{0.33\textwidth}
		\textbf{Network interaction model}

		\medskip

		\begin{itemize}
			\item Endogenous growth
			\item Interactions inducing growth through gravity potential
			\item Static physical network taken into account (geographical shortest path with topography)
		\end{itemize}

	\end{column}
	\vrule{}
	\begin{column}{0.33\textwidth}
		%\vspace{-1cm}
		\textbf{Favaro-Pumain model}

		\medskip

		\begin{itemize}
			\item Endogenous growth
			\item Innovation emerge and diffuse in cities
			\item Growth rates adapted according to utility of innovation and level of adaptation
		\end{itemize}

	\end{column}
	\vrule{}
	\begin{column}{0.33\textwidth}
		\textbf{Marius model}

		\medskip

		\begin{itemize}
			\item Cities produce economic goods
			\item Economic exchanges are estimated according to gravity flows
			\item Populations grow depending on final economic balances
		\end{itemize}

	\end{column}
\end{columns}

}

\sframe{Implementation and experiments}{

%\footnotesize
\justify

\textbf{Implementation}

Models implemented in \texttt{scala}, using the \texttt{spatialdata} library

\cite{raimbault2020scala}

\url{https://github.com/openmole/spatialdata}

\bigskip

\textbf{Model exploration and validation}

Integrated seamlessly into the OpenMOLE workflow engine

\cite{reuillon2013openmole} for model exploration, sensitivity analysis and validation

\url{https://openmole.org/}

\begin{center}
\includegraphics[width=0.1\linewidth]{figures/iconOM.png}
\includegraphics[width=0.4\linewidth]{figures/openmole.png}
\end{center}

}

\sframe{Calibration of dynamical models}{

\includegraphics[width=\textwidth]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig5.png}

}

\sframe{Indian urban system: direct interactions}{

% systems on which the behavior is particularly interesting ?

\centering
\includegraphics[width=0.85\textwidth]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig6.png}

}

\sframe{Brazilian urban system: multiple factors}{

\centering
\includegraphics[width=0.9\textwidth,height=0.8\textheight]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig7.png}

\footnotesize
\textit{Importance of topography; innovation processes mostly.}

}

\sframe{China: similar models}{

\centering
\includegraphics[width=0.9\textwidth,height=0.8\textheight]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig8.png}

\footnotesize
\textit{No clear best model: other processes in play ? (strong top-down planning)}

}

\sframe{Worldwide calibration of models}{

% new results with the worldwide run

\centering

\includegraphics[height=0.9\textheight]{../../../../Papers/EvolutionaryTheory/paper/figures/Fig9.png}

}

\section{Discussion}

\sframe{Towards multi-scale models}{

\justify

\textit{Crucial aspect of processes at multiple scales and feedback between these \cite{pumain2008socio}; need to be taken into account to build sustainable territorial policies \cite{rozenblat2018conclusion}}

\bigskip

$\rightarrow$ Empirical analysis of contradictory sustainability indicators on endogenous European mega-city regions \cite{raimbault2019multi}

\medskip

$\rightarrow$ A parsimonious multi-scalar urban growth model coupling spatial interactions \cite{raimbault2020indirect} at the macroscopic scale with reaction-diffusion models for urban form \cite{raimbault2018calibration} at the mesoscopic scale in \cite{raimbault:halshs-02351722}

\medskip

$\rightarrow$ A co-evolution model between cities and networks including physical transportation network \cite{raimbault2020hierarchy} (extended with governance processes - \textit{forthcoming talk this Wednesday at CCS2020})

}

\sframe{Discussion}{

% - robustesse des faits stylises / des conclusions theoriques 
% - complementarite
% - historique/politique - dependance au chemin ! pas inertie - adaptation aux contextes changeants - renforcement des hierarchies - temps tres long

\textbf{Synthesis}

$\rightarrow$ robustness of stylized facts, and of theoretical constructions

$\rightarrow$ complementarity of processes and models

$\rightarrow$ importance of the historical/political/geographical context, path-dependency

\medskip

\textbf{Open questions:}

$\rightarrow$ linking urban scaling and dynamical models

$\rightarrow$ endogenous consistent urban systems

$\rightarrow$ multiscale models

% applicable knowledge

\medskip

\textbf{Applications}

$\rightarrow$ Statistical predictability of city growth and size on short time periods

%$\rightarrow$ Largest metropolises are not ``monstruopolises''

$\rightarrow$ Transfer to practitioners: proactive adaptive strategies are necessary (imitation, or anticipation and risk), emulation (co-opetition)

$\rightarrow$ Robustness, variation and sustainability of urban systems (neither norm nor optimum)

}

\sframe{Conclusion}{

\justify

\vspace{-1cm}

$\rightarrow$ Robustness of results regarding data sources, multiple models. \textbf{Need for more systematic model exploration and sensitivity analysis.}

\smallskip

$\rightarrow$ Model complementarity. \textbf{Need for more integrated models.}

\smallskip

$\rightarrow$ Multiple perspectives on urban systems? \textbf{Need for more interdisciplinarity.}

\bigskip

%\footnotesize

\smallskip

\textbf{Open repository} for models and results at

\url{https://github.com/JusteRaimbault/UrbanGrowth}

\url{https://github.com/openmole/spatialdata}

\bigskip

\textbf{Paper} (open access) at \url{https://doi.org/10.3390/su12155954}

\bigskip

\textbf{Acknowledgments}: thanks to the \textit{European Grid Infrastructure} for access to the infrastructure.

}

\sframe{Reserve slides}{

\centering

\Large

\textbf{Reserve Slides}

}

\sframe{Rank-size by continents or trade areas}{

 \begin{center}
    \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    System & Population & Cities & Primacy & Rank-size & R2  \\\hline
Europe & 288Mio & 1067 & 1.45 & $0.93 \pm 0.003$  & 0.991 \\
America & 547Mio & 1521 & 1.02 & $1.02 \pm 0.002$ & 0.996 \\
Asia & 2143Mio & 7737 & 1.12 & $ 0.87 \pm 0.0004$ & 0.998 \\
Africa & 585Mio & 2876 & 1.70 & $0.78 \pm 0.0008$ & 0.997 \\
Oceania & 19Mio & 86 & 1.08 & $0.91 \pm 0.027$ & 0.926\\\hline
\end{tabular}
    \end{center}

    \medskip

    \begin{center}
    \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    System & Population & Cities & Primacy & Rank-size & R2  \\\hline
ASEAN & 293Mio & 874 & 1.67 & $0.92 \pm 0.003$ & 0.993 \\
MERCOSUR & 220Mio & 657 & 1.37 & $1.00 \pm 0.0016$ & 0.998 \\
COMESA & 252Mio & 1367 & 3.39 & $0.72 \pm 0.0014$ & 0.995 \\
EEA & 194Mio & 720 & 1.01 & $0.94 \pm 0.0026$ & 0.994 \\ \hline
\end{tabular}
    \end{center}

\medskip

$\rightarrow$ similar qualitative patterns, but different thematic questions can be tackled

}

\sframe{Scaling by continents or trade areas}{

\footnotesize

\begin{center}
    \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    System & Built-up area & GDP & Emissions \\\hline
Europe & $0.93 \pm 0.016$ (0.76) & $1.12 \pm 0.024$ (0.67) & $1.58 \pm 0.039$ (0.61) \\
America & $1.11 \pm 0.030$ (0.48) & $1.23 \pm 0.027$ (0.57) & $1.69 \pm 0.041$  (0.53) \\
Asia & $1.32 \pm 0.022$ (0.32) & $1.30 \pm 0.016$ (0.47) & $1.78 \pm 0.024$ (0.42) \\
Africa & $1.57 \pm 0.049$ (0.26) & $1.45 \pm 0.043$ (0.29) & $2.04 \pm 0.054$ (0.33) \\
Oceania & $2.56 \pm 0.44$ (0.28) & $1.95 \pm 0.32$ (0.33) & $2.97 \pm 0.44$ (0.34) \\\hline
\end{tabular}
\end{center}

\medskip

\begin{center}
    \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    System & Built-up area & GDP & Emissions \\\hline
ASEAN & $1.26 \pm 0.049$ (0.43) & $1.23 \pm 0.041$ (0.51) & $1.75 \pm 0.067$ (0.44) \\
MERCOSUR & $1.04 \pm 0.040$ (0.50) & $1.15 \pm 0.035$ (0.62) & $1.72 \pm 0.050$ (0.64) \\
COMESA & $1.65 \pm 0.074$ (0.26) & $1.52 \pm 0.072$ (0.26) & $1.93 \pm 0.085$ (0.28) \\
EEA & $0.93 \pm 0.015$ (0.84) & $1.15 \pm 0.019$ (0.83) & $1.50 \pm 0.037$ (0.69)\\\hline
\end{tabular}
\end{center}

}

\sframe{Models settings}{

$\rightarrow$ Work under Gibrat independence assumptions, i.e. $\Covb{P_i\left(t\right)}{P_j\left(t\right)} \textrm{=}0$. If $\vec{P}\left(t\textrm{+}1\right)\textrm{=}\mathbf{R}\cdot \vec{P}\left(t\right)$ where $\mathbf{R}$ is also independent, then $\Eb{\vec{P}\left(t\textrm{+}1\right)}\textrm{=}\Eb{\mathbf{R}}\cdot\Eb{\vec{P}}\left(t\right)$. Consider expectancies only (higher moments computable similarly)

\medskip

$\rightarrow$ With $\vec{\mu}\left(t\right)\textrm{=}\Eb{\vec{P}\left(t\right)}$, we generalize this approach by taking $\vec{\mu}\left(t\textrm{+}1\right)=f\left(\vec{\mu}\left(t\right)\right)$

}

\sframe{Network model}{

Direct network interaction model \cite{raimbault2018indirect}:

\medskip

Let $\vec{\mu}\left(t\right)\textrm{=}\Eb{\vec{P}\left(t\right)}$ cities population and $(d_{ij})$ distance matrix

\medskip

Model specified by

\[
f\left(\vec{\mu}\right) = r_0\cdot \mathbf{Id}\cdot \vec{\mu} \textrm{+} \mathbf{G}\cdot \mathbf{1} \textrm{+} \mathbf{N}
\]

 with 
\begin{itemize}
\item $G_{ij} \textrm{=} w_G\cdot \frac{V_{ij}}{<V_{ij}>}$ and $V_{ij} \textrm{=} \left(\frac{\mu_i\mu_j}{\sum{\mu_k}^2}\right)^{\gamma_G} \exp{\left(-d_{ij}/d_G\right)}$
\item $N_{i} = w_N \cdot \sum_{kl} \left(\frac{\mu_k\mu_l}{\sum\mu}\right)^{\gamma_N}\exp{\left(-d_{kl,i}\right)/d_N}$ where $d_{kl,i}$ is distance to shortest path between $k,l$ computed with slope impedance ($Z\textrm{=}\left(1\textrm{+}\alpha/\alpha_0\right)^{n_0}$ with $\alpha_0\simeq 3$)
\end{itemize}

}

\sframe{Innovation diffusion}{

Favaro-Pumain model \cite{favaro2011gibrat}:

\medskip

1) Diffuse innovations according to 
\[
\delta_{c,i,t} \textrm{=} \frac{\sum_j p_{c,j,t-1}^{s_c} \exp{\left(-\lambda_s d_{ij}\right)}}{\sum_c \sum_j p_{c,j,t-1}^{s_c} \exp{\left(-\lambda_s d_{ij}\right)}}
\]

\medskip

2) Update population with $G_{ij}$ (see network model) such that
\[
V_{ij} \textrm{=} \frac{p_i p_j}{\left(\sum_k p_k\right)^2} \exp{\left(-\lambda_m d_{ij} \prod_c \delta_{c,i}^{\phi_c}\right)}
\]
with $\phi_c \textrm{=} \sum_i p_{i,c}/\sum_{i,c} p_{i,c}$

\medskip

3) Introduce innovation with utility $s_{c \textrm{+} 1} \textrm{=} g_0 \cdot s_c$ in a randomly chosen city with a hierarchy parameter $\alpha_I$, if global adoption share $\phi_c$ is larger than a threshold $\theta_I$. Initial utility $s_0$ is a parameter. New innovation has an initial penetration rate $r_I$ in the city.

}

\sframe{Economic exchanges}{

Marius model \cite{10.1371/journal.pone.0138212}:

\medskip

Initial wealth as a power law of population (exponent $\alpha_W$)

\medskip

1) Update supply and demands as superlinear functions of population (exponents $\alpha_S,\alpha_D$)

\medskip

2) Exchange goods according to a gravity potential of interaction (distance decay $d_M$), supplies and demands; update wealth accordingly

\medskip

3) Update population such that population difference is a power law of wealth difference (economic multiplier $e_M$ and exponent $\alpha_P$)

}

\sframe{Benchmarked models}{

\begin{enumerate}
	\item Gibrat model: 1 param. $r_0$
	\item Direct interaction model (geographical distance): 4 param. $r_0,w_G,\gamma_G,d_G$
	\item Physical network interaction model (topographical distance: 4 param. $r_0,w_G,\gamma_G,d_G$
	\item Innovation diffusion model (simplified): 4 param. $r_0,w_I,\lambda_s,\lambda_m$ (other parameters at default values from \cite{favaro2011gibrat})
	\item Innovation diffusion model (full): 9 param. $r_0,w_I,\lambda_s,\lambda_m,s_0,g_0,r_I,\alpha_I,\theta_I$
	\item Restricted Marius model: 4 param. $e_M,\alpha_S,\alpha_D,d_M$
	\item Marius model: 6 param. $e_M,\alpha_S,\alpha_D,d_M,\alpha_W,\alpha_P$
\end{enumerate}

}

%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{biblio}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2021-01-29 Docs/Papers/HandbookEntropy/chapter_v6_20201214.tex
handbook_entropy

--- 2020-10-28 Docs/Papers/HandbookEntropy/arxiv/v1/chapter.tex
@@
 
-\title{Spatial dynamics of complex urban systems within an evolutionary theory frame}
+\title{\textbf{Handbook on ``Entropy, Complexity, and Spatial Dynamics: The Rebirth of Theory?''}\medskip\\
+Chapter by Juste Raimbault\textsuperscript{1} and Denise Pumain\textsuperscript{2}\medskip\\
+\textbf{Spatial dynamics of complex urban systems within an evolutionary theory frame}}
 
@@
 
-\author{
-Juste Raimbault\textsuperscript{1,2,3*},
-Denise Pumain\textsuperscript{3}\bigskip\\
-\textsuperscript{1} Centre for Advanced Spatial Analysis, UCL, London, United Kingdom
-\\
-\textsuperscript{2} UPS CNRS 3611 ISC-PIF, Paris, France
-\\
-\textsuperscript{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France\bigskip\\
-* \texttt{juste.raimbault@polytechnique.edu}
-}
+%\author{
+%Juste Raimbault\textsuperscript{1,2,3*},
+%Denise Pumain\textsuperscript{3}\bigskip\\
+%\textsuperscript{1} Centre for Advanced Spatial Analysis, UCL, London, United Kingdom
+%\\
+%\textsuperscript{2} UPS CNRS 3611 ISC-PIF, Paris, France
+%\\
+%\textsuperscript{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France\bigskip\\
+%* \texttt{juste.raimbault@polytechnique.edu}
+%}
+
+\author{\textsuperscript{1}Research Fellow, Center for Advanced Spatial Analysis,\\
+University College London. ORCID: 0000-0003-0768-9480\\
+\textsuperscript{2} Emeritus Professor, Universit{\'e} Paris I Panth{\'e}on-Sorbonne,\\
+UMR G{\'e}ographie-cit{\'e}s. ORCID : 0000-0002-0730-4563}
 
@@
 
-	\textbf{Keywords:} spatial dynamics; complex systems; system of cities; evolutionary theory; urban growth; simulation
+	\textbf{Keywords: spatial dynamics; complex systems; system of cities; evolutionary theory; urban growth; simulation}
 \end{abstract}
@@
 
-The context of this chapter, namely ``Entropy, Complexity, and Spatial Dynamics: The Rebirth of Theory'' is challenging.
+The title chosen for this book: ``Entropy, Complexity, and Spatial Dynamics: The Rebirth of Theory'' is challenging.
 The problem is twofold: first, it means demonstrating how the insights provided by concepts and models associated with the notion of complexity have contributed to theories on spatial dynamics; second, it questions whether this is a rebirth of the theoretical approach in this field, which would therefore presume its relative abandon before.
@@
 We may add that many scholars did not abandon the former urban theories but on contrary tested, revised and completed them during the last four decades \citep{pumain1997pour,pumain1998urban,pumain2003approche,pumain2020theories,batty2013new}.
-The major theoretical challenge of that period was three folds: first, to shift from idiographic postures to nomothetic ones; second, from static views toward dynamic and evolutionary; and third, to really transfer concepts and models from natural sciences to build a socially relevant knowledge.
+The major theoretical challenge of that period was three fold: first, to shift from idiographic postures to nomothetic ones; second, from static views toward dynamic and evolutionary; and third, to really transfer concepts and models from natural sciences to build a socially relevant knowledge.
 Elements of knowledge from ancient theories can consolidate when situated in their geo-historical contexts, when tests of these theoretical propositions are carried out on subsets of empirical data that are well identified in time and space, and when the means are provided to articulate the observations made at multiple levels of geographic scales.
@@
 This theory is abductive in the sense that it is constructed by frequent to-and-fro movements between empirical observations, logical propositions and mathematical and computer models.
-This theory is not entirely new as it integrates elements of knowledge already well identified by specialists in urban issues, such as geographers, economists, historians and archaeologists, sociologists, town planners and architects.
+This theory is not entirely new as it integrates elements of knowledge already well identified by specialists in urban issues, such as geographers, economists, historians, archaeologists, sociologists, town planners and architects.
 Its originality stems from progress in international comparisons made through using new harmonized databases, methods for validating the results of computer simulation models, and putting these results into the perspective of complex systems theories.
@@
 The interest of this model of spatially distributed urban growth in integrated systems of cities is to predict that the result of such a growth process is always a lognormal distribution.
-Thus, the model provides a first statistical explanation to the ``mystery'' of the Zipf's law still recently mentioned by the economist Paul \cite{krugman1996confronting}.
+Thus, the model provides a first statistical explanation to the ``mystery'' of the Zipf's law \citep{zipf1941national} still recently mentioned by the economist Paul \cite{krugman1996confronting}.
 A huge literature is dedicated to Zipf’s law and Pareto distributions that are observed in many natural and social systems, to the extent that these ``universal'' laws, together with fractal spatial patterns, have been sometimes considered as the ``signature'' of complexity in different systems.
@@
 The resulting cacophony is sometimes disentangled in courageous articles, such as for instance \citep{cottineau2017metazipf,cottineau2020metametazipf} in a meta-review of hundreds of references, but certainly represents a waste of time and efforts that could be avoided.
-A recent paper seemed promising in announcing the observation of worldwide trends in urbanization from 1950 to 2030 using some 1857 cities larger than 300 000 inhabitants from 155 countries \citep{egidi2020long}.
+A recent paper seemed promising in announcing the observation of worldwide trends in urbanization from 1950 to 2030 using some 1857 cities larger than 300,000 inhabitants from 155 countries \citep{egidi2020long}.
 The paper offers surprising results such as a systematic ``inverse relationship'' between urban growth and city size in all regions ``up to the late 1990s''.
@@
 Although the recent development of networks and the accelerating circulation of information seem to blur the three-level conceptual definition of these geographical objects, we believe that this representation remains useful and relevant for guiding urban research and applications.
-An interesting contribution by C.
-\cite{roth2006reconstruction} discusses the emergentist view compared to the reductionist and suggests that different modes of access and not only bottom- up interactions could be part of the generative processes of the features constituting one level – which is the view point from statistical mechanics in envisaging self-organizing systems.
+An interesting contribution by Camille \cite{roth2006reconstruction} discusses the emergentist view compared to the reductionist and suggests that different modes of access and not only bottom- up interactions could be part of the generative processes of the features constituting one level – which is the view point from statistical mechanics in envisaging self-organizing systems.
 This rejoins our theory for which urban interactions shaping cities and systems of cities are mostly but not exclusively conceived as emanating from one low level toward the higher level, in societal processes that take a long time, often longer than a human life.
@@
 While trying to identify major processes guiding the evolution of cities within system of cities, we stressed the role of information and innovation \citep{lane2009complexity}.
-Innovation when defined as a socially accepted invention may represent a form of social organization or a belief or a practice as well as a technological device or a new service.
+Innovation when defined as a socially accepted invention may represent a form of social organization or a belief or a practice as well as a technological device or a new service \citep{broitman2020endogenous}.
 We shall develop further in the models presented below how innovation and its more or less always hierarchical diffusion among cities within systems of cities appears as a major dynamic process for explaining the expansion of urbanization and the persistency of cities’ relative situation in urban systems.
@@
 
+%\bibliographystyle{apalike}
+%\bibliography{biblio}
+
 \begin{thebibliography}{}
@@
   38(3):731--755.
+
+\bibitem[Broitman and Czamanski, 2020]{broitman2020endogenous}
+Broitman, D. and Czamanski, D. (2020).
+\newblock Endogenous growth in a spatial economy: The impact of globalization
+  on innovations and convergence.
+\newblock {\em International Regional Science Review}, page 0160017620946095.
 
@@
 
+\bibitem[Zipf, 1941]{zipf1941national}
+Zipf, G. (1941).
+\newblock {\em National unity and disunity}.
+\newblock Bloomington Principia Press (Indiana).
+
 \end{thebibliography}

### 2021-01-29 Docs/Papers/HandbookEntropy/figures.sh

@@ -8,10 +8,12 @@ VERTICALPADDING=10
 
 convert figuresraw/coevol_leg.png -gravity center -extent 3500x figuresraw/tmpleg.png
 montage figuresraw/coevol_1.png figuresraw/coevol_2.png figuresraw/coevol_3.png figuresraw/coevol_4.png -tile 2x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x -border 2 -bordercolor Black figuresraw/tmp1.png
-montage figuresraw/tmp1.png figuresraw/tmpleg.png -tile 1x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x figures/Fig1.jpg
-rm figuresraw/tmp1.png figuresraw/tmpleg.png
+montage figuresraw/tmp1.png figuresraw/tmpleg.png -tile 1x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x figures/Fig1tmp.jpg
+convert figures/Fig1tmp.jpg -set colorspace Gray -separate -average Fig1.jpg
+rm figuresraw/tmp1.png figuresraw/tmpleg.png figures/Fig1tmp.jpg
 
 convert figuresraw/innovation_leg.png -gravity center -extent 3500x figuresraw/tmpleg.png
 montage figuresraw/innovation_1.png figuresraw/innovation_2.png figuresraw/innovation_3.png figuresraw/innovation_4.png -tile 2x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x -border 2 -bordercolor Black figuresraw/tmp2.png
-montage figuresraw/tmp2.png figuresraw/tmpleg.png -tile 1x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x figures/Fig2.jpg
-rm figuresraw/tmp2.png figuresraw/tmpleg.png
+montage figuresraw/tmp2.png figuresraw/tmpleg.png -tile 1x2 -geometry +"$HORIZONTALPADDING"+"$VERTICALPADDING" -resize "$WIDTH"x figures/Fig2tmp.jpg
+convert figures/Fig2tmp.jpg -set colorspace Gray -separate -average Fig2.jpg
+rm figuresraw/tmp2.png figuresraw/tmpleg.png figures/Fig2tmp.jpg

### 2021-01-29 Docs/Papers/Multiscale/paper.tex
paper_evolth
citygrwoth_abstract
paper multiscale

@@
 
-\title{Multiscalar modeling of urban growth}
+\title{Strong coupling between scales in a multi-scalar model of urban dynamics}
+
+\author{Juste Raimbault$^{1,2,3,\ast}$\medskip\\
+$^{1}$ Center for Advanced Spatial Analysis, University College London\\
+$^{2}$ UPS CNRS 3611 ISC-PIF\\
+$^{3}$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\medskip\\
+$^{\ast}$ \texttt{juste.raimbault@polytechnique.edu}
+}
+
 \date{}
@@
 	Urban evolution processes occur at different scales, with intricate interactions between levels and relatively distinct type of processes.
-To what extent actual urban dynamics include an actual strong coupling between scales, in the sense of both top-down and bottom-up feedbacks, remains an open issue with important practical application for the management of urban growth.
-We introduce in this paper a multi-scalar simulation model of urban growth, coupling an interaction model at the macroscopic scale with morphogenesis models for the evolution of urban form at the scale of metropolitan areas.
+To what extent actual urban dynamics include an actual strong coupling between scales, in the sense of both top-down and bottom-up feedbacks, remains an open issue with important practical implications for the sustainable management of territories.
+We introduce in this paper a multi-scalar simulation model of urban growth, coupling a system of cities interaction model at the macroscopic scale with morphogenesis models for the evolution of urban form at the scale of metropolitan areas.
+Strong coupling between scales is achieved through an update of model parameters at each scale depending on trajectories at the other scale.
+The model is applied and explored on synthetic systems of cities.
+Simulation results show a non-trivial effect of the strong coupling.
+As a consequence, an optimal action on policy parameters such as containing urban sprawl is shifted.
+We also run a multi-objective optimization algorithm on the model, showing showing that compromise between scales are captured.
+Our approach opens new research directions towards more operational urban dynamics models including a strong feedback between scales.
+	\medskip\\
+\textbf{Keywords: } Urban dynamics; Systems of cities; Urban morphogenesis; Multi-scalar modeling; Strong coupling
 \end{abstract}
+
+%  relevant target CEUS?
+https://www.elsevier.com/journals/computers-environment-and-urban-systems/0198-9715/guide-for-authors => special issue expired; IJGI 31/01:  https://www.mdpi.com/journal/ijgi/special_issues/spatio_temporal_models_geo
 
@@
 
+% abstract CCS
+% 
+%The modeling of urban growth is a crucial issue for the design of sustainable territorial policies, through the understanding of past urbanization processes and the forecasting of future urban trajectories.
+Several models have been proposed at different scales and integrating differ- ent dimensions of urban systems, such as land-use transport interaction models [1] or systems of cities models [2].
+While multi-scalar models are recognized as crucial for the study of such systems [3], they remain in practice unexplored.
+%This contribution introduces a parsimonious multi-scalar model for systems of cities, based on simple dimensions (mainly populations) with stylized processes, but yielding an effective strong coupling between the metropolitan mesoscopic scale and the macroscopic scale of the system of cities.
+The model couples the spatial interaction model of [4] for the macro scale with the reaction-diffusion model for urban form studied by [5].
+More precisely, urban areas viewed as a population grid are embedded into the macroscopic interaction model.
+To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved with the reaction-diffusion models at a given speed conditionally to the population variations; (iv) changes in urban form influence macroscopic interaction ranges (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.
+%The model is applied on synthetic systems of cities typical of a continental range (500km, hierarchy around 1, 20 cities), with initial local population grid configurations as monocen- tric.
+Parameter space is explored with the OpenMOLE model exploration software [6], eased by the implementation of the model in scala [7].
+First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a “transit-oriented development” scenario (negative feedback of population growth on diffusion) to a “sprawl” scenario (positive feedback).
+Similarly, mesoscopic urban form indicators are significantly influenced by the coupling process.
+%Further work will consist in more targeted simulation experiments, including specific ex- ploration algorithms such as diversity search for model regimes [6], to test the model as a proof-of-concept of models for policies.
+Such a model can also be calibrated on real city sys- tems and urban form trajectories, to extrapolate coupling parameters that would be difficult to obtain otherwise.
+Our contribution is thus a first step towards multi-scalar simulation models for systems of cities.
+
+% multilevel models : check https://scholar.google.fr/citations?user=x1BwVXoAAAAJ&hl=fr
+% + Morvan, G. (2014). Bibliographic database on multi-level and multi-scale agent-based modeling. Transportation Research Procedia, 2, 495-500.
+
+%\cite{louail2010comparer}
+
 \section{Introduction}
 
-The modeling of urban growth is a crucial issue for the design of sustainable territorial policies, through the understanding of past urbanization processes and the forecasting of future urban trajectories.
-Several models have been proposed at different scales and integrating different dimensions of urban systems, such as land-use transport interaction models \cite{wegener2004land} or systems of cities models \cite{pumain2017urban}.
-While multi-scalar models are recognized as crucial for the study of such systems \cite{Rozenblat2018}, they remain in practice unexplored.
+The modeling of urban growth and more generally the dynamics of urban systems is central to the design of sustainable territorial policies, through the understanding of past urbanisation processes and the anticipation of future urban trajectories.
+The design of sustainable future cities requires an historical knowledge of how past cities came to be and evolved \cite{batty2018inventing}.
+Several models have been proposed at different scales and integrating different dimensions of urban systems, such as models of land-use change at a mesoscopic scale or systems of cities models at a macroscopic scale \cite{pumain2017urban}.
 
-At the scale of a metropolitan area, Land-use Transport Interaction models \cite{wegener2004land} are for example a privileged tool to anticipate the answer of spatial distributions of activities (mostly residential location and economic activities) to an evolution of the accessibility landscape permitted by new transportation infrastructures.
-At the same scale, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}.
+At the scale of a metropolitan area, Land-use Transport Interaction models \cite{wegener2004land} are for example a widely used tool to estimate the dynamics of spatial distributions of activities (mostly residential location and economic activities) in response to an evolution of the accessibility landscape caused by new transportation infrastructures \cite{raimbault:halshs-02265423}.
+In a similar context, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}.
 At the smaller scale of the system of cities, macroscopic models of urban growth have focused on reproducing the distribution of city sizes, either through economic processes as e.g.
@@
 Territorial dynamics, and more particularly urban dynamics, have according to \cite{pumain1997pour} an intrinsic multi-scalar nature, with successive autonomous levels of emergence from individual microscopic agents to the mesoscopic scale of the city and the macroscopic scale of the system of cities.
-Furthermore, the need for sustainable territorial policies would imply the construction of multi-scalar models to take into account issues associated to each relevant scale \cite{Rozenblat2018}.
+While models at each scale with distinct ontologies are useful to answer their own questions, an explicit account of inter-scale feedbacks, both top-down and bottom-up, would allow testing policies and interventions distributed and differentiated across scales while not neglecting the interactions between scales \cite{wegener2018multi}.
+Indeed, the need for sustainable territorial policies would imply the construction of multi-scalar models to take simultaneously into account issues associated to each relevant scale \cite{Rozenblat2018,raimbault:halshs-02284933}.
 
-This paper contributes to that open question by introducing a multi-scale model of urban growth which focuses on the spatial structure of processes rather than on their multi-dimensionality.
-Therefore, we take into account only population variables, but both at the macroscopic scale of the system of cities in the legacy of \cite{pumain2017urban} and at the mesoscopic scale of the metropolitan area with an urban morphogenesis model.
-The coupling of these scales is a crucial novel feature of our model.
-We describe in the following stylized facts justifying the approach, describe the model, and summarize preliminary results from its exploration and calibration.
+% rq: no def nor clarification of what is meant by "scale"!
 
-\section{Multi-scale model}
+Multi-scalar models of urban dynamics are however still at their beginnings.
+\cite{murcio2015urban} consider population flows at different spatial ranges from the urban area to the country, but does not incorporate distinct ontologies and processes for the different scales.
+\cite{batty2005agents} however suggests that a similar formalism can be applied to urban processes at different scales.
+Multi-level statistical models capture some information at imbricated scales \cite{shu2020modelling}, although they can not be used as dynamical simulation models.
+Similarly, multi-level cellular automata (CA) models for urban growth include factors influencing urban expansion at multiple scales \cite{xu2019directional}.
+\cite{cheng2003modelling} propose a general framework for such approaches.
+\cite{torrens2001cellular} suggest that hybrid models coupling CA with other formalisms is a crucial development in the field.
+\cite{white2006modeling} introduces a CA with variable grid size to account for heterogeneities across scales.
+\cite{zhu2020dynamic} couple an agent-based model with a CA at multiple scales.
+\cite{yu2018modeling} embed a local CA into a regional intercity model and a macroscopic potential model.
+\cite{ford2019multi} couple at different scales an urban development model with a flooding risk model to forecast the future impact of extreme climate events on the London metropolitan region.
+\cite{xu2020urban} develop an agent-based model of urban expansion with both macro and micro agents.
+\cite{raimbault:halshs-02013006} suggests that integrating network dynamics at the link level in a macroscopic urban system models is a way to implement a multi-scale model, as done by \cite{raimbault2020hierarchy} which explores hierarchy properties of cities and networks in this context.
+
+In disciplines neighbour to urban modeling, methods have been developed for multi-scale models.
+For example in spatial epidemiology, \cite{banos2015importance} combines agent-based modeling for local diffusion dynamics with differential equations at the population scale.
+The NetLogo software for agent-based modeling includes a specific extension for multi-scale modeling \cite{hjorth2020levelspace}.
+Multi-scale models have also been used for the simulation of crowd dynamics \cite{crociani2016multi}.
+The study of traffic is also made more accurate by coupling macroscopic and microscopic models \cite{boulet2020coupling}.
+The management of ecosystems  requires integrating across actors and scales \cite{belem2013organizational}.
+These non-exhaustive illustrations highlight how the integration of scales is a crucial feature and issue in the understanding of complex systems \cite{chavalarias2009french}.
+
+This paper contributes to the open question of multi-scalar models of urban dynamics by introducing a new simulation model which integrates a strong coupling between the mesoscopic scale and the macroscopic scale.
+The dynamics within each scale influence the other and reciprocally in an iterative way.
+More precisely, the model is simple in its components as we focus on the spatial structure of processes rather than on their multi-dimensionality.
+Therefore, we take into account only population variables, but both at the macroscopic scale of the system of cities and at the mesoscopic scale of the metropolitan area with an urban morphogenesis model.
+Our contribution is novel regarding previous works in particular regarding the following points: (i) the stylised model explicitly couples distinct scales and ontologies in a strong manner, most models operating only a weak coupling between scales (i.e.
+no reciprocal and dynamical feedbacks); (ii) the behavior of the model is systematically studied on synthetic systems of cities using model exploration methods.
+
+% rq positioning: not large model but simple whih can be understood (cf requiem for large scale models)
+
+The rest of this paper is organised as follows: we first describe the model; we then develop its exploration on synthetic systems of cities, and optimization using a genetic algorithm; we finally discuss developments and implications of this work.
+
+% We describe in the following stylized facts justifying the approach => this could (should?) be an important development/basis for the work?
+
+%Strong coupling and multi-scalarity 
+%   Distinct processes and objects within each different scale
+%   Weak inter-scale coupling, such as progressive resolution for land-use model, does not considers emergence and autonomous scales
+% An integrated model, or strongly coupled, would be \textit{a new model extending the two coupled models in the sense that it includes them in some parameter settings or limit conditions}
+%Urban multi-scalar complexity must be captured by a strongly coupled model
+%no strongly coupled multi-scalar model in the literature (few examples such as  \cite{murcio2015urban} but without distinct ontologies)
+%need to however consider ``simple'' models to be able to understand their behavior and extract knowledge from them
+%\textbf{Research objective: }
+%\textit{Investigate a strong coupling of a simple urban system interaction network at the macroscopic scale with an urban morphogenesis model at the mesoscopic scale}
+
+\section{Multi-scale urban dynamics model}
 
 \subsection{Rationale}
+
+% ! reformulate - this is the CCS abstract
 
@@
 To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved with the reaction-diffusion models at a given speed conditionally to the population variations; (iv) changes in urban form influence macroscopic interaction ranges (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.
+
+% TODO figure
+% \includegraphics{figures/multiscale_morph.pdf}
 
@@
 
+\subsection{Synthetic setup}
+
+Model applied on synthetic systems of cities:
+
+\begin{itemize}
+	\item random positions and rank-size hierarchy ($\alpha \textrm{ = } 1.0$ and $P_0 \textrm{ = } 100,000$)
+	\item countrywide urban system scale: 500km and 20 cities
+	\item initial population grids as monocentric (grid of size 50 and center cell density 1000 units)
+	\item simulated for 20 macroscopic time steps (order of magnitude of half a century)
+\end{itemize}
+
 \subsection{Indicators}
@@
 Model behavior is characterized using the following indicators:
+
+\textit{Indicators at the macroscopic scale: distributions of population, accessibilities, centralities (summarized by average, hierarchy, entropy)}
+
+\textit{Indicators at the mesoscopic scale: urban form captured by Moran index, average distance, hierarchy, entropy}
 
@@
 
+\textit{Performance constraints: simulate $N$ mesoscopic morphogenesis models in parallel (macroscopic interactions are efficient as based on matrices)}
+
+ model implemented in \texttt{scala} and integrated within a broader library (including implementations of \cite{raimbault2018indirect} \cite{raimbault2018calibration}
+
+\cite{favaro2011gibrat} \cite{cottineau2015modular})
+
+\textit{Large number of parameters and output indicators}
+
+Model: \texttt{https://github.com/JusteRaimbault/UrbanGrowth-model}
+  Project: \texttt{https://github.com/JusteRaimbault/UrbanGrowth}
+  Simulation data: \texttt{https://doi.org/10.7910/DVN/IRHMQK}
+
+\subsection{Statistical consistency}
+
+% one factor sampling
+
+%(min around 1 for 3 indicators, above 5 for 2 and around 0.5 for 2)}
+
+\medskip
+
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/allindics_hist.png}
+\end{center}
+
+\textit{For all indicators, median sharpe ratios computing for a parameter point across repetitions are all larger than 1.6}
+
 \subsection{Grid exploration}
@@
 Similarly, mesoscopic urban form indicators are significantly influenced by the coupling process.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
+\end{figure}
+
+\textit{U-shape behavior of both macroscopic and mesoscopic indicators as a function of $\delta d$}
 
@@
 %  -> first thing to check : with/ without the model
+
+\begin{figure}
+\includegraphics[height=0.75\textheight]{figures/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
+\end{figure}
+
+\textit{Non-trivial influence of coupled feedbacks on the different scales}
+
+\subsection{Impact of policy parameters}
+
+\begin{figure}
+\includegraphics[height=0.7\textheight]{figures/deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+\end{figure}
+
+\textit{Mesoscopic centralization appears at a $\delta \beta$ critical value for low $\delta \alpha$; influenced by upward feedback}
+
+\begin{figure}
+\includegraphics[height=0.7\textheight]{figures/deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+\end{figure}
+
+\textit{Mesoscopic hierarchy has a U-shape of $\delta \beta$ in negative $\delta \alpha$, but has a plateau for positive values} 
+%$\rightarrow$ sprawl is needed in a sense
+
+\subsection{Multiscale optimization}
+
+\begin{figure}
+	\includegraphics[width=0.8\textheight]{figures/pareto_macroHierarchy-mesoHierarchy.png}
+\end{figure}
+
+\textit{Pareto front obtained with Genetic Algorithm optimization for two contradictory objective of macrosopic and mesoscopic hierarchies}
 
@@
 
+\textbf{Theoretical and practical implications}
+
+$\rightarrow$ model effectively captures an interaction between downward and upward feedback: weak emergence \cite{bedau2002downward}
+
+$\rightarrow$ coupling ``simple three parameters models'' yield a complicated and complex simulation model: necessity of complexity and simulation models to understand urban complexity? 
+% 'suggest in a way failure of reductionist empistemology (in the sense of a la Barthelemy) - at least for grabbing the multi-scalar nature of systems' (seems a tautology though
+
+$\rightarrow$ progressive integration towards models for policy?
+
+\textbf{Developments}
+
+$\rightarrow$ diversity search algorithm to find e.g. regimes with the strongest effect of feedback
+
+$\rightarrow$ parametrization on real systems; possibly calibration
+
+\cite{raimbault2019ilus}
+
+\section{Conclusion}
+
+A first step towards \textit{strongly} multi-scalar models to capture urban complexity towards policy models
+
+ Towards integrative models and theories for urban systems \cite{raimbault2019methods}
+
+\textbf{Acknowledgments}: thanks to the \textit{European Grid Infrastructure} for access to the infrastructure.
+
 \bibliographystyle{apalike}

### 2021-01-31 Docs/Papers/Multiscale/arxiv/paper.tex
arxiv

\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[margin=2cm]{geometry}
\usepackage{multirow}

\title{Strong coupling between scales in a multi-scalar model of urban dynamics}

\author{Juste Raimbault$^{1,2,3,\ast}$\medskip\\
$^{1}$ Center for Advanced Spatial Analysis, University College London\\
$^{2}$ UPS CNRS 3611 ISC-PIF\\
$^{3}$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\medskip\\
$^{\ast}$ \texttt{juste.raimbault@polytechnique.edu}
}

\date{}

\begin{document}

\maketitle

\begin{abstract}
	Urban evolution processes occur at different scales, with intricate interactions between levels and relatively distinct type of processes. To what extent actual urban dynamics include an actual strong coupling between scales, in the sense of both top-down and bottom-up feedbacks, remains an open issue with important practical implications for the sustainable management of territories. We introduce in this paper a multi-scalar simulation model of urban growth, coupling a system of cities interaction model at the macroscopic scale with morphogenesis models for the evolution of urban form at the scale of metropolitan areas. Strong coupling between scales is achieved through an update of model parameters at each scale depending on trajectories at the other scale. The model is applied and explored on synthetic systems of cities. Simulation results show a non-trivial effect of the strong coupling. As a consequence, an optimal action on policy parameters such as containing urban sprawl is shifted. We also run a multi-objective optimization algorithm on the model, showing showing that compromise between scales are captured. Our approach opens new research directions towards more operational urban dynamics models including a strong feedback between scales.
	\medskip\\
\textbf{Keywords: } Urban dynamics; Systems of cities; Urban morphogenesis; Multi-scalar modeling; Strong coupling
\end{abstract}

%  relevant target CEUS? https://www.elsevier.com/journals/computers-environment-and-urban-systems/0198-9715/guide-for-authors => special issue expired; IJGI 31/01:  https://www.mdpi.com/journal/ijgi/special_issues/spatio_temporal_models_geo

% Systems of cities
%Multi-scalar model
%Spatial interaction
%Reaction-diffusion

% abstract CCS
% 
%The modeling of urban growth is a crucial issue for the design of sustainable territorial policies, through the understanding of past urbanization processes and the forecasting of future urban trajectories. Several models have been proposed at different scales and integrating differ- ent dimensions of urban systems, such as land-use transport interaction models [1] or systems of cities models [2]. While multi-scalar models are recognized as crucial for the study of such systems [3], they remain in practice unexplored.
%This contribution introduces a parsimonious multi-scalar model for systems of cities, based on simple dimensions (mainly populations) with stylized processes, but yielding an effective strong coupling between the metropolitan mesoscopic scale and the macroscopic scale of the system of cities. The model couples the spatial interaction model of [4] for the macro scale with the reaction-diffusion model for urban form studied by [5]. More precisely, urban areas viewed as a population grid are embedded into the macroscopic interaction model. To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved with the reaction-diffusion models at a given speed conditionally to the population variations; (iv) changes in urban form influence macroscopic interaction ranges (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.
%The model is applied on synthetic systems of cities typical of a continental range (500km, hierarchy around 1, 20 cities), with initial local population grid configurations as monocen- tric. Parameter space is explored with the OpenMOLE model exploration software [6], eased by the implementation of the model in scala [7]. First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a “transit-oriented development” scenario (negative feedback of population growth on diffusion) to a “sprawl” scenario (positive feedback). Similarly, mesoscopic urban form indicators are significantly influenced by the coupling process.
%Further work will consist in more targeted simulation experiments, including specific ex- ploration algorithms such as diversity search for model regimes [6], to test the model as a proof-of-concept of models for policies. Such a model can also be calibrated on real city sys- tems and urban form trajectories, to extrapolate coupling parameters that would be difficult to obtain otherwise. Our contribution is thus a first step towards multi-scalar simulation models for systems of cities.

% multilevel models : check https://scholar.google.fr/citations?user=x1BwVXoAAAAJ&hl=fr
% + Morvan, G. (2014). Bibliographic database on multi-level and multi-scale agent-based modeling. Transportation Research Procedia, 2, 495-500.

%\cite{louail2010comparer}

\section{Introduction}

The modeling of urban growth and more generally the dynamics of urban systems is central to the design of sustainable territorial policies, through the understanding of past urbanisation processes and the anticipation of future urban trajectories. The design of sustainable future cities requires an historical knowledge of how past cities came to be and evolved \cite{batty2018inventing}. Several models have been proposed at different scales and integrating different dimensions of urban systems, such as models of land-use change at a mesoscopic scale or systems of cities models at a macroscopic scale \cite{pumain2017urban}.

At the scale of a metropolitan area, Land-use Transport Interaction models \cite{wegener2004land} are for example a widely used tool to estimate the dynamics of spatial distributions of activities (mostly residential location and economic activities) in response to an evolution of the accessibility landscape caused by new transportation infrastructures \cite{raimbault:halshs-02265423}. In a similar context, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}. At the smaller scale of the system of cities, macroscopic models of urban growth have focused on reproducing the distribution of city sizes, either through economic processes as e.g. \cite{gabaix1999zipf}, or from a geographical point of view focusing on interactions between cities \cite{favaro2011gibrat}.

Territorial dynamics, and more particularly urban dynamics, have according to \cite{pumain1997pour} an intrinsic multi-scalar nature, with successive autonomous levels of emergence from individual microscopic agents to the mesoscopic scale of the city and the macroscopic scale of the system of cities. While models at each scale with distinct ontologies are useful to answer their own questions, an explicit account of inter-scale feedbacks, both top-down and bottom-up, would allow testing policies and interventions distributed and differentiated across scales while not neglecting the interactions between scales \cite{wegener2018multi}. Indeed, the need for sustainable territorial policies would imply the construction of multi-scalar models to take simultaneously into account issues associated to each relevant scale \cite{Rozenblat2018,raimbault:halshs-02284933}.

% rq: no def nor clarification of what is meant by "scale"!

Multi-scalar models of urban dynamics are however still at their beginnings. \cite{murcio2015urban} consider population flows at different spatial ranges from the urban area to the country, but does not incorporate distinct ontologies and processes for the different scales. \cite{batty2005agents} however suggests that a similar formalism can be applied to urban processes at different scales. Multi-level statistical models capture some information at imbricated scales \cite{shu2020modelling}, although they can not be used as dynamical simulation models. Similarly, multi-level cellular automata (CA) models for urban growth include factors influencing urban expansion at multiple scales \cite{xu2019directional}. \cite{cheng2003modelling} propose a general framework for such approaches. \cite{torrens2001cellular} suggest that hybrid models coupling CA with other formalisms is a crucial development in the field. \cite{white2006modeling} introduces a CA with variable grid size to account for heterogeneities across scales. \cite{zhu2020dynamic} couple an agent-based model with a CA at multiple scales. \cite{yu2018modeling} embed a local CA into a regional intercity model and a macroscopic potential model. \cite{ford2019multi} couple at different scales an urban development model with a flooding risk model to forecast the future impact of extreme climate events on the London metropolitan region. \cite{xu2020urban} develop an agent-based model of urban expansion with both macro and micro agents. \cite{raimbault:halshs-02013006} suggests that integrating network dynamics at the link level in a macroscopic urban system models is a way to implement a multi-scale model, as done by \cite{raimbault2020hierarchy} which explores hierarchy properties of cities and networks in this context.

In disciplines neighbour to urban modeling, methods have been developed for multi-scale models. For example in spatial epidemiology, \cite{banos2015importance} combines agent-based modeling for local diffusion dynamics with differential equations at the population scale. The NetLogo software for agent-based modeling includes a specific extension for multi-scale modeling \cite{hjorth2020levelspace}. Multi-scale models have also been used for the simulation of crowd dynamics \cite{crociani2016multi}. The study of traffic is also made more accurate by coupling macroscopic and microscopic models \cite{boulet2020coupling}. The management of ecosystems  requires integrating across actors and scales \cite{belem2013organizational}. These non-exhaustive illustrations highlight how the integration of scales is a crucial feature and issue in the understanding of complex systems \cite{chavalarias2009french}.

This paper contributes to the open question of multi-scalar models of urban dynamics by introducing a new simulation model which integrates a strong coupling between the mesoscopic scale and the macroscopic scale. The dynamics within each scale influence the other and reciprocally in an iterative way. More precisely, the model is simple in its components as we focus on the spatial structure of processes rather than on their multi-dimensionality. Therefore, we take into account only population variables, but both at the macroscopic scale of the system of cities and at the mesoscopic scale of the metropolitan area with an urban morphogenesis model. Our contribution is novel regarding previous works in particular regarding the following points: (i) the stylised model explicitly couples distinct scales and ontologies in a strong manner, most models operating only a weak coupling between scales (i.e. no reciprocal and dynamical feedbacks); (ii) the behavior of the model is systematically studied on synthetic systems of cities using model exploration methods.

% rq positioning: not large model but simple whih can be understood (cf requiem for large scale models)

The rest of this paper is organised as follows: we first describe the model; we then develop its exploration on synthetic systems of cities, and optimization using a genetic algorithm; we finally discuss developments and implications of this work.

% We describe in the following stylized facts justifying the approach => this could (should?) be an important development/basis for the work?

%Strong coupling and multi-scalarity 
%   Distinct processes and objects within each different scale
%   Weak inter-scale coupling, such as progressive resolution for land-use model, does not considers emergence and autonomous scales
% An integrated model, or strongly coupled, would be \textit{a new model extending the two coupled models in the sense that it includes them in some parameter settings or limit conditions}
%Urban multi-scalar complexity must be captured by a strongly coupled model
%no strongly coupled multi-scalar model in the literature (few examples such as  \cite{murcio2015urban} but without distinct ontologies)
%need to however consider ``simple'' models to be able to understand their behavior and extract knowledge from them
%\textbf{Research objective: }
%\textit{Investigate a strong coupling of a simple urban system interaction network at the macroscopic scale with an urban morphogenesis model at the mesoscopic scale}

\section{Multi-scale urban dynamics model}

\subsection{Rationale}

% ! reformulate - this is the CCS abstract

The main characteristic of our model is a strong coupling between the macroscopic scale of a system of cities with the mesoscopic scale of metropolitan areas. We consider simple urban dimensions with population variables only to describe cities, and use stylised processes. The system of cities evolves following a spatial interaction model as described by \cite{raimbault2020indirect}, assuming that interaction flows between cities will increase their attractivity and thus their population growth rate. Population updates are not done directly following migration flows as in stochastic urban models such as in \cite{james2018zipf}, but with the hypothesis of hidden variables (generally scaling with population) capturing for example economic dimensions, which influence population growth rate. At the mesoscopic scale, we consider the spatial distribution of population within an urban area and focus on the growth of urban form. To achieve this, the reaction-diffusion model of urban morphogenesis described by \cite{raimbault2018calibration} allows capturing concentration forces and dispersion forces, and has been shown to reproduce a large range of existing urban forms. Furthermore, \cite{raimbault2020comparison} showed that this model had a good coverage of the morphological space when compared to other models of urban morphogenesis.

To couple the scales, we assume that (i) macroscopic performance of a city will influence the choices made by planers in terms of land-use, and thus parameters of the morphogenesis mode locally - empirical evidence of such links have been suggested in the literature \cite{joy2015toronto} and we stylise them here; (ii) urban form influences the inner workings of a city through congestion and the fostering of exchanges between agents (individuals, firms), and finally the global insertion of the city - thus spatial interaction parameters are updated as a function of local urban form. This last link is the most discutable in terms of empirical evidence and implies many other dimensions than population only. We however work on a stylised explicative model which must keep a simple structure but on which a systematic knowledge of model behavior can be developed. We show in Fig.~\ref{fig:fig1} a schematic description of model structure.

\begin{figure}
%  figure: more details?
 \includegraphics[width=\linewidth]{multiscale_morph.pdf}
 \caption{Schematic representation of model architecture. Urban areas with their own urban form are inserted into the macroscopic spatial interaction model, and strong coupling is achieved through top-down and bottom-up feedbacks.\label{fig:fig1}}
\end{figure}

\subsection{Formalization}

We consider $N$ urban areas, represented at the macroscopic scale by their population $P_j(t)$ at time $t$, and at the mesoscopic scale by a population grid $p_{kl}^{(j)}(t)$. The model runs for a total number $t_f$ of time steps, and we will assume that $\Delta t = 1$ for the sake of simplicity (the formulas can be generalized for arbitrary values of the time step, for example when running on real data with irregular time sampling).

The simulation model operates iteratively, and at each time step goes sequentially through the following steps:
\begin{enumerate}
	\item populations of urban areas are evolved at the macroscopic scale with the spatial interaction model (Equations \ref{eq:eq1} to \ref{eq:eq3} below);
	\item for each urban area, parameters of the moprhogenesis model are updated given parameters capturing typical scenarios: transit-oriented development or sprawl for the diffusion parameter, metropolization or uniformization for the aggregation parameter (Equations \ref{eq:eq4} to \ref{eq:eq6});
	\item local urban form are evolved conditionally to a fixed population increase and following the mescoscopic parameters;
	\item congestions and interactions within areas are synthesised following a cost function, which is used to update the macroscopic interaction parameters (Equations \ref{eq:eq7} and \ref{eq:eq8}).
\end{enumerate}

Details of each stage are described below

\subsubsection{Macroscopic model for populations}

Following \cite{raimbault2020indirect} and \cite{favaro2011gibrat}, aggregated populations variables $P_i(t)$ are evolved according to

\begin{equation}
	\label{eq:eq1}
		P_i(t+1) = P_i(t) \left(1 + \Delta t \cdot \left(g_i + \frac{w_i}{N} \cdot \sum_j \frac{V_{ij}}{<V_{ij}>} \right) \right)
\end{equation}

The gravity interaction potential $V_{ij}$ between cities is given by

\begin{equation}
	\label{eq:eq2}
		V_{ij} = \left(\frac{P_i P_j}{(\sum_k p_k)^2}\right)^{\gamma_G} \cdot \exp \left(- \frac{d_{ij}}{d_i} \right)
\end{equation}

We write the population variations as

\begin{equation}
	\label{eq:eq3}
	\Delta P_i (t) = P_i (t + 1) - P_i (t)
\end{equation}

\subsubsection{Top-down feedback}

The population variations at the macroscopic scale then influence the mesoscopic parameters to capture the top-down feedback. Therefore, the parameters of \cite{raimbault2018calibration}, namely $\beta_i$ of diffusion and $\alpha_i$ of aggregation are both updated. First, the mesoscopic growth rate is adjusted to the population growth uniformly over the time interval such that $N_G^{(i)} (t + 1) = \Delta P_i / t_m$, where $t_m$ is a global model parameter controlling the temporal granularity of local population increases.

The sprawl parameter evolves according to a fixed multiplier and the relative population increase following

			\begin{equation}
			\label{eq:eq4}
				\beta_i (t+1) = \beta_i (t) \cdot \left(1 + \delta\beta \cdot \frac{\Delta P_i (t)}{\max_k  \Delta P_k (t)}\right)
			\end{equation}

where the multiplier parameter $\delta\beta$ allows testing different scenarios: a negative value corresponds to transit-oriented development while a positive value corresponds to an uncontrolled sprawl.

The aggregation parameter evolves in a similar way but as a function of accessibility increase, following the rationale that more accessible place with concentrate more activities, as

\begin{equation}
	\label{eq:eq5}
	\alpha_i (t+1) = \alpha_i (t) \cdot \left(1 + \delta\alpha \cdot \frac{\Delta Z_i (t)}{\max_k  \Delta Z_k (t)}\right)
\end{equation}

where the multiplier parameter $\delta\alpha$ allows switching between a metropolisation scenario (more aggregation) and an uniformisation scenario (less aggregation). In that context, the accessibility is computed as

\begin{equation}
	\label{eq:eq5}
	Z_i = \sum_j \frac{P_j}{\sum_k P_k} \cdot \exp( - d_{ij} / d_i)
\end{equation}

Change in the level of sprawl ($\beta_i$) depends on the population pressure only, while aggregation depends on accessibility since it is linked to metropolisation processes.

%\textit{Note: the linear scale for these two parameters may not be relevant depending on the distribution of increments ?} $\rightarrow$ to be tested

\subsubsection{Evolution of urban form at the mesoscopic scale}

Mesoscopic grids population grids are evolved separately within each urban area, using the updated morphogenesis model parameters. We follow here the same exact steps described in \cite{raimbault2018calibration}, with parameters $N_G^{(i)}$ total population increase, $\beta_i$ diffusion level, $\alpha_i$ aggregation level, $t_m$ time steps and $n_d$ number of diffusions. The model adds iteratively population to grid cells, by distributing it with a probability following a preferential attachment to population, and diffusing it.

In terms of implementation, small population differences in the end between macroscopic and mesoscopic levels (due to rounding in computing the number of steps) is corrected by adjusting the macroscopic increments by the effective mesoscopic increments (which are assumed to be more precise).

\begin{table}
	\caption{Summary of model parameters\label{tab:parameters}}
	\centering
	\begin{tabular}{|c|c|c|c|}
	\hline
		Type & Parameter & Process & Range \\\hline
		\multirow{4}{*}{Macro} & $g_i = g_0$ & Endogenous growth & $\left[0 ; 0.5\right]$ \\
		& $w_i = w_G$ & Interactions weight & $\left[0 ; 0.1\right]$ \\
		& $\gamma_i = \gamma_G$ & Interactions hierarchy & $\left[0 ; 5.0\right]$ \\
		& $d_i$ & Interactions decay & $\left[0 ; 1000\right]$ \\ \hline
		\multirow{4}{*}{Meso} & $\alpha_i$ & Aggregation & $\left[0 ; 5\right]$ \\
		& $\beta_i$ & Diffusion & $\left[0 ; 0.5\right]$ \\
		& $t_m$ & Urban growth speed & $\left[0 ; 20\right]$ \\
		& $n_d$ & Diffusion & $\left[0 ; 5\right]$ \\ \hline
		\multirow{4}{*}{Multiscale} & $\delta\alpha$ & Downward feedback & $\left[-0.5 ; 0.5\right]$ \\
		& $\delta\beta$ & Downward feedback & $\left[-0.5 ; 0.5\right]$ \\
		& $\delta d$ & Upward feedback & $\left[-0.5 ; 0.5\right]$ \\\hline
		& $\lambda$ & Cost of congestion & $\left[0 ; 2\right]$ \\\hline
	\end{tabular}
\end{table}

\subsubsection{Bottom-up feedback}

Finally, bottom-up feedback is captured by updating the macroscopic parameters. For the sake of simplicity, only interaction decays are updated, assuming that patterns of urban form play a role in the global insertion of the city. More precisely, we compute gravity flows within the area, and aggregate their value as an economic activity with a squared negative externality interpreted as a congestion with a cost parameter $\lambda$, following

\begin{equation}
\label{eq:eq6}
		U_i = \sum_{kl} \left( \frac{P_k P_l}{P^2} \cdot \frac{1}{d_{kl}} - \lambda \left(\frac{P_k P_l}{P^2} \cdot \frac{1}{d_{kl}}\right)^2 \right)
\end{equation}

This utility $U_i$ is used to update the interaction decays following

\begin{equation}
\label{eq:eq7}
	d_i (t+1) = d_i (t) \left( 1 + \delta d \cdot \frac{U_i}{\max_k \left|U_k\right|} \right)
\end{equation}

where the multiplier parameter $\delta d$ allows controlling for the influence of local performance on global insertion.

\subsection{Parameters}

The Table~\ref{tab:parameters} summarizes model parameters. For clarity, we do not include parameters linked to synthetic model setup, and distinguish them by the scale to which they are attached.

% ! provide nominal values

\subsection{Synthetic setup}

The system is initialized with synthetic systems of cities. A number $N=20$ cities are distributed randomly into a regional urban system, as a square world of width $w=500km$ (reference unit for the decay parameter). City populations follow a Zipf law with hierarchy parameter $\alpha_0 = 1$ (may be modified in experiments), and $P_0 (0) = 100,000$ the initial size of the largest city. Within each city, we consider initial population grids as monocentric (grid of size $W=50$ and center cell density $P_m = 1000$ population units, following an exponential kernel of width one fifth of the world). %check param + necessary experiment: influecne of other initial configs: at least kernel mixture?
The model is simulated for 20 macroscopic time steps (order of magnitude of half a century), each representing $t_m$ mesoscopic time steps (parameter changed in experiments).

%%%
% ! fig placement

\begin{figure}[h!]
	\includegraphics[width=\linewidth]{allindics_hist.png}
	\caption{Statistical distribution of indicators for a sample of 10 parameter values.\label{fig:fig2}}
\end{figure}

\begin{figure}[h!]
	\includegraphics[width=\linewidth]{onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
	\caption{Variations of indicators as a function of $\delta d$, when all other parameters are fixed to their nominal value.\label{fig:fig3}}
\end{figure}

%%

\subsection{Indicators}

Model behavior is characterized using the following indicators: at the  macroscopic scale: distributions of population, accessibilities, and centralities (summarized by average, hierarchy, entropy), following \cite{raimbault2020unveiling}. At the mesoscopic scale, urban form is captured following the indicators used by \cite{raimbault2018calibration}, i.e. Moran index, average distance between individuals, hierarchy of population distribution, and entropy of population distribution.

\section{Results}

\subsection{Implementation}

Model implementation is done under some performance constraints: $N$ mesoscopic morphogenesis models must be simulated in parallel at each macroscopic time step. On the contrary, macroscopic interactions are efficient to implement as they are based on matrices computations. The model is implemented in \texttt{scala} and integrated into a broader library for spatial sensitivity analysis and dynamical models for settlement systems, namely the \texttt{spatialdata} library \cite{raimbault2020scala}. This library in particular already includes implementations of the models used in \cite{raimbault2020indirect} and \cite{raimbault2018calibration}, but also of other urban interaction models such as \cite{favaro2011gibrat}. The relative large number of parameters and output indicators imposes the use of appropriate model exploration methods to get a grasp of model behavior. Therefore, the parameter space is explored with the OpenMOLE model exploration software \cite{reuillon2013openmole}. Standalone source code for the model is available on an open git repository at \url{https://github.com/JusteRaimbault/UrbanGrowth-model}, while the repository for results is at \url{https://github.com/JusteRaimbault/UrbanGrowth}. Simulation data used in this paper is available on the dataverse repository at \url{https://doi.org/10.7910/DVN/IRHMQK}.

\subsection{Statistical consistency}

% one factor sampling
%(min around 1 for 3 indicators, above 5 for 2 and around 0.5 for 2)}

The first numerical experiment is aimed at checking the statistical consistency of model output across stochastic repetitions. We run a one-factor-at-a-time sampling on all parameters with 5 steps for each within the domain provided in Table~\ref{tab:parameters}, and 500 stochastic repetitions for each parameter.% check these values
Obtained statistical distribution are shown in Fig.~\ref{fig:fig2} for a sample of 10 parameters. Different indicators exhibit various distribution shapes, but are always unimodal and generally rather localized. This means different parameter values can be discriminated when considering summary statistics. To estimate the role of noise, we compute Sharpe ratios (rate between average and standard deviations) for each indicators and parameter values. Median values of these across parameters are always larger than 1.6, confirming a reasonable influence of noise on model outputs. We run experiments with 50 replications for each parameter value in the following.

\begin{figure}[h!]
	\includegraphics[width=\linewidth]{deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
	\caption{Macroscopic hierarchy of populations as a function of mesoscopic time steps $t_m$, for different values of interaction decay $d_G$ (colour), and of feedback parameters $\delta d$ (columns) and $\delta \beta$ (rows).\label{fig:fig4}}
\end{figure}

\begin{figure}[h!]
	\includegraphics[width=\linewidth]{deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
	\caption{Relative evolution of average mescoscopic Moran, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig5}}
\end{figure}

\begin{figure}[h!]
	\includegraphics[width=\textwidth]{deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
	\caption{Relative evolution of average mescoscopic population hierarchy, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig6}}
\end{figure}

\begin{figure}[h!]
	\begin{center}
	\includegraphics[width=0.6\linewidth]{pareto_macroHierarchy-mesoHierarchy.png}
	\end{center}
	\caption{Pareto front obtained with a NSGA2 genetic algorithm, to minimize simultaneously hierarchy at the macroscopic and mesoscopic scales. Point size gives the number of stochastic samples and colour the value of $\delta \beta$.\label{fig:fig7}}
\end{figure}

\subsection{Grid exploration}

% note : we should control here compare with / without feedback ?
%  -> first thing to check : with/ without the model
%  ~ done when \delta are set to 0 -> could compare systematically to baseline where all feedbacks are deactivated?

% 20190429_193158_MULTISCALE_GRID_GRID.csv
% not the one factor?

% We fix the congestion cost for bottom-up feedback to a rather strong effect with $\lambda = 1$.
% -> in the grid exp?

We then turn to a grid exploration of the parameter space. %precise successive experiments, step values and varied parameters
Results show an important role of the strong coupling between scales. This is shown for example by the variation of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario (negative feedback of population growth on diffusion $\delta \beta$) to a ``sprawl'' scenario (positive feedback $\delta \beta$) as shown in Fig.~\ref{fig:fig4} below.

We show in Fig.~\ref{fig:fig3} that both macroscopic and mesoscopic indicators have a U-shaped behavior as a function of the bottom-up feedback parameter $\delta d$. In terms of macroscopic trajectories, we see that hierarchies of accessibilities, closeness centrality and population are maximal in negative value when the mechanism is deactivated, what means that the feedback process moderates the hierarchisation of the global urban system. For the urban form, this feedback also mitigates the aggregation of population.

% 20190506_135221_MULTISCALE_TARGETEDGRID_GRID

% fig 4

The strong coupling of scales, in particular the interaction between bottom-up and top-down feedbacks, has a non-trivial influence on model behavior, as shown in Fig.~\ref{fig:fig4} for the macroscopic hierarchy of populations. This allows investigating the influence of the macroscopic interaction decay and the mesoscopic evolution speed, under stylised scenarios for the top-down feedback. We exhibit interaction effects between $\delta \beta$ and $\delta d$, as quantitative and qualitative behavior as a function of $t_m$ changes across panels. Interaction range $d_G$ has on the contrary no significant effect. We find that increasing $t_m$ decreases macroscopic hierarchy when top-down feedback is deactivated ($\delta \beta=0$), but a cancelling of this effect when it is activated. This may have policy implications in terms of controlling urban sprawl by acting or not on $\delta \beta$, and its influence on the overall urban system.

\subsection{Impact of policy parameters}

We then proceed to a targeted experiment by detailing more the grid for $\delta \beta$ and $\delta \alpha$. The aim is to investigate the influence of policies taken at the level of urban areas, but which can also be constrained by a higher level of governance. We show in Fig.~\ref{fig:fig5} the influence of $\delta \beta$ on urban form. In particular, the average variation of Moran index captures the centralisation of urban areas. This centralisation exhibits a minimal value when $\delta \alpha$ is low, consistent across values of $\lambda$. This implicates that when decentralisation policies are implemented by acting on aggregation ($\delta \alpha$), the policy on urban sprawl will have an unexpected effect on centralisation. This is an illustration of complex and possibly negative interaction between policies.
% influenced by upward feedback? No, rather stable when \lambda = 0 -> why? - check influence of lambda

When investigating population hierarchy the mesoscopic scale, i.e. inequalities of population distribution within the urban area, we find in Fig.~\ref{fig:fig6} a U-shaped behavior as a function of $\delta \beta$ for negative $\delta \alpha$ values, but has a plateau for positive values of $\delta \alpha$, showing also an interaction effect between policies. In terms of hierarchy, a maximal negative value may be wanted to limit inequalities (slopes are negative), and this optimum is reached for a slightly positive value of $\delta \beta$ (left columns in Fig.~\ref{fig:fig6}). This means that sprawl may be needed in some configurations to optimise this objective.

\subsection{Multiscale optimization}

We finally turn to an optimization objective to investigate if compromises have to be made when targets are simultaneously aimed for at different scales. We therefore use a NSGA2 algorithm implemented in OpenMOLE, with a population of 200 individuals, run for 100,000 generations. Objectives are the average mesoscopic hierarchy which can be minimized for more compact and thus sustainable cities, at the macroscopic population hierarchy. Convergence is reached in terms of hypervolume and points in the Pareto front.

We show in Fig.~\ref{fig:fig7} the Pareto front obtained after convergence of the algorithm. This confirms that the two objectives are indeed contradictory and that compromises between scales have to be made when trying to optimise both simultaneously. Most points correspond to policies favouring sprawl, at the exception of a small set of points with low macroscopic hierarchy values, which discourage sprawl ($\delta \beta < 0$). Systems with the most equalities between cities within the system of cities correspond to the most hierarchical urban form, implying inequalities at the mesoscopic scales.

\section{Discussion}

\subsection{Theoretical and practical implications}

Our model effectively captures an interaction between downward and upward feedback. In that sense, it illustrates weak emergence in its proper sense \cite{bedau2002downward}. This suggests this view can be effectively operationalised into models of urban dynamics.

In terms of model complexity and in relation to urban complexity in itself, we show that coupling simple models (each having only three parameters) already yields a complicated and complex simulation model. It also requires additional ontologies to achieve the coupling (in our case the policy processes on one side and the congestion-performance process on the other side). Therefore, we suggest this is additional evidence for a necessity of complexity and simulation models to understand urban complexity. This implies in some way the failure of reductionist epistemologies, at least to grab the multi-scalar nature of urban systems.
% 'suggest in a way failure of reductionist empistemology (in the sense of a la Barthelemy) - at least for grabbing the multi-scalar nature of systems' (seems a tautology though

Although a stylised and highly simplified model, we were already able to integrate parameters linked to policies. Indeed, governance processes are intrinsically multi-scalar \cite{liao2017ouverture}. Therefore, developing such models is a first step towards a progressive integration models for policy and sustainable management of territories.

\subsection{Developments}

%\subsection{Multi-modeling and concurrent processes}

This model is only a first structural sketch with very restrictive assumption, in particular regarding the downward and upward feedbacks on submodel parameters. There may be no link between urban form and global insertion, or it may be due to other processes, be expressed as an other functional form. An important stage before shifting to robust knowledge will consist in (i) reviewing and making a typology of such potential processes across scales; (ii) including most in a multi-modeling fashion to compare possible concurrent mechanisms.

Future work also implies the use of more elaborated model exploration and validation methods. For example, using the diversity search integrated in OpenMOLE \cite{reuillon2013openmole} would allow unveiling the diversity of possible future urban trajectories, and possibly unexpected novel pathways towards sustainability, finding for example regimes with the strongest effect of feedback parameters on indicators related to sustainability.
% spatial sensitivity - influence of synthetic system of cities

Regarding an application of the model on real systems of cities, a calibration on empirical trajectories would be needed. However, such data is difficult to obtain, and some investigations should be made on the relevant calibration targets (in particular at which scale).
% parametrization on real systems; possibly calibration \cite{raimbault2019ilus}

%We also did not include explicitly transportation networks in this model.

Our model is a first step towards multi-scalar models of urban dynamics which effectively include a strong coupling between processes at different scales. While remaining highly stylised, we showed an important role of the coupling and unveiled for example interaction effects between different types of policies. This type of approach will need in the future to be further developed, towards multi-scalar models for sustainable policy making \cite{Rozenblat2018}. This works paves the way for such more complex models.

% conclusion not needed
%\section{Conclusion}
%This paper provides a first step towards strongly multi-scalar models to capture urban complexity towards policy models. 
%Towards integrative models and theories for urban systems \cite{raimbault2019methods}

\section*{Acknowledgments}

The author acknowledges the funding of their institutions and the EPSRC project number EP/M023583/1. Results obtained in this paper were computed on the vo.complex-system.eu virtual organisation of the European Grid Infrastructure (http://www.egi.eu). We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.

\begin{thebibliography}{}

\bibitem[Banos et~al., 2015]{banos2015importance}
Banos, A., Corson, N., Gaudou, B., Laperri{\`e}re, V., and Coyrehourcq, S.~R.
  (2015).
\newblock The importance of being hybrid for spatial epidemic models: a
  multi-scale approach.
\newblock {\em Systems}, 3(4):309--329.

\bibitem[Batty, 2005]{batty2005agents}
Batty, M. (2005).
\newblock Agents, cells, and cities: new representational models for simulating
  multiscale urban dynamics.
\newblock {\em Environment and Planning A}, 37(8):1373--1394.

\bibitem[Batty, 2018]{batty2018inventing}
Batty, M. (2018).
\newblock {\em Inventing future cities}.
\newblock MIT press.

\bibitem[Bedau, 2002]{bedau2002downward}
Bedau, M. (2002).
\newblock Downward causation and the autonomy of weak emergence.
\newblock {\em Principia: an international journal of epistemology},
  6(1):5--50.

\bibitem[Belem and M{\"u}ller, 2013]{belem2013organizational}
Belem, M. and M{\"u}ller, J.~P. (2013).
\newblock An organizational model for multi-scale and multi-formalism
  simulation: Application in carbon dynamics simulation in west-african
  savanna.
\newblock {\em Simulation Modelling Practice and Theory}, 32:83--98.

\bibitem[Boulet et~al., 2020]{boulet2020coupling}
Boulet, X., Zargayouna, M., Scemama, G., and Leurent, F. (2020).
\newblock Coupling multi-agent and macroscopic simulators of traffic.
\newblock In {\em Agents and Multi-agent Systems: Technologies and Applications
  2019}, pages 323--332. Springer.

\bibitem[Chavalarias et~al., 2009]{chavalarias2009french}
Chavalarias, D., Bourgine, P., Perrier, E., Amblard, F., Arlabosse, F., Auger,
  P., Baillon, J.-B., Barreteau, O., Baudot, P., Bouchaud, E., et~al. (2009).
\newblock French roadmap for complex systems 2008-2009.

\bibitem[Cheng and Masser, 2003]{cheng2003modelling}
Cheng, J. and Masser, I. (2003).
\newblock Modelling urban growth patterns: a multiscale perspective.
\newblock {\em Environment and Planning A}, 35(4):679--704.

\bibitem[Clarke et~al., 2007]{clarke2007decade}
Clarke, K.~C., Gazulis, N., Dietzel, C., and Goldstein, N.~C. (2007).
\newblock A decade of sleuthing: Lessons learned from applications of a
  cellular automaton land use change model.
\newblock {\em Classics in IJGIS: twenty years of the international journal of
  geographical information science and systems}, pages 413--427.

\bibitem[Crociani et~al., 2016]{crociani2016multi}
Crociani, L., L{\"a}mmel, G., and Vizzari, G. (2016).
\newblock Multi-scale simulation for crowd management: a case study in an urban
  scenario.
\newblock In {\em International Conference on Autonomous Agents and Multiagent
  Systems}, pages 147--162. Springer.

\bibitem[Favaro and Pumain, 2011]{favaro2011gibrat}
Favaro, J.-M. and Pumain, D. (2011).
\newblock Gibrat revisited: An urban growth model incorporating spatial
  interaction and innovation cycles.
\newblock {\em Geographical Analysis}, 43(3):261--286.

\bibitem[Ford et~al., 2019]{ford2019multi}
Ford, A., Barr, S., Dawson, R., Virgo, J., Batty, M., and Hall, J. (2019).
\newblock A multi-scale urban integrated assessment framework for climate
  change studies: A flooding application.
\newblock {\em Computers, Environment and Urban Systems}, 75:229--243.

\bibitem[Gabaix, 1999]{gabaix1999zipf}
Gabaix, X. (1999).
\newblock Zipf's law for cities: an explanation.
\newblock {\em Quarterly journal of Economics}, pages 739--767.

\bibitem[Hjorth et~al., 2020]{hjorth2020levelspace}
Hjorth, A., Head, B., Brady, C., Wilensky, U., et~al. (2020).
\newblock Levelspace: A netlogo extension for multi-level agent-based modeling.
\newblock {\em Journal of Artificial Societies and Social Simulation},
  23(1):1--4.

\bibitem[James et~al., 2018]{james2018zipf}
James, C., Azaele, S., Maritan, A., and Simini, F. (2018).
\newblock Zipf's and taylor's laws.
\newblock {\em Physical Review E}, 98(3):032408.

\bibitem[Joy and Vogel, 2015]{joy2015toronto}
Joy, M. and Vogel, R.~K. (2015).
\newblock Toronto’s governance crisis: A global city under pressure.
\newblock {\em Cities}, 49:35--52.

\bibitem[Liao and Gaudin, 2017]{liao2017ouverture}
Liao, L. and Gaudin, J.-P. (2017).
\newblock L’ouverture au march{\'e} en chine (ann{\'e}es 1980-2000) et le
  d{\'e}veloppement {\'e}conomique local: une forme de gouvernance
  multi-niveaux?
\newblock {\em Cybergeo: European Journal of Geography}.

\bibitem[Murcio et~al., 2015]{murcio2015urban}
Murcio, R., Morphet, R., Gershenson, C., and Batty, M. (2015).
\newblock Urban transfer entropy across scales.
\newblock {\em PLoS One}, 10(7):e0133780.

\bibitem[Pumain, 1997]{pumain1997pour}
Pumain, D. (1997).
\newblock Pour une th{\'e}orie {\'e}volutive des villes.
\newblock {\em Espace g{\'e}ographique}, 26(2):119--134.

\bibitem[Pumain and Reuillon, 2017]{pumain2017urban}
Pumain, D. and Reuillon, R. (2017).
\newblock {\em Urban Dynamics and Simulation Models}.
\newblock Springer International.

\bibitem[Raimbault, 2018]{raimbault2018calibration}
Raimbault, J. (2018).
\newblock Calibration of a density-based model of urban morphogenesis.
\newblock {\em PloS one}, 13(9):e0203516.

\bibitem[Raimbault, 2019a]{raimbault:halshs-02265423}
Raimbault, J. (2019a).
\newblock {Evolving accessibility landscapes: mutations of transportation
  networks in China}.
\newblock In Aveline-Dubach, N., editor, {\em {PATHWAYS OF SUSTAINABLE URBAN
  DEVELOPMENT ACROSS CHINA: THE CASES OF HANGZHOU, DATONG AND ZHUHAI}}, pages
  89--108. {Imago Editor}.
\newblock ebook.

\bibitem[Raimbault, 2019b]{raimbault:halshs-02284933}
Raimbault, J. (2019b).
\newblock {Towards integrated urban models for sustainable policies}.
\newblock In {\em {Urban analytics at the interface between environment, human
  health and behaviour}}, Shanghai, China. {Tongji University and The Alan
  Turing Institute}.

\bibitem[Raimbault, 2019c]{raimbault:halshs-02013006}
Raimbault, J. (2019c).
\newblock {Towards multi-scalar models for the co-evolution of transportation
  networks and territories}.
\newblock In {\em {Theo Quant 2019}}, Besan{\c c}on, France. {Universit{\'e} de
  Franche-Comt{\'e}}.

\bibitem[Raimbault, 2020a]{raimbault2020comparison}
Raimbault, J. (2020a).
\newblock A comparison of simple models for urban morphogenesis.
\newblock {\em arXiv preprint arXiv:2008.13277}.

\bibitem[Raimbault, 2020b]{raimbault2020hierarchy}
Raimbault, J. (2020b).
\newblock Hierarchy and co-evolution processes in urban systems.
\newblock {\em arXiv preprint arXiv:2001.11989}.

\bibitem[Raimbault, 2020c]{raimbault2020indirect}
Raimbault, J. (2020c).
\newblock Indirect evidence of network effects in a system of cities.
\newblock {\em Environment and Planning B: Urban Analytics and City Science},
  47(1):138--155.

\bibitem[Raimbault, 2020d]{raimbault2020unveiling}
Raimbault, J. (2020d).
\newblock Unveiling co-evolutionary patterns in systems of cities: a systematic
  exploration of the simpopnet model.
\newblock In {\em Theories and Models of Urbanization}, pages 261--278.
  Springer.

\bibitem[Raimbault et~al., 2020]{raimbault2020scala}
Raimbault, J., Perret, J., and Reuillon, R. (2020).
\newblock A scala library for spatial sensitivity analysis.
\newblock {\em arXiv preprint arXiv:2007.10667}.

\bibitem[Reuillon et~al., 2013]{reuillon2013openmole}
Reuillon, R., Leclaire, M., and Rey-Coyrehourcq, S. (2013).
\newblock Openmole, a workflow engine specifically tailored for the distributed
  exploration of simulation models.
\newblock {\em Future Generation Computer Systems}, 29(8):1981--1990.

\bibitem[Rozenblat and Pumain, 2018]{Rozenblat2018}
Rozenblat, C. and Pumain, D. (2018).
\newblock {\em Toward a Methodology for Multi-scalar Urban System Policies},
  pages 385--393.
\newblock Springer Singapore, Singapore.

\bibitem[Shu et~al., 2020]{shu2020modelling}
Shu, B., Zhu, S., Qu, Y., Zhang, H., Li, X., and Carsjens, G.~J. (2020).
\newblock Modelling multi-regional urban growth with multilevel logistic
  cellular automata.
\newblock {\em Computers, Environment and Urban Systems}, 80:101457.

\bibitem[Torrens and O'Sullivan, 2001]{torrens2001cellular}
Torrens, P.~M. and O'Sullivan, D. (2001).
\newblock Cellular automata and urban simulation: where do we go from here?

\bibitem[Wegener and F{\"u}rst, 2004]{wegener2004land}
Wegener, M. and F{\"u}rst, F. (2004).
\newblock Land-use transport interaction: state of the art.
\newblock {\em Available at SSRN 1434678}.

\bibitem[Wegener and Spiekermann, 2018]{wegener2018multi}
Wegener, M. and Spiekermann, K. (2018).
\newblock Multi-level urban models: Integration across space, time and
  policies.

\bibitem[White, 2006]{white2006modeling}
White, R. (2006).
\newblock Modeling multi-scale processes in a cellular automata framework.
\newblock In {\em Complex artificial environments}, pages 165--177. Springer.

\bibitem[Xu and Gao, 2019]{xu2019directional}
Xu, T. and Gao, J. (2019).
\newblock Directional multi-scale analysis and simulation of urban expansion in
  auckland, new zealand using logistic cellular automata.
\newblock {\em Computers, Environment and Urban Systems}, 78:101390.

\bibitem[Xu et~al., 2020]{xu2020urban}
Xu, T., Gao, J., Coco, G., and Wang, S. (2020).
\newblock Urban expansion in auckland, new zealand: a gis simulation via an
  intelligent self-adapting multiscale agent-based model.
\newblock {\em International Journal of Geographical Information Science},
  34(11):2136--2159.

\bibitem[Yu et~al., 2018]{yu2018modeling}
Yu, Y., He, J., Tang, W., and Li, C. (2018).
\newblock Modeling urban collaborative growth dynamics using a multiscale
  simulation model for the wuhan urban agglomeration area, china.
\newblock {\em ISPRS International Journal of Geo-Information}, 7(5):176.

\bibitem[Zhu and Tian, 2020]{zhu2020dynamic}
Zhu, J. and Tian, S. (2020).
\newblock A dynamic urban lake area evolution model based on multilevel grid,
  cellular automata, and multiagent system.
\newblock {\em Complexity}, 2020.

\end{thebibliography}

\end{document}

### 2021-01-31 Docs/Papers/Multiscale/paper/paper.tex
arxiv

--- 2021-01-31 Docs/Papers/Multiscale/arxiv/paper.tex
@@
-\documentclass[11pt]{article}
+\documentclass[sustainability,article,submit,moreauthors,pdftex]{Definitions/mdpi}
 
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage[margin=2cm]{geometry}
-\usepackage{multirow}
+%=================================================================
+\firstpage{1} 
+\makeatletter 
+\setcounter{page}{\@firstpage} 
+\makeatother
+\pubvolume{xx}
+\issuenum{x}
+\articlenumber{x}
+\pubyear{2020}
+\copyrightyear{2020}
+%\externaleditor{Academic Editor: name}
+\history{Received: date; Accepted: date; Published: date}
+%\updates{yes} % If there is an update available, un-comment this line
+%% MDPI internal command: uncomment if new journal that already uses continuous page numbers 
+%\continuouspages{yes}
 
-\title{Strong coupling between scales in a multi-scalar model of urban dynamics}
+%=================================================================
+% Full title of the paper (Capitalized)
+\Title{Strong coupling between scales in a multi-scalar model of urban dynamics}
 
-\author{Juste Raimbault$^{1,2,3,\ast}$\medskip\\
-$^{1}$ Center for Advanced Spatial Analysis, University College London\\
-$^{2}$ UPS CNRS 3611 ISC-PIF\\
-$^{3}$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\medskip\\
-$^{\ast}$ \texttt{juste.raimbault@polytechnique.edu}
-}
+% Author Orchid ID: enter ID or remove command
+\newcommand{\orcidauthorA}{0000-0003-0768-9480} % Add \orcidA{} behind the author's name
+%\newcommand{\orcidauthorB}{0000-0000-000-000X} % Add \orcidB{} behind the author's name
 
-\date{}
+% Authors, for the paper (add full first names)
+\Author{Juste Raimbault $^{1,2,3,*}$\orcidA{}}
 
-\begin{document}
+% Authors, for metadata in PDF
+\AuthorNames{Juste Raimbault}
 
-\maketitle
+% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
+\address{%
+$^{1}$ \quad CASA, University College London\\
+$^{2}$ \quad Institut des Syst{\`e}mes Complexes Paris-Ile-de-France, CNRS\\
+$^{3}$ \quad UMR G{\'e}ographie-cit{\'e}s, CNRS}
 
-\begin{abstract}
-	Urban evolution processes occur at different scales, with intricate interactions between levels and relatively distinct type of processes.
+% Contact information of the corresponding author
+\corres{Correspondence: juste.raimbault@polytechnique.edu}
+
+% Current address and/or shared authorship
+% \dagger,\ddagger
+%\firstnote{Current address: Affiliation 3} 
+%\secondnote{These authors contributed equally to this work.}
+% The commands \thirdnote{} till \eighthnote{} are available for further notes
+
+%Dear Editor,
+
+%I am pleased to submit an original research article entitled "Strong coupling between scales in a multi-scalar model of urban dynamics" for consideration for publication in IJGI.
+This article introduces a novel modeling study on the dynamics of urban systems, with a particular on feedbacks between scales.
+The subject is in my opinion furthermore particularly suited for the special issue "Spatio-Temporal Models and Geo-Technologies".
+
+%A preliminary version of the methodology and results were presented as a main track talk at the Conference on Complex Systems 2019.
+A preprint was submitted on arXiv at the same date of this submission.
+
+%I had no previous contact with journal editors regarding this manuscript.
+This manuscript has not been published and is not under consideration for publication elsewhere.
+I have no conflicts of interest to disclose.
+I do not have any opposed reviewer.
+
+%Thank you for your consideration.
+
+%Yours faithfully,
+
+%Juste Raimbault
+%University College London
+
+\abstract{Urban evolution processes occur at different scales, with intricate interactions between levels and relatively distinct type of processes.
 To what extent actual urban dynamics include an actual strong coupling between scales, in the sense of both top-down and bottom-up feedbacks, remains an open issue with important practical implications for the sustainable management of territories.
@@
 We also run a multi-objective optimization algorithm on the model, showing showing that compromise between scales are captured.
-Our approach opens new research directions towards more operational urban dynamics models including a strong feedback between scales.
-	\medskip\\
-\textbf{Keywords: } Urban dynamics; Systems of cities; Urban morphogenesis; Multi-scalar modeling; Strong coupling
-\end{abstract}
+Our approach opens new research directions towards more operational urban dynamics models including a strong feedback between scales.}
 
-%  relevant target CEUS?
-https://www.elsevier.com/journals/computers-environment-and-urban-systems/0198-9715/guide-for-authors => special issue expired; IJGI 31/01:  https://www.mdpi.com/journal/ijgi/special_issues/spatio_temporal_models_geo
+% Keywords
+\keyword{Urban dynamics; Systems of cities; Urban morphogenesis; Multi-scalar modeling; Strong coupling}
 
-% Systems of cities
-%Multi-scalar model
-%Spatial interaction
-%Reaction-diffusion
-
-% abstract CCS
-% 
-%The modeling of urban growth is a crucial issue for the design of sustainable territorial policies, through the understanding of past urbanization processes and the forecasting of future urban trajectories.
-Several models have been proposed at different scales and integrating differ- ent dimensions of urban systems, such as land-use transport interaction models [1] or systems of cities models [2].
-While multi-scalar models are recognized as crucial for the study of such systems [3], they remain in practice unexplored.
-%This contribution introduces a parsimonious multi-scalar model for systems of cities, based on simple dimensions (mainly populations) with stylized processes, but yielding an effective strong coupling between the metropolitan mesoscopic scale and the macroscopic scale of the system of cities.
-The model couples the spatial interaction model of [4] for the macro scale with the reaction-diffusion model for urban form studied by [5].
-More precisely, urban areas viewed as a population grid are embedded into the macroscopic interaction model.
-To evolve populations and local urban forms, one time step consists of (i) population differences are computed by the interaction model; (ii) top-down feedback modifies parameters of mesoscopic models, given control parameters to capture typical scenarios (transit-oriented development or sprawl for diffusion, metropolization or uniformization for aggregation); (iii) local urban form are evolved with the reaction-diffusion models at a given speed conditionally to the population variations; (iv) changes in urban form influence macroscopic interaction ranges (capturing the impact of local activity on global insertion), by integrating gravity flows in the area with a squared cost function making a compromise between congestion and flows.
-%The model is applied on synthetic systems of cities typical of a continental range (500km, hierarchy around 1, 20 cities), with initial local population grid configurations as monocen- tric.
-Parameter space is explored with the OpenMOLE model exploration software [6], eased by the implementation of the model in scala [7].
-First results show a strong impact of the strong meso-macro coupling, such as for example a qualitative inversion of the behavior as a function of interaction range of macroscopic indicators trajectories when switching from a “transit-oriented development” scenario (negative feedback of population growth on diffusion) to a “sprawl” scenario (positive feedback).
-Similarly, mesoscopic urban form indicators are significantly influenced by the coupling process.
-%Further work will consist in more targeted simulation experiments, including specific ex- ploration algorithms such as diversity search for model regimes [6], to test the model as a proof-of-concept of models for policies.
-Such a model can also be calibrated on real city sys- tems and urban form trajectories, to extrapolate coupling parameters that would be difficult to obtain otherwise.
-Our contribution is thus a first step towards multi-scalar simulation models for systems of cities.
-
-% multilevel models : check https://scholar.google.fr/citations?user=x1BwVXoAAAAJ&hl=fr
-% + Morvan, G. (2014). Bibliographic database on multi-level and multi-scale agent-based modeling. Transportation Research Procedia, 2, 495-500.
-
-%\cite{louail2010comparer}
+%\setcounter{secnumdepth}{4}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %  figure: more details?
- \includegraphics[width=\linewidth]{multiscale_morph.pdf}
+ \includegraphics[width=\linewidth]{figures/multiscale_morph.pdf}
  \caption{Schematic representation of model architecture.
@@
 \begin{figure}[h!]
-	\includegraphics[width=\linewidth]{allindics_hist.png}
+	\includegraphics[width=\linewidth]{figures/allindics_hist.png}
 	\caption{Statistical distribution of indicators for a sample of 10 parameter values.\label{fig:fig2}}
@@
 \begin{figure}[h!]
-	\includegraphics[width=\linewidth]{onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
+	\includegraphics[width=\linewidth]{figures/onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
 	\caption{Variations of indicators as a function of $\delta d$, when all other parameters are fixed to their nominal value.\label{fig:fig3}}
@@
 \begin{figure}[h!]
-	\includegraphics[width=\linewidth]{deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
+	\includegraphics[width=\linewidth]{figures/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
 	\caption{Macroscopic hierarchy of populations as a function of mesoscopic time steps $t_m$, for different values of interaction decay $d_G$ (colour), and of feedback parameters $\delta d$ (columns) and $\delta \beta$ (rows).\label{fig:fig4}}
@@
 \begin{figure}[h!]
-	\includegraphics[width=\linewidth]{deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+	\includegraphics[width=\linewidth]{figures/deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
 	\caption{Relative evolution of average mescoscopic Moran, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig5}}
@@
 \begin{figure}[h!]
-	\includegraphics[width=\textwidth]{deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+	\includegraphics[width=\textwidth]{figures/deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
 	\caption{Relative evolution of average mescoscopic population hierarchy, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig6}}
@@
 	\begin{center}
-	\includegraphics[width=0.6\linewidth]{pareto_macroHierarchy-mesoHierarchy.png}
+	\includegraphics[width=0.6\linewidth]{figures/pareto_macroHierarchy-mesoHierarchy.png}
 	\end{center}
@@
 
-% conclusion not needed
-%\section{Conclusion}
-%This paper provides a first step towards strongly multi-scalar models to capture urban complexity towards policy models. 
-%Towards integrative models and theories for urban systems \cite{raimbault2019methods}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\vspace{6pt}
 
-\section*{Acknowledgments}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\authorcontributions{}
 
-The author acknowledges the funding of their institutions and the EPSRC project number EP/M023583/1.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\funding{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\acknowledgments{The author acknowledges the funding of their institutions and the EPSRC project number EP/M023583/1.
 Results obtained in this paper were computed on the vo.complex-system.eu virtual organisation of the European Grid Infrastructure (http://www.egi.eu).
-We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.
+We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.}
 
-\begin{thebibliography}{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\conflictsofinterest{}
 
-\bibitem[Banos et~al., 2015]{banos2015importance}
-Banos, A., Corson, N., Gaudou, B., Laperri{\`e}re, V., and Coyrehourcq, S.~R.
-  (2015).
-\newblock The importance of being hybrid for spatial epidemic models: a
-  multi-scale approach.
-\newblock {\em Systems}, 3(4):309--329.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\reftitle{References}
 
-\bibitem[Batty, 2005]{batty2005agents}
-Batty, M. (2005).
-\newblock Agents, cells, and cities: new representational models for simulating
-  multiscale urban dynamics.
-\newblock {\em Environment and Planning A}, 37(8):1373--1394.
+\externalbibliography{yes}
+\bibliography{biblio.bib}
 
-\bibitem[Batty, 2018]{batty2018inventing}
-Batty, M. (2018).
-\newblock {\em Inventing future cities}.
-\newblock MIT press.
+%% for journal Sci
+%\reviewreports{\\
+%Reviewer 1 comments and authors’ response\\
+%Reviewer 2 comments and authors’ response\\
+%Reviewer 3 comments and authors’ response
+%}
 
-\bibitem[Bedau, 2002]{bedau2002downward}
-Bedau, M. (2002).
-\newblock Downward causation and the autonomy of weak emergence.
-\newblock {\em Principia: an international journal of epistemology},
-  6(1):5--50.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\end{document}
 
-\bibitem[Belem and M{\"u}ller, 2013]{belem2013organizational}
-Belem, M. and M{\"u}ller, J.~P. (2013).
-\newblock An organizational model for multi-scale and multi-formalism
-  simulation: Application in carbon dynamics simulation in west-african
-  savanna.
-\newblock {\em Simulation Modelling Practice and Theory}, 32:83--98.
+%\begin{figure}[H]
+%\centering
+%\includegraphics[width=2 cm]{Definitions/logo-mdpi}
+%\caption{This is a figure, Schemes follow the same formatting.
+If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
+(\textbf{b}) Description of what is contained in the second panel.
+Figures should be placed in the main text near to the first time they are cited.
+A caption on a single line should be centered.}
+%\end{figure}
 
-\bibitem[Boulet et~al., 2020]{boulet2020coupling}
-Boulet, X., Zargayouna, M., Scemama, G., and Leurent, F. (2020).
-\newblock Coupling multi-agent and macroscopic simulators of traffic.
-\newblock In {\em Agents and Multi-agent Systems: Technologies and Applications
-  2019}, pages 323--332. Springer.
+%\begin{table}[H]
+%\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
+%\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+%\begin{tabular}{ccc}
+%\toprule
+%\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
+%\midrule
+%entry 1		& data			& data\\
+%entry 2		& data			& data\\
+%\bottomrule
+%\end{tabular}
+%\end{table}
 
-\bibitem[Chavalarias et~al., 2009]{chavalarias2009french}
-Chavalarias, D., Bourgine, P., Perrier, E., Amblard, F., Arlabosse, F., Auger,
-  P., Baillon, J.-B., Barreteau, O., Baudot, P., Bouchaud, E., et~al. (2009).
-\newblock French roadmap for complex systems 2008-2009.
-
-\bibitem[Cheng and Masser, 2003]{cheng2003modelling}
-Cheng, J. and Masser, I. (2003).
-\newblock Modelling urban growth patterns: a multiscale perspective.
-\newblock {\em Environment and Planning A}, 35(4):679--704.
-
-\bibitem[Clarke et~al., 2007]{clarke2007decade}
-Clarke, K.~C., Gazulis, N., Dietzel, C., and Goldstein, N.~C. (2007).
-\newblock A decade of sleuthing: Lessons learned from applications of a
-  cellular automaton land use change model.
-\newblock {\em Classics in IJGIS: twenty years of the international journal of
-  geographical information science and systems}, pages 413--427.
-
-\bibitem[Crociani et~al., 2016]{crociani2016multi}
-Crociani, L., L{\"a}mmel, G., and Vizzari, G. (2016).
-\newblock Multi-scale simulation for crowd management: a case study in an urban
-  scenario.
-\newblock In {\em International Conference on Autonomous Agents and Multiagent
-  Systems}, pages 147--162. Springer.
-
-\bibitem[Favaro and Pumain, 2011]{favaro2011gibrat}
-Favaro, J.-M. and Pumain, D. (2011).
-\newblock Gibrat revisited: An urban growth model incorporating spatial
-  interaction and innovation cycles.
-\newblock {\em Geographical Analysis}, 43(3):261--286.
-
-\bibitem[Ford et~al., 2019]{ford2019multi}
-Ford, A., Barr, S., Dawson, R., Virgo, J., Batty, M., and Hall, J. (2019).
-\newblock A multi-scale urban integrated assessment framework for climate
-  change studies: A flooding application.
-\newblock {\em Computers, Environment and Urban Systems}, 75:229--243.
-
-\bibitem[Gabaix, 1999]{gabaix1999zipf}
-Gabaix, X. (1999).
-\newblock Zipf's law for cities: an explanation.
-\newblock {\em Quarterly journal of Economics}, pages 739--767.
-
-\bibitem[Hjorth et~al., 2020]{hjorth2020levelspace}
-Hjorth, A., Head, B., Brady, C., Wilensky, U., et~al. (2020).
-\newblock Levelspace: A netlogo extension for multi-level agent-based modeling.
-\newblock {\em Journal of Artificial Societies and Social Simulation},
-  23(1):1--4.
-
-\bibitem[James et~al., 2018]{james2018zipf}
-James, C., Azaele, S., Maritan, A., and Simini, F. (2018).
-\newblock Zipf's and taylor's laws.
-\newblock {\em Physical Review E}, 98(3):032408.
-
-\bibitem[Joy and Vogel, 2015]{joy2015toronto}
-Joy, M. and Vogel, R.~K. (2015).
-\newblock Toronto’s governance crisis: A global city under pressure.
-\newblock {\em Cities}, 49:35--52.
-
-\bibitem[Liao and Gaudin, 2017]{liao2017ouverture}
-Liao, L. and Gaudin, J.-P. (2017).
-\newblock L’ouverture au march{\'e} en chine (ann{\'e}es 1980-2000) et le
-  d{\'e}veloppement {\'e}conomique local: une forme de gouvernance
-  multi-niveaux?
-\newblock {\em Cybergeo: European Journal of Geography}.
-
-\bibitem[Murcio et~al., 2015]{murcio2015urban}
-Murcio, R., Morphet, R., Gershenson, C., and Batty, M. (2015).
-\newblock Urban transfer entropy across scales.
-\newblock {\em PLoS One}, 10(7):e0133780.
-
-\bibitem[Pumain, 1997]{pumain1997pour}
-Pumain, D. (1997).
-\newblock Pour une th{\'e}orie {\'e}volutive des villes.
-\newblock {\em Espace g{\'e}ographique}, 26(2):119--134.
-
-\bibitem[Pumain and Reuillon, 2017]{pumain2017urban}
-Pumain, D. and Reuillon, R. (2017).
-\newblock {\em Urban Dynamics and Simulation Models}.
-\newblock Springer International.
-
-\bibitem[Raimbault, 2018]{raimbault2018calibration}
-Raimbault, J. (2018).
-\newblock Calibration of a density-based model of urban morphogenesis.
-\newblock {\em PloS one}, 13(9):e0203516.
-
-\bibitem[Raimbault, 2019a]{raimbault:halshs-02265423}
-Raimbault, J. (2019a).
-\newblock {Evolving accessibility landscapes: mutations of transportation
-  networks in China}.
-\newblock In Aveline-Dubach, N., editor, {\em {PATHWAYS OF SUSTAINABLE URBAN
-  DEVELOPMENT ACROSS CHINA: THE CASES OF HANGZHOU, DATONG AND ZHUHAI}}, pages
-  89--108. {Imago Editor}.
-\newblock ebook.
-
-\bibitem[Raimbault, 2019b]{raimbault:halshs-02284933}
-Raimbault, J. (2019b).
-\newblock {Towards integrated urban models for sustainable policies}.
-\newblock In {\em {Urban analytics at the interface between environment, human
-  health and behaviour}}, Shanghai, China. {Tongji University and The Alan
-  Turing Institute}.
-
-\bibitem[Raimbault, 2019c]{raimbault:halshs-02013006}
-Raimbault, J. (2019c).
-\newblock {Towards multi-scalar models for the co-evolution of transportation
-  networks and territories}.
-\newblock In {\em {Theo Quant 2019}}, Besan{\c c}on, France. {Universit{\'e} de
-  Franche-Comt{\'e}}.
-
-\bibitem[Raimbault, 2020a]{raimbault2020comparison}
-Raimbault, J. (2020a).
-\newblock A comparison of simple models for urban morphogenesis.
-\newblock {\em arXiv preprint arXiv:2008.13277}.
-
-\bibitem[Raimbault, 2020b]{raimbault2020hierarchy}
-Raimbault, J. (2020b).
-\newblock Hierarchy and co-evolution processes in urban systems.
-\newblock {\em arXiv preprint arXiv:2001.11989}.
-
-\bibitem[Raimbault, 2020c]{raimbault2020indirect}
-Raimbault, J. (2020c).
-\newblock Indirect evidence of network effects in a system of cities.
-\newblock {\em Environment and Planning B: Urban Analytics and City Science},
-  47(1):138--155.
-
-\bibitem[Raimbault, 2020d]{raimbault2020unveiling}
-Raimbault, J. (2020d).
-\newblock Unveiling co-evolutionary patterns in systems of cities: a systematic
-  exploration of the simpopnet model.
-\newblock In {\em Theories and Models of Urbanization}, pages 261--278.
-  Springer.
-
-\bibitem[Raimbault et~al., 2020]{raimbault2020scala}
-Raimbault, J., Perret, J., and Reuillon, R. (2020).
-\newblock A scala library for spatial sensitivity analysis.
-\newblock {\em arXiv preprint arXiv:2007.10667}.
-
-\bibitem[Reuillon et~al., 2013]{reuillon2013openmole}
-Reuillon, R., Leclaire, M., and Rey-Coyrehourcq, S. (2013).
-\newblock Openmole, a workflow engine specifically tailored for the distributed
-  exploration of simulation models.
-\newblock {\em Future Generation Computer Systems}, 29(8):1981--1990.
-
-\bibitem[Rozenblat and Pumain, 2018]{Rozenblat2018}
-Rozenblat, C. and Pumain, D. (2018).
-\newblock {\em Toward a Methodology for Multi-scalar Urban System Policies},
-  pages 385--393.
-\newblock Springer Singapore, Singapore.
-
-\bibitem[Shu et~al., 2020]{shu2020modelling}
-Shu, B., Zhu, S., Qu, Y., Zhang, H., Li, X., and Carsjens, G.~J. (2020).
-\newblock Modelling multi-regional urban growth with multilevel logistic
-  cellular automata.
-\newblock {\em Computers, Environment and Urban Systems}, 80:101457.
-
-\bibitem[Torrens and O'Sullivan, 2001]{torrens2001cellular}
-Torrens, P.~M. and O'Sullivan, D. (2001).
-\newblock Cellular automata and urban simulation: where do we go from here?
-
-\bibitem[Wegener and F{\"u}rst, 2004]{wegener2004land}
-Wegener, M. and F{\"u}rst, F. (2004).
-\newblock Land-use transport interaction: state of the art.
-\newblock {\em Available at SSRN 1434678}.
-
-\bibitem[Wegener and Spiekermann, 2018]{wegener2018multi}
-Wegener, M. and Spiekermann, K. (2018).
-\newblock Multi-level urban models: Integration across space, time and
-  policies.
-
-\bibitem[White, 2006]{white2006modeling}
-White, R. (2006).
-\newblock Modeling multi-scale processes in a cellular automata framework.
-\newblock In {\em Complex artificial environments}, pages 165--177. Springer.
-
-\bibitem[Xu and Gao, 2019]{xu2019directional}
-Xu, T. and Gao, J. (2019).
-\newblock Directional multi-scale analysis and simulation of urban expansion in
-  auckland, new zealand using logistic cellular automata.
-\newblock {\em Computers, Environment and Urban Systems}, 78:101390.
-
-\bibitem[Xu et~al., 2020]{xu2020urban}
-Xu, T., Gao, J., Coco, G., and Wang, S. (2020).
-\newblock Urban expansion in auckland, new zealand: a gis simulation via an
-  intelligent self-adapting multiscale agent-based model.
-\newblock {\em International Journal of Geographical Information Science},
-  34(11):2136--2159.
-
-\bibitem[Yu et~al., 2018]{yu2018modeling}
-Yu, Y., He, J., Tang, W., and Li, C. (2018).
-\newblock Modeling urban collaborative growth dynamics using a multiscale
-  simulation model for the wuhan urban agglomeration area, china.
-\newblock {\em ISPRS International Journal of Geo-Information}, 7(5):176.
-
-\bibitem[Zhu and Tian, 2020]{zhu2020dynamic}
-Zhu, J. and Tian, S. (2020).
-\newblock A dynamic urban lake area evolution model based on multilevel grid,
-  cellular automata, and multiagent system.
-\newblock {\em Complexity}, 2020.
-
-\end{thebibliography}
-
-\end{document}
+%\begin{listing}[H]
+%\caption{Title of the listing}
+%\rule{\textwidth}{1pt}
+%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
+%\rule{\textwidth}{1pt}
+%\end{listing}

### 2021-01-31 Docs/Papers/Multiscale/paper/paper_old.tex
arxiv

--- 2021-01-31 Docs/Papers/Multiscale/arxiv/paper.tex
@@
 
-\begin{thebibliography}{}
-
-\bibitem[Banos et~al., 2015]{banos2015importance}
-Banos, A., Corson, N., Gaudou, B., Laperri{\`e}re, V., and Coyrehourcq, S.~R.
-  (2015).
-\newblock The importance of being hybrid for spatial epidemic models: a
-  multi-scale approach.
-\newblock {\em Systems}, 3(4):309--329.
-
-\bibitem[Batty, 2005]{batty2005agents}
-Batty, M. (2005).
-\newblock Agents, cells, and cities: new representational models for simulating
-  multiscale urban dynamics.
-\newblock {\em Environment and Planning A}, 37(8):1373--1394.
-
-\bibitem[Batty, 2018]{batty2018inventing}
-Batty, M. (2018).
-\newblock {\em Inventing future cities}.
-\newblock MIT press.
-
-\bibitem[Bedau, 2002]{bedau2002downward}
-Bedau, M. (2002).
-\newblock Downward causation and the autonomy of weak emergence.
-\newblock {\em Principia: an international journal of epistemology},
-  6(1):5--50.
-
-\bibitem[Belem and M{\"u}ller, 2013]{belem2013organizational}
-Belem, M. and M{\"u}ller, J.~P. (2013).
-\newblock An organizational model for multi-scale and multi-formalism
-  simulation: Application in carbon dynamics simulation in west-african
-  savanna.
-\newblock {\em Simulation Modelling Practice and Theory}, 32:83--98.
-
-\bibitem[Boulet et~al., 2020]{boulet2020coupling}
-Boulet, X., Zargayouna, M., Scemama, G., and Leurent, F. (2020).
-\newblock Coupling multi-agent and macroscopic simulators of traffic.
-\newblock In {\em Agents and Multi-agent Systems: Technologies and Applications
-  2019}, pages 323--332. Springer.
-
-\bibitem[Chavalarias et~al., 2009]{chavalarias2009french}
-Chavalarias, D., Bourgine, P., Perrier, E., Amblard, F., Arlabosse, F., Auger,
-  P., Baillon, J.-B., Barreteau, O., Baudot, P., Bouchaud, E., et~al. (2009).
-\newblock French roadmap for complex systems 2008-2009.
-
-\bibitem[Cheng and Masser, 2003]{cheng2003modelling}
-Cheng, J. and Masser, I. (2003).
-\newblock Modelling urban growth patterns: a multiscale perspective.
-\newblock {\em Environment and Planning A}, 35(4):679--704.
-
-\bibitem[Clarke et~al., 2007]{clarke2007decade}
-Clarke, K.~C., Gazulis, N., Dietzel, C., and Goldstein, N.~C. (2007).
-\newblock A decade of sleuthing: Lessons learned from applications of a
-  cellular automaton land use change model.
-\newblock {\em Classics in IJGIS: twenty years of the international journal of
-  geographical information science and systems}, pages 413--427.
-
-\bibitem[Crociani et~al., 2016]{crociani2016multi}
-Crociani, L., L{\"a}mmel, G., and Vizzari, G. (2016).
-\newblock Multi-scale simulation for crowd management: a case study in an urban
-  scenario.
-\newblock In {\em International Conference on Autonomous Agents and Multiagent
-  Systems}, pages 147--162. Springer.
-
-\bibitem[Favaro and Pumain, 2011]{favaro2011gibrat}
-Favaro, J.-M. and Pumain, D. (2011).
-\newblock Gibrat revisited: An urban growth model incorporating spatial
-  interaction and innovation cycles.
-\newblock {\em Geographical Analysis}, 43(3):261--286.
-
-\bibitem[Ford et~al., 2019]{ford2019multi}
-Ford, A., Barr, S., Dawson, R., Virgo, J., Batty, M., and Hall, J. (2019).
-\newblock A multi-scale urban integrated assessment framework for climate
-  change studies: A flooding application.
-\newblock {\em Computers, Environment and Urban Systems}, 75:229--243.
-
-\bibitem[Gabaix, 1999]{gabaix1999zipf}
-Gabaix, X. (1999).
-\newblock Zipf's law for cities: an explanation.
-\newblock {\em Quarterly journal of Economics}, pages 739--767.
-
-\bibitem[Hjorth et~al., 2020]{hjorth2020levelspace}
-Hjorth, A., Head, B., Brady, C., Wilensky, U., et~al. (2020).
-\newblock Levelspace: A netlogo extension for multi-level agent-based modeling.
-\newblock {\em Journal of Artificial Societies and Social Simulation},
-  23(1):1--4.
-
-\bibitem[James et~al., 2018]{james2018zipf}
-James, C., Azaele, S., Maritan, A., and Simini, F. (2018).
-\newblock Zipf's and taylor's laws.
-\newblock {\em Physical Review E}, 98(3):032408.
-
-\bibitem[Joy and Vogel, 2015]{joy2015toronto}
-Joy, M. and Vogel, R.~K. (2015).
-\newblock Toronto’s governance crisis: A global city under pressure.
-\newblock {\em Cities}, 49:35--52.
-
-\bibitem[Liao and Gaudin, 2017]{liao2017ouverture}
-Liao, L. and Gaudin, J.-P. (2017).
-\newblock L’ouverture au march{\'e} en chine (ann{\'e}es 1980-2000) et le
-  d{\'e}veloppement {\'e}conomique local: une forme de gouvernance
-  multi-niveaux?
-\newblock {\em Cybergeo: European Journal of Geography}.
-
-\bibitem[Murcio et~al., 2015]{murcio2015urban}
-Murcio, R., Morphet, R., Gershenson, C., and Batty, M. (2015).
-\newblock Urban transfer entropy across scales.
-\newblock {\em PLoS One}, 10(7):e0133780.
-
-\bibitem[Pumain, 1997]{pumain1997pour}
-Pumain, D. (1997).
-\newblock Pour une th{\'e}orie {\'e}volutive des villes.
-\newblock {\em Espace g{\'e}ographique}, 26(2):119--134.
-
-\bibitem[Pumain and Reuillon, 2017]{pumain2017urban}
-Pumain, D. and Reuillon, R. (2017).
-\newblock {\em Urban Dynamics and Simulation Models}.
-\newblock Springer International.
-
-\bibitem[Raimbault, 2018]{raimbault2018calibration}
-Raimbault, J. (2018).
-\newblock Calibration of a density-based model of urban morphogenesis.
-\newblock {\em PloS one}, 13(9):e0203516.
-
-\bibitem[Raimbault, 2019a]{raimbault:halshs-02265423}
-Raimbault, J. (2019a).
-\newblock {Evolving accessibility landscapes: mutations of transportation
-  networks in China}.
-\newblock In Aveline-Dubach, N., editor, {\em {PATHWAYS OF SUSTAINABLE URBAN
-  DEVELOPMENT ACROSS CHINA: THE CASES OF HANGZHOU, DATONG AND ZHUHAI}}, pages
-  89--108. {Imago Editor}.
-\newblock ebook.
-
-\bibitem[Raimbault, 2019b]{raimbault:halshs-02284933}
-Raimbault, J. (2019b).
-\newblock {Towards integrated urban models for sustainable policies}.
-\newblock In {\em {Urban analytics at the interface between environment, human
-  health and behaviour}}, Shanghai, China. {Tongji University and The Alan
-  Turing Institute}.
-
-\bibitem[Raimbault, 2019c]{raimbault:halshs-02013006}
-Raimbault, J. (2019c).
-\newblock {Towards multi-scalar models for the co-evolution of transportation
-  networks and territories}.
-\newblock In {\em {Theo Quant 2019}}, Besan{\c c}on, France. {Universit{\'e} de
-  Franche-Comt{\'e}}.
-
-\bibitem[Raimbault, 2020a]{raimbault2020comparison}
-Raimbault, J. (2020a).
-\newblock A comparison of simple models for urban morphogenesis.
-\newblock {\em arXiv preprint arXiv:2008.13277}.
-
-\bibitem[Raimbault, 2020b]{raimbault2020hierarchy}
-Raimbault, J. (2020b).
-\newblock Hierarchy and co-evolution processes in urban systems.
-\newblock {\em arXiv preprint arXiv:2001.11989}.
-
-\bibitem[Raimbault, 2020c]{raimbault2020indirect}
-Raimbault, J. (2020c).
-\newblock Indirect evidence of network effects in a system of cities.
-\newblock {\em Environment and Planning B: Urban Analytics and City Science},
-  47(1):138--155.
-
-\bibitem[Raimbault, 2020d]{raimbault2020unveiling}
-Raimbault, J. (2020d).
-\newblock Unveiling co-evolutionary patterns in systems of cities: a systematic
-  exploration of the simpopnet model.
-\newblock In {\em Theories and Models of Urbanization}, pages 261--278.
-  Springer.
-
-\bibitem[Raimbault et~al., 2020]{raimbault2020scala}
-Raimbault, J., Perret, J., and Reuillon, R. (2020).
-\newblock A scala library for spatial sensitivity analysis.
-\newblock {\em arXiv preprint arXiv:2007.10667}.
-
-\bibitem[Reuillon et~al., 2013]{reuillon2013openmole}
-Reuillon, R., Leclaire, M., and Rey-Coyrehourcq, S. (2013).
-\newblock Openmole, a workflow engine specifically tailored for the distributed
-  exploration of simulation models.
-\newblock {\em Future Generation Computer Systems}, 29(8):1981--1990.
-
-\bibitem[Rozenblat and Pumain, 2018]{Rozenblat2018}
-Rozenblat, C. and Pumain, D. (2018).
-\newblock {\em Toward a Methodology for Multi-scalar Urban System Policies},
-  pages 385--393.
-\newblock Springer Singapore, Singapore.
-
-\bibitem[Shu et~al., 2020]{shu2020modelling}
-Shu, B., Zhu, S., Qu, Y., Zhang, H., Li, X., and Carsjens, G.~J. (2020).
-\newblock Modelling multi-regional urban growth with multilevel logistic
-  cellular automata.
-\newblock {\em Computers, Environment and Urban Systems}, 80:101457.
-
-\bibitem[Torrens and O'Sullivan, 2001]{torrens2001cellular}
-Torrens, P.~M. and O'Sullivan, D. (2001).
-\newblock Cellular automata and urban simulation: where do we go from here?
-
-\bibitem[Wegener and F{\"u}rst, 2004]{wegener2004land}
-Wegener, M. and F{\"u}rst, F. (2004).
-\newblock Land-use transport interaction: state of the art.
-\newblock {\em Available at SSRN 1434678}.
-
-\bibitem[Wegener and Spiekermann, 2018]{wegener2018multi}
-Wegener, M. and Spiekermann, K. (2018).
-\newblock Multi-level urban models: Integration across space, time and
-  policies.
-
-\bibitem[White, 2006]{white2006modeling}
-White, R. (2006).
-\newblock Modeling multi-scale processes in a cellular automata framework.
-\newblock In {\em Complex artificial environments}, pages 165--177. Springer.
-
-\bibitem[Xu and Gao, 2019]{xu2019directional}
-Xu, T. and Gao, J. (2019).
-\newblock Directional multi-scale analysis and simulation of urban expansion in
-  auckland, new zealand using logistic cellular automata.
-\newblock {\em Computers, Environment and Urban Systems}, 78:101390.
-
-\bibitem[Xu et~al., 2020]{xu2020urban}
-Xu, T., Gao, J., Coco, G., and Wang, S. (2020).
-\newblock Urban expansion in auckland, new zealand: a gis simulation via an
-  intelligent self-adapting multiscale agent-based model.
-\newblock {\em International Journal of Geographical Information Science},
-  34(11):2136--2159.
-
-\bibitem[Yu et~al., 2018]{yu2018modeling}
-Yu, Y., He, J., Tang, W., and Li, C. (2018).
-\newblock Modeling urban collaborative growth dynamics using a multiscale
-  simulation model for the wuhan urban agglomeration area, china.
-\newblock {\em ISPRS International Journal of Geo-Information}, 7(5):176.
-
-\bibitem[Zhu and Tian, 2020]{zhu2020dynamic}
-Zhu, J. and Tian, S. (2020).
-\newblock A dynamic urban lake area evolution model based on multilevel grid,
-  cellular automata, and multiagent system.
-\newblock {\em Complexity}, 2020.
-
-\end{thebibliography}
+\bibliographystyle{apalike}
+\bibliography{biblio}
 

### 2021-04-12 Docs/Papers/Multiscale/paper/paper.tex
submission_ijgi
revision_multiscale

@@
-\documentclass[sustainability,article,submit,moreauthors,pdftex]{Definitions/mdpi}
+\documentclass[ijgi,article,submit,moreauthors,pdftex]{Definitions/mdpi}
 
@@
 The design of sustainable future cities requires an historical knowledge of how past cities came to be and evolved \cite{batty2018inventing}.
-Several models have been proposed at different scales and integrating different dimensions of urban systems, such as models of land-use change at a mesoscopic scale or systems of cities models at a macroscopic scale \cite{pumain2017urban}.
+Several models have been developed at different scales and integrating different urban systems dimensions.
+Some typical examples are models of land-use change at a mesoscopic scale \cite{iacono2008models}, or systems of cities models at a macroscopic scale \cite{pumain2017urban}.
 
@@
 
-This paper contributes to the open question of multi-scalar models of urban dynamics by introducing a new simulation model which integrates a strong coupling between the mesoscopic scale and the macroscopic scale.
+This paper contributes to the open question of multi-scalar models of urban dynamics by introducing a new simulation model that integrates a strong coupling between the mesoscopic scale and the macroscopic scale.
 The dynamics within each scale influence the other and reciprocally in an iterative way.
@@
 To achieve this, the reaction-diffusion model of urban morphogenesis described by \cite{raimbault2018calibration} allows capturing concentration forces and dispersion forces, and has been shown to reproduce a large range of existing urban forms.
-Furthermore, \cite{raimbault2020comparison} showed that this model had a good coverage of the morphological space when compared to other models of urban morphogenesis.
+Furthermore, \cite{raimbault2020comparison} showed that this model had good coverage of the morphological space when compared to other models of urban morphogenesis.
 
 To couple the scales, we assume that (i) macroscopic performance of a city will influence the choices made by planers in terms of land-use, and thus parameters of the morphogenesis mode locally - empirical evidence of such links have been suggested in the literature \cite{joy2015toronto} and we stylise them here; (ii) urban form influences the inner workings of a city through congestion and the fostering of exchanges between agents (individuals, firms), and finally the global insertion of the city - thus spatial interaction parameters are updated as a function of local urban form.
-This last link is the most discutable in terms of empirical evidence and implies many other dimensions than population only.
+This last link is the most % discutable % TODO slippery?
+ in terms of empirical evidence and implies many other dimensions than population only.
 We however work on a stylised explicative model which must keep a simple structure but on which a systematic knowledge of model behavior can be developed.
@@
 
-\begin{figure}
+\begin{figure}[H]
 %  figure: more details?
@@
 We consider $N$ urban areas, represented at the macroscopic scale by their population $P_j(t)$ at time $t$, and at the mesoscopic scale by a population grid $p_{kl}^{(j)}(t)$.
-The model runs for a total number $t_f$ of time steps, and we will assume that $\Delta t = 1$ for the sake of simplicity (the formulas can be generalized for arbitrary values of the time step, for example when running on real data with irregular time sampling).
+The model runs for a total number $t_f$ of time steps, and we will assume that $\Delta t = 1$ for the sake of simplicity (the formulas can be generalised for arbitrary values of the time step, for example when running on real data with irregular time sampling).
 
@@
 	\item populations of urban areas are evolved at the macroscopic scale with the spatial interaction model (Equations \ref{eq:eq1} to \ref{eq:eq3} below);
-	\item for each urban area, parameters of the moprhogenesis model are updated given parameters capturing typical scenarios: transit-oriented development or sprawl for the diffusion parameter, metropolization or uniformization for the aggregation parameter (Equations \ref{eq:eq4} to \ref{eq:eq6});
+	\item for each urban area, parameters of the moprhogenesis model are updated given parameters capturing typical scenarios: transit-oriented development or sprawl for the diffusion parameter, metropolisation or uniformisation for the aggregation parameter (Equations \ref{eq:eq4} to \ref{eq:eq6});
 	\item local urban form are evolved conditionally to a fixed population increase and following the mescoscopic parameters;
@@
 
-\begin{table}
-	\caption{Summary of model parameters\label{tab:parameters}}
-	\centering
-	\begin{tabular}{|c|c|c|c|}
-	\hline
-		Type & Parameter & Process & Range \\\hline
-		\multirow{4}{*}{Macro} & $g_i = g_0$ & Endogenous growth & $\left[0 ; 0.5\right]$ \\
-		& $w_i = w_G$ & Interactions weight & $\left[0 ; 0.1\right]$ \\
-		& $\gamma_i = \gamma_G$ & Interactions hierarchy & $\left[0 ; 5.0\right]$ \\
-		& $d_i$ & Interactions decay & $\left[0 ; 1000\right]$ \\ \hline
-		\multirow{4}{*}{Meso} & $\alpha_i$ & Aggregation & $\left[0 ; 5\right]$ \\
-		& $\beta_i$ & Diffusion & $\left[0 ; 0.5\right]$ \\
-		& $t_m$ & Urban growth speed & $\left[0 ; 20\right]$ \\
-		& $n_d$ & Diffusion & $\left[0 ; 5\right]$ \\ \hline
-		\multirow{4}{*}{Multiscale} & $\delta\alpha$ & Downward feedback & $\left[-0.5 ; 0.5\right]$ \\
-		& $\delta\beta$ & Downward feedback & $\left[-0.5 ; 0.5\right]$ \\
-		& $\delta d$ & Upward feedback & $\left[-0.5 ; 0.5\right]$ \\\hline
-		& $\lambda$ & Cost of congestion & $\left[0 ; 2\right]$ \\\hline
-	\end{tabular}
-\end{table}
-
 \subsubsection{Bottom-up feedback}
@@
 
+\begin{table}[H]
+	\caption{Summary of model parameters\label{tab:parameters}}
+	\centering
+	\begin{tabular}{|c|c|c|c|}
+	\hline
+		Type & Parameter & Process & Range \\\hline
+		\multirow{4}{*}{Macro} & $g_i = g_0$ & Endogenous growth & $\left[0 ; 0.5\right]$ \\
+		& $w_i = w_G$ & Interactions weight & $\left[0 ; 0.1\right]$ \\
+		& $\gamma_i = \gamma_G$ & Interactions hierarchy & $\left[0 ; 5.0\right]$ \\
+		& $d_i$ & Interactions decay & $\left[0 ; 1000\right]$ \\ \hline
+		\multirow{4}{*}{Meso} & $\alpha_i$ & Aggregation & $\left[0 ; 5\right]$ \\
+		& $\beta_i$ & Diffusion & $\left[0 ; 0.5\right]$ \\
+		& $t_m$ & Urban growth speed & $\left[0 ; 20\right]$ \\
+		& $n_d$ & Diffusion & $\left[0 ; 5\right]$ \\ \hline
+		\multirow{4}{*}{Multiscale} & $\delta\alpha$ & Downward feedback & $\left[-0.5 ; 0.5\right]$ \\
+		& $\delta\beta$ & Downward feedback & $\left[-0.5 ; 0.5\right]$ \\
+		& $\delta d$ & Upward feedback & $\left[-0.5 ; 0.5\right]$ \\\hline
+		& $\lambda$ & Cost of congestion & $\left[0 ; 2\right]$ \\\hline
+	\end{tabular}
+\end{table}
+
 % ! provide nominal values
@@
 The model is simulated for 20 macroscopic time steps (order of magnitude of half a century), each representing $t_m$ mesoscopic time steps (parameter changed in experiments).
-
-%%%
-% ! fig placement
-
-\begin{figure}[h!]
-	\includegraphics[width=\linewidth]{figures/allindics_hist.png}
-	\caption{Statistical distribution of indicators for a sample of 10 parameter values.\label{fig:fig2}}
-\end{figure}
-
-\begin{figure}[h!]
-	\includegraphics[width=\linewidth]{figures/onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
-	\caption{Variations of indicators as a function of $\delta d$, when all other parameters are fixed to their nominal value.\label{fig:fig3}}
-\end{figure}
 
@@
 
+\begin{figure}[H]
+	\includegraphics[width=\linewidth]{figures/allindics_hist.png}
+	\caption{Statistical distribution of indicators for a sample of 10 parameter values.\label{fig:fig2}}
+\end{figure}
+
 % one factor sampling
@@
 We run experiments with 50 replications for each parameter value in the following.
-
-\begin{figure}[h!]
-	\includegraphics[width=\linewidth]{figures/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
-	\caption{Macroscopic hierarchy of populations as a function of mesoscopic time steps $t_m$, for different values of interaction decay $d_G$ (colour), and of feedback parameters $\delta d$ (columns) and $\delta \beta$ (rows).\label{fig:fig4}}
-\end{figure}
-
-\begin{figure}[h!]
-	\includegraphics[width=\linewidth]{figures/deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
-	\caption{Relative evolution of average mescoscopic Moran, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig5}}
-\end{figure}
-
-\begin{figure}[h!]
-	\includegraphics[width=\textwidth]{figures/deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
-	\caption{Relative evolution of average mescoscopic population hierarchy, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig6}}
-\end{figure}
-
-\begin{figure}[h!]
-	\begin{center}
-	\includegraphics[width=0.6\linewidth]{figures/pareto_macroHierarchy-mesoHierarchy.png}
-	\end{center}
-	\caption{Pareto front obtained with a NSGA2 genetic algorithm, to minimize simultaneously hierarchy at the macroscopic and mesoscopic scales.
-Point size gives the number of stochastic samples and colour the value of $\delta \beta$.\label{fig:fig7}}
-\end{figure}
 
@@
 
+\begin{figure}[H]
+	\includegraphics[width=\linewidth]{figures/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
+	\caption{Macroscopic hierarchy of populations as a function of mesoscopic time steps $t_m$, for different values of interaction decay $d_G$ (colour), and of feedback parameters $\delta d$ (columns) and $\delta \beta$ (rows).\label{fig:fig4}}
+\end{figure}
+
 We show in Fig.~\ref{fig:fig3} that both macroscopic and mesoscopic indicators have a U-shaped behavior as a function of the bottom-up feedback parameter $\delta d$.
@@
 For the urban form, this feedback also mitigates the aggregation of population.
+
+\begin{figure}[H]
+	\includegraphics[width=\linewidth]{figures/onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
+	\caption{Variations of indicators as a function of $\delta d$, when all other parameters are fixed to their nominal value.\label{fig:fig3}}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=\linewidth]{figures/deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+	\caption{Relative evolution of average mescoscopic Moran, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig5}}
+\end{figure}
 
@@
 
+\begin{figure}[H]
+	\includegraphics[width=\textwidth]{figures/deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+	\caption{Relative evolution of average mescoscopic population hierarchy, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig6}}
+\end{figure}
+
 When investigating population hierarchy the mesoscopic scale, i.e.
@@
 This means that sprawl may be needed in some configurations to optimise this objective.
+
+\begin{figure}[H]
+	\begin{center}
+	\includegraphics[width=0.6\linewidth]{figures/pareto_macroHierarchy-mesoHierarchy.png}
+	\end{center}
+	\caption{Pareto front obtained with a NSGA2 genetic algorithm, to minimize simultaneously hierarchy at the macroscopic and mesoscopic scales.
+Point size gives the number of stochastic samples and colour the value of $\delta \beta$.\label{fig:fig7}}
+\end{figure}
 

### 2021-04-13 Docs/Communications/2021/FRCCS/abstract.tex
skeleton_abstract_frccs

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Integrating and validating urban simulation models}
\subtitle{}

%\titlerunning{Short form of title}        % if too long for running head

\author{Juste Raimbault}

%\authorrunning{Short form of author list} % if too long for running head

\institute{J. Raimbault \at
              Center for Advanced Spatial Analysis, University College London\\
              \email{j.raimbault@ucl.ac.uk}
}

%\date{Received: date / Accepted: date}
\date{}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}

\keywords{Urban simulation models \and Model coupling \and Model validation}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}

The integration of environmental dimensions within simulation models for territorial systems from theoretical and quantitative geography is an other important implementation project. For example, \cite{viguie2014downscaling} use an urban growth model to evaluate the local impact of climate change, but does not make associated aspects endogenous, such as energy price or the energy efficiency of the urban structure. Negative externalities such as congestion and pollution, directly linked to emissions, have a feedback for example on the location of activities. More generally, systematic links between a precise description of urban structure and energy efficiency, in a dynamical and endogenous way, remains to be established. Similarly, numerous studies in ecology which establish the impact of anthropic habitat disturbances would particularly benefit from a coupling with urban growth models, for example for a better management of the interface between the city and nature within the new urban regimes that are urban mega-regions \cite{hall2006polycentric}.

At the macroscopic scale, several couplings between models of systems of cities and ecological models or from environmental science are possible and desirable. In terms of territorial planning at small scales, an estimation of the impact of flows from interactions between cities on traversed areas can for example yield compromises between the distribution of flows and their environmental impact, and establish recommendations to minimise the global impact of the urban system, with. the constraint of keeping a reasonable resilience and economic performance. In the same way, the questions of production, storage and distribution of energy are endogenous to territorial systems, and the dynamics of infrastructures and associated entities can. be integrated within models of systems of cities. This research direction can in fact be generalised to any type of resource, and the SimpopLocal model \cite{pumain2017simpoplocal} and also the Marius model family \cite{cottineau2015modular} have integrated these aspects without however making them endogenous nor making them a central component of models. This aspect is linked to the previous point, given the difficulty to endogenise energy within macroeconomic models.

This research project, inspired by sciences of complexity and a principally geographical viewpoint, proposes the construction of bridges, coupling, and interdisciplinary dialogues, i.e. the construction of integrated theories. A way to understand such approaches is described by the complex systems roadmap \cite{bourgine2009french}. It combines horizontal integration (fundamental transversal questions at the intersection of different types of complex systems) with vertical integration (multiple levels coupled within multi-scale models). 
%\cite{raimbault2017applied} has furthermore proposed that the integration of knowledge domains (defined as components of knowledge construction processes, including theories, models and empirical facts, but also data, tools and methods) as typical of complex approaches.
%Considering modeling as a crucial element of knowledge production, this project is thus situated within approaches in theoretical and quantitative geography \citep{cuyala2014analyse} and geosimulation \citep{benenson2004geosimulation} in the case of simulation models.

\section{Horizontal integration: coupling urban models and dimensions}

\section{Vertical integration: constructing multi-scale models}

Finally, the construction of multi-scale models will be in itself a crucial aspect, since it is a dimension in itself of the integration expected for the theories and models constructed. Following \cite{rozenblat2018conclusion}, given the multiple levels of articulation and interdependency that systems of cities have reached, the management and planning must necessarily be multi-scalar in order to take into account the geographical particularities while ensuring a global consistence which yield limited inequalities between territories.

%A possible concrete entry for a first exploration of multi-scale models, already aforementioned, would be the coupling of the evolutionary theory of cities with scaling theories. The first focuses on particularities of territorial entities while the second aims at producing universal laws, and both provide consistent explanations for scaling laws. A coupling strategy would be, relying on consolidated and comparable dataset, to (i) determine modular decompositions of territorial systems and corresponding scales, and the quantification of the universality of laws within these sub-systems by the study of scaling laws within and between subsystems; (ii) modeling this multi-scale system by. coupling. models of urban growth, which would be validated by scaling properties.

Moreover, a considerable methodological work would be necessary to elaborate coupling techniques between scales (for example as the hybrid modeling coupling agent-based models with differential equations for an epidemiological model \cite{banos2015coupling}), to determine the relevance of levels to be included and avoid ``ontological dead-ends'' \cite{roth2006reconstruction}, and to determine the nature of retroactions between scales and their necessity.

\section{Model exploration and validation}

Working to integrate models and theories necessarily necessitates a fine understanding of the modeling process itself, but also of stylised dynamics produced by simulation models. In the case of geographical models, such a knowledge has for example been developed within the Geodivercity ERC project \cite{pumain2017urban}, which lead to the conception of new methods answering to specific questions coming from geographical concerns (for example necessity and sufficiency of processes in a multi-modeling context \cite{reuillon2015}, exploration. of the feasible space of model outputs \cite{10.1371/journal.pone.0138212}).
%More generally in social sciences, modeling and simulation driven by new practices including model coupling, the use of high-performance computing for model exploration, open science practices, could be at the origin of new types of integrated knowledge \cite{banos2013pour}.

Besides, the development of methods and tools to improve the extraction of knowledge from simulation models, and an epistemological investigation on modeling practices, are crucial within this direction which can thus be understood as a theoretical and methodological direction. Indeed, methods to validate simulation models mostly remain to be developed. Methods for the exploration, sensitivity analysis, and validation, are essential for robust application of models, but also yield a better complementarity with other types of approaches since they can establish when modeling is not relevant.

\section{Discussion: towards evidence-based multi-scalar sustainable territorial planning}

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
\bibliographystyle{spphys}       % APS-like style for physics
\bibliography{biblio.bib}   % name your BibTeX data base

\end{document}
% end of file template.tex

%
%% For one-column wide figures use
%\begin{figure}
%% Use the relevant command to insert your figure file.
%% For example, with the graphicx package use
%  \includegraphics{example.eps}
%% figure caption is below the figure
%\caption{Please write your figure caption here}
%\label{fig:1}       % Give a unique label
%\end{figure}
%%
%% For two-column wide figures use
%\begin{figure*}
%% Use the relevant command to insert your figure file.
%% For example, with the graphicx package use
%  \includegraphics[width=0.75\textwidth]{example.eps}
%% figure caption is below the figure
%\caption{Please write your figure caption here}
%\label{fig:2}       % Give a unique label
%\end{figure*}
%%
%% For tables use
%\begin{table}
%% table caption is above the table
%\caption{Please write your table caption here}
%\label{tab:1}       % Give a unique label
%% For LaTeX tables use
%\begin{tabular}{lll}
%\hline\noalign{\smallskip}
%first & second & third  \\
%\noalign{\smallskip}\hline\noalign{\smallskip}
%number & number & number \\
%number & number & number \\
%\noalign{\smallskip}\hline
%\end{tabular}
%\end{table}

### 2021-06-07 Models/MesoBenchmark/pseResults.R
pse_reactiondiff

@@ -14,10 +14,12 @@ dir.create(resdir)
 resdirs = list(
   #'expmixture' = '20200828_1515_PSE_EXPMIXTURE_res'
   'expmixture' = 'openmole/pse/20201125_1115_PSE_EXPMIXTURE_REPLICATIONS_GRID',
-  'reactiondiffusion' = 'openmole/pse/20200828_1624_PSE_REACTIONDIFFUSION_res',
+  #'reactiondiffusion' = 'openmole/pse/20200828_1624_PSE_REACTIONDIFFUSION_res',
+  'reactiondiffusion' = 'openmole/pse/20210531_0803_PSE_REACTIONDIFFUSION_REPLICATIONS_LOCAL_res/',
   # 'correlatedperco' = '20200829_1330_PSE_CORRELATEDPERCO_res'
   'correlatedperco' = 'openmole/pse/20201124_1816_PSE_CORRELATEDPERCO_REPLICATIONS_GRID/',
-  'gravity' = 'openmole/pse/20200829_1526_PSE_GRAVITY_res'
+  #'gravity' = 'openmole/pse/20200829_1526_PSE_GRAVITY_res'
+  'gravity' = 'openmole/pse/20210507_1744_PSE_GRAVITY_REPLICATIONS_LOCAL/'
 )
 
 stochastic = FALSE

### 2021-06-25 Docs/Papers/Multiscale/paper/paper.tex
.
bib
revisions

@@
 \newcommand{\orcidauthorA}{0000-0003-0768-9480} % Add \orcidA{} behind the author's name
-%\newcommand{\orcidauthorB}{0000-0000-000-000X} % Add \orcidB{} behind the author's name
 
@@
 \address{%
-$^{1}$ \quad CASA, University College London\\
+$^{1}$ \quad Center for Advanced Spatial Analysis, University College London\\
 $^{2}$ \quad Institut des Syst{\`e}mes Complexes Paris-Ile-de-France, CNRS\\
@@
 
+\cite{lockerby2015asynchronous} physics multiscale simulations
++ multiscale traffic models (Nancy?)
+
+\cite{vinci2021population} : link macro migration - meso structure (polycentrism)
+
+def of scale \cite{atkinson2000spatial}
+
+\cite{jantz2005analysis}
+
 \section{Introduction}
@@
 
+% TODO emphasize more the research problem
+
 This paper contributes to the open question of multi-scalar models of urban dynamics by introducing a new simulation model that integrates a strong coupling between the mesoscopic scale and the macroscopic scale.
@@
 
-% ! reformulate - this is the CCS abstract
+% TODO ! reformulate - this is the CCS abstract
 
@@
 To couple the scales, we assume that (i) macroscopic performance of a city will influence the choices made by planers in terms of land-use, and thus parameters of the morphogenesis mode locally - empirical evidence of such links have been suggested in the literature \cite{joy2015toronto} and we stylise them here; (ii) urban form influences the inner workings of a city through congestion and the fostering of exchanges between agents (individuals, firms), and finally the global insertion of the city - thus spatial interaction parameters are updated as a function of local urban form.
-This last link is the most % discutable % TODO slippery?
- in terms of empirical evidence and implies many other dimensions than population only.
+This last link is the most questionable assumption in terms of empirical evidence and implies many other dimensions than population only.
 We however work on a stylised explicative model which must keep a simple structure but on which a systematic knowledge of model behavior can be developed.
@@
 \begin{figure}[H]
-%  figure: more details?
+% TODO figure: more details?
  \includegraphics[width=\linewidth]{figures/multiscale_morph.pdf}
- \caption{Schematic representation of model architecture.
+ \caption{Schematic representation of the model architecture.
 Urban areas with their own urban form are inserted into the macroscopic spatial interaction model, and strong coupling is achieved through top-down and bottom-up feedbacks.\label{fig:fig1}}
@@
 	\item populations of urban areas are evolved at the macroscopic scale with the spatial interaction model (Equations \ref{eq:eq1} to \ref{eq:eq3} below);
-	\item for each urban area, parameters of the moprhogenesis model are updated given parameters capturing typical scenarios: transit-oriented development or sprawl for the diffusion parameter, metropolisation or uniformisation for the aggregation parameter (Equations \ref{eq:eq4} to \ref{eq:eq6});
-	\item local urban form are evolved conditionally to a fixed population increase and following the mescoscopic parameters;
+	\item for each urban area, parameters of the morphogenesis model are updated given parameters capturing typical scenarios: transit-oriented development or sprawl for the diffusion parameter, metropolisation or uniformisation for the aggregation parameter (Equations \ref{eq:eq4} to \ref{eq:eq6});
+	\item local urban form are evolved conditionally to a fixed population increase and following the mesoscopic parameters;
 	\item congestions and interactions within areas are synthesised following a cost function, which is used to update the macroscopic interaction parameters (Equations \ref{eq:eq7} and \ref{eq:eq8}).
@@
 \begin{equation}
-	\label{eq:eq5}
+	\label{eq:eq6}
 	Z_i = \sum_j \frac{P_j}{\sum_k P_k} \cdot \exp( - d_{ij} / d_i)
@@
 \begin{equation}
-\label{eq:eq6}
+\label{eq:eq7}
 		U_i = \sum_{kl} \left( \frac{P_k P_l}{P^2} \cdot \frac{1}{d_{kl}} - \lambda \left(\frac{P_k P_l}{P^2} \cdot \frac{1}{d_{kl}}\right)^2 \right)
@@
 \begin{equation}
-\label{eq:eq7}
+\label{eq:eq8}
 	d_i (t+1) = d_i (t) \left( 1 + \delta d \cdot \frac{U_i}{\max_k \left|U_k\right|} \right)
@@
 
-The Table~\ref{tab:parameters} summarizes model parameters. For clarity, we do not include parameters linked to synthetic model setup, and distinguish them by the scale to which they are attached.
+The Table~\ref{tab:parameters} summarises model parameters. For clarity, we do not include parameters linked to synthetic model setup, and distinguish them by the scale to which they are attached.
+% TODO ref table beginning section
 
@@
 		& $\delta\beta$ & Downward feedback & $\left[-0.5 ; 0.5\right]$ \\
-		& $\delta d$ & Upward feedback & $\left[-0.5 ; 0.5\right]$ \\\hline
+		& $\delta d$ & Upward feedback & $\left[-0.5 ; 0.5\right]$ \\
 		& $\lambda$ & Cost of congestion & $\left[0 ; 2\right]$ \\\hline
@@
 
+% TODO extend indicator descritpion
+
 \section{Results}
@@
 Therefore, the parameter space is explored with the OpenMOLE model exploration software \cite{reuillon2013openmole}.
-Standalone source code for the model is available on an open git repository at \url{https://github.com/JusteRaimbault/UrbanGrowth-model}, while the repository for results is at \url{https://github.com/JusteRaimbault/UrbanGrowth}.
-Simulation data used in this paper is available on the dataverse repository at \url{https://doi.org/10.7910/DVN/IRHMQK}.
+Standalone source code for the model is available on an open git repository as \cite{urbangrowthmodel}.
+An other open repository for the analysis of results is also available as \cite{urbangrowth}.
+Simulation data used in this paper is available on the dataverse repository as \cite{dataverse}.
 
@@
 
+% TODO improve discussion - link with literature: related results, interpretation regarding previous studies
+
+% TODO limitations of the work
+
 \subsection{Theoretical and practical implications}
+
+% TODO implication of results in a broad context
 

### 2022-09-28 Docs/Papers/Multiscale/paper/paper.tex
.
template

@@
-\documentclass[ijgi,article,submit,moreauthors,pdftex]{Definitions/mdpi}
+\documentclass[referee,lineno,pdflatex,sn-apa]{sn-jnl}
 
-%=================================================================
-\firstpage{1} 
-\makeatletter 
-\setcounter{page}{\@firstpage} 
-\makeatother
-\pubvolume{xx}
-\issuenum{x}
-\articlenumber{x}
-\pubyear{2020}
-\copyrightyear{2020}
-%\externaleditor{Academic Editor: name}
-\history{Received: date; Accepted: date; Published: date}
-%\updates{yes} % If there is an update available, un-comment this line
-%% MDPI internal command: uncomment if new journal that already uses continuous page numbers 
-%\continuouspages{yes}
+\jyear{2022}%
 
-%=================================================================
-% Full title of the paper (Capitalized)
-\Title{Strong coupling between scales in a multi-scalar model of urban dynamics}
+\raggedbottom
 
-% Author Orchid ID: enter ID or remove command
-\newcommand{\orcidauthorA}{0000-0003-0768-9480} % Add \orcidA{} behind the author's name
+\begin{document}
 
-% Authors, for the paper (add full first names)
-\Author{Juste Raimbault $^{1,2,3,*}$\orcidA{}}
+\title[Multi-scalar urban dynamics model]{Strong coupling between scales in a multi-scalar model of urban dynamics}
 
-% Authors, for metadata in PDF
-\AuthorNames{Juste Raimbault}
+\author*[1,2,3,4]{\pfx{Dr} \fnm{Juste} \sur{Raimbault} \dgr{PhD}}\email{juste.raimbault@ign.fr}
+% orcid 0000-0003-0768-9480
 
-% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
-\address{%
-$^{1}$ \quad Center for Advanced Spatial Analysis, University College London\\
-$^{2}$ \quad Institut des Syst{\`e}mes Complexes Paris-Ile-de-France, CNRS\\
-$^{3}$ \quad UMR G{\'e}ographie-cit{\'e}s, CNRS}
-
-% Contact information of the corresponding author
-\corres{Correspondence: juste.raimbault@polytechnique.edu}
-
-% Current address and/or shared authorship
-% \dagger,\ddagger
-%\firstnote{Current address: Affiliation 3} 
-%\secondnote{These authors contributed equally to this work.}
-% The commands \thirdnote{} till \eighthnote{} are available for further notes
+\affil*[1]{\orgdiv{LASTIG}, \orgname{Univ. Gustave Eiffel, IGN-ENSG}} %, \orgaddress{\street{73 avenue de Paris}, \city{Saint-Mand{\'e}}, \postcode{93160}, \country{France}}
+\affil[2]{\orgdiv{Center for Advanced Spatial Analysis}, \orgname{University College London}} % , \orgaddress{\street{90 Tottenham Court Road}, \city{London}, \postcode{}, \country{UK}}
+\affil[3]{\orgdiv{UPS CNRS 3611 ISC-PIF}, \orgname{CNRS}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
+\affil[4]{\orgdiv{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s}, \orgname{Universit{\'e} Paris 1}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
 
@@
 
-%I am pleased to submit an original research article entitled "Strong coupling between scales in a multi-scalar model of urban dynamics" for consideration for publication in IJGI.
-This article introduces a novel modeling study on the dynamics of urban systems, with a particular on feedbacks between scales.
-The subject is in my opinion furthermore particularly suited for the special issue "Spatio-Temporal Models and Geo-Technologies".
+%I am pleased to submit an original research article entitled "Strong coupling between scales in a multi-scalar model of urban dynamics" for consideration for publication in Urban Informatics.
+This article introduces a novel modeling study on the dynamics of urban systems, with a particular focus on feedbacks between scales.
 
 %A preliminary version of the methodology and results were presented as a main track talk at the Conference on Complex Systems 2019.
-A preprint was submitted on arXiv at the same date of this submission.
+An older preprint was deposited on arXiv with the identifier arXiv:2101.12725.
 
@@
 %Juste Raimbault
-%University College London
+%IGN-ENSG
 
@@
 % Keywords
-\keyword{Urban dynamics; Systems of cities; Urban morphogenesis; Multi-scalar modeling; Strong coupling}
+\keywords{Urban dynamics, Systems of cities, Urban morphogenesis, Multi-scalar modelling, Strong coupling}
 
-%\setcounter{secnumdepth}{4}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\maketitle
+
+\cite{feng2022effects}
+
+\cite{venter2019mobility} mobility-congestion: eco vitality: feedback
+
+\cite{Ludeschere1922872118} nw sci / climate numerical sims: sort of multiscale
 
@@
 
-The modeling of urban growth and more generally the dynamics of urban systems is central to the design of sustainable territorial policies, through the understanding of past urbanisation processes and the anticipation of future urban trajectories.
-The design of sustainable future cities requires an historical knowledge of how past cities came to be and evolved \cite{batty2018inventing}.
-Several models have been developed at different scales and integrating different urban systems dimensions.
-Some typical examples are models of land-use change at a mesoscopic scale \cite{iacono2008models}, or systems of cities models at a macroscopic scale \cite{pumain2017urban}.
+Modelling urban growth and more generally the dynamics of urban systems is central to the design of sustainable territorial policies, enabling the understanding of past urbanisation processes and the anticipation of future urban trajectories.
+The design of sustainable future cities requires an historical knowledge of how past cities came to be and evolved \citep{batty2018inventing}.
+Several models have been developed at different scales and integrating multiple urban systems dimensions.
+Some typical examples are models of land-use change at a mesoscopic scale \citep{iacono2008models}, or systems of cities models at a macroscopic scale \citep{pumain2017urban}.
 
-At the scale of a metropolitan area, Land-use Transport Interaction models \cite{wegener2004land} are for example a widely used tool to estimate the dynamics of spatial distributions of activities (mostly residential location and economic activities) in response to an evolution of the accessibility landscape caused by new transportation infrastructures \cite{raimbault:halshs-02265423}.
-In a similar context, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \cite{clarke2007decade}.
+At the scale of a metropolitan area, Land-use Transport Interaction models \citep{wegener2004land} are a widely used tool to estimate the dynamics of spatial distributions of activities (mostly residential location and economic activities) in response to an evolution of the accessibility landscape caused by new transportation infrastructures \citep{raimbault:halshs-02265423}.
+In a similar context, cellular automata models of urban growth or land-use change study more generally land-use transitions with a high spatial resolution, and are mostly data-driven \citep{clarke2007decade}.
 At the smaller scale of the system of cities, macroscopic models of urban growth have focused on reproducing the distribution of city sizes, either through economic processes as e.g.
-\cite{gabaix1999zipf}, or from a geographical point of view focusing on interactions between cities \cite{favaro2011gibrat}.
+\citep{gabaix1999zipf}, or from a geographical point of view focusing on interactions between cities \citep{favaro2011gibrat}.
 
 Territorial dynamics, and more particularly urban dynamics, have according to \cite{pumain1997pour} an intrinsic multi-scalar nature, with successive autonomous levels of emergence from individual microscopic agents to the mesoscopic scale of the city and the macroscopic scale of the system of cities.
-While models at each scale with distinct ontologies are useful to answer their own questions, an explicit account of inter-scale feedbacks, both top-down and bottom-up, would allow testing policies and interventions distributed and differentiated across scales while not neglecting the interactions between scales \cite{wegener2018multi}.
+While models at each scale with distinct ontologies are useful to answer their own questions, an explicit account of inter-scale feedbacks, both top-down and bottom-up, would allow testing policies and interventions distributed and differentiated across scales while not neglecting the interactions between scales \citep{wegener2018multi}.
 Indeed, the need for sustainable territorial policies would imply the construction of multi-scalar models to take simultaneously into account issues associated to each relevant scale \cite{Rozenblat2018,raimbault:halshs-02284933}.
@@
 
-In disciplines neighbour to urban modeling, methods have been developed for multi-scale models.
-For example in spatial epidemiology, \cite{banos2015importance} combines agent-based modeling for local diffusion dynamics with differential equations at the population scale.
-The NetLogo software for agent-based modeling includes a specific extension for multi-scale modeling \cite{hjorth2020levelspace}.
+In disciplines close to urban modelling, methods have been developed for multi-scale models.
+For example in spatial epidemiology, \cite{banos2015importance} combines agent-based modelling for local diffusion dynamics with differential equations at the population scale.
+The NetLogo software for agent-based modelling includes a specific extension for multi-scale modelling \cite{hjorth2020levelspace}.
 Multi-scale models have also been used for the simulation of crowd dynamics \cite{crociani2016multi}.
@@
 Our contribution is novel regarding previous works in particular regarding the following points: (i) the stylised model explicitly couples distinct scales and ontologies in a strong manner, most models operating only a weak coupling between scales (i.e.
-no reciprocal and dynamical feedbacks); (ii) the behavior of the model is systematically studied on synthetic systems of cities using model exploration methods.
+no reciprocal and dynamical feedbacks); (ii) the behaviour of the model is systematically studied on synthetic systems of cities using model exploration methods.
 
@@
 
-% We describe in the following stylized facts justifying the approach => this could (should?) be an important development/basis for the work?
-
-%Strong coupling and multi-scalarity 
-%   Distinct processes and objects within each different scale
-%   Weak inter-scale coupling, such as progressive resolution for land-use model, does not considers emergence and autonomous scales
-% An integrated model, or strongly coupled, would be \textit{a new model extending the two coupled models in the sense that it includes them in some parameter settings or limit conditions}
-%Urban multi-scalar complexity must be captured by a strongly coupled model
-%no strongly coupled multi-scalar model in the literature (few examples such as  \cite{murcio2015urban} but without distinct ontologies)
-%need to however consider ``simple'' models to be able to understand their behavior and extract knowledge from them
-%\textbf{Research objective: }
-%\textit{Investigate a strong coupling of a simple urban system interaction network at the macroscopic scale with an urban morphogenesis model at the mesoscopic scale}
-
 \section{Multi-scale urban dynamics model}
@@
 
-% TODO ! reformulate - this is the CCS abstract
-
-The main characteristic of our model is a strong coupling between the macroscopic scale of a system of cities with the mesoscopic scale of metropolitan areas.
+The central feature of our urban simulation model is a strong coupling between the macroscopic scale of a system of cities with the mesoscopic scale of metropolitan areas.
 We consider simple urban dimensions with population variables only to describe cities, and use stylised processes.
-The system of cities evolves following a spatial interaction model as described by \cite{raimbault2020indirect}, assuming that interaction flows between cities will increase their attractivity and thus their population growth rate.
+The system of cities evolves following a spatial interaction model as described by \cite{raimbault2020indirect}, assuming that interaction flows between cities will increase their attractiveness and thus their population growth rate.
 Population updates are not done directly following migration flows as in stochastic urban models such as in \cite{james2018zipf}, but with the hypothesis of hidden variables (generally scaling with population) capturing for example economic dimensions, which influence population growth rate.
@@
 This last link is the most questionable assumption in terms of empirical evidence and implies many other dimensions than population only.
-We however work on a stylised explicative model which must keep a simple structure but on which a systematic knowledge of model behavior can be developed.
+We however work on a stylised explicative model which must keep a simple structure but on which a systematic knowledge of model behaviour can be developed.
 We show in Fig.~\ref{fig:fig1} a schematic description of model structure.
 
-\begin{figure}[H]
-% TODO figure: more details?
+%%%%%%%%%%%%%%%%%%
+\begin{figure}[h]
  \includegraphics[width=\linewidth]{figures/multiscale_morph.pdf}
@@
 \end{figure}
+%%%%%%%%%%%%%%%%%%
 
-\subsection{Formalization}
+\subsection{Formalisation}
 
@@
 \begin{equation}
-\label{eq:eq8}
-	d_i (t+1) = d_i (t) \left( 1 + \delta d \cdot \frac{U_i}{\max_k \left|U_k\right|} \right)
+	d_i (t+1) = d_i (t) \left( 1 + \delta d \cdot \frac{U_i}{\max_k \lvert U_k \rvert } \right) \label{eq:eq8}
 \end{equation}
@@
 
-\begin{table}[H]
+\begin{table}[h]
 	\caption{Summary of model parameters\label{tab:parameters}}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 	\includegraphics[width=\linewidth]{figures/allindics_hist.png}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 	\includegraphics[width=\linewidth]{figures/onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 	\includegraphics[width=\linewidth]{figures/deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 	\includegraphics[width=\textwidth]{figures/deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 	\begin{center}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\vspace{6pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\authorcontributions{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\funding{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{The author acknowledges the funding of their institutions and the EPSRC project number EP/M023583/1.
+% acknowledgments
+%The author acknowledges the funding of their institutions and the EPSRC project number EP/M023583/1.
 Results obtained in this paper were computed on the vo.complex-system.eu virtual organisation of the European Grid Infrastructure (http://www.egi.eu).
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\conflictsofinterest{}
+\bibliography{biblio}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\reftitle{References}
-
-\externalbibliography{yes}
-\bibliography{biblio.bib}
-
-%% for journal Sci
-%\reviewreports{\\
-%Reviewer 1 comments and authors’ response\\
-%Reviewer 2 comments and authors’ response\\
-%Reviewer 3 comments and authors’ response
-%}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{document}
 
-%\begin{figure}[H]
+%\begin{table}[h]
+%\begin{center}
+%\begin{minipage}{\textwidth}
+%\caption{Example of a lengthy table which is set to full textwidth}\label{tab2}
+%\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
+%\toprule%
+%& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]} & \multicolumn{3}{@{}c@{}}{Element 2\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
+%Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
+%\midrule
+%Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
+%Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
+%\botrule
+%\end{tabular*}
+%\footnotetext{Note: This is an example of table footnote.
+This is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
+%\footnotetext[1]{Example for a first table footnote.}
+%\footnotetext[2]{Example for a second table footnote.}
+%\end{minipage}
+%\end{center}
+%\end{table}
+%
+%
+%\begin{figure}[h]%
 %\centering
-%\includegraphics[width=2 cm]{Definitions/logo-mdpi}
-%\caption{This is a figure, Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
-(\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.}
+%\includegraphics[width=0.9\textwidth]{fig.eps}
+%\caption{This is a widefig. This is an example of long caption this is an example of long caption  this is an example of long caption this is an example of long caption}\label{fig1}
 %\end{figure}
-
-%\begin{table}[H]
-%\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
-%\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-%\begin{tabular}{ccc}
-%\toprule
-%\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-%\midrule
-%entry 1		& data			& data\\
-%entry 2		& data			& data\\
-%\bottomrule
-%\end{tabular}
-%\end{table}
-
-%\begin{listing}[H]
-%\caption{Title of the listing}
-%\rule{\textwidth}{1pt}
-%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
-%\rule{\textwidth}{1pt}
-%\end{listing}
+%

### 2022-09-28 Docs/Papers/Multiscale/paper/paper.tex
.

@@
 \affil*[1]{\orgdiv{LASTIG}, \orgname{Univ. Gustave Eiffel, IGN-ENSG}} %, \orgaddress{\street{73 avenue de Paris}, \city{Saint-Mand{\'e}}, \postcode{93160}, \country{France}}
-\affil[2]{\orgdiv{Center for Advanced Spatial Analysis}, \orgname{University College London}} % , \orgaddress{\street{90 Tottenham Court Road}, \city{London}, \postcode{}, \country{UK}}
+\affil[2]{\orgdiv{Center for Advanced Spatial Analysis}, \orgname{UCL}} % , \orgaddress{\street{90 Tottenham Court Road}, \city{London}, \postcode{}, \country{UK}}
 \affil[3]{\orgdiv{UPS CNRS 3611 ISC-PIF}, \orgname{CNRS}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
@@
 While models at each scale with distinct ontologies are useful to answer their own questions, an explicit account of inter-scale feedbacks, both top-down and bottom-up, would allow testing policies and interventions distributed and differentiated across scales while not neglecting the interactions between scales \citep{wegener2018multi}.
-Indeed, the need for sustainable territorial policies would imply the construction of multi-scalar models to take simultaneously into account issues associated to each relevant scale \cite{Rozenblat2018,raimbault:halshs-02284933}.
+Indeed, the need for sustainable territorial policies would imply the construction of multi-scalar models to take simultaneously into account issues associated to each relevant scale \citep{Rozenblat2018,raimbault:halshs-02284933}.
 
@@
 \cite{batty2005agents} however suggests that a similar formalism can be applied to urban processes at different scales.
-Multi-level statistical models capture some information at imbricated scales \cite{shu2020modelling}, although they can not be used as dynamical simulation models.
-Similarly, multi-level cellular automata (CA) models for urban growth include factors influencing urban expansion at multiple scales \cite{xu2019directional}.
+Multi-level statistical models capture some information at imbricated scales \citep{shu2020modelling}, although they can not be used as dynamical simulation models.
+Similarly, multi-level cellular automata (CA) models for urban growth include factors influencing urban expansion at multiple scales \citep{xu2019directional}.
 \cite{cheng2003modelling} propose a general framework for such approaches.

### 2022-10-26 Docs/Papers/Multiscale/paper/paper.tex
figs_paper

@@
 
-\author*[1,2,3,4]{\pfx{Dr} \fnm{Juste} \sur{Raimbault} \dgr{PhD}}\email{juste.raimbault@ign.fr}
-% orcid 0000-0003-0768-9480
+\author{}
+%\author*[1,2,3,4]{\pfx{Dr} \fnm{Juste} \sur{Raimbault} }\email{juste.raimbault@ign.fr}% \dgr{PhD} % orcid 0000-0003-0768-9480
 
-\affil*[1]{\orgdiv{LASTIG}, \orgname{Univ. Gustave Eiffel, IGN-ENSG}} %, \orgaddress{\street{73 avenue de Paris}, \city{Saint-Mand{\'e}}, \postcode{93160}, \country{France}}
-\affil[2]{\orgdiv{Center for Advanced Spatial Analysis}, \orgname{UCL}} % , \orgaddress{\street{90 Tottenham Court Road}, \city{London}, \postcode{}, \country{UK}}
-\affil[3]{\orgdiv{UPS CNRS 3611 ISC-PIF}, \orgname{CNRS}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
-\affil[4]{\orgdiv{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s}, \orgname{Universit{\'e} Paris 1}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
+%\affil*[1]{\orgdiv{LASTIG}, \orgname{Univ. Gustave Eiffel, IGN-ENSG}} %, \orgaddress{\street{73 avenue de Paris}, \city{Saint-Mand{\'e}}, \postcode{93160}, \country{France}}
+%\affil[2]{\orgdiv{Center for Advanced Spatial Analysis}, \orgname{UCL}} % , \orgaddress{\street{90 Tottenham Court Road}, \city{London}, \postcode{}, \country{UK}}
+%\affil[3]{\orgdiv{UPS CNRS 3611 ISC-PIF}, \orgname{CNRS}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
+%\affil[4]{\orgdiv{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s}, \orgname{Universit{\'e} Paris 1}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
 
@@
 \begin{figure}[h]
- \includegraphics[width=\linewidth]{figures/multiscale_morph.pdf}
+ \includegraphics[width=\linewidth]{figures/Fig1.jpg}
  \caption{Schematic representation of the model architecture.
@@
 We follow here the same exact steps described in \cite{raimbault2018calibration}, with parameters $N_G^{(i)}$ total population increase, $\beta_i$ diffusion level, $\alpha_i$ aggregation level, $t_m$ time steps and $n_d$ number of diffusions.
-The model adds iteratively population to grid cells, by distributing it with a probability following a preferential attachment to population, and diffusing it.
+The model iteratively adds population to grid cells by distributing it with a probability following a preferential attachment to population, and then diffuses it in space.
 
@@
 
-The Table~\ref{tab:parameters} summarises model parameters. For clarity, we do not include parameters linked to synthetic model setup, and distinguish them by the scale to which they are attached.
-% TODO ref table beginning section
+Model parameters are summarised in Table~\ref{tab:parameters}. For clarity, we do not include parameters linked to synthetic model setup, and distinguish them by the scale to which they correspond.
 
@@
 
-Model behavior is characterized using the following indicators: at the  macroscopic scale: distributions of population, accessibilities, and centralities (summarized by average, hierarchy, entropy), following \cite{raimbault2020unveiling}.
+Model behaviour is characterised using the following indicators: at the  macroscopic scale: distributions of population, accessibilities, and centralities (summarized by average, hierarchy, entropy), following \cite{raimbault2020unveiling}.
 At the mesoscopic scale, urban form is captured following the indicators used by \cite{raimbault2018calibration}, i.e.
@@
 
-Model implementation is done under some performance constraints: $N$ mesoscopic morphogenesis models must be simulated in parallel at each macroscopic time step.
+Model implementation has to meet some performance constraints: $N$ mesoscopic morphogenesis models must be simulated in parallel at each macroscopic time step.
 On the contrary, macroscopic interactions are efficient to implement as they are based on matrices computations.
@@
 \begin{figure}[h]
-	\includegraphics[width=\linewidth]{figures/allindics_hist.png}
+	\includegraphics[width=\linewidth]{figures/Fig2.jpg}
 	\caption{Statistical distribution of indicators for a sample of 10 parameter values.\label{fig:fig2}}
@@
 Results show an important role of the strong coupling between scales.
-This is shown for example by the variation of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario (negative feedback of population growth on diffusion $\delta \beta$) to a ``sprawl'' scenario (positive feedback $\delta \beta$) as shown in Fig.~\ref{fig:fig4} below.
+This is shown for example by the variation of macroscopic indicators trajectories when switching from a ``transit-oriented development'' scenario (negative feedback of population growth on diffusion $\delta \beta$) to a ``sprawl'' scenario (positive feedback $\delta \beta$) as shown in Fig.~\ref{fig:fig3} below.
 
 \begin{figure}[H]
-	\includegraphics[width=\linewidth]{figures/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png}
-	\caption{Macroscopic hierarchy of populations as a function of mesoscopic time steps $t_m$, for different values of interaction decay $d_G$ (colour), and of feedback parameters $\delta d$ (columns) and $\delta \beta$ (rows).\label{fig:fig4}}
+	\includegraphics[width=\linewidth]{figures/Fig3.jpg}
+	\caption{Macroscopic hierarchy of populations as a function of mesoscopic time steps $t_m$, for different values of interaction decay $d_G$ (colour), and of feedback parameters $\delta d$ (columns) and $\delta \beta$ (rows).\label{fig:fig3}}
 \end{figure}
 
-We show in Fig.~\ref{fig:fig3} that both macroscopic and mesoscopic indicators have a U-shaped behavior as a function of the bottom-up feedback parameter $\delta d$.
+We show in Fig.~\ref{fig:fig4} that both macroscopic and mesoscopic indicators have a U-shaped behaviour as a function of the bottom-up feedback parameter $\delta d$.
 In terms of macroscopic trajectories, we see that hierarchies of accessibilities, closeness centrality and population are maximal in negative value when the mechanism is deactivated, what means that the feedback process moderates the hierarchisation of the global urban system.
@@
 \begin{figure}[h]
-	\includegraphics[width=\linewidth]{figures/onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png}
-	\caption{Variations of indicators as a function of $\delta d$, when all other parameters are fixed to their nominal value.\label{fig:fig3}}
+	\includegraphics[width=\linewidth]{figures/Fig4.jpg}
+	\caption{Variations of indicators as a function of $\delta d$, when all other parameters are fixed to their nominal value.\label{fig:fig4}}
 \end{figure}
@@
 \begin{figure}[h]
-	\includegraphics[width=\linewidth]{figures/deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+	\includegraphics[width=\linewidth]{figures/Fig5.jpg}
 	\caption{Relative evolution of average mescoscopic Moran, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig5}}
@@
 When investigating population hierarchy the mesoscopic scale, i.e.
-inequalities of population distribution within the urban area, we find in Fig.~\ref{fig:fig6} a U-shaped behavior as a function of $\delta \beta$ for negative $\delta \alpha$ values, but has a plateau for positive values of $\delta \alpha$, showing also an interaction effect between policies.
+inequalities of population distribution within the urban area, we find in Fig.~\ref{fig:fig6} a U-shaped behaviour as a function of $\delta \beta$ for negative $\delta \alpha$ values, but has a plateau for positive values of $\delta \alpha$, showing also an interaction effect between policies.
 In terms of hierarchy, a maximal negative value may be wanted to limit inequalities (slopes are negative), and this optimum is reached for a slightly positive value of $\delta \beta$ (left columns in Fig.~\ref{fig:fig6}).
@@
 There may be no link between urban form and global insertion, or it may be due to other processes, be expressed as an other functional form.
-An important stage before shifting to robust knowledge will consist in (i) reviewing and making a typology of such potential processes across scales; (ii) including most in a multi-modeling fashion to compare possible concurrent mechanisms.
+An important stage before shifting to robust knowledge will consist in (i) reviewing and making a typology of such potential processes across scales; (ii) including most in a multi-modelling fashion to compare possible concurrent mechanisms.
 

### 2022-10-26 Docs/Papers/Multiscale/paper/figures.sh

@@ -1,0 +1,29 @@
+
+JPGQUALITY=50
+PDFRESOLUTION=300
+WIDTH=2000
+HORIZONTALPADDING=10
+VERTICALPADDING=10
+
+
+# Fig 1
+convert -density $PDFRESOLUTION -background white -alpha remove figuresraw/multiscale_morph.pdf -resize "$WIDTH"x -quality $JPGQUALITY figures/Fig1.jpg
+
+# Fig 2
+convert figuresraw/allindics_hist.png -resize "$WIDTH"x -quality $JPGQUALITY figures/Fig2.jpg
+
+# Fig 3
+convert figuresraw/deltaHierarchymacroPopulations-mesoTimeSteps_colorMacroInteractionDecay_facetmesoMacroDecayUpdateMax-macroMesoBetaUpdateMax.png -resize "$WIDTH"x -quality $JPGQUALITY figures/Fig3.jpg
+
+# Fig 4
+convert figuresraw/onefactor_allindics_mesoMacroDecayUpdateMax_errorbars.png -resize "$WIDTH"x -quality $JPGQUALITY figures/Fig4.jpg
+
+# Fig 5
+convert figuresraw/deltamesoMorans-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png -resize "$WIDTH"x -quality $JPGQUALITY figures/Fig5.jpg
+
+# Fig 6
+convert figuresraw/deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png -resize "$WIDTH"x -quality $JPGQUALITY figures/Fig6.jpg
+
+# Fig 7
+convert figuresraw/pareto_macroHierarchy-mesoHierarchy.png -resize "$WIDTH"x -quality $JPGQUALITY figures/Fig7.jpg
+

### 2022-10-26 Docs/Papers/Multiscale/paper/paper.tex
paper

@@
 
-The system is initialized with synthetic systems of cities.
+We setup the model with synthetic systems of cities.
+This allows obtaining conclusions independently of geographical contingencies \cite{raimbault2019space}.
+Setup on real configurations is left for future studies at this stage.
+
 A number $N=20$ cities are distributed randomly into a regional urban system, as a square world of width $w=500km$ (reference unit for the decay parameter).
@@
 Within each city, we consider initial population grids as monocentric (grid of size $W=50$ and center cell density $P_m = 1000$ population units, following an exponential kernel of width one fifth of the world).
-%check param + necessary experiment: influecne of other initial configs: at least kernel mixture?
 The model is simulated for 20 macroscopic time steps (order of magnitude of half a century), each representing $t_m$ mesoscopic time steps (parameter changed in experiments).
+
+%For the mesoscopic monocentric kernel: check param + necessary experiment: influecne of other initial configs: at least kernel mixture?
 
@@
 
-Model behaviour is characterised using the following indicators: at the  macroscopic scale: distributions of population, accessibilities, and centralities (summarized by average, hierarchy, entropy), following \cite{raimbault2020unveiling}.
+Model behaviour is characterised using aggregated indicators at each scale considered.
+At the  macroscopic scale, we follow \cite{raimbault2020unveiling} and consider distributions of population, accessibilities, and centralities, all three summarised by their respective average, hierarchy, and entropy.
 At the mesoscopic scale, urban form is captured following the indicators used by \cite{raimbault2018calibration}, i.e.
@@
 
-% TODO extend indicator descritpion
+% extend indicator description? not necessary, refs?
 
@@
 The relative large number of parameters and output indicators imposes the use of appropriate model exploration methods to get a grasp of model behavior.
-Therefore, the parameter space is explored with the OpenMOLE model exploration software \cite{reuillon2013openmole}.
+Therefore, the parameter space is explored with the OpenMOLE model exploration software \citep{reuillon2013openmole}.
 Standalone source code for the model is available on an open git repository as \cite{urbangrowthmodel}.
@@
 %(min around 1 for 3 indicators, above 5 for 2 and around 0.5 for 2)}
+% % check these values for steps and repetitions?
 
 The first numerical experiment is aimed at checking the statistical consistency of model output across stochastic repetitions.
-We run a one-factor-at-a-time sampling on all parameters with 5 steps for each within the domain provided in Table~\ref{tab:parameters}, and 500 stochastic repetitions for each parameter.% check these values
+We run a one-factor-at-a-time sampling on all parameters with 5 steps for each within the domain provided in Table~\ref{tab:parameters}, and 500 stochastic repetitions for each parameter.
 Obtained statistical distribution are shown in Fig.~\ref{fig:fig2} for a sample of 10 parameters.
 Different indicators exhibit various distribution shapes, but are always unimodal and generally rather localized.
-This means different parameter values can be discriminated when considering summary statistics.
+Therefore, different parameter values can be discriminated when considering summary statistics.
 To estimate the role of noise, we compute Sharpe ratios (rate between average and standard deviations) for each indicators and parameter values.
@@
 
-We then turn to a grid exploration of the parameter space. %precise successive experiments, step values and varied parameters
+ %precise successive experiments, step values and varied parameters
+
+We then turn to a grid exploration of the parameter space.
 Results show an important role of the strong coupling between scales.
@@
 
-The strong coupling of scales, in particular the interaction between bottom-up and top-down feedbacks, has a non-trivial influence on model behavior, as shown in Fig.~\ref{fig:fig4} for the macroscopic hierarchy of populations.
-This allows investigating the influence of the macroscopic interaction decay and the mesoscopic evolution speed, under stylised scenarios for the top-down feedback.
-We exhibit interaction effects between $\delta \beta$ and $\delta d$, as quantitative and qualitative behavior as a function of $t_m$ changes across panels.
+The strong coupling of scales, in particular the interaction between bottom-up and top-down feedbacks, has a non-trivial influence on model behaviour, as shown in Fig.~\ref{fig:fig4} for the macroscopic hierarchy of populations.
+We study therein the influence of the macroscopic interaction decay and the mesoscopic evolution speed, under stylised scenarios for the top-down feedback.
+We observe interaction effects between $\delta \beta$ and $\delta d$, as quantitative and qualitative behaviour as a function of $t_m$ changes across panels.
 Interaction range $d_G$ has on the contrary no significant effect.
@@
 In particular, the average variation of Moran index captures the centralisation of urban areas.
-This centralisation exhibits a minimal value when $\delta \alpha$ is low, consistent across values of $\lambda$.
-This implicates that when decentralisation policies are implemented by acting on aggregation ($\delta \alpha$), the policy on urban sprawl will have an unexpected effect on centralisation.
-This is an illustration of complex and possibly negative interaction between policies.
+It exhibits a minimal value when $\delta \alpha$ is low, consistent across values of $\lambda$.
+As a consequence, when decentralisation policies are implemented by acting on aggregation ($\delta \alpha$), the policy on urban sprawl will have an unexpected effect on centralisation.
+This illustrates complex and possibly negative interaction between policies.
+
 % influenced by upward feedback? No, rather stable when \lambda = 0 -> why? - check influence of lambda
@@
 \begin{figure}[h]
-	\includegraphics[width=\textwidth]{figures/deltamesoSlopes-macroMesoAlphaUpdateMax_colormacroMesoBetaUpdateMax_facetmesoMacroCongestionCost-mesoMacroDecayUpdateMax_mesoBeta0_11.png}
+	\includegraphics[width=\textwidth]{figures/Fig6.jpg}
 	\caption{Relative evolution of average mescoscopic population hierarchy, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig6}}
@@
 
-When investigating population hierarchy the mesoscopic scale, i.e.
+When investigating population hierarchy at the mesoscopic scale, i.e.
 inequalities of population distribution within the urban area, we find in Fig.~\ref{fig:fig6} a U-shaped behaviour as a function of $\delta \beta$ for negative $\delta \alpha$ values, but has a plateau for positive values of $\delta \alpha$, showing also an interaction effect between policies.
-In terms of hierarchy, a maximal negative value may be wanted to limit inequalities (slopes are negative), and this optimum is reached for a slightly positive value of $\delta \beta$ (left columns in Fig.~\ref{fig:fig6}).
-This means that sprawl may be needed in some configurations to optimise this objective.
+In terms of hierarchy, a maximal negative value may be wanted to limit inequalities (slopes are negative), and this optimum is reached for a slightly positive value of $\delta \beta$ (left columns in Fig.~\ref{fig:fig6}), what means that sprawl may be needed in some configurations to optimise this objective.
 
@@
 	\begin{center}
-	\includegraphics[width=0.6\linewidth]{figures/pareto_macroHierarchy-mesoHierarchy.png}
+	\includegraphics[width=0.6\linewidth]{figures/Fig7.jpg}
 	\end{center}
@@
 We finally turn to an optimization objective to investigate if compromises have to be made when targets are simultaneously aimed for at different scales.
-We therefore use a NSGA2 algorithm implemented in OpenMOLE, with a population of 200 individuals, run for 100,000 generations.
+We therefore use a NSGA2 algorithm (Non-dominated Sorting Genetic Algorithm)  implemented in OpenMOLE, with a population of 200 individuals, run for 100,000 generations.
 Objectives are the average mesoscopic hierarchy which can be minimized for more compact and thus sustainable cities, at the macroscopic population hierarchy.
@@
 We show in Fig.~\ref{fig:fig7} the Pareto front obtained after convergence of the algorithm.
-This confirms that the two objectives are indeed contradictory and that compromises between scales have to be made when trying to optimise both simultaneously.
+The shape of the front confirms that the two objectives are indeed contradictory and that compromises between scales have to be made when trying to optimise both simultaneously.
 Most points correspond to policies favouring sprawl, at the exception of a small set of points with low macroscopic hierarchy values, which discourage sprawl ($\delta \beta < 0$).
@@
 
-% TODO improve discussion - link with literature: related results, interpretation regarding previous studies
+%  improve discussion - link with literature: related results, interpretation regarding previous studies
 
-% TODO limitations of the work
+%  limitations of the work
 
@@
 
-% TODO implication of results in a broad context
+% implication of results in a broad context
 
@@
 In that sense, it illustrates weak emergence in its proper sense \cite{bedau2002downward}.
-This suggests this view can be effectively operationalised into models of urban dynamics.
+We confirm therefore that this approach can be effectively operationalised into models of urban dynamics.
 
@@
 However, such data is difficult to obtain, and some investigations should be made on the relevant calibration targets (in particular at which scale).
-% parametrization on real systems; possibly calibration \cite{raimbault2019ilus}
+% parametrization on real systems; possibly calibration \cite{ilus}
 

### 2022-11-03 Docs/Papers/Multiscale/paper/paper.tex
revision_paper

@@
 We show in Fig.~\ref{fig:fig4} that both macroscopic and mesoscopic indicators have a U-shaped behaviour as a function of the bottom-up feedback parameter $\delta d$.
-In terms of macroscopic trajectories, we see that hierarchies of accessibilities, closeness centrality and population are maximal in negative value when the mechanism is deactivated, what means that the feedback process moderates the hierarchisation of the global urban system.
-For the urban form, this feedback also mitigates the aggregation of population.
+Regarding macroscopic indicators, we see that hierarchies of accessibilities, closeness centrality and population are all minimal when the corresponding  feedback mechanism is deactivated ($\delta d = 0$), what means that the feedback process moderates the hierarchisation of the global urban system.
+Regarding the urban form at the mesoscopic scale, this feedback also mitigates the intra-urban hierarchy of population (since mesoscopic slope has a similar behaviour).
 
+%%%%%%%%%%%%%%%%%%
 \begin{figure}[h]
@@
 \end{figure}
+%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%
 \begin{figure}[h]
 	\includegraphics[width=\linewidth]{figures/Fig5.jpg}
-	\caption{Relative evolution of average mescoscopic Moran, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig5}}
+	\caption{Relative evolution of average mesoscopic Moran, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig5}}
 \end{figure}
-
-% 20190506_135221_MULTISCALE_TARGETEDGRID_GRID
-
-% fig 4
+%%%%%%%%%%%%%%%%%%
 
@@
 	\includegraphics[width=\textwidth]{figures/Fig6.jpg}
-	\caption{Relative evolution of average mescoscopic population hierarchy, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (colour), and of feedback parameters $\delta d$ (columns) and $\lambda$ (rows).\label{fig:fig6}}
+	\caption{Relative evolution of average mescoscopic population hierarchy, as a function of feedback parameter $\delta \beta$, for different values of feedback parameter $\delta \alpha$ (columns), and of feedback parameters $\delta d$ (colour) and $\lambda$ (rows).\label{fig:fig6}}
 \end{figure}
@@
 When investigating population hierarchy at the mesoscopic scale, i.e.
-inequalities of population distribution within the urban area, we find in Fig.~\ref{fig:fig6} a U-shaped behaviour as a function of $\delta \beta$ for negative $\delta \alpha$ values, but has a plateau for positive values of $\delta \alpha$, showing also an interaction effect between policies.
-In terms of hierarchy, a maximal negative value may be wanted to limit inequalities (slopes are negative), and this optimum is reached for a slightly positive value of $\delta \beta$ (left columns in Fig.~\ref{fig:fig6}), what means that sprawl may be needed in some configurations to optimise this objective.
+inequalities of population distribution within the urban area, we find in Fig.~\ref{fig:fig6} a U-shaped behaviour as a function of $\delta \beta$ for negative $\delta \alpha$ values.
+This behaviour disappear for positive values of $\delta \alpha$, showing also an interaction effect between policies.
+A maximal value of this indicator may be desirable to limit inequalities of population distribution within urban areas, and this optimum is reached for a slightly positive value of $\delta \beta$ (left columns in Fig.~\ref{fig:fig6}), what means that a slight amount of urban sprawl may be needed in some configurations to optimise this objective.
 
@@
 We finally turn to an optimization objective to investigate if compromises have to be made when targets are simultaneously aimed for at different scales.
-We therefore use a NSGA2 algorithm (Non-dominated Sorting Genetic Algorithm)  implemented in OpenMOLE, with a population of 200 individuals, run for 100,000 generations.
+We therefore use a NSGA2 algorithm - Non-dominated Sorting Genetic Algorithm - implemented in OpenMOLE, with a population of 200 individuals, run for 100,000 generations.
 Objectives are the average mesoscopic hierarchy which can be minimized for more compact and thus sustainable cities, at the macroscopic population hierarchy.
@@
 
-% implication of results in a broad context
+In a broad perspective, our model effectively captures an interaction between downward and upward feedback.
+In that sense, it illustrates weak emergence in its proper sense \citep{bedau2002downward}.
+We confirm therefore that this approach can be effectively operationalised into models of urban dynamics.
+Considering with more details the simulation results we obtained, % TODO bilan des faits importants
 
-Our model effectively captures an interaction between downward and upward feedback.
-In that sense, it illustrates weak emergence in its proper sense \cite{bedau2002downward}.
-We confirm therefore that this approach can be effectively operationalised into models of urban dynamics.
-
-In terms of model complexity and in relation to urban complexity in itself, we show that coupling simple models (each having only three parameters) already yields a complicated and complex simulation model.
-It also requires additional ontologies to achieve the coupling (in our case the policy processes on one side and the congestion-performance process on the other side).
-Therefore, we suggest this is additional evidence for a necessity of complexity and simulation models to understand urban complexity.
-This implies in some way the failure of reductionist epistemologies, at least to grab the multi-scalar nature of urban systems.
-% 'suggest in a way failure of reductionist empistemology (in the sense of a la Barthelemy) - at least for grabbing the multi-scalar nature of systems' (seems a tautology though
+Considering urban complexity and how it can be captured within simulation models, we show that coupling simple models (each having only three parameters) already yields a complicated and complex simulation model.
+It also requires additional ontologies to achieve the coupling (in our case the policy processes for the bottom-up feedback and the congestion-performance process for the top-down feedback).
+We suggest this is additional evidence for a necessity of complex simulation models to understand urban complexity.
+In some way, reductionist epistemologies will thus fail to grab the multi-scalar nature of urban systems.
 
 Although a stylised and highly simplified model, we were already able to integrate parameters linked to policies.
-Indeed, governance processes are intrinsically multi-scalar \cite{liao2017ouverture}.
-Therefore, developing such models is a first step towards a progressive integration models for policy and sustainable management of territories.
+Indeed, governance processes are intrinsically multi-scalar \citep{liao2017ouverture}, and developing such models is a first step towards a progressive integration of models for policy and sustainable management of territories.
 
@@
 
-%\subsection{Multi-modeling and concurrent processes}
-
-This model is only a first structural sketch with very restrictive assumption, in particular regarding the downward and upward feedbacks on submodel parameters.
-There may be no link between urban form and global insertion, or it may be due to other processes, be expressed as an other functional form.
-An important stage before shifting to robust knowledge will consist in (i) reviewing and making a typology of such potential processes across scales; (ii) including most in a multi-modelling fashion to compare possible concurrent mechanisms.
+The model we studied is only a first structural sketch with very restrictive assumptions,s in particular regarding the downward and upward feedbacks on submodel parameters.
+There may be no such direct link between urban form and global performance of urban areas, or it may be due to other processes, or captured with others functional forms.
+An important stage before shifting to robust knowledge will consist in (i) reviewing and making a typology of such potential processes across scales; (ii) including most in a multi-modelling framework to compare the plausibility of possible concurrent mechanisms.
 
@@
 
+\section{Conclusion}
+
 Our model is a first step towards multi-scalar models of urban dynamics which effectively include a strong coupling between processes at different scales.
 While remaining highly stylised, we showed an important role of the coupling and unveiled for example interaction effects between different types of policies.
-This type of approach will need in the future to be further developed, towards multi-scalar models for sustainable policy making \cite{Rozenblat2018}.
+This type of approach will need in the future to be further developed, towards multi-scalar models for sustainable policy making \citep{Rozenblat2018}.
 This works paves the way for such more complex models.

### 2022-11-03 Docs/Papers/Multiscale/paper/paper.tex
.

@@
 Future work also implies the use of more elaborated model exploration and validation methods.
-For example, using the diversity search integrated in OpenMOLE \cite{reuillon2013openmole} would allow unveiling the diversity of possible future urban trajectories, and possibly unexpected novel pathways towards sustainability, finding for example regimes with the strongest effect of feedback parameters on indicators related to sustainability.
+For example, using the diversity search integrated in OpenMOLE \citep{reuillon2013openmole} would allow unveiling the diversity of possible future urban trajectories, and possibly unexpected novel pathways towards sustainability, finding for example regimes with the strongest effect of feedback parameters on indicators related to sustainability.
 % spatial sensitivity - influence of synthetic system of cities

### 2024-10-01 Docs/Papers/MesoBenchmark/paper/paper.tex
revision_multiscale
revisions
.
bib
papers

@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
+\documentclass[referee,lineno,pdflatex,sn-apa]{sn-jnl}
 
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-%\documentclass[10pt]{article}
-%\usepackage[margin=2cm]{geometry}
+\jyear{2024}%
 
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-\usepackage{setspace} 
-\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
-%% END MACROS SECTION
+\raggedbottom
 
 \begin{document}
-\vspace*{0.2in}
 
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{A comparison of simple models for urban morphogenesis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Juste Raimbault\textsuperscript{1,2,3,*}
-\bigskip\\
-\textsuperscript{1} Center for Advanced Spatial Analysis, University College London
-\\
-\textsuperscript{2} UPS CNRS 3611 ISC-PIF
-\\
-\textsuperscript{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
-\bigskip\\
-* juste.raimbault@polytechnique.edu
+\title[A comparison of simple models for urban morphogenesis]{A comparison of simple models for urban morphogenesis}
 
-\end{flushleft}
+\author{}
+%\author*[1,2,3,4]{\pfx{Dr} \fnm{Juste} \sur{Raimbault} }\email{juste.raimbault@ign.fr}% \dgr{PhD} % orcid 0000-0003-0768-9480
 
-% Please keep the abstract below 300 words
-\section*{Abstract}
-\vspace{1cm}
-The spatial distribution of population and activities within urban areas, or urban form at the mesoscopic scale, is the outcome of multiple antagonist processes.
+%\affil*[1]{\orgdiv{LASTIG}, \orgname{Univ. Gustave Eiffel, IGN-ENSG}} %, \orgaddress{\street{73 avenue de Paris}, \city{Saint-Mand{\'e}}, \postcode{93160}, \country{France}}
+%\affil[2]{\orgdiv{Center for Advanced Spatial Analysis}, \orgname{UCL}} % , \orgaddress{\street{90 Tottenham Court Road}, \city{London}, \postcode{}, \country{UK}}
+%\affil[3]{\orgdiv{UPS CNRS 3611 ISC-PIF}, \orgname{CNRS}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
+%\affil[4]{\orgdiv{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s}, \orgname{Universit{\'e} Paris 1}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
+
+% Target journal : Computational Urban Science https://link.springer.com/journal/43762/submission-guidelines
+
+% TODO Reviews Plos One
+
+\abstract{The spatial distribution of population and activities within urban areas, or urban form at the mesoscopic scale, is the outcome of multiple antagonist processes.
 We propose in this paper to benchmark different models of urban morphogenesis, to systematically compare the urban forms they can produce.
@@
 Applying a diversity search algorithm, we estimate the feasible space of each model within a space of urban form indicators, in comparison of empirical values for worldwide urban areas.
-We find a complementarity of the different types of processes, advocating for a plurality of urban models.
+We find a complementarity of the different types of processes, advocating for a plurality of urban models.}
 
-% urban morphogenesis; reaction-diffusion; correlated percolation; gravity model; novelty search
+\keywords{Urban morphogenesis; Reaction-diffusion; Correlated percolation; Gravity model; Novelty search}
 
-\linenumbers
+\maketitle
+
+Batty, M. (2001). Polynucleated urban landscapes. Urban studies, 38(4), 635-655. model to include?
+
+\cite{xu2021emergence}
+
+\cite{strano2021agglomeration}
+
+\cite{rybski2021comparing}
+
+\cite{Rybski2021Modeling} coupling inside model - more to benchmark - multimodeling
+
+\cite{bithellsystematic} benchmark abm covid
 
 \cite{jun2020modling} entropy and urban growth
+
+\cite{Nicholse2100769118} reproducibility - benchmarks - sequence of studies
 
@@
 Area of circles gives population.\label{fig:fig1}}
+% TODO redo maps with open resources https://www.naturalearthdata.com/downloads/110m-cultural-vectors/
 \end{figure}
@@
 
-%\section*{Supporting information}
-
-%\section*{Acknowledgments}
-
-\nolinenumbers
-
 \bibliography{biblio}

### 2024-10-01 Docs/Papers/Multiscale/paper/paper.tex
.
papers

@@
 
-\jyear{2022}%
+\jyear{2024}%
 
@@
 %\affil[4]{\orgdiv{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s}, \orgname{Universit{\'e} Paris 1}} % , \orgaddress{\street{Place Nationale}, \city{Paris}, \postcode{75013}, \country{France}}
+
+% Target: Urban Informatics
+% https://www.springer.com/journal/44212/submission-guidelines
+% check Response_to_Reviewers submission IJGI
 
@@
 
-\abstract{Urban evolution processes occur at different scales, with intricate interactions between levels and relatively distinct type of processes.
-To what extent actual urban dynamics include an actual strong coupling between scales, in the sense of both top-down and bottom-up feedbacks, remains an open issue with important practical implications for the sustainable management of territories.
+\abstract{Urban dynamics occur at different scales, with intricate interactions between levels and relatively distinct type of processes.
+To what extent observed urban trajectories include an actual strong coupling between scales, in the sense of both top-down and bottom-up feedbacks, remains an open issue with important practical implications for the sustainable management of territories.
 We introduce in this paper a multi-scalar simulation model of urban growth, coupling a system of cities interaction model at the macroscopic scale with morphogenesis models for the evolution of urban form at the scale of metropolitan areas.
@@
 
-Modelling urban growth and more generally the dynamics of urban systems is central to the design of sustainable territorial policies, enabling the understanding of past urbanisation processes and the anticipation of future urban trajectories.
+Modelling urban growth and more generally the dynamics of urban systems is central to the design of sustainable territorial policies, enabling the understanding of past urbanisation processes and the anticipation of future urban trajectories \citep{pumain2017urban}.
 The design of sustainable future cities requires an historical knowledge of how past cities came to be and evolved \citep{batty2018inventing}.
 Several models have been developed at different scales and integrating multiple urban systems dimensions.
-Some typical examples are models of land-use change at a mesoscopic scale \citep{iacono2008models}, or systems of cities models at a macroscopic scale \citep{pumain2017urban}.
+Some typical examples are models of land-use change at a mesoscopic scale \citep{iacono2008models}, or systems of cities models at a macroscopic scale \citep{raimbault2020empowering}.
 
@@
 
-In disciplines close to urban modelling, methods have been developed for multi-scale models.
+In disciplines that are neighbour to urban modelling, methods have been developed for multi-scale models.
 For example in spatial epidemiology, \cite{banos2015importance} combines agent-based modelling for local diffusion dynamics with differential equations at the population scale.
@@
 The study of traffic is also made more accurate by coupling macroscopic and microscopic models \cite{boulet2020coupling}.
-The management of ecosystems  requires integrating across actors and scales \cite{belem2013organizational}.
-These non-exhaustive illustrations highlight how the integration of scales is a crucial feature and issue in the understanding of complex systems \cite{chavalarias2009french}.
+The management of ecosystems  requires integrating across actors and scales \citep{belem2013organizational}.
+These non-exhaustive illustrations highlight how the integration of scales is a crucial feature and issue in the understanding of complex systems \citep{chavalarias2009french}.
+
+In other disciplines such as physics or engineering, the methodology and application of multi-scale modelling is much more mature.
+% TODO examples physics
 
@@
 At the  macroscopic scale, we follow \cite{raimbault2020unveiling} and consider distributions of population, accessibilities, and centralities, all three summarised by their respective average, hierarchy, and entropy.
-At the mesoscopic scale, urban form is captured following the indicators used by \cite{raimbault2018calibration}, i.e.
+At the mesoscopic scale, urban form is captured using the indicators proposed by \cite{raimbault2018calibration}, i.e.
 Moran index, average distance between individuals, hierarchy of population distribution, and entropy of population distribution.
@@
 On the contrary, macroscopic interactions are efficient to implement as they are based on matrices computations.
-The model is implemented in \texttt{scala} and integrated into a broader library for spatial sensitivity analysis and dynamical models for settlement systems, namely the \texttt{spatialdata} library \cite{raimbault2020scala}.
+The model is implemented in \texttt{scala} and integrated into a broader library for spatial sensitivity analysis and dynamical models for settlement systems, namely the \texttt{spatialdata} library \citep{raimbault2020scala}.
 This library in particular already includes implementations of the models used in \cite{raimbault2020indirect} and \cite{raimbault2018calibration}, but also of other urban interaction models such as \cite{favaro2011gibrat}.
@@
 Simulation data used in this paper is available on the dataverse repository as \cite{dataverse}.
+% TODO anon repos
 

### 2024-10-08 Docs/Papers/Multiscale/paper/paper.tex
.

@@
 
-\title[Multi-scalar urban dynamics model]{Strong coupling between scales in a multi-scalar model of urban dynamics}
+\title[Multi-scalar model of urban dynamics]{Strong coupling between scales in a multi-scalar model of urban dynamics}
 
@@
 
-\cite{lockerby2015asynchronous} physics multiscale simulations
-+ multiscale traffic models (Nancy?)
+% multiscale traffic models (Nancy?) -> ? - not found -> more lit joueiai2014multiscale ok
 
@@
 Multi-scale models have also been used for the simulation of crowd dynamics \cite{crociani2016multi}.
-The study of traffic is also made more accurate by coupling macroscopic and microscopic models \cite{boulet2020coupling}.
+The study of traffic is also made more accurate by coupling macroscopic and microscopic models \citep{boulet2020coupling,joueiai2014multiscale}.
+Urban climate \citep{rasheed2009multiscale}
 The management of ecosystems requires integrating across actors and scales \citep{belem2013organizational}.
-These non-exhaustive illustrations highlight how the integration of scales is a crucial feature and issue in the understanding of complex systems \citep{chavalarias2009french}.
+These non-exhaustive illustrations highlight how the integration of scales is a crucial feature but difficult issue to adress for the understanding of urban and social complex systems \citep{chavalarias2009french}.
 
-In other disciplines such as physics or engineering, the methodology and application of multi-scale modelling is much more mature.
+In other disciplines such as physics or engineering, the methodology and application of multi-scale modelling has been studied more extensively. \cite{lockerby2015asynchronous}
 % TODO examples physics
@@
 % TODO emphasize more the research problem
+% rq positioning: not large model but simple whih can be understood (cf requiem for large scale models)
+% \citep{batty2014can}
 
@@
 no reciprocal and dynamical feedbacks); (ii) the behaviour of the model is systematically studied on synthetic systems of cities using model exploration methods.
-
-% rq positioning: not large model but simple whih can be understood (cf requiem for large scale models)
 
